---
title: "Tutorial 1: Importing Data from DATASUS"
author: "climasus4r Team"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    code_download: true
    self_contained: true
    df_print: paged
---

```{css, echo=FALSE}
/* -- Variaveis de Cores Modernas -- */
:root {
  --primary-color: #1D9E75;
  --dark-green: #0f3d2e;
  --bg-light: #f8faf9;
  --text-main: #1a1a1a;
  --code-bg: #fdfdfd;
  --warning-orange: #ff9800;
  --info-blue: #2196f3;
}
html { scroll-behavior: smooth; }
body { font-family: 'Source Serif 4', Georgia, serif; font-size: 18px; color: var(--text-main); background-color: #fff; line-height: 1.6; }
h1, h2, h3, h4 { font-family: 'Inter', sans-serif; font-weight: 700; color: var(--dark-green); }
h1.title { font-size: 2.8em; border-bottom: 4px solid var(--primary-color); padding-bottom: 10px; margin-bottom: 1.5rem; }
h2 { margin-top: 2rem; border-left: 4px solid var(--primary-color); padding-left: 1rem; }
pre.r, pre.sourceCode { background-color: var(--code-bg); border: 1px solid #e1e1e1; border-left: 5px solid var(--primary-color) !important; border-radius: 8px; box-shadow: 0 2px 4px rgba(0,0,0,0.05); padding: 1rem; }
.callout { padding: 1.2em 1.5em; margin: 1.5em 0; border-radius: 8px; border-left: 6px solid; background-color: var(--bg-light); }
.callout-note { background-color: #f0f7ff; border-color: #007bff; }
.callout-tip { background-color: #e8f5e9; border-color: #1D9E75; }
.callout-warning { background-color: #fff3e0; border-color: #ff9800; }
.workflow-container { background: linear-gradient(135deg, #f8faf7 0%, #ffffff 100%); border-radius: 12px; padding: 1.5rem; margin: 1.5rem 0; border: 1px solid #dce8e1; overflow-x: auto; }
.workflow-steps { display: flex; flex-wrap: wrap; justify-content: center; gap: 0.5rem; font-family: 'Inter', monospace; font-size: 0.8rem; }
.workflow-step { background: var(--primary-color); color: white; padding: 0.5rem 1rem; border-radius: 20px; white-space: nowrap; }
.workflow-arrow { color: var(--primary-color); font-weight: bold; font-size: 1.2rem; }
.feature-grid { display: grid; grid-template-columns: repeat(auto-fit, minmax(280px, 1fr)); gap: 1.5rem; margin: 2rem 0; }
.feature-card { background: white; border-radius: 12px; padding: 1.5rem; border: 1px solid #e1e1e1; transition: transform 0.2s ease, box-shadow 0.2s ease; text-align: center; }
.feature-card:hover { transform: translateY(-4px); box-shadow: 0 8px 24px rgba(0,0,0,0.1); }
.feature-icon { font-size: 2rem; margin-bottom: 1rem; }
.feature-card h3 { margin-top: 0; color: var(--primary-color); font-size: 1.1rem; }
table { width: 100%; border-collapse: collapse; margin: 1.5rem 0; }
th, td { padding: 12px; text-align: left; border-bottom: 1px solid #e1e1e1; }
th { background-color: var(--bg-light); font-weight: 600; color: var(--dark-green); }
.github-feedback-btn { display: inline-flex; align-items: center; gap: 8px; background-color: #24292e; color: white; padding: 10px 20px; border-radius: 6px; text-decoration: none; font-family: sans-serif; transition: background-color 0.2s; margin-top: 1rem; }
.github-feedback-btn:hover { background-color: #2c3e50; color: white; text-decoration: none; }
@media (max-width: 768px) { body { font-size: 16px; } h1.title { font-size: 2rem; } .workflow-steps { flex-direction: column; align-items: center; } }
```


```{r setup, include=FALSE}
# ATENÇÃO — estratégia de pré-renderização.
# Este tutorial NÃO executa downloads durante a renderização. Todas as figuras e
# tabelas foram geradas UMA vez pelo script scripts/tutorial_01_importacao.R e
# são apenas carregadas aqui (PNG via include_graphics; RDS via readRDS).
# Por isso, os chunks de ensino abaixo usam eval = FALSE: o código é mostrado,
# mas não é executado no momento do knit.
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)

# Helpers para exibir artefatos de forma robusta: se o arquivo ainda não foi
# gerado (script de geração não executado), mostram uma nota em vez de quebrar
# o knit. Assim a vignette renderiza a qualquer momento.
fig_ou_nota <- function(path, nota) {
  if (file.exists(path)) knitr::include_graphics(path)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
tbl_ou_nota <- function(path, nota, caption = NULL) {
  if (file.exists(path)) knitr::kable(readRDS(path), caption = caption)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
.nota_gerar <- "Artefato indisponível — rode `Rscript scripts/tutorial_01_importacao.R` (a partir da raiz do pacote) para gerá-lo."
```

# Objective

By the end of this tutorial you will be able to **import microdata from the
DATASUS Health Information Systems** using a single function —
`sus_data_import()` — and to control its main features: the choice of
**system** (SIM, SIH, SINAN, SINASC, CNES, SIA) and its subsystems, the
**spatial scope** (by state, by a predefined region, or by municipality), the
difference between **annual and monthly** systems, the choice of processing
**backend**, and — essential for reproducible, large-volume work — **smart
caching** and **parallel processing**.

Import is the **first stage** of the `climasus4r` pipeline. Everything that
comes afterward (standardization, filtering, variable creation, aggregation,
spatial join, socioeconomic linkage, climate integration, and modeling) starts
from the raw base built here.

# Scientific context

Climate-health research in Brazil relies on the **Health Information Systems
(SIS)** maintained by the Ministry of Health and made available through
DATASUS. Each system records one type of event, with national coverage and
municipal granularity:

- **SIM** — Mortality Information System (death certificates);
- **SIH** — Hospital Information System (hospitalizations/AIH);
- **SINAN** — Notifiable Diseases Information System (notifiable diseases,
  such as dengue, zika, and chikungunya);
- **SINASC** — Live Birth Information System;
- **CNES** — National Registry of Health Establishments;
- **SIA** — Outpatient Information System.

Programmatic access to these `.dbc` files is mediated by the `microdatasus`
package (Saldanha, Bastos & Barcellos, 2019). `climasus4r` wraps this access in
`sus_data_import()`, adding **caching** (so the same data is not re-downloaded
on every run), **parallelism** (to speed up downloads across many
states/years), and multilingual messages. Importing reproducibly — recording
the system, year, and spatial scope — is what makes a study auditable and
replicable.

> **Guiding case.** Throughout this tutorial series we investigate
> **temperature and health outcomes in older adults (60+) in the São Paulo
> Metropolitan Region, 2014–2019**. In this first tutorial we assemble the
> **multi-system raw base**: deaths (SIM-DO), hospitalizations (SIH-RD), and,
> as a secondary case on climate × arboviruses, dengue (SINAN-DENGUE). The
> objects imported here are saved and will serve as input for Tutorial 2
> (Encoding and Standardization).

# Data used

| System | `system =` | Periodicity | Role in the case |
|---------|------------|---------------|---------------|
| SIM-DO (deaths) | `"SIM-DO"` | Annual | Primary outcome (mortality) |
| SIH-RD (hospitalizations) | `"SIH-RD"` | Monthly (requires `month`) | Hospital morbidity |
| SINAN-DENGUE | `"SINAN-DENGUE"` | Annual | Secondary case (climate × dengue) |

Spatial scope: the state of **São Paulo** (`uf = "SP"`), 2014–2019. All data
are **public and real**, obtained from DATASUS via `microdatasus`.

# Complete pipeline

The blocks below reproduce what the `scripts/tutorial_01_importacao.R` script
runs. Start by loading the package:

```{r carregar}
library(climasus4r)
```

## Importing an annual system (SIM-DO)

SIM is **annual**: the `month` argument is ignored. We request all six years at
once and leave caching and parallelism turned on.

```{r import-sim}
obitos <- sus_data_import(
  uf        = "SP",
  year      = 2014:2019,
  system    = "SIM-DO",
  backend   = "arrow",   # "arrow" (lazy, padrão) ou "tibble" (em memória)
  use_cache = TRUE,        # reaproveita downloads anteriores
  parallel  = TRUE,        # baixa anos em paralelo
  workers   = 4,
  lang      = "pt"
)
```

**View the data**
```{r}
dplyr::glimpse(obitos)
```


## Importing a monthly system (SIH-RD)

SIH is **monthly**: the `month` argument is **mandatory**. Here we request
only the winter months (June to August), when respiratory mortality in older
adults tends to be higher.

```{r import-sih}
internacoes <- sus_data_import(
  uf        = "SP",
  year      = 2014:2019,
  month     = 6:8,         # obrigatório para SIH, CNES e SIA
  system    = "SIH-RD",
  use_cache = TRUE,
  parallel  = TRUE,
  workers   = 4,
  lang      = "pt"
)
```

**View the data**
```{r}
dplyr::glimpse(internacoes)
```


## Importing notifications (SINAN-DENGUE)

```{r import-sinan}
dengue <- sus_data_import(
  uf        = "SP",
  year      = 2014:2019,
  system    = "SINAN-DENGUE",
  use_cache = TRUE,
  lang      = "pt"
)
```


**View the data**
```{r}
dplyr::glimpse(dengue)
```


## Spatial scope: by region or by municipality

Instead of listing states, you can use a **predefined group** in `region`
(macro-regions, biomes, or river basins) — useful for regional studies:

```{r import-region}
# Toda a macrorregião Sudeste (ES, MG, RJ, SP) de uma vez:
obitos_sudeste <- sus_data_import(
  region = "sudeste",
  year   = 2019,
  system = "SIM-DO",
  lang   = "pt"
)
```


**View the data**
```{r}
dplyr::glimpse(obitos_sudeste)
```


Or restrict to a single **municipality** by IBGE code (or by name in `city`).
The filter is resolved to the IBGE code **before** the download:

```{r import-muni}
# Apenas o município de São Paulo (código IBGE 3550308):
obitos_muni_sp <- sus_data_import(
  uf                = "SP",
  year              = 2014:2019,
  system            = "SIM-DO",
  municipality_code = "3550308",
  lang              = "pt"
)
```


**View the data**
```{r}
dplyr::glimpse(obitos_muni_sp)
```

## (Optional) Re-reading already exported data

After exporting the base (with `sus_data_export()`), you can read it back in
future sessions without downloading it again, using `sus_data_read()` — which
automatically detects the format:

```{r read, eval = FALSE}
obitos <- sus_data_read("output/sim_sp_2014_2019.parquet")
```

At the end of this stage we have `obitos`, `internacoes`, and `dengue` — the
**multi-system raw base** that will feed the next tutorials.

# Function features

## `sus_data_import()`

The single entry point for the import stage. It wraps
`microdatasus::fetch_datasus` (Saldanha, Bastos & Barcellos, 2019) with caching
and parallelism.

| Argument | Essential? | Description |
|-----------|------------|-----------|
| `uf` / `region` | yes (one of them) | State(s) (`"SP"`, `c("SP","RJ")`, `"all"`) **or** a predefined group (`"sudeste"`, `"amazonia_legal"`, …) |
| `year` | yes | 4-digit year(s) (e.g., `2014:2019`) |
| `system` | yes | System/subsystem (see systems table below) |
| `month` | monthly only | Month(s) 1–12; **mandatory** for SIH, CNES, and SIA; ignored for SIM, SINAN, and SINASC |
| `backend` | optional | `"arrow"` (lazy, default, very fast) or `"tibble"` (in memory) |
| `city` / `municipality_code` | optional | Restricts to municipalities (name or 6/7-digit IBGE code) |
| `use_cache` | recommended | `TRUE` reuses downloads; `cache_dir` sets the folder |
| `force_redownload` | optional | `TRUE` ignores the cache and re-downloads everything |
| `parallel` / `workers` | optional | Parallelizes state/year combinations; respects the user's `future::plan()` |
| `lang` / `verbose` | optional | Language (`"pt"`, `"en"`, `"es"`) and message level |

**Return:** a `tibble`/`data.frame` (or an Arrow object, if `backend = "arrow"`)
with the system's original variables, plus `source_system` and
`download_timestamp`, dates as `Date` objects, and UTF-8 text.

**Smart caching.** Calls are identified by a **SHA-256** key of the parameters
(state, year, month, system). Results are stored in `cache_dir`
(`~/.climasus4r_cache/data` by default) as compressed RDS and are reused
automatically. The cache expires after ~30 days for **dynamic** systems (CNES,
SIA, SIH) and ~365 days for **static** ones (SIM, SINAN, SINASC).

## `sus_data_read()`

Re-reads already exported files without going back to the source.
Auto-detects the format.

| Argument | Description |
|-----------|-----------|
| `path` | Path to a file, a vector of files, or a folder |
| `format` | `NULL` (auto-detects) or `"parquet"`, `"csv"`, `"dbc"`, `"rds"`, `"shapefile"`, … |
| `parallel` / `workers` | Parallel reading of many files |
| `read_metadata` | Recovers companion metadata as attributes |

**Return:** the table read (or an `sf` object, for spatial data); batch reads
are combined with `dplyr::bind_rows()`.

# Supported systems and subsystems

```{r tab-sistemas, eval=TRUE, echo=FALSE}
tbl_ou_nota("dados/import_sistemas.rds", .nota_gerar,
            caption = "Table 1. DATASUS systems accessible via sus_data_import().")
```

`region` shortcuts (download several states at once):

```{r tab-regioes, eval=TRUE, echo=FALSE}
tbl_ou_nota("dados/import_regioes.rds", .nota_gerar,
            caption = "Table 2. Examples of predefined groups accepted in region.")
```

Typical structure of SIM-DO right after import (underlying cause in
`CAUSABAS`, age coded in `IDADE`, municipality of residence in `CODMUNRES`):

```{r tab-estrutura, eval=TRUE, echo=FALSE}
tbl_ou_nota("dados/import_estrutura_sim.rds", .nota_gerar,
            caption = "Table 3. Sample of key columns from the imported SIM-DO.")
```

# Performance: caching and parallelism

Two features of `sus_data_import()` make importing reproducible and fast. The
**relative gain** (speedup) is the ratio between the time without the feature
and the time with it:

$$
\text{speedup} = \frac{t_{\text{sem recurso}}}{t_{\text{com recurso}}}
$$

Speedup values above 1 indicate acceleration. The figures below show **real
measurements** taken by the generation script.

```{r fig-cache, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/import_bench_cache.png",
            "Figura 1 indisponível: o benchmark de cache requer acesso ao DATASUS. Reexecute o script de geração com internet.")
```

**Figure 1.** Cache benchmark. The first call downloads the data from the
network; the second reads from the local cache, typically orders of magnitude
faster — without any change to the analysis code.

```{r fig-parallel, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/import_bench_parallel.png",
            "Figura 2 indisponível: o benchmark de paralelismo requer acesso ao DATASUS. Reexecute o script de geração com internet.")
```

**Figure 2.** Parallelism benchmark. Importing the six years with
`workers = 4` is faster than doing so in series (`workers = 1`). The gain
depends on the number of available cores; `sus_data_import()` warns if
`workers` exceeds the machine's cores.

# Illustrated results

```{r fig-registros, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/import_registros_por_sistema.png", .nota_gerar)
```

**Figure 3.** Number of records imported per system for São Paulo, 2014–2019
(SIH-RD was restricted to winter, June–August, to keep the example light). The
very different order of magnitude across systems helps in sizing memory and
download time.

```{r fig-serie, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/import_serie_obitos_sp.png", .nota_gerar)
```

**Figure 4.** Monthly series of deaths (all causes) from SIM-DO in São Paulo,
2014–2019. Winter seasonality is already visible in the raw base and will be
the starting point for the subsets (older adults, respiratory causes) in the
next tutorials.

# References

- Saldanha, R. de F., Bastos, R. R., & Barcellos, C. (2019). Microdatasus: pacote
  para download e pré-processamento de microdados do Departamento de Informática do
  SUS (DATASUS). *Cadernos de Saúde Pública*, 35(9), e00032419.
- Ministério da Saúde (Brasil). DATASUS — Departamento de Informática do SUS.
  Available at: <http://datasus.saude.gov.br>.

---

# Navigation

- [Back to tutorial hub](tutorials.html)
- **Next:** [Tutorial 2 — Encoding and Standardization](01-importacao.html)
