---
title: "Tutorial 2: Encoding and Standardization"
author: "climasus4r Team"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    code_download: true
    self_contained: true
    df_print: paged
---

```{css, echo=FALSE}
/* -- Variaveis de Cores Modernas -- */
:root {
  --primary-color: #1D9E75;
  --dark-green: #0f3d2e;
  --bg-light: #f8faf9;
  --text-main: #1a1a1a;
  --code-bg: #fdfdfd;
  --warning-orange: #ff9800;
  --info-blue: #2196f3;
}
html { scroll-behavior: smooth; }
body { font-family: 'Source Serif 4', Georgia, serif; font-size: 18px; color: var(--text-main); background-color: #fff; line-height: 1.6; }
h1, h2, h3, h4 { font-family: 'Inter', sans-serif; font-weight: 700; color: var(--dark-green); }
h1.title { font-size: 2.8em; border-bottom: 4px solid var(--primary-color); padding-bottom: 10px; margin-bottom: 1.5rem; }
h2 { margin-top: 2rem; border-left: 4px solid var(--primary-color); padding-left: 1rem; }
pre.r, pre.sourceCode { background-color: var(--code-bg); border: 1px solid #e1e1e1; border-left: 5px solid var(--primary-color) !important; border-radius: 8px; box-shadow: 0 2px 4px rgba(0,0,0,0.05); padding: 1rem; }
.callout { padding: 1.2em 1.5em; margin: 1.5em 0; border-radius: 8px; border-left: 6px solid; background-color: var(--bg-light); }
.callout-note { background-color: #f0f7ff; border-color: #007bff; }
.callout-tip { background-color: #e8f5e9; border-color: #1D9E75; }
.callout-warning { background-color: #fff3e0; border-color: #ff9800; }
.workflow-container { background: linear-gradient(135deg, #f8faf7 0%, #ffffff 100%); border-radius: 12px; padding: 1.5rem; margin: 1.5rem 0; border: 1px solid #dce8e1; overflow-x: auto; }
.workflow-steps { display: flex; flex-wrap: wrap; justify-content: center; gap: 0.5rem; font-family: 'Inter', monospace; font-size: 0.8rem; }
.workflow-step { background: var(--primary-color); color: white; padding: 0.5rem 1rem; border-radius: 20px; white-space: nowrap; }
.workflow-arrow { color: var(--primary-color); font-weight: bold; font-size: 1.2rem; }
.feature-grid { display: grid; grid-template-columns: repeat(auto-fit, minmax(280px, 1fr)); gap: 1.5rem; margin: 2rem 0; }
.feature-card { background: white; border-radius: 12px; padding: 1.5rem; border: 1px solid #e1e1e1; transition: transform 0.2s ease, box-shadow 0.2s ease; text-align: center; }
.feature-card:hover { transform: translateY(-4px); box-shadow: 0 8px 24px rgba(0,0,0,0.1); }
.feature-icon { font-size: 2rem; margin-bottom: 1rem; }
.feature-card h3 { margin-top: 0; color: var(--primary-color); font-size: 1.1rem; }
table { width: 100%; border-collapse: collapse; margin: 1.5rem 0; }
th, td { padding: 12px; text-align: left; border-bottom: 1px solid #e1e1e1; }
th { background-color: var(--bg-light); font-weight: 600; color: var(--dark-green); }
.github-feedback-btn { display: inline-flex; align-items: center; gap: 8px; background-color: #24292e; color: white; padding: 10px 20px; border-radius: 6px; text-decoration: none; font-family: sans-serif; transition: background-color 0.2s; margin-top: 1rem; }
.github-feedback-btn:hover { background-color: #2c3e50; color: white; text-decoration: none; }
@media (max-width: 768px) { body { font-size: 16px; } h1.title { font-size: 2rem; } .workflow-steps { flex-direction: column; align-items: center; } }
```


```{r setup, include=FALSE}
# ATENÇÃO — estratégia de pré-renderização.
# Este tutorial NÃO executa análise durante a renderização. Todas as figuras e
# tabelas foram geradas UMA vez pelo script
# scripts/tutorial_02_encoding_padronizacao.R e são apenas carregadas aqui
# (PNG via include_graphics; RDS via readRDS). Por isso, os chunks de ensino
# abaixo usam eval = FALSE: o código é mostrado, mas não é executado no knit.
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)

# Helpers para exibir artefatos de forma robusta: se o arquivo ainda não foi
# gerado (script de geração não executado), mostram uma nota em vez de quebrar
# o knit. Assim a vignette renderiza a qualquer momento.
fig_ou_nota <- function(path, nota) {
  if (file.exists(path)) knitr::include_graphics(path)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
tbl_ou_nota <- function(path, nota, caption = NULL) {
  if (file.exists(path)) knitr::kable(readRDS(path), caption = caption)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
.nota_gerar <- "Artefato indisponível — rode `Rscript scripts/tutorial_02_encoding_padronizacao.R` (a partir da raiz do pacote) para gerá-lo."
```

# Objective

By the end of this tutorial you will be able to **get the raw DATASUS dataset
ready for analysis**, running the two functions of the second pipeline stage:

1. **`sus_data_clean_encoding()`** — audits text columns and fixes
   **character encoding** issues (Latin1 text read as UTF-8, which turns
   "São Paulo" into unreadable sequences);
2. **`sus_data_standardize()`** — **translates column names** into readable
   labels and **decodes categorical values** (for example, `SEXO = 1` becomes
   `"Masculino"`, `RACACOR = 4` becomes `"Parda"`), and converts date columns
   to the `Date` type. All of this in a **multilingual** way
   (`"pt"`, `"en"`, `"es"`).

This is the **second stage** of the pipeline (`import → clean → stand → …`). It
transforms a table with cryptic names and numeric codes into an auditable
table that the following tutorials (filtering, variable creation,
aggregation) can handle unambiguously.

# Scientific context

DATASUS microdata are distributed in `.dbc` files inherited from legacy
systems, which store text in **Latin1 (ISO-8859-1)**. When those bytes are
read as **UTF-8**, accents and cedillas turn into "mojibake" — corrupted
characters that break name-based joins, groupings, and reports. The
`microdatasus` package (Saldanha, Bastos & Barcellos, 2019) already handles
much of the preprocessing, but `sus_data_clean_encoding()` acts as a
**final auditor**, ensuring every text column is valid UTF-8 before analysis.

Beyond encoding, fields from SIM, SIH, SINAN, and other systems arrive with
**abbreviated names** (`CAUSABAS`, `CODMUNRES`, `DTOBITO`) and **categories
coded as numbers** (Brazil, Ministry of Health). Working directly with these
codes is fragile and error-prone. Standardizing to a **controlled, readable
vocabulary** — and being able to produce it in three languages — is what
makes the analysis reproducible and communicable to readers from different
countries (FAIR principle of interoperability; Wilkinson et al., 2016).

> **Guiding case.** We are investigating **temperature and health outcomes in
> older adults (60+) in the São Paulo Metropolitan Region, 2014–2019**. In
> Tutorial 1 we assembled the raw multi-system dataset
> (`caso_base_importado.rds`). Here we take the **deaths (SIM-DO)** dataset,
> fix its encoding, and standardize names/categories (sex, race/color, ICD,
> dates). The result is saved as `caso_padronizado.rds`, the input for
> Tutorial 3 (Filtering).

# Data used

| Object | Source | Role in the case |
|--------|--------|---------------|
| `caso_base_importado$obitos` | SIM-DO (Tutorial 1) | Raw dataset to be cleaned and standardized |
| PT/EN/ES translation dictionary | `R/utils-i18n.R` | Maps codes → labels per system |

Scope: state of **São Paulo**, 2014–2019. If the `.rds` from Tutorial 1 does
not yet exist, the generation script falls back to a **minimal example**
(text deliberately in Latin1) so this stage remains illustrable offline.

# Complete pipeline

The blocks below reproduce what the script
`scripts/tutorial_02_encoding_padronizacao.R` runs. Start by loading the
package and the dataset from the previous tutorial:

```{r carregar}
library(climasus4r)

caso_base   <- readRDS("vignettes-pt/dados/caso_base_importado.rds")
obitos_bruto <- caso_base$obitos
```

## Step 1 — Fix the encoding

`sus_data_clean_encoding()` scans all **text** columns, detects the ones
containing invalid UTF-8 bytes, and converts them from **Latin1 to UTF-8**.
Numeric and date columns are ignored.

```{r clean-encoding}
obitos_utf8 <- sus_data_clean_encoding(
  df      = obitos_bruto,
  backend = "tibble",   # "arrow" (lazy, padrão) ou "tibble" (em memória)
  lang    = "pt",        # idioma das mensagens: "pt", "en", "es"
  verbose = TRUE         # imprime quais colunas foram corrigidas
)
```

## Step 2 — Standardize names and categories

`sus_data_standardize()` **automatically detects the system** (SIM, SIH,
SINAN, …), translates the column names to the controlled vocabulary, and
decodes categorical values in the chosen language. It also converts date
columns (prefix `data_` in Portuguese) to the `Date` type.

```{r standardize}
obitos_padr <- sus_data_standardize(
  df                 = obitos_utf8,
  lang               = "pt",   # "pt" / "en" / "es"
  translate_columns  = TRUE,   # CAUSABAS -> causa_basica, SEXO -> sexo, ...
  standardize_values = TRUE,   # 1 -> Masculino; 4 -> Parda; ...
  keep_original      = FALSE,  # TRUE mantém as colunas originais lado a lado
  backend            = "tibble",
  verbose            = TRUE
)
```

## Chaining the two functions

Since every function in this stage receives and returns a `climasus_df`, the
two chain naturally with the pipe — exactly the recommended way to use them:

```{r pipe-completo}
obitos_padr <- obitos_bruto |>
  sus_data_clean_encoding(backend = "tibble", lang = "pt") |>
  sus_data_standardize(lang = "pt", backend = "tibble")
```

In the end, `obitos_padr` is the **clean and standardized** dataset — with
`stage = "stand"` in its metadata — which we save for the next tutorial.

# Function characteristics

## `sus_data_clean_encoding()`

Final encoding auditor. Detects text columns with invalid UTF-8 bytes and
converts Latin1 → UTF-8.

| Argument | Essential? | Description |
|-----------|------------|-----------|
| `df` | yes | `data.frame`/`tibble`/`climasus_df` (or a collectible Arrow object) |
| `backend` | optional | `"arrow"` (lazy, default) or `"tibble"` (in memory); with `"arrow"`, falls back to tibble if lazy evaluation fails |
| `lang` | optional | Message language: `"pt"` (default), `"en"`, `"es"` |
| `verbose` | optional | `TRUE` (default) prints the columns checked and corrected |

**How it works.** For each text column, it tests `stringi::stri_enc_isutf8()`
on the non-null values; if **any** value is invalid UTF-8, the whole column
is considered "sick" and converted with `stringi::stri_conv(x, "latin1",
"utf8")`. With the `"arrow"` backend, a sample of up to 10,000 rows is used
to detect which columns need correction.

**Return:** a `climasus_df` with text columns in valid UTF-8,
`stage = "clean"`, and a history entry recording the correction.

## `sus_data_standardize()`

Translates column names and decodes categorical values, with automatic
detection of the DATASUS system.

| Argument | Essential? | Description |
|-----------|------------|-----------|
| `df` | yes | Output of `sus_data_import()` / `sus_data_clean_encoding()` |
| `lang` | optional | Language of **names and labels**: `"pt"` (default), `"en"`, `"es"` |
| `translate_columns` | optional | `TRUE` (default) translates column names |
| `standardize_values` | optional | `TRUE` (default) decodes categorical values |
| `keep_original` | optional | `TRUE` keeps the original columns alongside the translated ones |
| `backend` | optional | `"arrow"` (lazy, default) or `"tibble"` (in memory) |
| `verbose` | optional | `TRUE` (default) prints the standardization report |

**Internal steps.** (1) detects the system (`detect_health_system()` or
metadata); (2) translates column names using the system/language dictionary;
(3) decodes categorical values (e.g., `SEXO`, `RACACOR`, `LOCOCOR`) into
readable factors; (4) converts date columns to `Date` using explicit formats
(`%Y-%m-%d`, `%Y%m%d`, `%d%m%Y`), portably across operating systems.

**Return:** a `climasus_df` with `stage = "stand"`, `system` filled in, and
unique column names (`make.unique()`).

# Encoding detection: the heuristic

The decision rule of `sus_data_clean_encoding()` for flagging a column $c$ as
"to be corrected" can be written as:

$$
\text{corrigir}(c) \;=\;
\Bigl[\; \exists\, x \in c \;:\; \neg\,\mathrm{isUTF8}(x) \;\Bigr]
$$

that is, just **one** value with invalid UTF-8 bytes is enough for the whole
column to be converted from Latin1 to UTF-8. As a diagnostic, the generation
script measures the **fraction of invalid strings** per column, before and
after:

$$
f_{\text{inv}}(c) \;=\; \frac{1}{n_c}\sum_{i=1}^{n_c}
\mathbf{1}\!\left[\neg\,\mathrm{isUTF8}(x_i)\right]
$$

where $n_c$ is the number of non-null values in the column. After the
correction, $f_{\text{inv}}(c) = 0$ for every text column.

# Supported systems and dictionaries

`sus_data_standardize()` loads a different dictionary depending on the
detected system. Each system has dictionaries in all three languages
(`pt`, `en`, `es`):

- **SIM** — `get_translation_dict_{pt,en,es}()`
- **SIH** — `get_translation_dict_{pt,en,es}_sih()`
- **SINAN** — `get_translation_dict_{pt,en,es}_sinan()`
- **CNES** — `get_translation_dict_{pt,en,es}_cnes()`
- **SINASC** — `get_translation_dict_{pt,en,es}_sinasc()`
- **SIA** (and subtypes AM, AQ, AR, AD, PS) — `get_translation_dict_*_sia*()`

The table below shows a sample of the **PT dictionary for SIM** used to
decode the categorical values of the guiding case (sex, race/color, and place
of occurrence of death):

```{r tab-dicionario, eval=TRUE, echo=FALSE}
tbl_ou_nota("dados/enc_dicionario.rds", .nota_gerar,
            caption = "Table 1. Sample of the SIM translation dictionary (pt): numeric code → readable label.")
```

# Illustrated results

```{r fig-encoding, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/enc_pipeline_encoding.png", .nota_gerar)
```

**Figure 1.** Encoding diagnostic before and after `sus_data_clean_encoding()`.
Each bar is the percentage of strings with invalid UTF-8 bytes in a text
column. After the Latin1 → UTF-8 conversion, the fraction of invalid strings
drops to zero — without any change to the readable content.

```{r fig-sexo, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/enc_sexo.png", .nota_gerar)
```

**Figure 2.** Distribution of deaths by **sex** after standardization. The
numeric SIM codes (`1`, `2`) were decoded into readable labels (`Masculino`,
`Feminino`) by `sus_data_standardize()`, ready for use in charts and tables.

```{r fig-raca, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/enc_raca.png", .nota_gerar)
```

**Figure 3.** Distribution of deaths by **race/color** after standardization.
The codes (`1`–`5`, `9`) were translated into `Branca`, `Preta`, `Amarela`,
`Parda`, `Indígena`, and `Ignorado`. This decoding is a prerequisite for
health equity analyses in the following stages.

```{r tab-resumo, eval=TRUE, echo=FALSE}
tbl_ou_nota("dados/enc_resumo.rds", .nota_gerar,
            caption = "Table 2. Stage summary: dimensions, encoding, detected system, and pipeline stage.")
```

**Table 2.** Before/after summary of this stage. The metadata `stage`
advances to `stand`, signaling to the following tutorials that the dataset
is already standardized.

# References

- Brasil. Ministério da Saúde. *Manual de Instruções para o preenchimento da
  Declaração de Óbito*. Brasília: Ministério da Saúde. (Defines the codes of
  the SIM variables, such as `SEXO`, `RACACOR`, and `LOCOCOR`.)
- Saldanha, R. de F., Bastos, R. R., & Barcellos, C. (2019). Microdatasus: pacote
  para download e pré-processamento de microdados do Departamento de Informática do
  SUS (DATASUS). *Cadernos de Saúde Pública*, 35(9), e00032419.
- Wilkinson, M. D., et al. (2016). The FAIR Guiding Principles for scientific data
  management and stewardship. *Scientific Data*, 3, 160018.

---

# Navigation

- [Back to tutorial hub](tutorials.html)
- **Previous:** [Tutorial 1 — DATASUS Data Import](01-importacao.html)
- **Next:** [Tutorial 3 — Filtering (ICD-10 and demographics)](03-filtragem.html)

```
