---
title: "Tutorial 3: Filtering by Disease (ICD-10) and Demographics"
author: "climasus4r Team"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    code_download: true
    self_contained: true
    df_print: paged
---

```{css, echo=FALSE}
/* -- Variaveis de Cores Modernas -- */
:root {
  --primary-color: #1D9E75;
  --dark-green: #0f3d2e;
  --bg-light: #f8faf9;
  --text-main: #1a1a1a;
  --code-bg: #fdfdfd;
  --warning-orange: #ff9800;
  --info-blue: #2196f3;
}
html { scroll-behavior: smooth; }
body { font-family: 'Source Serif 4', Georgia, serif; font-size: 18px; color: var(--text-main); background-color: #fff; line-height: 1.6; }
h1, h2, h3, h4 { font-family: 'Inter', sans-serif; font-weight: 700; color: var(--dark-green); }
h1.title { font-size: 2.8em; border-bottom: 4px solid var(--primary-color); padding-bottom: 10px; margin-bottom: 1.5rem; }
h2 { margin-top: 2rem; border-left: 4px solid var(--primary-color); padding-left: 1rem; }
pre.r, pre.sourceCode { background-color: var(--code-bg); border: 1px solid #e1e1e1; border-left: 5px solid var(--primary-color) !important; border-radius: 8px; box-shadow: 0 2px 4px rgba(0,0,0,0.05); padding: 1rem; }
.callout { padding: 1.2em 1.5em; margin: 1.5em 0; border-radius: 8px; border-left: 6px solid; background-color: var(--bg-light); }
.callout-note { background-color: #f0f7ff; border-color: #007bff; }
.callout-tip { background-color: #e8f5e9; border-color: #1D9E75; }
.callout-warning { background-color: #fff3e0; border-color: #ff9800; }
.workflow-container { background: linear-gradient(135deg, #f8faf7 0%, #ffffff 100%); border-radius: 12px; padding: 1.5rem; margin: 1.5rem 0; border: 1px solid #dce8e1; overflow-x: auto; }
.workflow-steps { display: flex; flex-wrap: wrap; justify-content: center; gap: 0.5rem; font-family: 'Inter', monospace; font-size: 0.8rem; }
.workflow-step { background: var(--primary-color); color: white; padding: 0.5rem 1rem; border-radius: 20px; white-space: nowrap; }
.workflow-arrow { color: var(--primary-color); font-weight: bold; font-size: 1.2rem; }
.feature-grid { display: grid; grid-template-columns: repeat(auto-fit, minmax(280px, 1fr)); gap: 1.5rem; margin: 2rem 0; }
.feature-card { background: white; border-radius: 12px; padding: 1.5rem; border: 1px solid #e1e1e1; transition: transform 0.2s ease, box-shadow 0.2s ease; text-align: center; }
.feature-card:hover { transform: translateY(-4px); box-shadow: 0 8px 24px rgba(0,0,0,0.1); }
.feature-icon { font-size: 2rem; margin-bottom: 1rem; }
.feature-card h3 { margin-top: 0; color: var(--primary-color); font-size: 1.1rem; }
table { width: 100%; border-collapse: collapse; margin: 1.5rem 0; }
th, td { padding: 12px; text-align: left; border-bottom: 1px solid #e1e1e1; }
th { background-color: var(--bg-light); font-weight: 600; color: var(--dark-green); }
.github-feedback-btn { display: inline-flex; align-items: center; gap: 8px; background-color: #24292e; color: white; padding: 10px 20px; border-radius: 6px; text-decoration: none; font-family: sans-serif; transition: background-color 0.2s; margin-top: 1rem; }
.github-feedback-btn:hover { background-color: #2c3e50; color: white; text-decoration: none; }
@media (max-width: 768px) { body { font-size: 16px; } h1.title { font-size: 2rem; } .workflow-steps { flex-direction: column; align-items: center; } }
```


```{r setup, include=FALSE}
# ATENÇÃO — estratégia de pré-renderização.
# Este tutorial NÃO executa análise durante a renderização. Todas as figuras e
# tabelas foram geradas UMA vez pelo script scripts/tutorial_03_filtragem.R e
# são apenas carregadas aqui (PNG via include_graphics; RDS via readRDS).
# Por isso, os chunks de ensino abaixo usam eval = FALSE: o código é mostrado,
# mas não é executado no momento do knit.
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)

# Helpers para exibir artefatos de forma robusta: se o arquivo ainda não foi
# gerado (script de geração não executado), mostram uma nota em vez de quebrar
# o knit. Assim a vignette renderiza a qualquer momento.
fig_ou_nota <- function(path, nota) {
  if (file.exists(path)) knitr::include_graphics(path)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
tbl_ou_nota <- function(path, nota, caption = NULL) {
  if (file.exists(path)) knitr::kable(readRDS(path), caption = caption)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
.nota_gerar <- "Artefato indisponível — rode `Rscript scripts/tutorial_03_filtragem.R` (a partir da raiz do pacote) para gerá-lo."
```

# Objective

By the end of this tutorial you will be able to **subset already-standardized
health microdata** along two complementary axes:

1. **By disease (ICD-10)** — with `sus_data_filter_cid()`, selecting specific
   codes, ranges, entire chapters, or one of the package's **50+ predefined
   disease groups**;
2. **By demographic characteristics** — with `sus_data_filter_demographics()`,
   restricting by sex, race/color, **age group**, education, region, or
   municipality.

You will also get to know `sus_data_cid_select()`, an **interactive explorer**
that helps you discover disease-group names without needing to memorize
ICD-10 codes.

Filtering is the **third stage** of the pipeline (`import → clean → stand →
**filter_cid → filter_demo** → derive → aggregate → …`). It requires that the
data have already been standardized (stage `stand`); functions therefore abort
with a guiding message if they receive data at an earlier stage.

# Scientific context

**Climate and health** studies rarely analyze all deaths or hospitalizations
at once: interest centers on **climate-sensitive outcomes** in **vulnerable
subpopulations**. The literature is consistent in showing that thermal
extremes disproportionately affect **respiratory** causes (ICD-10 chapter J,
J00–J99) and **cardiovascular** causes (chapter I, I00–I99), especially among
**older adults** (Gasparrini et al., 2015; Bhaskaran et al., 2013).

The **International Classification of Diseases, 10th revision (ICD-10)**
organizes causes into chapters identified by an initial letter and into
three- to four-character codes (WHO, 2016). `climasus4r` translates this
structure into a flexible filtering syntax — single code, range
(`"J00-J99"`), chapter (`"J"`), or named group (`"respiratory"`) — and
automatically detects the appropriate cause-of-death column for each DATASUS
system (`CAUSABAS`/`causa_basica` in SIM, `DIAG_PRINC`/`diagnostico_principal`
in SIH).

Subsetting **before** aggregating is also good reproducibility practice: the
`climasus_df` object records in its history exactly which codes and
demographic criteria were applied, making the study auditable.

> **Driving case study.** We investigate **temperature and health outcomes in
> older adults (60+) in the São Paulo Metropolitan Region, 2014–2019**. In
> this tutorial, starting from the standardized base from Tutorial 2, we
> subset the **respiratory causes (J00–J99)** and **cardiovascular causes
> (I00–I99)** and restrict to the **older population (60 years or more)**
> residing in the RMSP (São Paulo Metropolitan Region). The filtered object
> feeds into Tutorial 4 (Variable Creation and Aggregation).

# Data used

Input: `vignettes-pt/dados/caso_padronizado.rds` — the **standardized** SIM-DO
base produced in Tutorial 2 (canonical columns such as `causa_basica`,
`age_years`, `sex`, `residence_municipality_code`). The script loads this
`.rds` file with a `file.exists()` guard; if it does not yet exist (Tutorial 2
not run), it falls back to a **minimal synthetic example** solely to produce
the didactic artifacts.

| Subset | Criterion | Function |
|---------|----------|--------|
| Respiratory causes | ICD-10 J00–J99 (`disease_group = "respiratory"`) | `sus_data_filter_cid()` |
| Cardiovascular causes | ICD-10 I00–I99 (`icd_codes = "I00-I99"`) | `sus_data_filter_cid()` |
| Older adults | `age_range = c(60, Inf)` | `sus_data_filter_demographics()` |
| RMSP | `municipality_code = <IBGE codes>` | `sus_data_filter_demographics()` |

# Full pipeline

The blocks below reproduce what the `scripts/tutorial_03_filtragem.R` script
executes. Start by loading the package and the standardized base from the
previous tutorial:

```{r carregar}
library(climasus4r)

# Saída do Tutorial 2 (Encoding & Padronização):
caso_padronizado <- readRDS("vignettes-pt/dados/caso_padronizado.rds")
obitos_std <- caso_padronizado$obitos   # SIM-DO padronizado (estágio "stand")
```

## Explore disease groups

`sus_data_cid_select()` helps you discover **which groups exist** and which
are **climate-sensitive**, without opening the source code. In interactive
mode (`output = "browser"`) it opens an HTML interface; in `"console"` mode it
invisibly returns a `data.frame` with the groups:

```{r cid-select}
# Interface interativa no navegador (padrão):
sus_data_cid_select(lang = "pt")

# Apenas os grupos sensíveis ao clima:
sus_data_cid_select(lang = "pt", filter_climate = TRUE)

# Sem abrir o navegador — devolve o data.frame de grupos:
grupos <- sus_data_cid_select(lang = "pt", output = "console", verbose = FALSE)
```

## Filter by ICD-10

There are two equivalent ways to subset the respiratory and cardiovascular
causes: by **named group** (more readable) or by **code range** (more
explicit). The cause-of-death column (`causa_basica`, in SIM-DO) is
**automatically detected** from the `climasus_df` metadata.

```{r filter-cid}
# (a) Respiratório (J00-J99) por grupo de doença:
resp <- sus_data_filter_cid(
  df            = obitos_std,
  disease_group = "respiratory",   # equivale a icd_codes = "J00-J99"
  match_type    = "starts_with",   # "J18" casa "J18", "J18.0", "J18.9"
  backend       = "tibble",        # "arrow" (lazy) ou "tibble" (em memória)
  lang          = "pt"
)

# (b) Cardiovascular (I00-I99) por intervalo de códigos:
cardio <- sus_data_filter_cid(
  df         = obitos_std,
  icd_codes  = "I00-I99",          # equivale a disease_group = "cardiovascular"
  match_type = "starts_with",
  backend    = "tibble",
  lang       = "pt"
)
```

## Filter by demographics (older adults 60+ in the RMSP)

We chain the demographic subset on top of the respiratory subset. The
argument `age_range = c(60, Inf)` requires the `age_years` column; the
spatial subset uses **IBGE codes** for the municipalities of the São Paulo
Metropolitan Region:

```{r filter-demo}
# Municípios da RMSP (códigos IBGE de 7 dígitos):
RMSP <- c("3550308", "3548708", "3509502", "3548500",
          "3547809", "3530607", "3534401", "3505708")

resp_idosos <- sus_data_filter_demographics(
  df                = resp,
  age_range         = c(60, Inf),   # idosos: 60 anos ou mais
  municipality_code = RMSP,         # recorte espacial: RMSP
  drop_ignored      = FALSE,        # TRUE removeria sexo/raça "Ignorado"/NA
  backend           = "tibble",
  use_cache         = TRUE,
  lang              = "pt"
)

cardio_idosos <- sus_data_filter_demographics(
  df                = cardio,
  age_range         = c(60, Inf),
  municipality_code = RMSP,
  backend           = "tibble",
  lang              = "pt"
)
```

You can also filter by **city name** (tolerant of accents and typos), by
**sex/race/education**, or by predefined **region**:

```{r filter-demo-extra}
# Por nome de cidade (em vez de código) e por sexo:
mulheres_sp <- sus_data_filter_demographics(
  df   = resp,
  sex  = "Feminino",
  city = "Sao Paulo",   # aceita acento; faz correspondência aproximada
  lang = "pt"
)

# Por macrorregião (atalho que resolve para um conjunto de UFs):
sudeste <- sus_data_filter_demographics(resp, region = "sudeste", lang = "pt")
```

At the end of this step we have `resp_idosos` and `cardio_idosos` — the
**subsets of interest** that will be saved and will feed into Tutorial 4.

# Function features

## `sus_data_filter_cid()`

Filters by ICD-10. Requires `icd_codes` **or** `disease_group` (if both are
supplied, `disease_group` takes precedence). The cause-of-death column is
automatically detected by the system (`SIM → causa_basica`,
`SIH → diagnostico_principal`).

| Argument | Essential? | Description |
|-----------|------------|-----------|
| `df` | yes | `climasus_df` at stage `stand` or later |
| `icd_codes` | one of the two | Code(s)/range(s): `"J18.9"`, `c("I21","I22")`, `"J00-J99"` |
| `disease_group` | one of the two | Named group: `"respiratory"`, `"cardiovascular"`, `"dengue"`, … |
| `icd_column` | optional | Cause column; `NULL` = autodetection |
| `match_type` | optional | `"starts_with"` (default), `"exact"`, `"range"` |
| `backend` | optional | `"arrow"` (lazy, default) or `"tibble"` (in memory) |
| `lang` / `verbose` | optional | Language (`"pt"`/`"en"`/`"es"`) and messages |

**Returns:** a filtered `climasus_df` (stage `filter_cid`), preserving all
original columns and recording in its history the codes/groups applied.

## `sus_data_filter_demographics()`

Filters by demographic and geographic characteristics. All criteria are
optional; if none are supplied, the data are returned unchanged. Combinations
are applied in sequence (intersection across criteria); `city` and
`municipality_code` are combined by **union**.

| Argument | Description |
|-----------|-----------|
| `df` | `climasus_df` at stage `stand` or later |
| `sex` | `"Masculino"`/`"Feminino"` (accepts pt/en/es) |
| `race` | IBGE categories: `"Branca"`, `"Preta"`, `"Parda"`, … |
| `age_range` | Vector `c(min, max)`; use `Inf` for no upper limit. Requires `age_years` |
| `education` | Education levels |
| `region` | Predefined group of states/UFs (`"sudeste"`, `"semi_arido"`, …) |
| `city` | Municipality name(s) (tolerant of accents/typos) |
| `municipality_code` | 6- or 7-digit IBGE code(s) |
| `drop_ignored` | `TRUE` removes `NA`/"Ignored" codes from demographic variables |
| `backend`, `use_cache`, `cache_dir` | Backend and spatial cache (relevant with `city`) |
| `lang` / `verbose` | Language and messages |

**Returns:** a filtered `climasus_df` (stage `filter_demo`). Internally,
municipality codes are expanded to cover **both** formats (DATASUS's 6 digits
and IBGE's 7 digits), so the filter works regardless of the format stored.

## `sus_data_cid_select()`

Disease-group explorer. **Does not alter the data** — it only helps discover
group names to use in `sus_data_filter_cid()`.

| Argument | Description |
|-----------|-----------|
| `lang` | `"pt"` (default), `"en"`, `"es"` |
| `output` | `"browser"` (interactive HTML, default) or `"console"` (summary in R) |
| `filter_climate` | `TRUE` shows only climate-sensitive groups |
| `verbose` | Informational messages |

**Returns:** invisibly, a `data.frame` with `name`, `icd_codes`,
`climate_factors`, `description`, `category`, and `climate_sensitive`.

# ICD-10 code syntax

The `icd_codes` syntax is expanded internally by `process_icd_codes()`. A
**range** `"J00-J99"` is expanded into the sequence of prefixes
$J00, J01, \dots, J99$. Formally, for a range of the same letter $L$ between
numbers $a$ and $b$:

$$
\mathrm{expand}(L\,a\text{–}L\,b) = \{\, L\,n : n \in \{a, a+1, \dots, b\} \,\}
$$

The `match_type` parameter defines how each code is compared against the
cause column:

- **`"exact"`** — exact equality (`"I10"` matches only `"I10"`);
- **`"starts_with"`** (default) — prefix (`"I10"` matches `"I10"`, `"I10.0"`,
  `"I10.9"`), which reflects the ICD-10 hierarchy;
- **`"range"`** — matching by any prefix of the expanded range.

The fraction of records retained at each filter is simply

$$
\text{retido (\%)} = 100 \times \frac{n_{\text{filtrado}}}{n_{\text{original}}}\,.
$$

# Illustrated results

```{r tab-grupos, eval=TRUE, echo=FALSE}
tbl_ou_nota("dados/filt_grupos_doenca.rds", .nota_gerar,
            caption = "Table 1. ICD-10 disease groups of interest in the driving case study (output of sus_data_cid_select).")
```

```{r tab-resumo, eval=TRUE, echo=FALSE}
tbl_ou_nota("dados/filt_resumo.rds", .nota_gerar,
            caption = "Table 2. Filtering funnel: records retained at each step.")
```

```{r fig-funil, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/filt_funil_cid.png", .nota_gerar)
```

**Figure 1.** Filtering funnel. Each bar shows how many records remain after
the standardized base, the ICD-10 subset (respiratory, J00–J99), and the
demographic subset (older adults 60+ residing in the RMSP). The narrowing
illustrates how the analysis object is progressively restricted to the
study's scope.

```{r fig-comp, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/filt_resp_vs_cardio.png", .nota_gerar)
```

**Figure 2.** Deaths in older adults (60+) in the RMSP by major cause group,
after the combined filtering (ICD-10 + demographics). The comparison between
the respiratory (J) and cardiovascular (I) chapters is the starting point for
the climate-sensitivity analysis in the next tutorials.

```{r fig-piramide, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/filt_piramide_idade.png", .nota_gerar)
```

**Figure 3.** Age distribution of respiratory deaths. The filter
`age_range = c(60, Inf)` retains only the 60–79 and 80+ age groups, making
explicit the effect of the demographic subset on the population analyzed.

```{r fig-serie, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/filt_serie_idosos.png", .nota_gerar)
```

**Figure 4.** Monthly series of respiratory deaths in older adults (60+) in
the RMSP, 2014–2019, after filtering. The winter seasonality — expected for
respiratory causes — already appears in the subsetted object and will be
compared against temperature in the climate and modeling stages.

# References

- Bhaskaran, K., Gasparrini, A., Hajat, S., Smeeth, L., & Armstrong, B. (2013).
  Time series regression studies in environmental epidemiology.
  *International Journal of Epidemiology*, 42(4), 1187–1195.
- Gasparrini, A., Guo, Y., Hashizume, M., et al. (2015). Mortality risk
  attributable to high and low ambient temperature: a multicountry observational
  study. *The Lancet*, 386(9991), 369–375.
- Organização Mundial da Saúde. (2016). *CID-10: Classificação Estatística
  Internacional de Doenças e Problemas Relacionados à Saúde*, 10ª revisão.
- Ministério da Saúde (Brasil). DATASUS — CID-10. Available at:
  <http://datasus.saude.gov.br/cid10>.

---

# Navigation

- [Back to tutorial hub](tutorials.html)
- **Previous:** [Tutorial 2 — Encoding and Standardization](02-encoding-padronizacao.html)
- **Next:** [Tutorial 4 — Variable Creation and Aggregation](04-variaveis-agregacao.html)
