---
title: "Tutorial 5: Spatial Join"
author: "climasus4r Team"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    code_download: true
    self_contained: true
    df_print: paged
---

```{css, echo=FALSE}
/* -- Variaveis de Cores Modernas -- */
:root {
  --primary-color: #1D9E75;
  --dark-green: #0f3d2e;
  --bg-light: #f8faf9;
  --text-main: #1a1a1a;
  --code-bg: #fdfdfd;
  --warning-orange: #ff9800;
  --info-blue: #2196f3;
}
html { scroll-behavior: smooth; }
body { font-family: 'Source Serif 4', Georgia, serif; font-size: 18px; color: var(--text-main); background-color: #fff; line-height: 1.6; }
h1, h2, h3, h4 { font-family: 'Inter', sans-serif; font-weight: 700; color: var(--dark-green); }
h1.title { font-size: 2.8em; border-bottom: 4px solid var(--primary-color); padding-bottom: 10px; margin-bottom: 1.5rem; }
h2 { margin-top: 2rem; border-left: 4px solid var(--primary-color); padding-left: 1rem; }
pre.r, pre.sourceCode { background-color: var(--code-bg); border: 1px solid #e1e1e1; border-left: 5px solid var(--primary-color) !important; border-radius: 8px; box-shadow: 0 2px 4px rgba(0,0,0,0.05); padding: 1rem; }
.callout { padding: 1.2em 1.5em; margin: 1.5em 0; border-radius: 8px; border-left: 6px solid; background-color: var(--bg-light); }
.callout-note { background-color: #f0f7ff; border-color: #007bff; }
.callout-tip { background-color: #e8f5e9; border-color: #1D9E75; }
.callout-warning { background-color: #fff3e0; border-color: #ff9800; }
.workflow-container { background: linear-gradient(135deg, #f8faf7 0%, #ffffff 100%); border-radius: 12px; padding: 1.5rem; margin: 1.5rem 0; border: 1px solid #dce8e1; overflow-x: auto; }
.workflow-steps { display: flex; flex-wrap: wrap; justify-content: center; gap: 0.5rem; font-family: 'Inter', monospace; font-size: 0.8rem; }
.workflow-step { background: var(--primary-color); color: white; padding: 0.5rem 1rem; border-radius: 20px; white-space: nowrap; }
.workflow-arrow { color: var(--primary-color); font-weight: bold; font-size: 1.2rem; }
.feature-grid { display: grid; grid-template-columns: repeat(auto-fit, minmax(280px, 1fr)); gap: 1.5rem; margin: 2rem 0; }
.feature-card { background: white; border-radius: 12px; padding: 1.5rem; border: 1px solid #e1e1e1; transition: transform 0.2s ease, box-shadow 0.2s ease; text-align: center; }
.feature-card:hover { transform: translateY(-4px); box-shadow: 0 8px 24px rgba(0,0,0,0.1); }
.feature-icon { font-size: 2rem; margin-bottom: 1rem; }
.feature-card h3 { margin-top: 0; color: var(--primary-color); font-size: 1.1rem; }
table { width: 100%; border-collapse: collapse; margin: 1.5rem 0; }
th, td { padding: 12px; text-align: left; border-bottom: 1px solid #e1e1e1; }
th { background-color: var(--bg-light); font-weight: 600; color: var(--dark-green); }
.github-feedback-btn { display: inline-flex; align-items: center; gap: 8px; background-color: #24292e; color: white; padding: 10px 20px; border-radius: 6px; text-decoration: none; font-family: sans-serif; transition: background-color 0.2s; margin-top: 1rem; }
.github-feedback-btn:hover { background-color: #2c3e50; color: white; text-decoration: none; }
@media (max-width: 768px) { body { font-size: 16px; } h1.title { font-size: 2rem; } .workflow-steps { flex-direction: column; align-items: center; } }
```


```{r setup, include=FALSE}
# ATENÇÃO — estratégia de pré-renderização.
# Este tutorial NÃO executa downloads de geometrias durante a renderização.
# Todas as figuras e tabelas foram geradas UMA vez pelo script
# scripts/tutorial_05_juncao_espacial.R e são apenas carregadas aqui
# (PNG via include_graphics; RDS via readRDS). Por isso, os chunks de ensino
# abaixo usam eval = FALSE: o código é mostrado, mas não é executado no knit.
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)

# Helpers para exibir artefatos de forma robusta: se o arquivo ainda não foi
# gerado (script de geração não executado), mostram uma nota em vez de quebrar
# o knit. Assim a vignette renderiza a qualquer momento.
fig_ou_nota <- function(path, nota) {
  if (file.exists(path)) knitr::include_graphics(path)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
tbl_ou_nota <- function(path, nota, caption = NULL) {
  if (file.exists(path)) knitr::kable(readRDS(path), caption = caption)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
.nota_gerar <- "Artefato indisponível — rode `Rscript scripts/tutorial_05_juncao_espacial.R` (a partir da raiz do pacote) para gerá-lo."
```

# Objective

By the end of this tutorial you will be able to **assign a geometry to each
health record** — linking the time series built in the previous stage to
**Brazil's official geographic units** (municipalities, biomes, health
regions, metropolitan regions, urban areas, indigenous lands, postal codes)
with `sus_spatial_join()`, and to **integrate gridded environmental data**
into the health pipeline with `sus_grid_join()`.

The spatial join is the **fifth stage** of the `climasus4r` pipeline. It
turns an "ordinary" table (health × date × municipal code) into an object of
class **`sf`** with explicit geometry — a requirement for mapping,
neighborhood computations and — critical to our driving case — merging with
socioeconomic indicators (Tutorial 6) and gridded climate data (Tutorial 8).

# Scientific context

Climate–health studies are, by nature, **spatial**: the exposure
(temperature, pollution) and the outcome (deaths, hospitalizations) occur in
concrete places. Linking a health record to its geographic unit allows
mapping inequalities, controlling for contextual confounders, and defining
the **unit of analysis**. This choice is not neutral: the **Modifiable
Areal Unit Problem** (MAUP; Openshaw, 1984) shows that statistical results
can change depending on the scale (municipality vs. census tract) and the
partitioning (health regions vs. metropolitan regions) adopted.

In Brazil, official geometries come from **IBGE** and are accessed through
the `geobr` package (Pereira et al., 2019). `climasus4r` wraps this access in
`sus_spatial_join()`, adding geometry **caching** and **code normalization** —
a sensitive point, since the IBGE municipal code has **7 digits** (including
a check digit), while the Health Information Systems usually record **6
digits**. The function aligns the two representations automatically. The
coordinate reference system adopted is **SIRGAS 2000** (EPSG:4674), the
official Brazilian standard, reprojected to WGS84 (EPSG:4326) when
centroids are computed.

> **Geographer's hat.** Three concerns run through this stage: (i) the
> **CRS** (always use SIRGAS 2000 / WGS84 consistently); (ii) the difference
> between **6- and 7-digit** IBGE codes (resolved by the function); and
> (iii) the **MAUP** — the scale and the delineation chosen affect
> inference, so document them.

> **Driving case.** We are investigating **temperature and outcomes in
> older adults (60+) in the São Paulo Metropolitan Region, 2014–2019**. At
> this stage, we link the respiratory-death series (output of Tutorial 4) to
> the **municipalities of the São Paulo Metropolitan Region (RMSP)**,
> obtaining the spatial object that Tutorial 6 (Socioeconomic Linkage) will
> use.

# Data used

| Input | Source | Role in the case |
|--------|--------|---------------|
| `caso_serie.rds` | Tutorial 4 (series aggregated by municipality × date) | Health input |
| Municipal geometries | IBGE via `geobr::read_municipality()` | RMSP polygons |
| Climate grid (demonstration) | synthetic (real climate enters in T8) | Illustrates `sus_grid_join()` |

The input is the object **`dados/caso_serie.rds`** (loaded with an
`file.exists()` guard; in its absence, the script uses a minimal synthetic
example). The output is **`dados/caso_espacial.rds`**, which feeds
Tutorial 6.

# Full pipeline

The blocks below reproduce what the script
`scripts/tutorial_05_juncao_espacial.R` executes. Start by loading the
package:

```{r carregar}
library(climasus4r)
```

## Load the health series (output of Tutorial 4)

```{r carregar-entrada}
# A etapa anterior salvou a série agregada (óbitos respiratórios × município × data).
caso_serie  <- readRDS("dados/caso_serie.rds")
serie_saude <- if (is.data.frame(caso_serie)) caso_serie else caso_serie$serie
```

## Link the series to municipalities with `sus_spatial_join()`

`sus_spatial_join()` **automatically detects** the municipal code column
(`join_col = NULL`), downloads the geometries from `geobr` (with caching)
and returns an **`sf`** object at `stage = "spatial"`. The default level is
`"munic"`:

```{r spatial-join-munic}
caso_espacial <- sus_spatial_join(
  df        = serie_saude,
  level     = "munic",     # município (padrão)
  join_col  = NULL,         # auto-detecção da coluna geográfica
  lang      = "pt",
  use_cache = TRUE,         # reaproveita geometrias já baixadas
  cache_dir = "~/.climasus4r_cache/spatial",
  verbose   = TRUE
)
```

The function accepts many **other** geographic **levels** — simply change
`level`:

```{r spatial-join-niveis}
# Bioma (chapéu ambiental):
sf_bioma <- sus_spatial_join(serie_saude, level = "biomes", lang = "pt")

# Região metropolitana:
sf_metro <- sus_spatial_join(serie_saude, level = "metro_area", lang = "pt")

# Geocodificação por CEP (somente sistemas SIH e CNES, gera POINTs):
sf_cep   <- sus_spatial_join(df_sih, level = "cep", lang = "pt")
```

## Integrate gridded environmental data with `sus_grid_join()`

`sus_grid_join()` is the **bridge** between the environmental pipeline
(`sus_grid_*`, which produces objects at `stage = "climate"`) and the health
pipeline. It merges the two tables on the space-time key
`c("code_muni", "date")` (real climate data is introduced in Tutorial 8;
here the usage is illustrative):

```{r grid-join}
# grid_clim é a saída de uma função sus_grid_*() (ERA5, CHIRPS, ...), stage = "climate"
caso_clima <- sus_grid_join(
  health_data = caso_espacial,
  grid_data   = grid_clim,
  by          = c("code_muni", "date"),  # chave espaço-temporal
  type_out    = NULL,                      # herda o type do grid
  lang        = "pt",
  verbose     = TRUE
)

sus_meta(caso_clima, "stage")  # "climate"
```

> **Temporal granularity.** If health data are **monthly** and the grid is
> **daily**, joining on `date` produces many `NA`s. In that case, aggregate
> the grid first (with `sus_climate_aggregate()`), or, for **annual** grids
> (e.g., PRODES), join only on `by = c("code_muni")` — the annual value is
> replicated for every month.

At the end of this stage we have `caso_espacial` — the health series **with
geometry**, saved for Tutorial 6.

# Function characteristics

## `sus_spatial_join()`

Links health data to Brazil's official geometries (`geobr`) or geocodes
postal codes (CEP). Requires a `climasus_df` at minimum `stage`
`"stand"` and returns an `sf` object at `stage = "spatial"`.

| Argument | Essential? | Description |
|-----------|------------|-----------|
| `df` | yes | `climasus_df` (at least standardized) with a geographic identifier column |
| `level` | yes | Geographic unit. `"munic"` (default); `"biomes"`, `"health_region"`, `"amazon"`, `"semiarid"`, `"conservation_units"`, `"disaster_risk_area"`, `"indigenous_land"`, `"urban_area"`, `"metro_area"`, `"urban_concentrations"`, `"pop_arrangements"`, `"schools"`, `"health_facilities"`, `"neighborhood"`, `"cep"` |
| `join_col` | optional | Column with the geographic code; `NULL` (default) auto-detects |
| `use_cache` / `cache_dir` | recommended | Geometry cache at `~/.climasus4r_cache/spatial` |
| `lang` / `verbose` | optional | Language (`"pt"`, `"en"`, `"es"`) and progress messages |

**System restrictions:** `"cep"` and `"neighborhood"` only for **SIH/CNES**;
`"health_facilities"` only for **CNES**.

**Return:** an `sf` object (`climasus_df`, `stage = "spatial"`, `type = level`)
with the original columns plus `geometry` — **POLYGON/MULTIPOLYGON** for
area levels, **POINT** for `"cep"`/`"health_facilities"`.

**Code normalization.** Accepts municipal codes with **6 or 7 digits** (the
check digit is handled automatically); CEP is left-padded with zeros
(8 characters); all keys are coerced to text. The area join is an
`inner_join` — codes with no matching geometry are **dropped** (see
Figure 3).

## `sus_grid_join()`

Merges the output of any `sus_grid_*()` (at `stage = "climate"`) with a
health `climasus_df` (at `stage` `"spatial"`, `"aggregate"`, or `"census"`),
producing a `climasus_df` at `stage = "climate"`.

| Argument | Essential? | Description |
|-----------|------------|-----------|
| `health_data` | yes | Health `climasus_df` with `code_muni` and `date` |
| `grid_data` | yes | `climasus_df` at `stage = "climate"` (output of `sus_grid_*`) |
| `by` | optional | Join key; default `c("code_muni", "date")`; use `c("code_muni")` for an annual grid |
| `type_out` | optional | Overrides the resulting `type`; `NULL` inherits it from the grid |
| `lang` / `verbose` | optional | Language and messages |

**Return:** `climasus_df` (`stage = "climate"`) with the health columns plus
the environmental variables; uses **`left_join`**, so health rows with no
matching grid receive `NA`. The `sf` geometry is dropped before the join
(the join is tabular, on the key). The function **warns** about column-name
collisions and about **temporal-granularity mismatches**.

# Formalization

The area join can be seen as a mapping from the municipal key
$k$ (IBGE code) to the geometry $G(k)$ of the set of geometries
$\mathcal{G}$:

$$
\text{join}(r) =
\begin{cases}
(r,\, G(k_r)) & \text{se } k_r \in \mathcal{G} \quad (\text{inner: descarta o resto}) \\[4pt]
\varnothing   & \text{caso contrário}
\end{cases}
$$

where $r$ is a health record and $k_r$ its municipal code. The **match
rate** (quality control in Figure 3) is

$$
\text{cobertura} = \frac{\bigl|\{\,k : k \in \mathcal{K}_{\text{entrada}} \cap \mathcal{K}_{\text{geobr}}\,\}\bigr|}
{\bigl|\mathcal{K}_{\text{entrada}}\bigr|},
$$

the fraction of input codes $\mathcal{K}_{\text{entrada}}$ that find a
matching geometry in $\mathcal{K}_{\text{geobr}}$. For `sus_grid_join()`,
the fraction of rows with missing climate is
$\frac{1}{n}\sum_{i=1}^{n} \mathbf{1}\{x_i = \text{NA}\}$,
directly reported in the completion message.

# Illustrated results

```{r fig-mapa, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/geo_mapa_obitos.png",
            "Figura 1 indisponível: o mapa requer sf + geobr e acesso à internet. Reexecute o script de geração.")
```

**Figure 1.** Choropleth map of respiratory deaths by municipality after the
spatial join (RMSP, 2014–2019). Each polygon now carries the health outcome
— the basis for mapping inequalities and cross-referencing with contextual
indicators.

```{r fig-serie, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/geo_serie.png", .nota_gerar)
```

**Figure 2.** Monthly series preserved after the join. The spatial join
**adds geometry without altering the temporal structure**: the time
dimension remains intact for modeling (Tutorial 9).

```{r fig-cobertura, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/geo_cobertura.png", .nota_gerar)
```

**Figure 3.** Coverage diagnostic of the join. Because `sus_spatial_join()`
uses `inner_join`, municipal codes with no matching geometry in `geobr` are
dropped. Checking this rate is an essential quality control (it avoids
silent loss of records).

```{r tab-resumo, eval=TRUE, echo=FALSE}
tbl_ou_nota("dados/geo_resumo.rds", .nota_gerar,
            caption = "Tabela 1. Resumo da junção espacial (nível, CRS, linhas antes/depois, geometria).")
```

```{r tab-amostra, eval=TRUE, echo=FALSE}
tbl_ou_nota("dados/geo_amostra.rds", .nota_gerar,
            caption = "Tabela 2. Amostra das colunas-chave do objeto sf resultante (geometria omitida).")
```

# References

- Openshaw, S. (1984). *The Modifiable Areal Unit Problem*. CATMOG 38. Norwich:
  Geo Books.
- Pereira, R. H. M., Gonçalves, C. N., et al. (2019). *geobr: Loads Shapefiles
  of Official Spatial Data Sets of Brazil*. GitHub: <https://github.com/ipeaGIT/geobr>.
- Pereira, R. H. M., & Barbosa, R. J. (2023). *censobr: Download Data from
  Brazil's Population Census*. R package v0.4.0.
  DOI: 10.32614/CRAN.package.censobr.
- IBGE — Instituto Brasileiro de Geografia e Estatística. Geocentric reference
  system SIRGAS 2000.

---

# Navigation

- [Back to tutorial hub](tutorials.html)
- **Previous:** [Tutorial 4 — Variable Creation and Aggregation]()
- **Next:** [Tutorial 6 — Socioeconomic Linkage](06-socioeconomico.html)

```
