---
title: "Tutorial 6: Linking Socioeconomic Data"
author: "climasus4r Team"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    code_download: true
    self_contained: true
    df_print: paged
---

```{css, echo=FALSE}
/* -- Variaveis de Cores Modernas -- */
:root {
  --primary-color: #1D9E75;
  --dark-green: #0f3d2e;
  --bg-light: #f8faf9;
  --text-main: #1a1a1a;
  --code-bg: #fdfdfd;
  --warning-orange: #ff9800;
  --info-blue: #2196f3;
}
html { scroll-behavior: smooth; }
body { font-family: 'Source Serif 4', Georgia, serif; font-size: 18px; color: var(--text-main); background-color: #fff; line-height: 1.6; }
h1, h2, h3, h4 { font-family: 'Inter', sans-serif; font-weight: 700; color: var(--dark-green); }
h1.title { font-size: 2.8em; border-bottom: 4px solid var(--primary-color); padding-bottom: 10px; margin-bottom: 1.5rem; }
h2 { margin-top: 2rem; border-left: 4px solid var(--primary-color); padding-left: 1rem; }
pre.r, pre.sourceCode { background-color: var(--code-bg); border: 1px solid #e1e1e1; border-left: 5px solid var(--primary-color) !important; border-radius: 8px; box-shadow: 0 2px 4px rgba(0,0,0,0.05); padding: 1rem; }
.callout { padding: 1.2em 1.5em; margin: 1.5em 0; border-radius: 8px; border-left: 6px solid; background-color: var(--bg-light); }
.callout-note { background-color: #f0f7ff; border-color: #007bff; }
.callout-tip { background-color: #e8f5e9; border-color: #1D9E75; }
.callout-warning { background-color: #fff3e0; border-color: #ff9800; }
.workflow-container { background: linear-gradient(135deg, #f8faf7 0%, #ffffff 100%); border-radius: 12px; padding: 1.5rem; margin: 1.5rem 0; border: 1px solid #dce8e1; overflow-x: auto; }
.workflow-steps { display: flex; flex-wrap: wrap; justify-content: center; gap: 0.5rem; font-family: 'Inter', monospace; font-size: 0.8rem; }
.workflow-step { background: var(--primary-color); color: white; padding: 0.5rem 1rem; border-radius: 20px; white-space: nowrap; }
.workflow-arrow { color: var(--primary-color); font-weight: bold; font-size: 1.2rem; }
.feature-grid { display: grid; grid-template-columns: repeat(auto-fit, minmax(280px, 1fr)); gap: 1.5rem; margin: 2rem 0; }
.feature-card { background: white; border-radius: 12px; padding: 1.5rem; border: 1px solid #e1e1e1; transition: transform 0.2s ease, box-shadow 0.2s ease; text-align: center; }
.feature-card:hover { transform: translateY(-4px); box-shadow: 0 8px 24px rgba(0,0,0,0.1); }
.feature-icon { font-size: 2rem; margin-bottom: 1rem; }
.feature-card h3 { margin-top: 0; color: var(--primary-color); font-size: 1.1rem; }
table { width: 100%; border-collapse: collapse; margin: 1.5rem 0; }
th, td { padding: 12px; text-align: left; border-bottom: 1px solid #e1e1e1; }
th { background-color: var(--bg-light); font-weight: 600; color: var(--dark-green); }
.github-feedback-btn { display: inline-flex; align-items: center; gap: 8px; background-color: #24292e; color: white; padding: 10px 20px; border-radius: 6px; text-decoration: none; font-family: sans-serif; transition: background-color 0.2s; margin-top: 1rem; }
.github-feedback-btn:hover { background-color: #2c3e50; color: white; text-decoration: none; }
@media (max-width: 768px) { body { font-size: 16px; } h1.title { font-size: 2rem; } .workflow-steps { flex-direction: column; align-items: center; } }
```


```{r setup, include=FALSE}
# ATENÇÃO — estratégia de pré-renderização.
# Este tutorial NÃO executa downloads do censo durante a renderização. Todas as
# figuras e tabelas foram geradas UMA vez pelo script
# scripts/tutorial_06_socioeconomico.R e são apenas carregadas aqui (PNG via
# include_graphics; RDS via readRDS). Por isso, os chunks de ensino abaixo usam
# eval = FALSE: o código é mostrado, mas não é executado no momento do knit.
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)

# Helpers para exibir artefatos de forma robusta: se o arquivo ainda não foi
# gerado (script de geração não executado), mostram uma nota em vez de quebrar
# o knit. Assim a vignette renderiza a qualquer momento.
fig_ou_nota <- function(path, nota) {
  if (file.exists(path)) knitr::include_graphics(path)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
tbl_ou_nota <- function(path, nota, caption = NULL) {
  if (file.exists(path)) knitr::kable(readRDS(path), caption = caption)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
.nota_gerar <- "Artefato indisponível — rode `Rscript scripts/tutorial_06_socioeconomico.R` (a partir da raiz do pacote) para gerá-lo."
```

# Objective

By the end of this tutorial you will be able to **link Brazilian Demographic
Census variables to health data that has already been spatialized** and
**compute standardized socioeconomic and epidemiological indicators**. You
will learn to:

- **discover** which census variables are available with
  `sus_census_select()`;
- **download and aggregate** census microdata (or census tracts) to the
  municipal level and join them to the health data with `sus_census_join()`
  (via the `censobr` package);
- **compute** indicators such as aging, dependency, and sanitation coverage —
  with **confidence intervals** — using `sus_socio_compute_indicators()`;
- **query** the full indicator catalog with `sus_socio_list_indicators()`.

The socioeconomic linkage is the **sixth stage** of the `climasus4r` pipeline.
It requires the data to have already gone through the **spatial join**
(Tutorial 5), since the linkage key is the municipality's **7-digit IBGE
code** (`code_muni_7`).

# Scientific context

Climate and health studies can rarely treat the population as homogeneous.
**Social vulnerability** — poverty, low education, poor sanitation, an aging
age structure — modulates the relationship between environmental exposure
(temperature, pollution) and health outcomes, acting both as a **confounder**
and as an **effect modifier** (Gouveia, Hajat & Armstrong, 2003; Hajat,
O'Connor & Kosatsky, 2010). Elderly people living in households lacking
adequate infrastructure, for example, suffer more during heat waves.

In Brazil, the official source for these variables is IBGE's **Demographic
Census**, accessible programmatically through the `censobr` package (Pereira &
Barbosa, 2023). `climasus4r` wraps this access in `sus_census_join()`,
downloading the chosen dataset, **filtering only the municipalities present in
the health data before loading into memory** (Arrow/Parquet optimization),
aggregating the microdata to the municipal level, and performing the join on
the geographic key.

From these variables, `sus_socio_compute_indicators()` computes a set of
**standardized indicators** (demographic, socioeconomic, mortality, morbidity,
maternal-and-child health, and health-resource indicators), each with its
**formula**, **multiplier**, and **uncertainty method** declared in an
internal catalog.

> **Methodological caveat (ecological bias).** Municipal indicators describe
> the *aggregate*, not the individual: associations observed between a
> municipality's mean vulnerability and its health outcomes **do not**
> automatically transfer to individuals (ecological fallacy). The
> **aggregation scale** also matters — the same phenomenon can change sign
> depending on the unit (modifiable areal unit problem, MAUP). Use these
> indicators as area-level covariates, not as individual attributes.

> **Guiding case.** We investigate **temperature and health outcomes in the
> elderly (60+) in the São Paulo Metropolitan Region, 2014–2019**. In
> Tutorial 5 we spatialized respiratory (J00–J99) and cardiovascular
> (I00–I99) deaths by municipality. Here we link **2010 Census** variables and
> compute vulnerability indicators (aging, dependency, sewage and water
> coverage) that will serve as **covariates/effect modifiers** in the
> climate–health modeling of the upcoming tutorials.

# Data used

| Item | Value | Role in the case |
|------|-------|---------------|
| Input | `dados/caso_espacial.rds` (Tutorial 5) | Spatialized deaths by municipality (RMSP) |
| Census dataset | `"population"` (2010 Census) | Population by municipality (elderly, young, working-age) |
| Linkage key | `code_muni_7` (IBGE, 7 digits) | Links health × census |
| Indicators | aging, dependency, sewage, water | Municipal vulnerability |
| Output | `dados/caso_socio.rds` | Input for Tutorial 7 (INMET climate) |

All data are **public and real** (IBGE/Census via `censobr`).

# Full pipeline

The blocks below reproduce what the script
`scripts/tutorial_06_socioeconomico.R` executes. Start by loading the package
and the chained input:

```{r carregar}
library(climasus4r)

# Saída do Tutorial 5: óbitos espacializados por município (stage "spatial",
# com a coluna code_muni_7 — chave do vínculo).
caso_espacial <- readRDS("dados/caso_espacial.rds")
```

## Step 1 — Discover census variables (`sus_census_select()`)

Before downloading the census, it is worth **discovering the codes** of the
variables. In interactive use, `output = "browser"` opens a searchable HTML
interface; in scripts, prefer `output = "codes"`, which **returns a vector of
codes** without opening the browser:

```{r select}
# Códigos das variáveis de população do Censo 2010 (uso programático):
vars_pop <- sus_census_select(
  dataset = "population",  # "all", "population", "households", "families", ...
  year    = 2010,           # 2010 (padrão) ou 2000
  output  = "codes",        # "browser" (padrão), "console" ou "codes"
  lang    = "pt",
  verbose = FALSE
)
head(vars_pop)
```

## Step 2 — Link the census (`sus_census_join()`)

`sus_census_join()` downloads the chosen dataset via `censobr`, **filters only
the municipalities present** in `code_muni_7`, aggregates the microdata to the
municipal level (`aggregation_fun`), and performs the `left_join`. It requires
the data to be at stage `>= "spatial"`:

```{r join}
caso_enriquecido <- sus_census_join(
  df              = caso_espacial,
  dataset         = "population",     # "population"/"households"/"families"/
                                       # "mortality"/"emigration"/"tracts"
  year            = 2010,             # 2010 (padrão) ou 2000
  census_vars     = head(vars_pop, 6),# NULL = todas (não recomendado)
  aggregation_fun = "sum",            # "sum","mean","median","min","max","sd",
                                       # "q25","q75","q95","q99"
  use_cache       = TRUE,             # cache local do censobr
  cache_dir       = "~/.climasus4r_cache/census",
  lang            = "pt",
  verbose         = TRUE
)
```

Each aggregated numeric variable receives the **suffix of the aggregation
mode** (e.g., `V0001` aggregated by sum becomes `V0001_sum`). If the input is
an `sf` object, the **geometry is preserved** in the output.

To use **census tract** data instead of microdata, switch to
`dataset = "tracts"` and choose the table in `tracts_dataset` (e.g.,
`"Basico"`, `"DomicilioRenda"`, `"Pessoa03"`):

```{r join-tracts}
caso_setores <- sus_census_join(
  df             = caso_espacial,
  dataset        = "tracts",
  tracts_dataset = "Basico",   # "Basico","Domicilio01/02","DomicilioRenda",
                                # "Pessoa01"..."Pessoa13","ResponsavelRenda", ...
  year           = 2010,
  lang           = "pt"
)
```

## Step 3 — Query the indicator catalog (`sus_socio_list_indicators()`)

Before computing, inspect what is available. Without a network, it returns a
`tibble` with `id`, `name`, `category`, `required_cols`, `formula`,
`multiplier`, `unit`, `uncertainty_method`, and `source`:

```{r list}
catalogo <- sus_socio_list_indicators(lang = "pt")

# Filtrar por categoria, se desejar:
sus_socio_list_indicators(lang = "pt", category = "mortality")
```

## Step 4 — Compute indicators (`sus_socio_compute_indicators()`)

Indicators are defined in **formula space** (names such as `pop_elderly`,
`total_hh`). Since the columns coming from the census have different names
(e.g., `V0001_sum`), use `col_mapping` to **connect** formula names to actual
columns:

```{r indicators}
caso_socio <- sus_socio_compute_indicators(
  df               = caso_enriquecido,
  indicators       = c("aging_index", "dependency_ratio",
                       "sewage_connection_rate", "water_connection_rate"),
  col_mapping = list(
    pop_elderly = "pop_65_plus_sum",  # aponte para as colunas agregadas reais
    pop_young   = "pop_0_14_sum",
    pop_working = "pop_15_64_sum",
    pop_total   = "V0001_sum",
    total_hh    = "V003_sum",
    hh_sewage   = "V0207_sum",
    hh_water    = "V0208_sum"
  ),
  confidence_level = 0.95,   # IC de Poisson (taxas) / Wilson (proporções)
  add_ci           = TRUE,   # adiciona colunas ind_<id>_low / ind_<id>_high
  lang             = "pt",
  verbose          = TRUE
)
```

Each computed indicator becomes a column `ind_<id>` (e.g.,
`ind_aging_index`), and — when the uncertainty method is not `"none"` and
`add_ci = TRUE` — also `ind_<id>_low` and `ind_<id>_high`. If
`indicators = NULL`, the function **automatically detects** all indicators
whose required columns are present (after `col_mapping`); indicators lacking
the necessary columns are **skipped with a warning**, never interrupting
execution.

At the end, `caso_socio` is saved (with `save_tbl()` in the script) and feeds
Tutorial 7 (Station-Based Climate — INMET).

# Function characteristics

## `sus_census_select()`

Explorer for census variables. Helps **discover codes** to use in
`sus_census_join(census_vars = ...)`.

| Argument | Essential? | Description |
|-----------|------------|-----------|
| `dataset` | recommended | `"all"` (default), `"population"`, `"households"`, `"families"`, `"mortality"`, `"emigration"`, `"tracts"` |
| `year` | optional | `2010` (default) or `2000` |
| `output` | recommended | `"browser"` (interactive HTML, default), `"console"`, or `"codes"` (vector) |
| `lang` | optional | `"pt"` (default), `"en"`, `"es"` |
| `verbose` | optional | Informative messages |

**Return:** with `"codes"`, a character vector of codes; with
`"browser"`/`"console"`, invisibly returns a `data.frame` (and opens/prints
the interface).

## `sus_census_join()`

Links census variables (microdata or tracts) to health data. Wrapper around
`censobr::read_population()`, `read_households()`, `read_tracts()`, etc.

| Argument | Essential? | Description |
|-----------|------------|-----------|
| `df` | yes | `climasus_df`/`sf` at stage `>= "spatial"`, with `code_muni_7` |
| `dataset` | yes | `"population"` (default), `"households"`, `"families"`, `"mortality"`, `"emigration"`, `"tracts"` |
| `tracts_dataset` | only `tracts` | Tract table: `"Basico"` (default), `"DomicilioRenda"`, `"Pessoa01..13"`, … |
| `year` | optional | `2010` (default) or `2000` |
| `census_vars` | recommended | Vector of codes (use `sus_census_select()`); `NULL` = all |
| `aggregation_fun` | optional | `"sum"` (default), `"mean"`, `"median"`, `"min"`, `"max"`, `"sd"`, `"q25"`, `"q75"`, `"q95"`, `"q99"` |
| `join_muni_col` | optional | IBGE code column; `NULL` detects SUS patterns |
| `use_cache` / `cache_dir` | recommended | `censobr` cache (`~/.climasus4r_cache/census`) |
| `translate_columns` / `standardize_values` | optional | Translates names and labels (default `TRUE`) |
| `lang` / `verbose` | optional | Language and messages |

**Return:** the same `df` (or `sf`) with additional census columns (suffixed
by the aggregation mode, e.g., `_sum`); metadata advances to `stage =
"census"`.

## `sus_socio_compute_indicators()`

Computes standardized indicators over the present columns, with optional
confidence intervals.

| Argument | Essential? | Description |
|-----------|------------|-----------|
| `df` | yes | `climasus_df` at stage `>= "spatial"` (output of `sus_census_join()`) |
| `indicators` | optional | Vector of IDs; `NULL` = all computable (auto-detection) |
| `col_mapping` | recommended | Named list: formula name → actual column (e.g., `list(total_hh = "V003_sum")`) |
| `confidence_level` | optional | Confidence level (default `0.95`) |
| `add_ci` | optional | `TRUE` (default) adds `ind_<id>_low`/`_high` |
| `lang` / `verbose` | optional | Language and messages |

**Return:** the `climasus_df` with columns `ind_<id>` (and `_low`/`_high`
where applicable); `sus_meta` updated to `type = "indicators"`.

## `sus_socio_list_indicators()`

Lists the full indicator catalog. **Does not depend on network access.**

| Argument | Description |
|-----------|-----------|
| `lang` | Language of the names: `"pt"` (default), `"en"`, `"es"` |
| `category` | Filter by category (e.g., `"demographic"`, `"mortality"`); `NULL` = all |

**Return:** a `tibble` with `id`, `name`, `category`, `required_cols`,
`formula`, `multiplier`, `unit`, `uncertainty_method`, `source`.

# Formulas

Indicators are computed as a ratio (rate/proportion) multiplied by a scaling
factor. In general form:

$$
\text{indicador} = \left(\frac{\text{numerador}}{\text{denominador}}\right) \times m
$$

where $m$ is the `multiplier` declared in the catalog. Examples used in the
guiding case:

$$
\text{Índice de Envelhecimento} = \frac{\text{pop}_{\text{idosa}}}{\text{pop}_{\text{jovem}}} \times 100
$$

$$
\text{Razão de Dependência} = \frac{\text{pop}_{\text{jovem}} + \text{pop}_{\text{idosa}}}{\text{pop}_{\text{ativa}}} \times 100
$$

$$
\text{Cobertura de Esgoto} = \frac{\text{domicílios com esgoto}}{\text{total de domicílios}} \times 100
$$

**Confidence intervals.** For **rates** based on event counts (mortality,
incidence, hospitalizations), the **exact Poisson interval (Garwood)** is
used, based on the Gamma distribution:

$$
\text{IC}_{\text{inf}} = \frac{1}{n}\,\chi^2_{\alpha/2;\,2k}\big/2, \qquad
\text{IC}_{\text{sup}} = \frac{1}{n}\,\chi^2_{1-\alpha/2;\,2(k+1)}\big/2
$$

For **proportions** (coverage rates, %), the **Wilson score interval** is
used:

$$
\hat{p} \pm \frac{z}{1 + z^2/n}\left[\frac{z}{2n} + \sqrt{\frac{\hat{p}(1-\hat{p})}{n} + \frac{z^2}{4n^2}}\right]
$$

with $z = z_{1-\alpha/2}$, $k$ the number of events, and $n$ the denominator.
Ratios with complex numerators or direct indices (e.g., Gini) do not receive a
CI (`uncertainty_method = "none"`).

# Illustrated results

```{r tab-catalogo, eval=TRUE, echo=FALSE}
tbl_ou_nota("dados/socio_catalogo.rds", .nota_gerar,
            caption = "Tabela 1. Amostra do catálogo de indicadores (sus_socio_list_indicators).")
```

```{r fig-categorias, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/socio_catalogo_categorias.png", .nota_gerar)
```

**Figure 1.** Number of indicators available per category in the
`sus_socio_compute_indicators()` catalog. Categories cover demographics,
socioeconomic vulnerability, mortality, morbidity, maternal-and-child health,
and health resources.

```{r tab-resumo, eval=TRUE, echo=FALSE}
tbl_ou_nota("dados/socio_resumo.rds", .nota_gerar,
            caption = "Tabela 2. Indicadores calculados por município (amostra).")
```

```{r fig-vulnerabilidade, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/socio_vulnerabilidade.png", .nota_gerar)
```

**Figure 2.** Municipal ranking by the Aging Index (elderly per 100 young
people, 2010 Census) in RMSP. More aged municipalities tend to concentrate
higher risk in the face of temperature extremes.

```{r fig-ic, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/socio_indicador_ic.png", .nota_gerar)
```

**Figure 3.** Sewage coverage by municipality with a 95% confidence interval
(Wilson score), as produced by `add_ci = TRUE`. The error bar conveys the
uncertainty of the proportion estimate.

```{r fig-dispersao, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/socio_dispersao_idosos.png", .nota_gerar)
```

**Figure 4.** Relationship between aging and sewage coverage across RMSP
municipalities. Crossing demographic and infrastructure vulnerability helps
raise hypotheses about confounding and effect modification in the
temperature–health relationship explored in the following tutorials.

# References

- Pereira, R. H. M., & Barbosa, R. J. (2023). *censobr: Download Data from
  Brazil's Population Census.* R package version 0.4.0.
  <https://CRAN.R-project.org/package=censobr>. DOI: 10.32614/CRAN.package.censobr.
- Gouveia, N., Hajat, S., & Armstrong, B. (2003). Socioeconomic differentials in
  the temperature–mortality relationship in São Paulo, Brazil. *International
  Journal of Epidemiology*, 32(3), 390–397.
- Hajat, S., O'Connor, M., & Kosatsky, T. (2010). Health effects of hot weather:
  from awareness of risk factors to effective health protection. *The Lancet*,
  375(9717), 856–863.
- IBGE — Instituto Brasileiro de Geografia e Estatística. *Censo Demográfico
  2010.* Rio de Janeiro: IBGE.

---

# Navigation

- [Back to tutorial hub](tutorials.html)
- **Previous:** [Tutorial 5 — Spatial Join](05-juncao-espacial.html)
- **Next:** [Tutorial 7 — Station-Based Climate (INMET)](07-clima-estacoes.html)
