---
title: "Tutorial 7: Station-Based Climate Integration (INMET)"
author: "climasus4r Team"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    code_download: true
    self_contained: true
    df_print: paged
---

```{css, echo=FALSE}
/* -- Variaveis de Cores Modernas -- */
:root {
  --primary-color: #1D9E75;
  --dark-green: #0f3d2e;
  --bg-light: #f8faf9;
  --text-main: #1a1a1a;
  --code-bg: #fdfdfd;
  --warning-orange: #ff9800;
  --info-blue: #2196f3;
}
html { scroll-behavior: smooth; }
body { font-family: 'Source Serif 4', Georgia, serif; font-size: 18px; color: var(--text-main); background-color: #fff; line-height: 1.6; }
h1, h2, h3, h4 { font-family: 'Inter', sans-serif; font-weight: 700; color: var(--dark-green); }
h1.title { font-size: 2.8em; border-bottom: 4px solid var(--primary-color); padding-bottom: 10px; margin-bottom: 1.5rem; }
h2 { margin-top: 2rem; border-left: 4px solid var(--primary-color); padding-left: 1rem; }
pre.r, pre.sourceCode { background-color: var(--code-bg); border: 1px solid #e1e1e1; border-left: 5px solid var(--primary-color) !important; border-radius: 8px; box-shadow: 0 2px 4px rgba(0,0,0,0.05); padding: 1rem; }
.callout { padding: 1.2em 1.5em; margin: 1.5em 0; border-radius: 8px; border-left: 6px solid; background-color: var(--bg-light); }
.callout-note { background-color: #f0f7ff; border-color: #007bff; }
.callout-tip { background-color: #e8f5e9; border-color: #1D9E75; }
.callout-warning { background-color: #fff3e0; border-color: #ff9800; }
.workflow-container { background: linear-gradient(135deg, #f8faf7 0%, #ffffff 100%); border-radius: 12px; padding: 1.5rem; margin: 1.5rem 0; border: 1px solid #dce8e1; overflow-x: auto; }
.workflow-steps { display: flex; flex-wrap: wrap; justify-content: center; gap: 0.5rem; font-family: 'Inter', monospace; font-size: 0.8rem; }
.workflow-step { background: var(--primary-color); color: white; padding: 0.5rem 1rem; border-radius: 20px; white-space: nowrap; }
.workflow-arrow { color: var(--primary-color); font-weight: bold; font-size: 1.2rem; }
.feature-grid { display: grid; grid-template-columns: repeat(auto-fit, minmax(280px, 1fr)); gap: 1.5rem; margin: 2rem 0; }
.feature-card { background: white; border-radius: 12px; padding: 1.5rem; border: 1px solid #e1e1e1; transition: transform 0.2s ease, box-shadow 0.2s ease; text-align: center; }
.feature-card:hover { transform: translateY(-4px); box-shadow: 0 8px 24px rgba(0,0,0,0.1); }
.feature-icon { font-size: 2rem; margin-bottom: 1rem; }
.feature-card h3 { margin-top: 0; color: var(--primary-color); font-size: 1.1rem; }
table { width: 100%; border-collapse: collapse; margin: 1.5rem 0; }
th, td { padding: 12px; text-align: left; border-bottom: 1px solid #e1e1e1; }
th { background-color: var(--bg-light); font-weight: 600; color: var(--dark-green); }
.github-feedback-btn { display: inline-flex; align-items: center; gap: 8px; background-color: #24292e; color: white; padding: 10px 20px; border-radius: 6px; text-decoration: none; font-family: sans-serif; transition: background-color 0.2s; margin-top: 1rem; }
.github-feedback-btn:hover { background-color: #2c3e50; color: white; text-decoration: none; }
@media (max-width: 768px) { body { font-size: 16px; } h1.title { font-size: 2rem; } .workflow-steps { flex-direction: column; align-items: center; } }
```


```{r setup, include=FALSE}
# ATENÇÃO — estratégia de pré-renderização.
# Este tutorial NÃO executa downloads nem treina modelos durante a renderização.
# Todas as figuras e tabelas foram geradas UMA vez pelo script
# scripts/tutorial_07_clima_estacoes.R e são apenas carregadas aqui (PNG via
# include_graphics; RDS via readRDS). Por isso, os chunks de ensino abaixo usam
# eval = FALSE: o código é mostrado, mas não é executado no momento do knit.
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)

# Helpers para exibir artefatos de forma robusta: se o arquivo ainda não foi
# gerado (script de geração não executado), mostram uma nota em vez de quebrar
# o knit. Assim a vignette renderiza a qualquer momento.
fig_ou_nota <- function(path, nota) {
  if (file.exists(path)) knitr::include_graphics(path)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
tbl_ou_nota <- function(path, nota, caption = NULL) {
  if (file.exists(path)) knitr::kable(readRDS(path), caption = caption)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
.nota_gerar <- "Artefato indisponível — rode `Rscript scripts/tutorial_07_clima_estacoes.R` (a partir da raiz do pacote) para gerá-lo."
```

# Objective

By the end of this tutorial you will be able to **bring the observed climate
dimension into the pipeline** using data from the **National Institute of
Meteorology (INMET)** weather stations. Specifically, you will learn to:

1. **Import** the hourly series of a station with `sus_climate_inmet()`;
2. **Fill gaps** in the series with machine learning models
   (XGBoost) per station, with `sus_climate_fill_inmet()`;
3. **Visualize** the gap-filling result (observed vs imputed) with
   `sus_climate_plot_fill()`;
4. **Compute the 30-year climatological normals** with `sus_climate_normals()`,
   which serve as a climatological baseline;
5. **Align** the climate series with the health series using
   epidemiologically grounded temporal strategies (same day, lag, moving
   window...) with `sus_climate_aggregate()`.

This is the **seventh stage** of the pipeline (`stage = "climate"`). It
connects the already-aggregated health series (Tutorial 4) to observed
climate, producing the exposure × outcome base that will feed the gridded
climate stage (Tutorial 8) and modelling (Tutorial 9).

# Scientific context

Environmental epidemiology relates **climate exposures** to **health
outcomes**. In Brazil, INMET's automatic stations provide hourly
observations of temperature, humidity, precipitation, pressure, and wind,
with growing national coverage since the 2000s. Unlike gridded reanalysis
products (Tutorial 8), station data measure the environment *in situ*, with
high temporal resolution but with **limited spatial representativeness**
around the station (Saldanha et al., 2019; Geirinhas et al., 2018).

Three methodological challenges run through this stage:

- **Quality and gaps.** Sensors fail; series have missing values and
  physical inconsistencies. Gap-filling with models such as XGBoost
  (Chen & Guestrin, 2016) recovers gaps from temporal patterns
  (lags, daily and seasonal cycles), preserving the series structure without
  introducing the crude biases of linear interpolation.
- **Climatological baseline.** The **climatological normals** (30-year
  averages) define the reference climate of each location — essential for
  contextualizing anomalies and defining the reference point of
  dose-exposure response models (Gasparrini, 2011).
- **Causal temporal alignment.** The effect of heat or cold on health is
  rarely instantaneous: there are **lags** (delayed effects) and
  **accumulation** (cumulative exposure). The choice of temporal strategy
  must reflect the hypothesized biological mechanism (Bhaskaran et al.,
  2013; Gasparrini et al., 2010), and must never use *future* climate to
  explain a present event (anticipation bias).

> **Guiding case.** We investigate **temperature and health outcomes in
> older adults (60+) in the São Paulo Metropolitan Region (RMSP), 2014–2019**.
> In this stage, we import the series of the RMSP reference station —
> **São Paulo – Mirante de Santana (code A701)** —, fill its gaps, compute
> the normals, and **align temperature with the daily health series**
> inherited from Tutorial 4. The resulting base (health × climate) is saved
> for Tutorial 8.

# Data used

| Source | Function | Role in the case |
|-------|--------|---------------|
| Hourly INMET series, station A701 (RMSP) | `sus_climate_inmet()` | Observed climate exposure |
| A701 series with gaps filled (XGBoost) | `sus_climate_fill_inmet()` | Continuous temperature series |
| Climatological normals 1991–2020 (SP) | `sus_climate_normals()` | Climatological baseline |
| Daily health series from T4 (`dados/caso_serie.rds`) | `sus_climate_aggregate()` | Outcome to be aligned |

**Chained input:** `dados/caso_serie.rds` (health series from Tutorial 4).
**Chained output:** `dados/caso_clima_estacao.rds` (health + aligned climate),
which Tutorial 8 will load. All data are **public and real** (INMET).

# Full pipeline

The blocks below reproduce what the script `scripts/tutorial_07_clima_estacoes.R`
executes. Load the package:

```{r carregar}
library(climasus4r)
```

## Import the station series (A701)

`sus_climate_inmet()` downloads INMET's hourly data, standardizes the
column names to canonical names (e.g., `tair_dry_bulb_c`, `rh_mean_porc`,
`rainfall_mm`) and applies physical quality control. We filter by the RMSP
reference station via `station_code`.

```{r import-inmet}
clima_horario <- sus_climate_inmet(
  years        = 2014:2019,
  uf           = "SP",
  station_code = "A701",    # São Paulo - Mirante de Santana
  use_cache    = TRUE,       # cache em dois níveis (memória + disco/Parquet)
  parallel     = TRUE,       # paraleliza entre anos e entre arquivos do ano
  workers      = 4,
  lang         = "pt",
  verbose      = TRUE
)
```

The return is a `climasus_df` at `stage = "climate"`, `type = "inmet"`, with
**hourly** frequency and dates in **UTC**.

## Fill gaps with XGBoost

`sus_climate_fill_inmet()` fits an XGBoost model **per station and per
variable**, automatically creating temporal features (hour, day, month,
cyclical transforms), lags (1, 2, 3, 12, 24, 168 periods) and rolling
statistics. We first assess accuracy (evaluation mode), then actually fill
the gaps (production mode):

```{r fill-avaliacao}
# Modo AVALIAÇÃO: cria lacunas artificiais (MCAR) e devolve métricas por estação
aval <- sus_climate_fill_inmet(
  df                = clima_horario,
  target_var        = "tair_dry_bulb_c",
  quality_threshold = 0.4,    # exclui estações com > 40% de faltantes
  run_evaluation    = TRUE,
  gap_percentage    = 0.2,    # 20% de lacunas artificiais
  parallel          = TRUE,
  workers           = 4,
  lang              = "pt"
)
aval$tair_dry_bulb_c$metrics   # RMSE, MAE, R², sMAPE, slope_bias, n_gaps
```

```{r fill-producao}
# Modo PRODUÇÃO: preenche as lacunas reais e adiciona a flag is_imputed_<var>
clima_preenchido <- sus_climate_fill_inmet(
  df                = clima_horario,
  target_var        = "tair_dry_bulb_c",
  quality_threshold = 0.4,
  run_evaluation    = FALSE,
  keep_features     = FALSE,   # retorna só as colunas originais + flag
  parallel          = TRUE,
  workers           = 4,
  lang              = "pt"
)
```

To impute **all** 17 canonical INMET variables at once, use
`target_var = "all"`.

## Visualize the gap-filling

`sus_climate_plot_fill()` automatically detects the mode (production vs
evaluation) and draws the observed series with the imputed points
highlighted. In static mode (`interactive = FALSE`) it returns a `ggplot2`
object:

```{r plot-fill}
sus_climate_plot_fill(
  df_filled   = clima_preenchido,
  df_original = clima_horario,
  target_var  = "tair_dry_bulb_c",
  interactive = FALSE,        # FALSE = ggplot2; TRUE = plotly interativo
  output_type = "plot",       # "plot", "table", "metrics" ou "all"
  lang        = "pt"
)
```

## Compute the climatological normals

`sus_climate_normals()` downloads INMET's 30-year normals. We choose the
most recent period and the temperature variables:

```{r normais}
normais <- sus_climate_normals(
  period     = "1991-2020",                       # ou "1961-1990", "1981-2010"
  target_var = c("t_max", "t_min", "t_mean_comp"),
  use_cache  = TRUE,
  lang       = "pt"
)

# Para explorar os códigos de variável disponíveis antes:
# sus_climate_normals_meta(period = "1991-2020", lang = "pt")
```

## Align climate and health (temporal strategies)

`sus_climate_aggregate()` links each health record to the nearest station
(Euclidean distance) and applies the chosen temporal strategy. We compare
**same-day** exposure (`exact`) with the **7-day moving window**
(`moving_window`, recommended for cumulative exposure):

```{r aggregate}
# Exposição aguda (mesmo dia)
saude_exact <- sus_climate_aggregate(
  health_data       = saude_df,             # série de saúde do T4 (com geometria)
  climate_data      = clima_preenchido,
  climate_var       = "tair_dry_bulb_c",
  temporal_strategy = "exact",
  lang              = "pt"
)

# Exposição cumulativa (média móvel retroativa de 7 dias: t-6 .. t)
saude_movel <- sus_climate_aggregate(
  health_data       = saude_df,
  climate_data      = clima_preenchido,
  climate_var       = "tair_dry_bulb_c",
  temporal_strategy = "moving_window",
  window_days       = 7,
  lang              = "pt"
)
```

The resulting series (health + aligned temperature) is saved to
`dados/caso_clima_estacao.rds` for Tutorial 8.

# Function characteristics

## `sus_climate_inmet()`

Imports, standardizes, and quality-controls INMET's hourly data.

| Argument | Essential? | Description |
|-----------|------------|-----------|
| `years` | recommended | Year(s), 2000–current year (e.g., `2014:2019`); `NULL` uses the last 2 |
| `uf` | recommended | State(s) (`"SP"`, `c("RJ","MG")`); `NULL` downloads all 27 states |
| `station_code` | optional | Filters stations by code (`"A701"`); `NULL` = all in the state |
| `use_cache` / `cache_dir` | recommended | Two-level cache (memory + Parquet/Zstd) |
| `parallel` / `workers` | optional | Parallelizes across years and across CSVs within a year |
| `lang` / `verbose` | optional | Language (`"pt"`/`"en"`/`"es"`) and progress |

**Return:** `climasus_df` (`stage = "climate"`, `type = "inmet"`), **hourly**
frequency, dates in **UTC**, canonical columns. **Automatic QC:** validation
of physical ranges, dew point consistency (Magnus formula), and zeroed
nighttime solar radiation.

## `sus_climate_fill_inmet()`

Imputes missing values with XGBoost, one model per station per variable.

| Argument | Essential? | Description |
|-----------|------------|-----------|
| `df` | yes | Output of `sus_climate_inmet()` |
| `target_var` | yes | Column(s) to impute, or `"all"` (the 17 canonical variables) |
| `quality_threshold` | recommended | Max. missing proportion per station (default `0.4`) |
| `run_evaluation` | optional | `TRUE` creates artificial gaps and returns metrics |
| `gap_percentage` | optional | Proportion of artificial gaps in evaluation (default `0.2`) |
| `keep_features` | optional | `TRUE` keeps engineered features (lags/rolling means) |
| `parallel` / `workers` | optional | Processes stations in parallel (`furrr`) |
| `datetime_col` / `station_col` | optional | Column names; auto-detected if `NULL` |

**Return (production):** the same tibble with `target_var` imputed and flag
`is_imputed_<var>`. **Return (evaluation):** named list with `$data` and
`$metrics` (RMSE, MAE, R², sMAPE, `slope_bias`, `n_gaps`) per station.

## `sus_climate_plot_fill()`

Visualizes the gap-filling; automatically detects production vs evaluation.

| Argument | Essential? | Description |
|-----------|------------|-----------|
| `df_filled` | yes | Output of `sus_climate_fill_inmet()` |
| `df_original` | production only | Raw series (from `sus_climate_inmet()`) |
| `target_var` | yes | Variable(s) to visualize |
| `interactive` | optional | `FALSE` → `ggplot2`; `TRUE` → `plotly` |
| `output_type` | optional | `"plot"`, `"table"`, `"metrics"` or `"all"` |
| `save_plot` | optional | Path to save (`.png` or `.html`) |

**Return:** the plot, the table, the metrics, or a list with all four
components. **Evaluation mode** draws a 2×2 diagnostic panel
(observed vs predicted, residual histogram, residuals over time and vs observed).

## `sus_climate_normals()`

Downloads INMET's 30-year climatological normals.

| Argument | Essential? | Description |
|-----------|------------|-----------|
| `period` | recommended | `"1961-1990"`, `"1981-2010"` or `"1991-2020"` (default) |
| `target_var` | optional | Variable codes (`c("t_max","t_min")`); `NULL` = all |
| `use_cache` / `cache_dir` | recommended | Disk cache of the catalog and files |
| `lang` / `verbose` | optional | Language and progress |

**Return:** `climasus_df` (`type = "normals"`), long format with `codigo`,
`nome_estacao`, `uf`, `mes`, `decada`, `valor`, `var_code` and multilingual
labels. Companion function: `sus_climate_normals_meta()` lists the available
codes.

## `sus_climate_aggregate()`

Links climate to health via the nearest station and applies the temporal
strategy.

| Argument | Essential? | Description |
|-----------|------------|-----------|
| `health_data` | yes | `climasus_df` with `date`, `code_muni` and geometry (`sf`) |
| `climate_data` | yes | Output of `sus_climate_fill_inmet()` |
| `climate_var` | recommended | Climate variable(s); `"all"` (default) |
| `temporal_strategy` | yes | One of 10 strategies (see below) |
| `window_days` | conditional | Window size (moving, degree-days, threshold, weighted) |
| `lag_days` | conditional | Lag(s) for `discrete_lag` / max for `distributed_lag` |
| `offset_days` | conditional | `c(W1, W2)` for `offset_window` |
| `temp_base` / `gdd_temp_var` | conditional | Thermal base and variable for `degree_days` |
| `threshold_value` / `threshold_direction` | conditional | Threshold for `*_exceedance` |
| `climate_region` | optional | `"auto"` (default), `"tropical"`, `"subtropical"`, `"temperate"` |
| `min_obs` | optional | Minimum proportion of valid observations in the window (default `0.7`) |

**Temporal strategies (`temporal_strategy`):** `exact`, `discrete_lag`,
`moving_window`, `offset_window`, `distributed_lag`, `degree_days`, `seasonal`,
`threshold_exceedance`, `cold_wave_exceedance`, `weighted_window`.

**Return:** `climasus_df` with the health data and the integrated climate
variables as new columns.

# Formulas

**Backward-looking moving average (`moving_window`, window of $W$ days).**
The cumulative exposure of day $t$ is the average of the observations from
$t-W+1$ to $t$ — it never uses the future (avoids anticipation bias):

$$
\bar{x}_t^{(W)} = \frac{1}{W} \sum_{\ell=0}^{W-1} x_{t-\ell}
$$

**Discrete lag (`discrete_lag`).** Exposure is the value observed exactly
$L$ days before the event:

$$
x_t^{(L)} = x_{t-L}
$$

**Degree-days (`degree_days`, base $T_b$).** Accumulation of the temperature
excess over the base across the window (thermal stress):

$$
\mathrm{GDD}_t = \sum_{\ell=0}^{W-1} \max\!\left(x_{t-\ell} - T_b,\; 0\right)
$$

**Threshold exceedance (`threshold_exceedance`).** Number of days in the
window in which temperature exceeds the threshold $\tau$ (heat wave proxy):

$$
N_t = \sum_{\ell=0}^{W-1} \mathbb{1}\!\left(x_{t-\ell} > \tau\right)
$$

# Illustrated results

```{r fig-serie, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/inmet_serie.png", .nota_gerar)
```

**Figure 1.** Daily mean temperature observed at station A701 (São Paulo –
Mirante de Santana), 2014–2019. The winter seasonality (lower temperatures
between June and August) is clear and will be the axis of alignment with
mortality in older adults.

```{r tab-resumo, eval=TRUE, echo=FALSE}
tbl_ou_nota("dados/inmet_resumo.rds", .nota_gerar,
            caption = "Tabela 1. Cobertura e lacunas da série importada (antes do preenchimento).")
```

```{r fig-fill, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/inmet_fill.png", .nota_gerar)
```

**Figure 2.** Result of the gap-filling (`sus_climate_plot_fill()`):
the observed series with the values imputed by XGBoost highlighted. The
imputation preserves the daily and seasonal cycles, rather than "flattening"
the missing stretches.

```{r tab-metricas, eval=TRUE, echo=FALSE}
tbl_ou_nota("dados/inmet_fill_metricas.rds", .nota_gerar,
            caption = "Tabela 2. Métricas de avaliação do preenchimento (lacunas artificiais MCAR): RMSE, MAE, R² e n. de lacunas por estação.")
```

```{r fig-normais, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/inmet_normais.png",
            "Figura 3 indisponível: as normais climatológicas exigem acesso ao portal do INMET. Reexecute o script de geração com internet.")
```

**Figure 3.** Annual cycle of the temperature climatological normals
(1991–2020) for the state of São Paulo. The normals provide the
climatological baseline against which anomalies and extremes are measured in
the following tutorials.

```{r fig-lag, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/inmet_lag.png", .nota_gerar)
```

**Figure 4.** Temperature exposure by temporal strategy in the winter of 2014
(RMSP): the **same-day** value (`exact`) oscillates more, while the **7-day
moving average** (`moving_window`) smooths the signal and captures cumulative
exposure — a choice that reflects the biological mechanism of interest in the
final model.

# References

- Bhaskaran, K., Gasparrini, A., Hajat, S., Smeeth, L., & Armstrong, B. (2013).
  Time series regression studies in environmental epidemiology.
  *International Journal of Epidemiology*, 42(4), 1187–1195.
- Chen, T., & Guestrin, C. (2016). XGBoost: A scalable tree boosting system.
  *Proceedings of the 22nd ACM SIGKDD*, 785–794.
- Gasparrini, A. (2011). Distributed lag linear and non-linear models in R: the
  package dlnm. *Journal of Statistical Software*, 43(8), 1–20.
- Gasparrini, A., Armstrong, B., & Kenward, M. G. (2010). Distributed lag
  non-linear models. *Statistics in Medicine*, 29(21), 2224–2234.
- Geirinhas, J. L., et al. (2018). Climatic and synoptic characterization of heat
  waves in Brazil. *International Journal of Climatology*, 38(4), 1760–1776.
- Saldanha, R. de F., Bastos, R. R., & Barcellos, C. (2019). Microdatasus:
  pacote para download e pré-processamento de microdados do DATASUS.
  *Cadernos de Saúde Pública*, 35(9), e00032419.
- INMET — Instituto Nacional de Meteorologia. Dados históricos.
  Available at: <https://portal.inmet.gov.br/dadoshistoricos>.

---

# Navigation

- [Back to tutorial hub](tutorials.html)
- **Previous:** [Tutorial 6 — Socioeconomic Linkage (Census)](06-socioeconomico.html)
- **Next:** [Tutorial 8 — Gridded Climate & Environmental Data (ERA5/CHIRPS)](08-clima-grade.html)
