---
title: "Tutorial 8: Gridded Climate and Environmental Data"
author: "climasus4r Team"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    code_download: true
    self_contained: true
    df_print: paged
---

```{css, echo=FALSE}
/* -- Variaveis de Cores Modernas -- */
:root {
  --primary-color: #1D9E75;
  --dark-green: #0f3d2e;
  --bg-light: #f8faf9;
  --text-main: #1a1a1a;
  --code-bg: #fdfdfd;
  --warning-orange: #ff9800;
  --info-blue: #2196f3;
}
html { scroll-behavior: smooth; }
body { font-family: 'Source Serif 4', Georgia, serif; font-size: 18px; color: var(--text-main); background-color: #fff; line-height: 1.6; }
h1, h2, h3, h4 { font-family: 'Inter', sans-serif; font-weight: 700; color: var(--dark-green); }
h1.title { font-size: 2.8em; border-bottom: 4px solid var(--primary-color); padding-bottom: 10px; margin-bottom: 1.5rem; }
h2 { margin-top: 2rem; border-left: 4px solid var(--primary-color); padding-left: 1rem; }
pre.r, pre.sourceCode { background-color: var(--code-bg); border: 1px solid #e1e1e1; border-left: 5px solid var(--primary-color) !important; border-radius: 8px; box-shadow: 0 2px 4px rgba(0,0,0,0.05); padding: 1rem; }
.callout { padding: 1.2em 1.5em; margin: 1.5em 0; border-radius: 8px; border-left: 6px solid; background-color: var(--bg-light); }
.callout-note { background-color: #f0f7ff; border-color: #007bff; }
.callout-tip { background-color: #e8f5e9; border-color: #1D9E75; }
.callout-warning { background-color: #fff3e0; border-color: #ff9800; }
.workflow-container { background: linear-gradient(135deg, #f8faf7 0%, #ffffff 100%); border-radius: 12px; padding: 1.5rem; margin: 1.5rem 0; border: 1px solid #dce8e1; overflow-x: auto; }
.workflow-steps { display: flex; flex-wrap: wrap; justify-content: center; gap: 0.5rem; font-family: 'Inter', monospace; font-size: 0.8rem; }
.workflow-step { background: var(--primary-color); color: white; padding: 0.5rem 1rem; border-radius: 20px; white-space: nowrap; }
.workflow-arrow { color: var(--primary-color); font-weight: bold; font-size: 1.2rem; }
.feature-grid { display: grid; grid-template-columns: repeat(auto-fit, minmax(280px, 1fr)); gap: 1.5rem; margin: 2rem 0; }
.feature-card { background: white; border-radius: 12px; padding: 1.5rem; border: 1px solid #e1e1e1; transition: transform 0.2s ease, box-shadow 0.2s ease; text-align: center; }
.feature-card:hover { transform: translateY(-4px); box-shadow: 0 8px 24px rgba(0,0,0,0.1); }
.feature-icon { font-size: 2rem; margin-bottom: 1rem; }
.feature-card h3 { margin-top: 0; color: var(--primary-color); font-size: 1.1rem; }
table { width: 100%; border-collapse: collapse; margin: 1.5rem 0; }
th, td { padding: 12px; text-align: left; border-bottom: 1px solid #e1e1e1; }
th { background-color: var(--bg-light); font-weight: 600; color: var(--dark-green); }
.github-feedback-btn { display: inline-flex; align-items: center; gap: 8px; background-color: #24292e; color: white; padding: 10px 20px; border-radius: 6px; text-decoration: none; font-family: sans-serif; transition: background-color 0.2s; margin-top: 1rem; }
.github-feedback-btn:hover { background-color: #2c3e50; color: white; text-decoration: none; }
@media (max-width: 768px) { body { font-size: 16px; } h1.title { font-size: 2rem; } .workflow-steps { flex-direction: column; align-items: center; } }
```


```{r setup, include=FALSE}
# ATENÇÃO — estratégia de pré-renderização.
# Este tutorial NÃO executa downloads durante a renderização. Todas as figuras e
# tabelas foram geradas UMA vez pelo script scripts/tutorial_08_clima_grade.R e
# são apenas carregadas aqui (PNG via include_graphics; RDS via readRDS).
# Por isso, os chunks de ensino abaixo usam eval = FALSE: o código é mostrado,
# mas não é executado no momento do knit.
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)

# Helpers para exibir artefatos de forma robusta: se o arquivo ainda não foi
# gerado (script de geração não executado), mostram uma nota em vez de quebrar
# o knit. Assim a vignette renderiza a qualquer momento.
fig_ou_nota <- function(path, nota) {
  if (file.exists(path)) knitr::include_graphics(path)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
tbl_ou_nota <- function(path, nota, caption = NULL) {
  if (file.exists(path)) knitr::kable(readRDS(path), caption = caption)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
.nota_gerar <- "Artefato indisponível — rode `Rscript scripts/tutorial_08_clima_grade.R` (a partir da raiz do pacote) para gerá-lo."
```

# Objective

By the end of this tutorial you will be able to **integrate gridded climate
and environmental data** (remote sensing and reanalysis) into your health
time series: import **ERA5-Land** (~10 km) with `sus_grid_era5()` and
aggregate it to municipal polygons; get to know the full `sus_grid_*` family
(CHIRPS precipitation, fire hotspots, PDSI/SMVI droughts, PRODES
deforestation, and CAMS/GHAP/MERRA-2 pollution); and derive **climate
indicators for epidemiology**: anomalies (`sus_climate_anomaly()`),
bioclimatic indices (`sus_climate_compute_indicators()`), heatwaves
(`sus_climate_compute_heatwaves()` + `sus_climate_plot_heatwaves()`), and the
drought indices **SPI** (`sus_climate_compute_spi()`) and **SPEI**
(`sus_climate_compute_spei()`).

This is the **eighth stage** of the pipeline: while Tutorial 7 built the
climate exposure from INMET **stations** (point measurements), here we build
an **alternative and complementary gridded** exposure — continuous spatial
coverage, with no observation gaps — that will feed the DLNM model in
Tutorial 9.

# Scientific context

## Station vs. pixel: two ways of measuring exposure

A weather station measures climate at a **point**; a gridded product
represents the **mean value of a pixel** (a fixed-area cell). **ERA5-Land**
(Muñoz-Sabater et al., 2021) is a **reanalysis**: a physical land-surface
model driven by observations, producing continuous hourly fields since 1950
at ~0.1° resolution (**~10 km** at the equator). For Brazil, the *ERA5-Land
Daily Aggregates for Latin America* project (Saldanha, Zenodo, CC-BY 4.0)
provides ready-made **daily** aggregates, requiring no API key — this is the
source used by `sus_grid_era5()`.

Three caveats for anyone working with remote sensing/reanalysis:

1. **Resolution and representativeness.** A ~10 km pixel is an areal
   average: it smooths out local extremes (urban heat islands, valley
   microclimates) that a point station would capture. On the other hand, it
   covers municipalities **without a station** and has no missing data. For
   a large municipality such as São Paulo (~1,500 km²), dozens of pixels are
   aggregated to the polygon — `climasus4r` does this with an **area-weighted**
   mean of the intersection (via `exactextractr`), which avoids edge bias.
2. **Physical units.** The original NetCDF files come in SI units:
   temperature in **Kelvin** (K → °C conversion: subtract 273.15),
   precipitation in **meters** (m → mm: multiply by 1,000), and pressure in
   **Pa** (Pa → hPa: divide by 100). `sus_grid_era5()` applies these
   conversions automatically and delivers canonical columns
   (`tair_dry_bulb_c`, `rainfall_mm`, ...).
3. **Cross-validation.** Reanalysis does not replace observation: it is
   recommended to compare the gridded series with the nearest INMET station
   (Tutorial 7) before choosing the model's exposure — systematic
   differences (bias) are expected and should be documented.

## From raw fields to epidemiological indicators

The climate-health literature rarely uses raw temperature directly:

- **Anomalies** (WMO, 2017) make cities with different climates comparable
  and highlight extreme days to correlate with mortality peaks (Gasparrini et
  al., 2017);
- **Heatwaves** defined by percentiles and duration (Perkins & Alexander,
  2013; Geirinhas et al., 2018) and by the *Excess Heat Factor* — EHF (Nairn
  & Fawcett, 2015) capture the acute effect of sustained heat;
- **SPI** (McKee et al., 1993) and **SPEI** (Vicente-Serrano et al., 2010)
  standardize drought/wetness across multiple time scales — relevant to
  diarrheal diseases, leptospirosis, and arboviral diseases;
- **Bioclimatic indices** (UTCI, Bröde et al., 2012; WBGT, Liljegren et al.,
  2008; HI, Rothfusz, 1990) combine temperature, humidity, wind, and
  radiation into a single measure of human thermal stress.

> **Guiding case.** We continue with **temperature and health outcomes in
> older adults (60+) in the São Paulo Metropolitan Region, 2014–2019**. In
> this tutorial we import ERA5-Land for the municipality of São Paulo and
> derive anomalies, heatwaves, and SPI/SPEI. The result
> (`caso_clima_grade.rds`) feeds into Tutorial 9 as an alternative/
> complementary exposure to the station-based one (`caso_clima_estacao.rds`,
> Tutorial 7).

# Data used

| Source | Function | Resolution | Variables | Role in the case |
|-------|--------|-----------|-----------|---------------|
| ERA5-Land (Zenodo) | `sus_grid_era5()` | ~10 km, daily | Mean, max, min temperature, precipitation | Primary gridded exposure |
| Health series (T4) | `dados/caso_serie.rds` | Municipality × day | Deaths J00–J99/I00–I99, 60+ | Temporal/spatial skeleton |

Chained input: `dados/caso_serie.rds` (Tutorial 4). Output:
`dados/caso_clima_grade.rds` (a list with the ERA5-Land daily series, the
aligned health+climate series, monthly SPI/SPEI, and the heatwave events).

The `sus_grid_*` family covers other environmental sources with the **same
grammar** (years/months, `municipalities` as an `sf` object, pixel→polygon
aggregation, caching):

| Function | Source | Theme |
|--------|-------|------|
| `sus_grid_chirps()` | CHIRPS (0.05°) | Precipitation (Funk et al., 2015) |
| `sus_grid_fires()` | INPE/Queimadas | Fire hotspots |
| `sus_grid_pdsi()` | PDSI | Drought (Palmer index) |
| `sus_grid_smvi()` | SMVI | Flash drought (soil moisture) |
| `sus_grid_prodes()` | PRODES/TerraBrasilis | Deforestation |
| `sus_grid_pollution_cams()` | CAMS/Copernicus | Air pollution (reanalysis) |
| `sus_grid_pollution_ghap()` | GHAP | High-resolution PM2.5 |
| `sus_grid_pollution_merra2()` | MERRA-2/NASA | Aerosols and pollutants |
| `sus_grid_join()` | — | Joins grids to the health base by `code_muni` × `date` |

# Full pipeline

The blocks below reproduce what `scripts/tutorial_08_clima_grade.R` executes.
Start by loading the package and the chained health series:

```{r carregar}
library(climasus4r)

caso_serie <- readRDS("dados/caso_serie.rds")  # série diária de saúde (T4)
```

## Importing ERA5-Land for the municipality of São Paulo

First we obtain the municipal polygon (an `sf` object via `geobr`) and then
call `sus_grid_era5()`. The function downloads the NetCDF files from Zenodo
(no API key needed), converts units (K → °C; m → mm), and aggregates each
pixel to the polygon with an area-weighted mean:

```{r import-era5}
library(geobr)

sp_geom <- read_municipality(code_muni = "SP", year = 2020)
sp_geom <- sp_geom[as.character(sp_geom$code_muni) == "3550308", ]  # São Paulo

era5_sp <- sus_grid_era5(
  years          = 2014:2019,
  months         = 1:12,
  vars           = c("t2m", "t2m_max", "t2m_min", "tp"),
  municipalities = sp_geom,    # sf -> agrega raster ao polígono
  agg_fun        = "mean",     # média areal (ponderada) do município
  use_cache      = TRUE,       # NetCDF reaproveitados em ~/.climasus4r_cache/era5
  parallel       = FALSE,
  lang           = "pt",
  verbose        = TRUE
)

sus_meta(era5_sp, "stage")  # "climate"
sus_meta(era5_sp, "type")   # "era5_land"
```

The return value is a `climasus_df` with `code_muni`, `date`, and one column
per variable: `tair_dry_bulb_c` (t2m), `tair_max_c` (t2m_max), `tair_min_c`
(t2m_min), and `rainfall_mm` (tp).

```{r tab-resumo, eval=TRUE, echo=FALSE}
tbl_ou_nota("dados/grid_resumo.rds", .nota_gerar,
            caption = "Table 1. Summary of the ERA5-Land daily series for São Paulo, 2014–2019.")
```

## Climate anomalies

`sus_climate_anomaly()` compares observations to INMET's **climatological
normals** (from `sus_climate_normals()`), with three methods — absolute,
relative, and standardized:

```{r anomalia}
normais <- sus_climate_normals(
  period     = "1991-2020",
  target_var = c("t_max", "t_min", "t_mean_comp", "precipitation")
)

anom <- sus_climate_anomaly(
  observed   = clima_estacao,        # climasus_df do T7 (estações INMET)
  normals    = normais,
  method     = "standardized",       # z-score: comparável entre cidades
  time_scale = "monthly",
  lang       = "pt"
)
```

Since the ERA5-Land series uses `code_muni` (rather than `station_code`
matched to the INMET normals), the generation script illustrates the **same
mathematics** using the monthly climatology derived from the series itself —
obs − month mean and the corresponding z-score (see the formulas below).

## Bioclimatic indicators (demonstration)

`sus_climate_compute_indicators()` computes WBGT, HI, THI, UTCI, PET,
degree-days (CDD/HDD/GDD), and others, from **hourly** data with temperature,
humidity (`rh_mean_porc`), radiation (`sr_kj_m2`), and wind (`ws_2_m_s`):

```{r indicadores}
clima_ind <- sus_climate_compute_indicators(
  df         = clima_horario,                  # T7 (INMET horário) ou ERA5 com td2m/u10/v10/sp
  indicators = c("wbgt", "hi", "utci"),
  region     = "atlantic_forest",              # limiares adaptados ao bioma
  lang       = "pt"
)
```

The minimal ERA5-Land series used in this tutorial (temperature and rain
only) does not carry humidity/wind/radiation; in a real workflow, apply the
function to the hourly INMET series (T7) or request `vars = "all"` in
`sus_grid_era5()` (includes `td2m`, `u10`, `v10`, `sp`).

## Heatwaves

`sus_climate_compute_heatwaves()` expects a `climasus_df` at
`stage = "climate"` with `date`, `station_code`, and temperature columns. We
adapt the gridded series to this contract (`station_code = code_muni`) and
apply three methods:

```{r ondas-calor}
hw <- sus_climate_compute_heatwaves(
  df             = era5_hw,                   # ERA5 adaptado ao contrato de estação
  method         = c("WHO", "WMO", "INMET"),  # ou "all" (inclui EHF, UTCI, WBGT, HI)
  baseline_start = "2014-01-01",              # período de referência dos percentis
  baseline_end   = "2019-12-31",
  percentile     = 90,                        # P90 para WHO/WMO
  lang           = "pt"
)

hw$events    # um evento por linha: início, fim, duração, pico, intensidade
hw$daily     # flags diárias hw_who, hw_wmo, hw_inmet, hw_any
hw$summary   # resumo anual por estação e método
```

A ready-made visualization comes from `sus_climate_plot_heatwaves()`:

```{r plot-hw}
sus_climate_plot_heatwaves(
  hw,
  type        = "trend",     # ou "timeline", "calendar", "intensity"
  interactive = FALSE,       # TRUE -> plotly
  lang        = "pt"
)
```

## SPI and SPEI: droughts and wet periods

Both functions require **monthly** data (one row per municipality × month,
with `date` on the 1st day of the month). We aggregate the daily series:
precipitation by **sum**, temperature by **mean**:

```{r spi-spei}
mensal <- era5_sp |>
  dplyr::mutate(date_m = as.Date(format(date, "%Y-%m-01"))) |>
  dplyr::summarise(
    rainfall_mm     = sum(rainfall_mm, na.rm = TRUE),
    tair_dry_bulb_c = mean(tair_dry_bulb_c, na.rm = TRUE),
    .by = c("code_muni", "date_m")
  ) |>
  dplyr::rename(date = date_m)

# SPI: apenas precipitação (gamma + zeros)
spi <- sus_climate_compute_spi(
  df     = mensal,
  var    = "rainfall_mm",        # padrão é "rainfall_chirps_mm" (sus_grid_chirps)
  scales = c(1L, 3L, 6L),        # série de 6 anos: escalas curtas
  min_n  = 24L,                  # mínimo de 24 meses para ajustar a distribuição
  lang   = "pt"
)

# SPEI: balanço hídrico P - PET (PET de Thornthwaite, só temperatura)
spei <- sus_climate_compute_spei(
  df         = spi,
  rain_var   = "rainfall_mm",
  pet_method = "thornthwaite",   # ou "column" com pet_var = "pet_mm"
  temp_var   = "tair_dry_bulb_c",
  scales     = c(1L, 3L, 6L),
  min_n      = 24L,
  lang       = "pt"
)

names(spei)  # ..., spi_1mo, spi_3mo, spi_6mo, spei_1mo, spei_3mo, spei_6mo
```

> **Methodological note.** The WMO recommends **≥ 30 years** of calibration
> for SPI/SPEI (`ref_start`/`ref_end`). With 6 years (72 months), the indices
> in this case are **illustrative**: in real research, calibrate with CHIRPS
> 1981–2010 (`sus_grid_chirps()`) or ERA5-Land since 1950.

## Chaining: aligning gridded climate to health and saving for T9

Finally, we join the daily ERA5-Land temperature to the health series by
`date` and save the gridded exposure bundle that Tutorial 9 will use:

```{r encadear}
saude_clima <- dplyr::left_join(
  caso_serie,
  era5_sp |> dplyr::select(date, tair_dry_bulb_c, tair_max_c,
                           tair_min_c, rainfall_mm),
  by = "date"
)

caso_clima_grade <- list(
  era5_diario = era5_sp,
  saude_clima = saude_clima,
  spi_spei    = spei,
  ondas_calor = hw$events
)
saveRDS(caso_clima_grade, "dados/caso_clima_grade.rds")
```

# Function characteristics

## `sus_grid_era5()`

The entry point for ERA5-Land: downloads daily aggregates from Zenodo
(Saldanha, 1950–2025) and aggregates pixels to municipal polygons.

| Argument | Essential? | Description |
|-----------|------------|-----------|
| `years` | yes | Years (1950–2025), e.g. `2014:2019` |
| `months` | optional | Months 1–12 (default: all) |
| `vars` | optional | ERA5 aliases: `"t2m"`, `"t2m_max"`, `"t2m_min"`, `"td2m"`, `"u10"`, `"v10"`, `"sp"`, `"tp"`, or `"all"` (default `c("t2m","tp")`) |
| `municipalities` | recommended | `sf` object with the polygons (detects `code_muni`/`CD_MUN`); `NULL` returns only the NetCDF paths |
| `agg_fun` | optional | Pixel→polygon aggregation via `exactextractr`: `"mean"` (default, area-weighted), `"sum"`, `"median"`, `"min"`, `"max"` |
| `use_cache` / `cache_dir` | recommended | Reuses downloaded NetCDF files (`~/.climasus4r_cache/era5`) |
| `parallel` / `workers` | optional | Parallel downloads; respects the user's `future::plan()` |
| `lang` / `verbose` | optional | Language (`"pt"`, `"en"`, `"es"`) and messages |

**Return:** `climasus_df` [`stage = "climate"`, `type = "era5_land"`] with
`code_muni`, `date`, and canonical columns already in practical units
(`tair_dry_bulb_c` in °C, `rainfall_mm` in mm, `patm_hpa` in hPa).

## `sus_climate_anomaly()`

| Argument | Description |
|-----------|-----------|
| `observed` | `climasus_df` from `sus_climate_inmet()` or `sus_climate_aggregate()` |
| `normals` | `climasus_df` from `sus_climate_normals()` (columns `codigo`, `mes`, `decada`, `var_code`, `valor`) |
| `vars` | Map of observed column → normals `var_code`, e.g. `c(tair_max_c = "t_max")`; `NULL` auto-detects |
| `method` | `"absolute"` (default), `"relative"`, `"standardized"`, or `"all"` |
| `time_scale` | `"monthly"` (default) or `"decadal"` (10-day periods) |
| `station_col` / `date_col` | Names of the station and date columns (`"station_code"`, `"date"`) |

**Return:** `climasus_df` [`stage = "climate"`, `type = "anomaly"`], one row
per station × year × month, with `{v}_obs`, `{v}_normal`, `{v}_anomaly`, and,
depending on the method, `{v}_anomaly_pct` and `{v}_anomaly_std`.

## `sus_climate_compute_indicators()`

| Argument | Description |
|-----------|-----------|
| `df` | Hourly data from `sus_climate_inmet()`/`sus_climate_fill_inmet()` |
| `indicators` | `"all"` (default) or codes: `"wbgt"`, `"hi"`, `"thi"`, `"wcet"`, `"wct"`, `"et"`, `"utci"`, `"pet"`, `"cdd"`, `"hdd"`, `"gdd"`, `"diurnal_range"`, `"vapor_pressure"`, `"heat_stress_risk"`, `"koppen_humidity"` |
| `region` | `"auto"` (default) or biome: `"amazon"`, `"cerrado"`, `"caatinga"`, `"atlantic_forest"`, `"pampa"`, `"pantanal"`, `"southeast"`, `"south"` |
| `confidence_flags` | `_flag_extreme`/`_flag_high`/`_flag_low` flags per indicator |
| `verify_physics` | Checks solar radiation against extraterrestrial limits |
| `compute_uncertainty` | 95% CI via Monte Carlo (200 simulations) |

**Return:** `climasus_df` with columns such as `wbgt_c`, `hi_c`, `utci_c` —
a prerequisite for the UTCI/WBGT/HI heatwave methods.

## `sus_climate_compute_heatwaves()`

| Argument | Description |
|-----------|-----------|
| `df` | `climasus_df` [`stage="climate"`, type `inmet`/`filled`/`indicators`] with `date`, `station_code`, and temperatures |
| `method` | `"WHO"`, `"WMO"`, `"INMET"`, `"EHF"`, `"UTCI"`, `"WBGT"`, `"HI"`, or `"all"` (methods missing a required column are skipped with a warning) |
| `baseline_start` / `baseline_end` | Reference period for the thresholds; `NULL` uses the whole series |
| `percentile` | Threshold percentile (default 90) |
| `min_duration` | Minimum consecutive days; `NULL` uses each method's default (WHO=3, WMO=5, INMET=5, EHF=3...) |

**Return:** a `climasus_hw` list with `events` (discrete events), `daily`
(daily flags `hw_<method>`, `hw_any`), and `summary` (annual summary).
Visualization via `sus_climate_plot_heatwaves(hw_result, type = "timeline" |
"calendar" | "intensity" | "trend", interactive, lang)`.

## `sus_climate_compute_spi()` and `sus_climate_compute_spei()`

| Argument | SPI | SPEI |
|-----------|-----|------|
| `df` | monthly, with `code_muni` + `date` | same |
| rainfall variable | `var` (default `"rainfall_chirps_mm"`) | `rain_var` |
| PET | — | `pet_method = "column"` (`pet_var`) or `"thornthwaite"` (`temp_var`) |
| `scales` | scales in months (default `c(1,3,6,12)`) | same |
| `ref_start` / `ref_end` | calibration period (WMO: ≥ 30 years) | same |
| `min_n` | minimum valid months per municipality (default 24) | same |

**Return:** the input `climasus_df` with columns `spi_{s}mo` /
`spei_{s}mo` (`type = "spi"` / `"spei"`). Distributions: gamma with a
zero-inflation mixture (SPI); 3-parameter log-logistic via L-moments (SPEI).

# Formulas

**Anomaly** (WMO, 2017). For variable $x$ in month $m$ of year $t$, with
climatological normal $\bar{x}_m$ and interannual standard deviation
$\sigma_m$:

$$
A_{t,m} = x_{t,m} - \bar{x}_m
\qquad\text{(absolute, °C)}
$$

$$
Z_{t,m} = \frac{x_{t,m} - \bar{x}_m}{\sigma_m}
\qquad\text{(standardized, z-score)}
$$

The `"relative"` method divides by $|\bar{x}_m|$ and multiplies by 100 (%,
customary for precipitation).

**SPI** (McKee et al., 1993). Precipitation accumulated over $s$ months,
$P_s$, is fitted to a gamma distribution (with probability $p_0$ of dry
months) and transformed to the standard normal:

$$
\mathrm{SPI}_s = \Phi^{-1}\!\big[\, p_0 + (1 - p_0)\, F_\gamma(P_s)\,\big]
$$

where $F_\gamma$ is the gamma CDF and $\Phi^{-1}$ the inverse standard normal.
Values $\le -1$ indicate moderate drought; $\le -1{,}5$ severe; $\le -2$
extreme.

**SPEI** (Vicente-Serrano et al., 2010). Replaces $P$ with the accumulated
**climatic water balance**, sensitive to warming:

$$
D = P - \mathrm{PET}, \qquad
\mathrm{SPEI}_s = \Phi^{-1}\!\big[\, F_{LL}\!\big(\textstyle\sum_s D\big)\,\big]
$$

with $F_{LL}$ the 3-parameter log-logistic CDF and PET estimated by
Thornthwaite (1948) when only temperature is available.

**EHF — Excess Heat Factor** (Nairn & Fawcett, 2015). Combines the anomaly
against the climate ($\mathrm{EHI}_{sig}$) and against the recent 30 days
(acclimatization, $\mathrm{EHI}_{acc}$):

$$
\mathrm{EHI}_{sig} = \bar{T}_3 - T_{95}, \qquad
\mathrm{EHI}_{acc} = \bar{T}_3 - \bar{T}_{30}
$$

$$
\mathrm{EHF} = \mathrm{EHI}_{sig} \times \max(1,\ \mathrm{EHI}_{acc})
$$

where $\bar{T}_3$ and $\bar{T}_{30}$ are the means of $T_{mean}$ over the last
3 and 30 days, and $T_{95}$ is the 95th percentile of the baseline. The
WHO/WMO methods, in turn, require $T_{max} > P_{90}(T_{max})$ (WMO also
$T_{min} > P_{90}(T_{min})$) for a minimum number of consecutive days.

# Illustrated results

```{r fig-serie, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/grid_serie.png", .nota_gerar)
```

**Figure 1.** Mean daily temperature from ERA5-Land (~10 km) aggregated to
the municipality of São Paulo, 2014–2019. Note the typical smoothness of the
areal value: averaging dozens of pixels dampens the local extremes a station
would record — this is the trade-off for complete spatial coverage.

```{r fig-anomalia, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/grid_anomalia.png", .nota_gerar)
```

**Figure 2.** Monthly temperature anomaly (observed − monthly normal), São
Paulo, 2014–2019. Orange bars (positive) mark months warmer than the local
climatology; it is exactly this signal — rather than the absolute
temperature — that travels between cities in multicenter analyses.

```{r tab-eventos, eval=TRUE, echo=FALSE}
tbl_ou_nota("dados/grid_eventos_hw.rds", .nota_gerar,
            caption = "Table 2. Detected heatwave events (sample): method, start, end, duration, and intensity.")
```

```{r fig-hw, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/grid_ondas_calor.png", .nota_gerar)
```

**Figure 3.** Heatwaves detected per year and method (WHO/WMO/INMET) in the
São Paulo ERA5-Land series, 2014–2019. More demanding methods (WMO: dual
threshold on Tmax **and** Tmin for ≥ 5 days) detect fewer events than the
WHO criterion (P90 of Tmax for ≥ 3 days) — the choice should reflect the
physiological mechanism of interest (daytime heat vs. warm nights without
relief).

```{r fig-spi, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/grid_spi_spei.png", .nota_gerar)
```

**Figure 4.** Monthly SPI-3 vs. SPEI-3 (dashed lines: ±1 = moderate
threshold). When temperature rises without a change in rainfall, SPEI drops
more than SPI — potential evapotranspiration amplifies the water deficit.
São Paulo's 2014–2015 water crisis is the kind of episode these indices
signal ahead of time for outcomes such as diarrheal diseases.

# References

- Bröde, P., et al. (2012). Deriving the operational procedure for the
  Universal Thermal Climate Index (UTCI). *International Journal of
  Biometeorology*, 56(3), 481–494.
- Funk, C., et al. (2015). The climate hazards infrared precipitation with
  stations — a new environmental record for monitoring extremes.
  *Scientific Data*, 2, 150066.
- Gasparrini, A., et al. (2017). Projections of temperature-related excess
  mortality under climate change scenarios. *The Lancet Planetary Health*,
  1(9), e360–e367.
- Geirinhas, J. L., et al. (2018). Climatic and synoptic characterization of
  heat waves in Brazil. *International Journal of Climatology*, 38(4),
  1760–1776.
- Liljegren, J. C., et al. (2008). Modeling the wet bulb globe temperature
  using standard meteorological measurements. *Journal of Occupational and
  Environmental Hygiene*, 5(10), 645–655.
- McKee, T. B., Doesken, N. J., & Kleist, J. (1993). The relationship of
  drought frequency and duration to time scales. *Proceedings of the 8th
  Conference on Applied Climatology*, AMS, 179–184.
- Muñoz-Sabater, J., et al. (2021). ERA5-Land: a state-of-the-art global
  reanalysis dataset for land applications. *Earth System Science Data*,
  13(9), 4349–4383.
- Nairn, J. R., & Fawcett, R. J. B. (2015). The Excess Heat Factor: a metric
  for heatwave intensity and its use in classifying heatwave severity.
  *International Journal of Environmental Research and Public Health*, 12(1),
  227–253.
- Perkins, S. E., & Alexander, L. V. (2013). On the measurement of heat waves.
  *Journal of Climate*, 26(13), 4500–4517.
- Rothfusz, L. P. (1990). The heat index "equation". NWS Technical Attachment
  SR/SSD 90-23.
- Saldanha, R. ERA5-Land Daily Aggregates for Latin America (1950–2025).
  Zenodo. <https://zenodo.org/doi/10.5281/zenodo.10013254>. CC-BY 4.0.
- Thornthwaite, C. W. (1948). An approach toward a rational classification of
  climate. *Geographical Review*, 38(1), 55–94.
- Vicente-Serrano, S. M., Beguería, S., & López-Moreno, J. I. (2010). A
  multiscalar drought index sensitive to global warming: the Standardized
  Precipitation Evapotranspiration Index. *Journal of Climate*, 23(7),
  1696–1718.
- WMO (2017). *WMO Guidelines on the Calculation of Climate Normals*
  (WMO-No. 1203). Geneva.

---

# Navigation

- [Back to tutorial hub](tutorials.html)
- **Previous:** [Tutorial 7 — Station-Based Climate (INMET)](07-clima-estacoes.html)
- **Next:** [Tutorial 9 — Epidemiological Modeling (DLNM)](09-modelagem.html)
