---
title: "Tutorial 9: Epidemiological Modeling (DLNM and Attributable Burden)"
author: "climasus4r Team"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    code_download: true
    self_contained: true
    df_print: paged
---

```{css, echo=FALSE}
/* -- Variaveis de Cores Modernas -- */
:root {
  --primary-color: #1D9E75;
  --dark-green: #0f3d2e;
  --bg-light: #f8faf9;
  --text-main: #1a1a1a;
  --code-bg: #fdfdfd;
  --warning-orange: #ff9800;
  --info-blue: #2196f3;
}
html { scroll-behavior: smooth; }
body { font-family: 'Source Serif 4', Georgia, serif; font-size: 18px; color: var(--text-main); background-color: #fff; line-height: 1.6; }
h1, h2, h3, h4 { font-family: 'Inter', sans-serif; font-weight: 700; color: var(--dark-green); }
h1.title { font-size: 2.8em; border-bottom: 4px solid var(--primary-color); padding-bottom: 10px; margin-bottom: 1.5rem; }
h2 { margin-top: 2rem; border-left: 4px solid var(--primary-color); padding-left: 1rem; }
pre.r, pre.sourceCode { background-color: var(--code-bg); border: 1px solid #e1e1e1; border-left: 5px solid var(--primary-color) !important; border-radius: 8px; box-shadow: 0 2px 4px rgba(0,0,0,0.05); padding: 1rem; }
.callout { padding: 1.2em 1.5em; margin: 1.5em 0; border-radius: 8px; border-left: 6px solid; background-color: var(--bg-light); }
.callout-note { background-color: #f0f7ff; border-color: #007bff; }
.callout-tip { background-color: #e8f5e9; border-color: #1D9E75; }
.callout-warning { background-color: #fff3e0; border-color: #ff9800; }
.workflow-container { background: linear-gradient(135deg, #f8faf7 0%, #ffffff 100%); border-radius: 12px; padding: 1.5rem; margin: 1.5rem 0; border: 1px solid #dce8e1; overflow-x: auto; }
.workflow-steps { display: flex; flex-wrap: wrap; justify-content: center; gap: 0.5rem; font-family: 'Inter', monospace; font-size: 0.8rem; }
.workflow-step { background: var(--primary-color); color: white; padding: 0.5rem 1rem; border-radius: 20px; white-space: nowrap; }
.workflow-arrow { color: var(--primary-color); font-weight: bold; font-size: 1.2rem; }
.feature-grid { display: grid; grid-template-columns: repeat(auto-fit, minmax(280px, 1fr)); gap: 1.5rem; margin: 2rem 0; }
.feature-card { background: white; border-radius: 12px; padding: 1.5rem; border: 1px solid #e1e1e1; transition: transform 0.2s ease, box-shadow 0.2s ease; text-align: center; }
.feature-card:hover { transform: translateY(-4px); box-shadow: 0 8px 24px rgba(0,0,0,0.1); }
.feature-icon { font-size: 2rem; margin-bottom: 1rem; }
.feature-card h3 { margin-top: 0; color: var(--primary-color); font-size: 1.1rem; }
table { width: 100%; border-collapse: collapse; margin: 1.5rem 0; }
th, td { padding: 12px; text-align: left; border-bottom: 1px solid #e1e1e1; }
th { background-color: var(--bg-light); font-weight: 600; color: var(--dark-green); }
.github-feedback-btn { display: inline-flex; align-items: center; gap: 8px; background-color: #24292e; color: white; padding: 10px 20px; border-radius: 6px; text-decoration: none; font-family: sans-serif; transition: background-color 0.2s; margin-top: 1rem; }
.github-feedback-btn:hover { background-color: #2c3e50; color: white; text-decoration: none; }
@media (max-width: 768px) { body { font-size: 16px; } h1.title { font-size: 2rem; } .workflow-steps { flex-direction: column; align-items: center; } }
```


```{r setup, include=FALSE}
# ATENÇÃO — estratégia de pré-renderização.
# Este tutorial NÃO ajusta modelos durante a renderização. Todas as figuras e
# tabelas foram geradas UMA vez pelo script scripts/tutorial_09_modelagem.R e
# são apenas carregadas aqui (PNG via include_graphics; RDS via readRDS).
# Por isso, os chunks de ensino abaixo usam eval = FALSE: o código é mostrado,
# mas não é executado no momento do knit.
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)

# Helpers para exibir artefatos de forma robusta: se o arquivo ainda não foi
# gerado (script de geração não executado), mostram uma nota em vez de quebrar
# o knit. Assim a vignette renderiza a qualquer momento.
fig_ou_nota <- function(path, nota) {
  if (file.exists(path)) knitr::include_graphics(path)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
tbl_ou_nota <- function(path, nota, caption = NULL) {
  if (file.exists(path)) knitr::kable(readRDS(path), caption = caption)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
.nota_gerar <- "Artefato indisponível — rode `Rscript scripts/tutorial_09_modelagem.R` (a partir da raiz do pacote) para gerá-lo."
```

# Objective

By the end of this tutorial you will be able to **turn a daily health series
aligned with climate into epidemiological inference**: estimate the
**non-linear and lagged** relationship between temperature and mortality with a
**DLNM** (`sus_mod_dlnm()`), quantify the **attributable burden** of heat and
cold (`sus_mod_af()`), and get to know the other tools in `climasus4r`'s
biostatistical toolkit — interrupted time series (`sus_mod_its()`),
case-crossover (`sus_mod_casecrossover()`), and multi-city meta-analysis
(`sus_mod_pool()`).

Modeling is the **last stage** of the pipeline. It consumes the output of the
previous stages (the aggregated health series from Tutorial 4 and the aligned
climate exposure from Tutorials 7 and 8) and produces **relative risk
estimates**, **attributable fractions**, and publication-ready visualizations.

# Scientific context

Modern environmental epidemiology treats the association between an
environmental exposure (temperature, pollutants) and a daily count outcome
(deaths, hospitalizations) as a problem that is both **non-linear** and
**lagged in time**. Extreme heat kills on the same day or within 1–2 days
(acute effect); cold kills over one to two weeks (delayed, prolonged effect).
Modeling only the day's temperature underestimates the effect of cold.

The **Distributed Lag Non-linear Model (DLNM)** of
**Gasparrini, Armstrong & Kenward (2010)** solves this with a **cross-basis**:
the outer product of two spline bases — one on the **exposure** dimension
(captures the U/J-shaped curve: risk rising with both cold and heat) and one
on the **lag** dimension (captures how the effect is distributed over the
days). The R implementation is the `dlnm` package (**Gasparrini, 2011**), on
top of which `climasus4r` builds `sus_mod_dlnm()`.

The fit follows the good practices for **time-series regression in
environmental epidemiology** systematized by **Bhaskaran et al. (2013)**: a
log-linear quasi-Poisson model (robust to overdispersion), control for
long-term trend and seasonality via a natural spline of time, and adjustment
for day of week. The **attributable fraction** decomposes the total burden of
events into heat and cold components (**Gasparrini & Armstrong, 2013**).

> **Guiding case.** We investigate **temperature and health outcomes in older
> adults (60+) in the São Paulo Metropolitan Region, 2014–2019**
> (respiratory J00–J99 and cardiovascular I00–I99). Here we join the **daily
> death series** (`caso_serie.rds`, Tutorial 4) with the **temperature from
> INMET station A701** (São Paulo – Mirante de Santana), already aligned
> (`caso_clima_estacao.rds`, Tutorial 7), build the distributed-lag structure,
> and fit the DLNM to estimate the exposure-response curve and the burden
> attributable to heat and cold.

# Data used

| Input | Source | Content | Role |
|---------|--------|----------|-------|
| `dados/caso_serie.rds` | Tutorial 4 | Daily death series (older adults 60+) | Outcome `n_obitos` |
| `dados/caso_clima_estacao.rds` | Tutorial 7 | Daily temperature from station A701 | Exposure `tair_dry_bulb_c` |

The two series are joined by **date**. The exposure is then expanded into
**distributed-lag** columns (`tair_dry_bulb_c_lag0`, …, `tair_dry_bulb_c_lag21`),
a structure that in a full workflow is produced by
`sus_climate_aggregate(temporal_strategy = "distributed_lag", lag_days = 21)`.
The script guards this stage with `requireNamespace()`/`tryCatch()` and, if
the inputs are missing, generates a **minimal, realistic synthetic series**
(with a V-shaped temperature-mortality signal) purely for illustration.

# Full pipeline

The blocks below reproduce what `scripts/tutorial_09_modelagem.R` executes.

```{r carregar}
library(climasus4r)
```

## Building the DLNM input (distributed lags)

`sus_mod_dlnm()` expects a `climasus_df` at stage `"climate"` and type
`"distributed_lag"`, with the `date` column, the outcome, and the
`{variable}_lag0 … _lagL` columns. In a real workflow this comes from
`sus_climate_aggregate()`:

```{r montar-dl}
df_dl <- sus_climate_aggregate(
  health_data       = caso_serie,          # health series (from Tutorial 4)
  climate_data      = caso_clima_estacao,  # aligned temperature (from Tutorial 7)
  climate_var       = "tair_dry_bulb_c",
  temporal_strategy = "distributed_lag",
  lag_days          = 21,
  lang              = "pt"
)
```

## Fitting the DLNM (`sus_mod_dlnm`)

Cross-basis with **exposure** on a natural spline with 4 degrees of freedom
(captures the U/J heat+cold shape) and **lag** on a natural spline with 3
degrees (smooth decay), **quasi-Poisson** family, and automatic seasonal
control (`dof_per_year = 6`, recommended for subtropical Brazil):

```{r dlnm}
fit <- sus_mod_dlnm(
  df           = df_dl,
  outcome_col  = "n_obitos",
  climate_col  = "tair_dry_bulb_c",
  lag_max      = 21,
  argvar       = list(fun = "ns", df = 4),  # exposure dimension
  arglag       = list(fun = "ns", df = 3),  # lag dimension
  family       = "quasipoisson",
  dof_per_year = 6L,                        # df/year for seasonality
  lang         = "pt"
)

print(fit)
fit$exposure_response   # cumulative RR on the percentile grid
fit$lag_response        # RR by lag at the 75th percentile
fit$diagnostics$disp_ratio
```

## Attributable fraction from heat and cold (`sus_mod_af`)

From the fitted DLNM, we decompose the total burden of deaths into **heat**
(above the comfort threshold) and **cold** (below it) components, with
confidence intervals from **Monte Carlo** simulation:

```{r af}
af <- sus_mod_af(
  fit  = fit,
  by   = "month",   # optional monthly breakdown
  nsim = 200L,      # use 5000L for publication-quality results
  lang = "pt"
)

print(af)
af$total        # total AF/AN, heat and cold
af$by_period    # monthly breakdown
```

## Visualizing (`sus_mod_plot_dlnm`)

The package offers ready-made scientific visualizations — exposure-response
curve, lag-response, 3-D surface, contour, lag slices, distribution, and time
series:

```{r plot}
sus_mod_plot_dlnm(fit, type = "overall", lang = "pt")  # exposure-response
sus_mod_plot_dlnm(fit, type = "contour", lang = "pt")  # log(RR) surface
sus_mod_plot_dlnm(fit, type = "lag",     lang = "pt")  # lag-response
```

# Function characteristics

## `sus_mod_dlnm()` — Distributed Lag Non-linear Model

| Argument | Essential? | Description |
|-----------|------------|-----------|
| `df` | yes | `climasus_df` at `stage = "climate"`, `type = "distributed_lag"` |
| `outcome_col` | yes | Daily count column (e.g., `"n_obitos"`) |
| `climate_col` | optional | Exposure base without the `_lagN` suffix; `NULL` auto-detects |
| `lag_max` | optional | Maximum lag; `NULL` detects it from the largest `_lagN` present |
| `argvar` | optional | **Exposure** basis (e.g., `list(fun = "ns", df = 4)`) |
| `arglag` | optional | **Lag** basis (e.g., `list(fun = "ns", df = 3)`) |
| `family` | optional | `"quasipoisson"` (default), `"poisson"`, `"negbin"` (→ quasi) |
| `ns_df` / `dof_per_year` | optional | Degrees of freedom of the time spline (seasonality) |
| `ref_value` | optional | Reference value (RR = 1); `NULL` uses the median |
| `covariates` | optional | Additional linear confounders (humidity, etc.) |

**Returns:** a `climasus_dlnm` object with `$model` (GLM), `$crossbasis`,
`$pred` (`crosspred` on a 100-point grid), `$exposure_response`,
`$lag_response`, `$models` (one-line summary), `$data_daily`, and
`$diagnostics` (dispersion, Ljung-Box autocorrelation, AIC).

## `sus_mod_af()` — Attributable Fraction and Number

| Argument | Description |
|-----------|-----------|
| `fit` | `climasus_dlnm` object from `sus_mod_dlnm()` |
| `threshold` | Threshold separating heat (above) and cold (below); `NULL` uses `ref_value` |
| `pred_at` | Percentiles of the ranges (default `c(0.75, 0.90, 0.95, 0.99)`) |
| `by` | Temporal breakdown: `"month"`, `"year"`, `"season"`, or `NULL` |
| `nsim` | Monte Carlo simulations for the CI (default `1000L`) |
| `alpha` | Significance level (default `0.05` → 95% CI) |

**Returns:** `climasus_af` with `$total` (total/heat/cold), `$by_quantile`,
`$by_period`, `$daily`, and `$custom`.

## `sus_mod_its()` — Interrupted Time Series

**Segmented** Poisson regression for the effect of interventions (policies,
extreme events, pandemics) on a count series (Bernal et al., 2017).

| Argument | Description |
|-----------|-----------|
| `data` | `climasus_df` (`aggregate`/`climate`) or `data.frame` with date + count |
| `outcome_col` / `date_col` | Outcome and date columns |
| `interruption_dates` | One or more intervention dates (within the range) |
| `harmonics` | Sinusoidal pairs for seasonality (default `2L`) |
| `family` | `"quasipoisson"` (default) or `"poisson"` |
| `counterfactual` | Projects the pre-intervention trend (default `TRUE`) |

**Returns:** `climasus_its` with `$effects` (level and trend change),
`$counterfactual`, and `$segments`.

## `sus_mod_casecrossover()` — Time-Stratified Case-Crossover

Each time stratum (month/week) acts as its own control, removing trend and
seasonality **by design** (Maclure, 1991; Armstrong et al., 2014).

| Argument | Description |
|-----------|-----------|
| `data` | `climasus_df` or `data.frame` with `date`, outcome, and exposure |
| `outcome_col` / `exposure_col` | Outcome and exposure (no auto-detection) |
| `stratum` | `"month"` (default), `"week"`, or an existing column |
| `lag` | Single integer lag or vector (moving average, e.g., `0:6`) |
| `method` | `"conditional_poisson"` (default) or `"clogit"` |
| `family` | `"quasipoisson"` (default) or `"poisson"` |

**Returns:** `climasus_casecrossover` with `$or_table` (OR, CI, p) and
`$diagnostics`.

## `sus_mod_pool()` — Multi-City Meta-Analysis

Combines DLNM estimates from several cities via **multivariate meta-analysis**
(REML via `mvmeta`), producing a pooled curve, heterogeneity (Q, I²), and
per-city BLUPs (Gasparrini et al., 2012).

| Argument | Description |
|-----------|-----------|
| `fits` | **Named** list of `climasus_dlnm` objects (same cross-basis) |
| `method` | `"reml"` (default), `"ml"`, or `"fixed"` |
| `blup` | Computes per-city BLUP predictions (default `TRUE`) |
| `pred_at` | Percentiles of the pooled table |

**Returns:** `climasus_pool` with `$exposure_response`, `$lag_response`,
`$heterogeneity`, `$city_table`, and `$blup_preds`.

# Statistical formulation

The DLNM models the daily count $Y_t$ as quasi-Poisson with mean $\mu_t$. The
log-linear predictor combines the **cross-basis** $\mathrm{cb}(x_t, \ell; \beta)$ —
the effect of the exposure history $x_{t}, x_{t-1}, \ldots, x_{t-L}$ — with
control for **seasonality** (natural spline of time) and **day of week**
(DOW):

$$
\log\,\mathbb{E}(Y_t) \;=\; \alpha
\;+\; \mathrm{cb}(x_t,\,\ell;\,\beta)
\;+\; \mathrm{ns}(\text{tempo})
\;+\; \mathrm{DOW}.
$$

The cross-basis is the outer product of a basis on exposure (`argvar`) and a
basis on lag (`arglag`). The **cumulative effect** across all lags, at
exposure level $x$ relative to the reference $x_0$, is

$$
RR_{\text{cum}}(x) \;=\;
\exp\!\left\{ \sum_{\ell=0}^{L} \big[ f(x,\ell) - f(x_0,\ell) \big] \right\}.
$$

The cumulative **attributable fraction** associated with the exposure history
follows the form (Gasparrini & Armstrong, 2013):

$$
AF \;=\; 1 - \exp\!\left( -\sum_{\ell=0}^{L} \beta_\ell\, x_{t-\ell} \right),
\qquad
AN_t \;=\; Y_t \left( 1 - \frac{1}{RR_t} \right),
$$

with the daily attributable number $AN_t$ summed over the period and the CI
obtained by Monte Carlo simulation of the cross-basis coefficients.

## Modeling decisions

- **Overdispersion.** Death counts typically have variance greater than the
  mean ($\hat\phi > 1$). The **quasi-Poisson** family inflates the standard
  error by $\sqrt{\hat\phi}$ and is recommended when the dispersion ratio
  exceeds ~1.5 (Armstrong, 2006). The negative binomial (NB) alternative is
  incompatible with `dlnm::crossbasis()`, and `climasus4r` redirects it to
  quasi-Poisson.
- **Choice of knots and lags.** For **exposure**, `ns` with 4 df captures the
  U/J shape (cold + heat) well; 3 df is more conservative, 5 df requires a
  large N. For **lag**, `ns` with 3 df models the smooth decay typical of the
  temperature-mortality relationship; a maximum lag of **21 days** covers the
  delayed, prolonged effect of cold (Gasparrini, 2011).
- **Seasonality.** The natural spline of time absorbs long-term trend and the
  annual cycle that confound the association. A value of ~`dof_per_year = 4`
  is recommended for 1–2 years and **6–8** for subtropical Brazil and
  diseases with strong seasonality (Bhaskaran et al., 2013). Underfitting
  leaves residual autocorrelation (Ljung-Box test in `$diagnostics`);
  overfitting consumes the signal.

# Illustrated results

```{r fig-surface, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/mod_dlnm_surface.png", .nota_gerar)
```

**Figure 1.** DLNM response surface: `log(RR)` of deaths in older adults as a
function of temperature (y-axis) and lag day (x-axis). Orange tones indicate
increased risk (heat, short lags); green tones, reduced risk. The
two-dimensional structure reveals when, after each exposure level, the effect
manifests.

```{r fig-overall, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/mod_dlnm_overall.png", .nota_gerar)
```

**Figure 2.** **Cumulative** exposure-response curve (RR and 95% CI) summed
over all lags, centered on the median temperature (reference, RR = 1). The
U/J shape — rising risk with both cold and heat — is the classic signature of
the temperature-mortality relationship.

```{r fig-lag, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/mod_dlnm_lag.png", .nota_gerar)
```

**Figure 3.** **Lag-response** curve at the 75th percentile of exposure:
lag-specific RR between temperature and death. It allows distinguishing the
acute effect (short lags) from the delayed, prolonged effect (long lags).

```{r fig-af, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/mod_af.png", .nota_gerar)
```

**Figure 4.** Attributable fraction of deaths (%) decomposed into **total**,
**heat**, and **cold**, with 95% Monte Carlo CI. In subtropical climates the
cold fraction tends to exceed the heat fraction, reflecting the prolonged
effect of low temperatures on older adults.

```{r tab-resumo, eval=TRUE, echo=FALSE}
tbl_ou_nota("dados/mod_resumo.rds", .nota_gerar,
            caption = "Table 1. Summary of the DLNM fit (specification, RR at P75, and diagnostics).")
```

```{r tab-af, eval=TRUE, echo=FALSE}
tbl_ou_nota("dados/mod_af_resumo.rds", .nota_gerar,
            caption = "Table 2. Attributable fraction (AF) and attributable number (AN) by thermal component.")
```

# References

- Armstrong, B. (2006). Models for the relationship between ambient temperature
  and daily mortality. *Epidemiology*, 17(6), 624–631.
- Bernal, J. L., Cummins, S., & Gasparrini, A. (2017). Interrupted time series
  regression for the evaluation of public health interventions: a tutorial.
  *International Journal of Epidemiology*, 46(1), 348–355.
  <https://doi.org/10.1093/ije/dyw098>
- Bhaskaran, K., Gasparrini, A., Hajat, S., Smeeth, L., & Armstrong, B. (2013).
  Time series regression studies in environmental epidemiology. *International
  Journal of Epidemiology*, 42(4), 1187–1195.
  <https://doi.org/10.1093/ije/dyt092>
- Gasparrini, A. (2011). Distributed lag linear and non-linear models in R: the
  package dlnm. *Journal of Statistical Software*, 43(8), 1–20.
  <https://doi.org/10.18637/jss.v043.i08>
- Gasparrini, A., Armstrong, B., & Kenward, M. G. (2010). Distributed lag
  non-linear models. *Statistics in Medicine*, 29(21), 2224–2234.
  <https://doi.org/10.1002/sim.3940>
- Gasparrini, A., & Armstrong, B. (2013). Reducing and meta-analysing estimates
  from distributed lag non-linear models. *BMC Medical Research Methodology*,
  14, 70. <https://doi.org/10.1186/1471-2288-14-70>
- Maclure, M. (1991). The case-crossover design: a method for studying transient
  effects on the risk of acute events. *American Journal of Epidemiology*,
  133(2), 144–153.

---

# Navigation

- [Back to tutorial hub](tutorials.html)
- **Previous:** [Tutorial 8 — Gridded Climate and Environmental Data](08-clima-grade.html)
- This is the **last tutorial** in the `climasus4r` pipeline series.
