---
title: "Bioclimatic Indices for Public Health"
subtitle: "Theory and INMET → Gap-filling → Thermal Indicators Pipeline | `sus_climate_compute_indicators()`"
author:
  - name: "Max Anjos"
  - name: "João Gobo"
  - name: "Thauã Menezes"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    code_download: true
    self_contained: true
    df_print: paged
---

```{css, echo=FALSE}
/* ── Variáveis de Cores Modernas ────────────────── */
:root {
  --primary-color: #1D9E75;
  --dark-green: #0f3d2e;
  --bg-light: #f8faf9;
  --text-main: #1a1a1a;
  --code-bg: #fdfdfd;
  --warning-orange: #ff9800;
  --info-blue: #2196f3;
}

html {
  scroll-behavior: smooth;
}

body {
  font-family: 'Source Serif 4', Georgia, serif;
  font-size: 18px;
  color: var(--text-main);
  background-color: #fff;
  line-height: 1.6;
}

/* ── Títulos com Inter-spacing ─────────────────── */
h1, h2, h3, h4 {
  font-family: 'Inter', sans-serif;
  font-weight: 700;
  color: var(--dark-green);
}

h1.title {
  font-size: 2.8em;
  border-bottom: 4px solid var(--primary-color);
  padding-bottom: 10px;
  margin-bottom: 1.5rem;
}

h2 {
  margin-top: 2rem;
  border-left: 4px solid var(--primary-color);
  padding-left: 1rem;
}

/* ── Code Blocks Estilizados ────────────────────── */
pre.r, pre.sourceCode {
  background-color: var(--code-bg);
  border: 1px solid #e1e1e1;
  border-left: 5px solid var(--primary-color) !important;
  border-radius: 8px;
  box-shadow: 0 2px 4px rgba(0,0,0,0.05);
  padding: 1rem;
}

/* ── Custom Callouts (Estilo Quarto) ────────────── */
.callout {
  padding: 1.2em 1.5em;
  margin: 1.5em 0;
  border-radius: 8px;
  border-left: 6px solid;
  background-color: var(--bg-light);
}

.callout-note {
  background-color: #f0f7ff;
  border-color: #007bff;
}

.callout-tip {
  background-color: #e8f5e9;
  border-color: #1D9E75;
}

.callout-warning {
  background-color: #fff3e0;
  border-color: #ff9800;
}

.callout-info {
  background-color: #e3f2fd;
  border-color: #2196f3;
}

/* ── Workflow container ─────────────────────────── */
.workflow-container {
  background: linear-gradient(135deg, #f8faf7 0%, #ffffff 100%);
  border-radius: 12px;
  padding: 1.5rem;
  margin: 1.5rem 0;
  border: 1px solid var(--fc-border);
  overflow-x: auto;
}

.workflow-steps {
  display: flex;
  flex-wrap: wrap;
  justify-content: center;
  gap: 0.5rem;
  font-family: 'Inter', monospace;
  font-size: 0.8rem;
}

.workflow-step {
  background: var(--primary-color);
  color: white;
  padding: 0.5rem 1rem;
  border-radius: 20px;
  white-space: nowrap;
}

.workflow-arrow {
  color: var(--primary-color);
  font-weight: bold;
  font-size: 1.2rem;
}

/* ── Feature cards ───────────────────────────────── */
.feature-grid {
  display: grid;
  grid-template-columns: repeat(auto-fit, minmax(280px, 1fr));
  gap: 1.5rem;
  margin: 2rem 0;
}

.feature-card {
  background: white;
  border-radius: 12px;
  padding: 1.5rem;
  border: 1px solid #e1e1e1;
  transition: transform 0.2s ease, box-shadow 0.2s ease;
  text-align: center;
}

.feature-card:hover {
  transform: translateY(-4px);
  box-shadow: 0 8px 24px rgba(0,0,0,0.1);
}

.feature-icon {
  font-size: 2rem;
  margin-bottom: 1rem;
}

.feature-card h3 {
  margin-top: 0;
  color: var(--primary-color);
  font-size: 1.1rem;
}

.feature-card p {
  font-size: 0.85rem;
  color: var(--text-main);
  margin-bottom: 0;
}

/* ── Table styling ───────────────────────────────── */
table {
  width: 100%;
  border-collapse: collapse;
  margin: 1.5rem 0;
}

th, td {
  padding: 12px;
  text-align: left;
  border-bottom: 1px solid #e1e1e1;
}

th {
  background-color: var(--bg-light);
  font-weight: 600;
  color: var(--dark-green);
}

tr:hover {
  background-color: var(--bg-light);
}

/* ── Figure styling ─────────────────────────────── */
.figure {
  margin: 1.5rem 0;
  text-align: center;
}

.figure img {
  border-radius: 8px;
  box-shadow: 0 4px 8px rgba(0,0,0,0.1);
  max-width: 100%;
  height: auto;
}

.figure .caption {
  font-size: 0.85rem;
  color: #666;
  margin-top: 8px;
  font-style: italic;
}

/* ── Botão GitHub aprimorado ───────────────────── */
.github-feedback-btn {
  display: inline-flex;
  align-items: center;
  gap: 8px;
  background-color: #24292e;
  color: white;
  padding: 10px 20px;
  border-radius: 6px;
  text-decoration: none;
  font-family: sans-serif;
  transition: background-color 0.2s;
  margin-top: 1rem;
}

.github-feedback-btn:hover {
  background-color: #2c3e50;
  color: white;
  text-decoration: none;
}

/* ── Responsive ─────────────────────────────────── */
@media (max-width: 768px) {
  body {
    font-size: 16px;
  }

  h1.title {
    font-size: 2rem;
  }

  .workflow-steps {
    flex-direction: column;
    align-items: center;
  }

  .workflow-arrow {
    transform: rotate(90deg);
  }

  .feature-grid {
    grid-template-columns: 1fr;
  }

  th, td {
    padding: 8px;
    font-size: 0.85rem;
  }
}

@media (prefers-reduced-motion: reduce) {
  .feature-card,
  .github-feedback-btn {
    transition: none;
  }
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = FALSE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.align = "center",
  out.width = "80%",
  fig.retina = 2
)
```


::: {.callout-box}
**About this tutorial**
This document demonstrates the full **climasus4r** pipeline for generating
bioclimatic indices from hourly INMET data: download → gap-filling →
computation of 15 indicators → visualization and regional categorization of
thermal risk.
:::

---

# Theoretical Foundations

## Why bioclimatic indices?

Air temperature (`T`) alone does not adequately represent human thermal stress. The thermoregulatory system responds to a **complex energy balance** that integrates:

*   **Air temperature** (`T`, °C)
*   **Relative humidity** (`RH`, %): affects the efficiency of sweat evaporation.
*   **Wind speed** (`WS`, m/s): modifies the convective boundary layer.
*   **Solar radiation** (`SR`, W/m²): adds radiant heat load.

Indices such as WBGT, UTCI, and PET translate this interaction into a single scalar value with direct physiological meaning, and are standard tools in occupational health, heat wave surveillance, and urban planning.

## 📊 Bioclimatic and Thermal Stress Indicators

| Code | Name | Input Variables | Domain of Validity | Main Reference |
|:---|:---|:---|:---|:---|
| `wbgt` | *Wet-Bulb Globe Temperature* | T, RH, SR, WS | General | Liljegren *et al.* (2008) |
| `hi` | *Heat Index* | T, RH | T ≥ 26.7 °C | Rothfusz (1990); NWS |
| `thi` | *Temperature-Humidity Index* | T, RH | General | Thom (1959) |
| `wcet` | *Wind Chill Equivalent Temp.* (CA) | T, WS | T ≤ 0 °C | Environment Canada (2001) |
| `wct` | *Wind Chill Temp.* (NWS) | T, WS | T ≤ 10 °C | NWS (2001) |
| `et` | *Effective Temperature* | T, RH, WS | General | Missenard (1937) *apud* Auliciems (2007) |
| `utci` | *Universal Thermal Climate Index* | T, RH, WS, SR | -60 to +50 °C | Bröde *et al.* (2012) |
| `pet` | *Physiological Equivalent Temp.* | T, RH, WS, SR | General | Matzarakis *et al.* (1999) |
| `cdd` | *Cooling Degree Days* | T | T > T_base | ASHRAE (2021) |
| `hdd` | *Heating Degree Days* | T | T < T_base | ASHRAE (2021) |
| `gdd` | *Growing Degree Days* | T | T_base < T < T_upper | McMaster & Wilhelm (1997) |
| `diurnal_range` | Diurnal Thermal Amplitude | T (min, max) | General | - |
| `vapor_pressure` | Actual Vapor Pressure | T, RH | General | Magnus-Tetens (1930) |
| `heat_stress_risk` | Thermal Stress Risk | WBGT, UTCI | Categorical | OSHA/ACGIH (2017) |
| `koppen_humidity` | Humidity Class (Köppen) | RH | Categorical | Köppen (1936) |


## Mathematical Formulas

### 1. WBGT (*Wet-Bulb Globe Temperature*)

Model implemented for outdoor environments under direct solar load.

<div class="equation-card">

**WBGT<sub>out</sub>** = 0.7 × T<sub>nwb</sub> + 0.2 × T<sub>g</sub> + 0.1 × T<sub>db</sub>

Where:

*   **T<sub>db</sub>** = Dry-bulb temperature (°C).
*   **T<sub>nwb</sub>** = Natural wet-bulb temperature (°C). For numerical robustness, the average of two estimates is used:
    *   *Liljegren*: `T_nwb_lilj = T * atan(0.16 * sqrt(e_a + 0.1)) + 3.0`
    *   *Bernard & Pourmoghani*: `T_nwb_bp = T + 0.33 * (RH/100) * exp(0.0514 * T) - 4.0`
    *   `T_nwb = (T_nwb_lilj + T_nwb_bp) / 2`
    *   Where `e_a` is the actual vapor pressure (kPa), calculated from T and RH.
*   **T<sub>g</sub>** = Globe temperature (°C), which models the radiant load:
    *   `T_g = T + 0.0144 * max(G_0, 0)^0.6 / WS^0.2 - 2.0`
    *   `G_0` is the solar irradiance in W/m², converted from SR (kJ/m²) via `SR / 3.6`.

**Weighting:** The coefficients 0.7, 0.2, and 0.1 represent the relative contribution of evaporative cooling (wet bulb), radiant load (globe), and sensible heat (dry bulb) to thermal stress in outdoor workers.
</div>

### 2. *Heat Index* (HI)

Implementation of the polynomial regression model from the *National Weather Service* (NWS), with adjustments for extreme humidity conditions.

<div class="equation-card">

**HI<sub>°F</sub>** = −42.379 + 2.04901523·T + 10.14333127·RH − 0.22475541·T·RH − 0.00683783·T² − 0.05481717·RH² + 0.00122874·T²·RH + 0.00085282·T·RH² − 0.00000199·T²·RH²

The final value is converted to Celsius: `HI = (HI°F - 32) * 5/9`.

**Validity Threshold Adapted by Biome:**
*   Standard: T ≥ 26.7 °C.
*   Tropical regions (Amazon, Atlantic Forest): T ≥ 24.0 °C, due to the acclimatization of local populations.

**Post-Calculation Adjustments:**
*   If RH < 13% and 26.7 ≤ T ≤ 44.4 °C: `HI = HI - ((13 - RH)/4) * sqrt((17 - |T - 95|) / 17)`.
*   If RH > 85% and 26.7 ≤ T ≤ 30.6 °C: `HI = HI + ((RH - 85)/10) * ((87 - T)/5)`.
</div>

### 3. *Universal Thermal Climate Index* (UTCI)

Polynomial approximation of the Fiala reference model, optimized for large-scale operational applications.

<div class="equation-card">

1.  **Mean Radiant Temperature (T<sub>mrt</sub>):** `T_mrt = T + 0.07 * G_0 - 1.5` (approximation for shortwave radiation).
2.  **Vapor Pressure (VP, kPa):** `VP = 0.6108 * exp(17.27 * T / (T + 237.3)) * (RH / 100)` @Tetens1930.
3.  **Wind Speed at 10 m (WS<sub>10m</sub>):** `WS_10m = max(WS * 1.5, 0.5)` (conversion from the standard 2 m height to 10 m).

**UTCI (°C) = T + 0.048 * ((VP*1000) - 1000)/100 - 0.29 * ln(WS_10m + 0.1) + 0.016 * ΔT_mrt + 0.010 * (ΔT_mrt²/10) - 6e-5 * ΔT_mrt² * WS_10m + δ<sub>regional</sub>**

Where `ΔT_mrt = T_mrt - T`. The term `δ<sub>regional</sub>` is an empirical correction calibrated for Brazilian biomes.
</div>

### 4. Degree Days (CDD, HDD, GDD)

<div class="equation-card">

*   **CDD** (*Cooling Degree Days*): `CDD = max(0, T - T_base_cdd)`. Regional base: 18–20 °C.
*   **HDD** (*Heating Degree Days*): `HDD = max(0, T_base_hdd - T)`. Regional base: 14–18 °C.
*   **GDD** (*Growing Degree Days*): `GDD = max(0, min(T, T_upper) - T_base_gdd)`. Regional base: 8–15 °C. Regional upper limit: 30–35 °C.

</div>

---

# Full Pipeline

## Step 1 · INMET Import

```{r step1-inmet, eval=FALSE}
library(climasus4r)
library(dplyr)
# ── Pipeline para 4 biomas representativos ───────────────────────────────────
# Amazônia (AM), Caatinga (CE), Mata Atlântica (SP), Pampa (RS)

df_am <- sus_climate_inmet(
  years   = 2023,
  uf      = "AM",          # Amazônia
  parallel = TRUE,
  workers  = 4,
  verbose  = TRUE
)

df_ce <- sus_climate_inmet(years = 2023, uf = "CE")   # Caatinga
df_sp <- sus_climate_inmet(years = 2023, uf = "SP")   # Mata Atlântica
df_rs <- sus_climate_inmet(years = 2023, uf = "RS")   # Pampa

# Inspecionar estrutura
head(df_am)
```

## Step 2 · Gap-Filling

```{r step2-fill, eval=FALSE}
# ── XGBoost gap-filling — todas as variáveis ─────────────────────────────────
df_am_filled <- df_am |>
  sus_climate_fill_inmet(
    target_var        = "all",
    quality_threshold = 0.4,
    parallel          = TRUE,
    workers           = 4,
    verbose           = TRUE
  )

df_ce_filled <- df_ce |> sus_climate_fill_inmet(target_var = "all")
df_sp_filled <- df_sp |> sus_climate_fill_inmet(target_var = "all")
df_rs_filled <- df_rs |> sus_climate_fill_inmet(target_var = "all")
```


## Step 3 · Computing the Indicators

```{r step3-indicators, eval=FALSE}
# ── Carregar função ───────────────────────────────────────────────────────────
# Em produção: já disponível via library(climasus4r)
# Aqui: source do arquivo do pacote
# source("sus_climate_compute_indicators.R")  # descomente se necessário

# ── Configuração de indicadores por bioma ────────────────────────────────────

# Amazônia — adaptação tropical, foco em estresse por calor e umidade
ind_am <- df_am_filled |>
  sus_climate_compute_indicators(
    region           = "amazon",
    confidence_flags = TRUE,
    verify_physics   = TRUE,
    verbose          = FALSE
  )

# Caatinga — semiárido, limiares rebaixados para RH
ind_ce <- df_ce_filled |>
  sus_climate_compute_indicators(
    region           = "caatinga",
    confidence_flags = TRUE,
    verbose          = FALSE
  )

# Mata Atlântica — subtropical úmido
ind_sp <- df_sp_filled |>
  sus_climate_compute_indicators(
    region           = "atlantic_forest",
    confidence_flags = TRUE,
    verbose          = FALSE
  )

# Pampa — subtropical temperado, wind chill relevante no inverno
ind_rs <- df_rs_filled |>
  sus_climate_compute_indicators(
    region           = "pampa",
    confidence_flags = TRUE,
    verbose          = FALSE
  )

# ── Combinar todos os biomas ──────────────────────────────────────────────────
ind_all <- bind_rows(
  mutate(ind_am, bioma = "Amazônia",      regiao = "Norte"),
  mutate(ind_ce, bioma = "Caatinga",      regiao = "Nordeste"),
  mutate(ind_sp, bioma = "Mata Atlântica",regiao = "Sudeste"),
  mutate(ind_rs, bioma = "Pampa",         regiao = "Sul")
)

glimpse(ind_all |> select(date, bioma, wbgt_c, hi_c, thi_c, utci_c,
                           pet_c, heat_stress_risk, koppen_humidity))
```

---

# Visualizations

## 4.1 Overview of the Indicators

```{r vis-overview, fig.height=9, eval=FALSE}
# ── Distribuições dos principais índices por bioma ───────────────────────────
library(tidyr)
library(ggplot2)
library(ggridges)
library(patchwork)
library(scales)


bioma_colors <- c(
  "Amazônia"       = "#1a7a4a",
  "Caatinga"       = "#c0392b",
  "Mata Atlântica" = "#2471a3",
  "Pampa"          = "#7d3c98"
)

ind_long <- ind_all |>
  select(date, bioma, wbgt_c, hi_c, thi_c, utci_c, pet_c, et_c) |>
  pivot_longer(
    cols      = c(wbgt_c, hi_c, thi_c, utci_c, pet_c, et_c),
    names_to  = "indice",
    values_to = "valor"
  ) |>
  filter(!is.na(valor)) |>
  mutate(indice = recode(indice,
    wbgt_c = "WBGT (°C)",
    hi_c   = "Heat Index (°C)",
    thi_c  = "THI (°C)",
    utci_c = "UTCI (°C)",
    pet_c  = "PET (°C)",
    et_c   = "ET (°C)"
  ))

ggplot(ind_long, aes(x = valor, y = bioma, fill = bioma)) +
  ggridges::geom_density_ridges(
    alpha = 0.75, scale = 0.9,
    quantile_lines = TRUE, quantiles = c(0.25, 0.5, 0.75),
    quantile_fun = function(x, ...) quantile(x, ...)
  ) +
  scale_fill_manual(values = bioma_colors, guide = "none") +
  facet_wrap(~indice, scales = "free_x", ncol = 3) +
  labs(
    title    = "Distribuição dos Índices Bioclimáticos por Bioma — 2023",
    subtitle = "Linhas internas = quartis (Q1, mediana, Q3)",
    x = "Valor do índice", y = NULL,
    caption  = "Fonte: INMET / climasus4r | Dados horários 2023"
  ) +
  theme(strip.background = element_rect(fill = "#eaf4fb", colour = "grey80"))
```

*Note: the block above uses `ggridges`; install it with `install.packages("ggridges")` if needed.*

```{r vis-overview-fallback, fig.height=8, eval=FALSE}
# Versão alternativa sem ggridges
ggplot(ind_long |> filter(!is.na(valor)),
       aes(x = valor, colour = bioma, fill = bioma)) +
  geom_density(alpha = 0.35, linewidth = 0.9) +
  scale_fill_manual(values  = bioma_colors, name = "Bioma") +
  scale_colour_manual(values = bioma_colors, name = "Bioma") +
  facet_wrap(~indice, scales = "free", ncol = 3) +
  labs(
    title   = "Distribuição dos Índices Bioclimáticos — 2023",
    x = "Valor do índice (°C equiv.)", y = "Densidade",
    caption = "Fonte: INMET / climasus4r"
  )
```

## 4.2 Seasonal Time Series

```{r vis-temporal, fig.height=8, eval=FALSE}
# ── Média diária de WBGT por bioma ───────────────────────────────────────────
diario <- ind_all |>
  mutate(data_d = as.Date(date)) |>
  group_by(bioma, data_d) |>
  summarise(
    wbgt_med  = mean(wbgt_c,  na.rm = TRUE),
    utci_med  = mean(utci_c,  na.rm = TRUE),    .groups   = "drop"
  )

# Bandas de risco WBGT (OSHA/ACGIH)
bandas_wbgt <- tibble(
  ymin  = c(-Inf, 20, 25, 28, 30, 33),
  ymax  = c(20,   25, 28, 30, 33, Inf),
  risco = factor(
    c("Nenhum","Baixo","Moderado","Alto","Muito Alto","Extremo"),
    levels = c("Nenhum","Baixo","Moderado","Alto","Muito Alto","Extremo")
  )
)
cores_risco <- c(
  "Nenhum"    = "#d5f5e3",
  "Baixo"     = "#a9dfbf",
  "Moderado"  = "#f9e79f",
  "Alto"      = "#f0b27a",
  "Muito Alto"= "#ec7063",
  "Extremo"   = "#922b21"
)

p1 <- ggplot() +
  geom_rect(
    data = bandas_wbgt,
    aes(xmin = as.Date("2022-12-31"), xmax = as.Date("2024-01-01"),
        ymin = ymin, ymax = ymax, fill = risco),
    alpha = 0.25
  ) +
  scale_fill_manual(values = cores_risco, name = "Risco WBGT") +
  #new_scale_fill() +
  geom_line(data = diario,
            aes(x = data_d, y = wbgt_med, colour = bioma), linewidth = 0.8) +
  scale_colour_manual(values = bioma_colors, name = "Bioma") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  labs(
    title    = "WBGT Médio Diário — Biomas Brasileiros 2023",
    subtitle = "Bandas de risco: OSHA/ACGIH (adaptadas por bioma)",
    x = NULL, y = "WBGT (°C)",
    caption  = "Zonas de risco: Baixo = 20–25°C | Moderado = 25–28°C | Alto = 28–30°C | Muito Alto = 30–33°C | Extremo > 33°C"
  )

# UTCI e WBGT sobrepostos para Caatinga
p2 <- diario |>
  filter(bioma == "Caatinga") |>
  pivot_longer(c(wbgt_med, utci_med),
               names_to = "var", values_to = "val") |>
  mutate(var = recode(var,
    wbgt_med = "WBGT", utci_med = "UTCI")) |>
  ggplot(aes(data_d, val, colour = var)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = c("WBGT"="#c0392b","UTCI"="#8e44ad","T ar"="#2980b9"),
                      name = "Índice") +
  scale_x_date(date_labels = "%b", date_breaks = "2 months") +
  labs(
    title    = "Caatinga — WBGT vs UTCI vs T ar",
    subtitle = "WBGT e UTCI consistentemente acima da temperatura seca",
    x = NULL, y = "°C"
  )

p1 / p2
```


## 4.4 Regional Risk Categories

```{r vis-risco, fig.height=7, eval=FALSE}
# ── Distribuição mensal das categorias de risco por bioma ────────────────────
risco_mensal <- ind_all |>
  mutate(mes = month(date, label = TRUE, abbr = TRUE, locale = "pt_BR.UTF-8")) |>
  group_by(bioma, mes) |>
  count(heat_stress_risk) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup() |>
  mutate(
    heat_stress_risk = factor(heat_stress_risk,
      levels = c("None","Low","Moderate","High","Very High","Extreme")),
    risco_label = recode(heat_stress_risk,
      "None"      = "Nenhum",
      "Low"       = "Baixo",
      "Moderate"  = "Moderado",
      "High"      = "Alto",
      "Very High" = "Muito Alto",
      "Extreme"   = "Extremo"
    ),
    risco_label = factor(risco_label,
      levels = c("Nenhum","Baixo","Moderado","Alto","Muito Alto","Extremo"))
  )

ggplot(risco_mensal, aes(mes, pct, fill = risco_label)) +
  geom_col(position = "stack", width = 0.85, colour = "white", linewidth = 0.2) +
  scale_fill_manual(
    values = c(
      "Nenhum"    = "#abebc6",
      "Baixo"     = "#82e0aa",
      "Moderado"  = "#f9e79f",
      "Alto"      = "#f0b27a",
      "Muito Alto"= "#ec7063",
      "Extremo"   = "#7b241c"
    ),
    name = "Risco\ntérmico"
  ) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  facet_wrap(~bioma, ncol = 2) +
  labs(
    title    = "Composição Mensal do Risco de Estresse Térmico (WBGT-base)",
    subtitle = "Adaptado por bioma: limiares OSHA/ACGIH + correção regional",
    x = NULL, y = "% horas por categoria",
    caption  = paste("Caatinga: pico de risco Extremo em set–out (seca)",
                     "| Amazônia: risco Alto-Muito Alto persistente",
                     "| Pampa: dominado por Nenhum no inverno",
                     sep = "\n")
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 4.5 Agricultural Degree Days

```{r vis-gdd, fig.height=5, eval=FALSE}
# ── Acumulado de GDD, CDD e HDD por bioma ────────────────────────────────────
grausdia <- ind_all |>
  mutate(data_d = as.Date(date)) |>
  group_by(bioma, data_d) |>
  summarise(
    cdd = sum(cdd_c, na.rm = TRUE),
    hdd = sum(hdd_c, na.rm = TRUE),
    gdd = sum(gdd_c, na.rm = TRUE),
    .groups = "drop"
  ) |>
  group_by(bioma) |>
  arrange(data_d) |>
  mutate(
    cdd_acum = cumsum(cdd),
    hdd_acum = cumsum(hdd),
    gdd_acum = cumsum(gdd)
  ) |>
  ungroup()

p_gdd1 <- grausdia |>
  select(bioma, data_d, cdd_acum, hdd_acum) |>
  pivot_longer(c(cdd_acum, hdd_acum), names_to = "tipo", values_to = "val") |>
  mutate(tipo = recode(tipo,
    cdd_acum = "CDD (resfriamento)",
    hdd_acum = "HDD (aquecimento)"
  )) |>
  ggplot(aes(data_d, val, colour = bioma, linetype = tipo)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = bioma_colors, name = "Bioma") +
  scale_linetype_manual(values = c("solid","dashed"), name = NULL) +
  scale_x_date(date_labels = "%b") +
  labs(
    title = "Graus-dia Acumulados — CDD e HDD",
    subtitle = "Base regional (ASHRAE adaptado): 18–20°C para CDD | 14–18°C para HDD",
    x = NULL, y = "Graus-dia acumulados"
  )

p_gdd2 <- grausdia |>
  ggplot(aes(data_d, gdd_acum, colour = bioma)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = bioma_colors, name = "Bioma") +
  scale_x_date(date_labels = "%b") +
  labs(
    title = "GDD Acumulados — Potencial Agrícola",
    subtitle = "Base 8–15°C (regional) | Limite superior 30–35°C",
    x = NULL, y = "GDD acumulados"
  )

p_gdd1 | p_gdd2
```


## 4.7 Thermal Amplitude and Vapor Pressure

```{r vis-diurnal-vapor, fig.height=5, eval=FALSE}
vapor_dtr <- ind_all |>
  mutate(
    mes = month(date),
    trimestre = case_when(
      mes %in% 1:3  ~ "T1 (Jan–Mar)",
      mes %in% 4:6  ~ "T2 (Abr–Jun)",
      mes %in% 7:9  ~ "T3 (Jul–Set)",
      TRUE          ~ "T4 (Out–Dez)"
    )
  ) |>
  filter(!is.na(diurnal_range_c), !is.na(vapor_pressure_kpa))

p_vap <- ggplot(vapor_dtr,
       aes(diurnal_range_c, vapor_pressure_kpa, colour = bioma)) +
  geom_point(alpha = 0.08, size = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.1) +
  scale_colour_manual(values = bioma_colors, name = "Bioma") +
  scale_y_continuous(labels = label_number(suffix = " kPa")) +
  labs(
    title    = "Amplitude Térmica Diurna × Pressão de Vapor Atual",
    subtitle = "Caatinga: alta amplitude + baixa pressão de vapor (aridez)\nAmazônia: baixa amplitude + alta pressão (carga evaporativa máxima)",
    x = "DTR — Amplitude Diurna (°C)", y = "Pressão de vapor ea (kPa)"
  ) +
  facet_wrap(~bioma, ncol = 2) +
  theme(legend.position = "none")

p_vap
```

## 4.8 Categorical Classifications

```{r vis-categorias, fig.height=5, eval=FALSE}
# ── Köppen Humidity e Heat Stress Risk — proporção anual por bioma ────────────

koppen_prop <- ind_all |>
  count(bioma, koppen_humidity) |>
  group_by(bioma) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup()

risk_anual <- ind_all |>
  count(bioma, heat_stress_risk) |>
  group_by(bioma) |>
  mutate(
    pct = n / sum(n) * 100,
    heat_stress_risk = factor(heat_stress_risk,
      levels = c("None","Low","Moderate","High","Very High","Extreme"))
  ) |>
  ungroup()

p_kopp <- ggplot(koppen_prop, aes(bioma, pct, fill = koppen_humidity)) +
  geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(
    values = c("Arid"="#f0b27a","Semi-arid"="#f9e79f",
               "Humid"="#82e0aa","Perhumid"="#2ecc71"),
    name = "Classe Köppen"
  ) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  coord_flip() +
  labs(
    title = "Classificação Köppen de Umidade",
    subtitle = "% horas anuais por classe",
    x = NULL, y = "% horas"
  )

p_risk <- ggplot(risk_anual, aes(bioma, pct, fill = heat_stress_risk)) +
  geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(
    values = c(
      "None"      = "#abebc6",
      "Low"       = "#82e0aa",
      "Moderate"  = "#f9e79f",
      "High"      = "#f0b27a",
      "Very High" = "#ec7063",
      "Extreme"   = "#7b241c"
    ),
    name = "Risco Térmico\n(WBGT-base)"
  ) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  coord_flip() +
  labs(
    title = "Risco de Estresse Térmico (OSHA/ACGIH)",
    subtitle = "% horas anuais por nível de risco",
    x = NULL, y = "% horas"
  )

p_kopp | p_risk
```

---

## Effect of the Regional Threshold on HI

```{r regional-hi-effect, fig.height=5, eval=FALSE}
# ── Demonstração do efeito do limiar regional no HI ──────────────────────────
t_seq  <- seq(22, 38, 0.2)
rh_val <- 65

calc_hi_vec <- function(airT, rh, min_temp) {
  tf <- (airT * 9/5) + 32
  hi_f <- -42.379 + 2.04901523*tf + 10.14333127*rh - 0.22475541*tf*rh -
           6.83783e-3*tf^2 - 5.481717e-2*rh^2 + 1.22874e-3*tf^2*rh +
           8.5282e-4*tf*rh^2 - 1.99e-6*tf^2*rh^2
  hi_c <- (hi_f - 32) * 5 / 9
  ifelse(airT < min_temp, NA, hi_c)
}

df_hi_reg <- tibble(
  airT = rep(t_seq, 3),
  bioma = rep(c("Sul/Pampa (26°C)","Padrão (26.7°C)","Amazônia (24°C)"), each = length(t_seq)),
  hi    = c(
    calc_hi_vec(t_seq, rh_val, 26.0),
    calc_hi_vec(t_seq, rh_val, 26.7),
    calc_hi_vec(t_seq, rh_val, 24.0)
  )
)

ggplot(df_hi_reg, aes(airT, hi, colour = bioma)) +
  geom_line(linewidth = 1.3, na.rm = TRUE) +
  geom_vline(xintercept = c(24.0, 26.0, 26.7),
             linetype = "dotted", colour = c("#1a7a4a","#7d3c98","#2471a3"),
             linewidth = 0.8) +
  annotate("text", x = c(24.2, 26.2, 26.9), y = 22,
           label = c("Amazônia\n24°C","Sul\n26°C","Padrão\n26.7°C"),
           hjust = 0, size = 3, colour = c("#1a7a4a","#7d3c98","#2471a3")) +
  scale_colour_manual(
    values = c("Sul/Pampa (26°C)"="#7d3c98",
               "Padrão (26.7°C)"="#2471a3",
               "Amazônia (24°C)"="#1a7a4a"),
    name = "Limiar regional"
  ) +
  labs(
    title    = "Impacto do Limiar Regional no Heat Index (RH = 65%)",
    subtitle = "Populações tropicais aclimatadas percebem desconforto a T mais baixas",
    x = "Temperatura do ar (°C)", y = "Heat Index (°C)",
    caption  = "Rothfusz (1990) — limiar padrão 26.7°C | adaptado: 24–26°C para trópicos"
  )
```

---

# Uncertainty & Confidence Flags

## Uncertainty Estimation (Monte Carlo)

```{r uncertainty-demo, eval=FALSE}
# ── Incerteza para São Paulo — 200 simulações ─────────────────────────────────
ind_sp_unc <- df_sp_filled |>
  sus_climate_compute_indicators(
    indicators          = c("wbgt", "hi", "utci"),
    region              = "atlantic_forest",
    compute_uncertainty = TRUE,      # ← 200 simulações MC
    confidence_flags    = TRUE,
    verbose             = TRUE
  )

# Colunas geradas: wbgt_c, wbgt_c_ci_low, wbgt_c_ci_high
# Interpretação: IC 95% propagando incertezas T±0.5°C, RH±3%, WS±0.3 m/s, SR±10%

ind_sp_unc |>
  filter(!is.na(wbgt_c_ci_low)) |>
  mutate(ci_width = wbgt_c_ci_high - wbgt_c_ci_low) |>
  summary()
```


## Confidence Flags

```{r flags-demo, fig.height=4, eval=FALSE}
# ── Visualizar os flags gerados ───────────────────────────────────────────────
flags_summary <- ind_all |>
  select(bioma, date, wbgt_c,
         wbgt_c_flag_extreme, wbgt_c_flag_high, wbgt_c_flag_low) |>
  group_by(bioma) |>
  summarise(
    pct_extreme = 100 * mean(wbgt_c_flag_extreme, na.rm = TRUE),
    pct_high    = 100 * mean(wbgt_c_flag_high,    na.rm = TRUE),
    pct_low     = 100 * mean(wbgt_c_flag_low,     na.rm = TRUE),
    .groups     = "drop"
  ) |>
  pivot_longer(-bioma, names_to = "flag", values_to = "pct") |>
  mutate(
    flag = recode(flag,
      pct_extreme = "Flag: Extremo (> 31°C)",
      pct_high    = "Flag: Alto (> 28°C)",
      pct_low     = "Flag: Frio (< 15°C)"
    )
  )

ggplot(flags_summary, aes(bioma, pct, fill = flag)) +
  geom_col(position = "dodge", width = 0.7,
           colour = "white", linewidth = 0.3) +
  scale_fill_manual(
    values = c("Flag: Extremo (> 31°C)"="#922b21",
               "Flag: Alto (> 28°C)"   ="#e67e22",
               "Flag: Frio (< 15°C)"  ="#2980b9"),
    name = NULL
  ) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(
    title    = "Frequência dos Flags de Confiança WBGT por Bioma",
    subtitle = "Flags: _flag_extreme | _flag_high | _flag_low — gerados automaticamente",
    x = NULL, y = "% horas com flag ativo",
    caption  = paste("Limiares adaptados por bioma:",
                     "Amazônia: extremo>31°C | Caatinga: extremo>33°C | Sul: frio<15°C",
                     sep = "\n")
  )
```

---

# Advanced Usage Patterns

## Full Integration into the SUS Pipeline

```{r pipeline-completo, eval=FALSE}
# ── Pipeline de produção para análise epidemiológica ─────────────────────────

library(climasus4r)

# 1. Download multi-UF, multi-ano
df_raw <- sus_climate_inmet(
  years    = 2020:2023,
  uf       = c("AM","PA","CE","BA","SP","MG","RS"),
  parallel = TRUE,
  workers  = 6
)

# 2. Preenchimento XGBoost
df_filled <- df_raw |>
  sus_climate_fill_inmet(
    target_var        = "all",
    quality_threshold = 0.40,
    parallel          = TRUE,
    workers           = 6
  )

# 3. Indicadores com adaptação regional automática
df_indicators <- df_filled |>
  sus_climate_compute_indicators(
    region              = "auto",       # detecta bioma por UF/latitude
    confidence_flags    = TRUE,
    verify_physics      = TRUE,
    compute_uncertainty = FALSE,        # TRUE para análise de sensibilidade
    keep_source_vars    = FALSE
  )

# 4. Resumo mensal por estação — para linkage com dados de mortalidade SIM
resumo_mensal <- df_indicators |>
  mutate(
    ano_mes = format(date, "%Y-%m"),
    station = station_code
  ) |>
  group_by(station, ano_mes) |>
  summarise(
    across(c(wbgt_c, hi_c, utci_c, pet_c, thi_c),
           list(med = ~mean(.x, na.rm=TRUE),
                p90 = ~quantile(.x, 0.90, na.rm=TRUE),
                p95 = ~quantile(.x, 0.95, na.rm=TRUE)),
           .names = "{.col}_{.fn}"),
    horas_risco_alto    = sum(heat_stress_risk %in% c("High","Very High","Extreme"),
                              na.rm = TRUE),
    horas_risco_extremo = sum(heat_stress_risk == "Extreme", na.rm = TRUE),
    cdd_acum            = sum(cdd_c, na.rm = TRUE),
    gdd_acum            = sum(gdd_c, na.rm = TRUE),
    .groups             = "drop"
  )
```

## Custom Thresholds for Research

```{r custom-thresholds, eval=FALSE}
# ── Thresholds customizados para estudo específico ───────────────────────────
ind_custom <- df_am_filled |>
  sus_climate_compute_indicators(
    region = "amazon",
    custom_thresholds = list(
      hi_min_temp = 22.0,   # limiar ainda mais conservador
      hi_min_rh   = 35,
      cdd_base    = 22,     # base climática local calibrada
      gdd_base    = 18,
      gdd_upper   = 33
    ),
    confidence_flags = TRUE
  )
```

## Selected Indicators Only

```{r selected-indicators, eval=FALSE}
# ── Cálculo seletivo — mais eficiente para análises focadas ──────────────────

# Só indicadores de saúde pública ocupacional
ind_ocupacional <- df_ce_filled |>
  sus_climate_compute_indicators(
    indicators       = c("wbgt", "utci", "heat_stress_risk"),
    region           = "caatinga",
    confidence_flags = TRUE,
    verbose          = FALSE
  )

# Só graus-dia para modelagem energética/agrícola
ind_energia <- df_rs_filled |>
  sus_climate_compute_indicators(
    indicators = c("cdd", "hdd", "gdd", "diurnal_range"),
    region     = "pampa",
    verbose    = FALSE
  )

# Só indicadores que não dependem de radiação solar (dados incompletos)
ind_sem_sr <- df_sp_filled |>
  sus_climate_compute_indicators(
    indicators = c("hi", "thi", "et", "vapor_pressure", "koppen_humidity"),
    region     = "atlantic_forest",
    verbose    = FALSE
  )

cat("Colunas geradas (ocupacional):", names(ind_ocupacional), "\n")
```

---

# Risk Category Reference Table

### 🌡️ Thermal Risk Classification by Bioclimatic Indices

| Index | Category | Threshold (°C) | Physiological Meaning | Source |
|--------|----------|-------------|--------------------------|-------|
| WBGT | None | < 20 | No perceptible thermal risk | OSHA/ACGIH |
| WBGT | Low | 20–25 | Mild stress during intense activity | OSHA/ACGIH |
| WBGT | Moderate | 25–28 | Rest breaks recommended for workers | OSHA/ACGIH |
| WBGT | High | 28–30 | Real risk of thermal disorders | OSHA/ACGIH |
| WBGT | Very High | 30–33 | Severe reduction in working time | OSHA/ACGIH |
| WBGT | Extreme | > 33 | Risk of heat collapse | OSHA/ACGIH |
| UTCI | Extreme cold | < −27 | Imminent hypothermia | Bröde et al. (2012) |
| UTCI | Neutral | 9–26 | Thermal comfort zone | Bröde et al. (2012) |
| UTCI | Mild heat | 26–32 | Mild heat stress | Bröde et al. (2012) |
| UTCI | Strong heat | 32–38 | Moderate heat stress | Bröde et al. (2012) |
| UTCI | Extreme heat | > 46 | Extreme thermal stress | Bröde et al. (2012) |
| PET | Extreme cold | < 4 | Thermal discomfort from cold | Matzarakis et al. (1999) |
| PET | Comfort | 18–23 | Neutral thermal zone | Matzarakis et al. (1999) |
| PET | Mild heat | 23–29 | Mild heat discomfort | Matzarakis et al. (1999) |
| PET | Extreme heat | > 41 | Very strong thermal stress | Matzarakis et al. (1999) |
| HI | Caution | 27–32 | Fatigue possible with prolonged exposure | NWS |
| HI | Extreme danger | > 54 | Heat stroke highly likely | NWS |
---

# Scientific Considerations and Limitations

::: {.warning-box}
**⚠️ Important Limitations to Report**

1. **UTCI and PET are approximations** — the full UTCI (Fiala) requires a
   polynomial with 210+ terms; this implementation may diverge by 3–5°C under
   extreme conditions (Bröde et al., 2012). For critical epidemiological
   studies, use `reticulate` + `pythermalcomfort` for the full UTCI.

2. **Nighttime solar radiation = 0** — NA values in `sr_kj_m2` during the
   night are imputed as 0 (no solar load). Nighttime WBGT reflects only
   temperature, humidity, and wind — physically correct.

3. **PET and clothing** — the simplified model assumes average seasonal
   clothing (clo 0.5–0.9); populations with very different habits (e.g.,
   miners, rural workers) require specific calibration.

4. **Regional adaptations are empirical** — the offsets (+1.5°C Amazon,
   −0.5°C Caatinga, etc.) reflect bibliographic consensus; local validation
   against globe thermometer measurements is recommended for publications.

5. **INMET data** — automatic stations have maintenance issues; XGBoost
   gap-filling reduces gaps but does not eliminate measurement uncertainty.
   Use `compute_uncertainty = TRUE` to propagate this uncertainty.
:::

---

# References

- Alduchov, O. A., & Eskridge, R. E. (1996).
  Improved Magnus form approximation of saturation vapor pressure. Journal of Applied Meteorology, 35, 601–609.

- Bernard, T. E., & Pourmoghani, M. (1999).
  Prediction of workplace wet bulb globe temperature. *AIHAJ, 60*(1), 32–37.

- Bröde, P., Fiala, D., Błażejczyk, K., Holmér, I., Jendritzky, G., Kampmann, B., et al. (2012).
  Deriving the operational procedure for the Universal Thermal Climate Index (UTCI). *International Journal of Biometeorology, 56*(4), 481–494.

- International Organization for Standardization. (1989).
  *ISO 7243: Hot environments—Estimation of the heat stress on working man based on the WBGT-index*.

- Joint Action Group for Temperature Indices. (2001). Development of the wind chill temperature index.

- Liljegren, J. C., Carhart, R. A., Lawday, P., Tschopp, S., & Sharp, R. (2008).
  Modeling the wet bulb globe temperature using standard meteorological measurements. Journal of Occupational and Environmental Hygiene, 5(10), 645–655.

- Magnus, G. (1844). Versuche über die spannkräfte des wasserdampfs.

- Matzarakis, A., Mayer, H., & Iziomon, M. G. (1999).
  Applications of a universal thermal index: Physiological equivalent temperature. International Journal of Biometeorology, 43, 76–84.

- Missenard, A. (1937).
  *Température sensible*.

- McMaster, G. S., & Wilhelm, W. W. (1997).
  Growing degree-days: One equation, two interpretations. Agricultural and Forest Meteorology, 87, 291–300.

- Missenard, A. (1937). Température sensible.

- National Weather Service. (n.d.).
  Heat index equation. https://www.wpc.ncep.noaa.gov/html/heatindex_equation.shtml

- Occupational Safety and Health Administration. (2017). Protecting workers from the effects of heat (OSHA 3154).

- Rothfusz, L. P. (1990).
  The heat index equation (Technical Attachment SR/SSD 90-23). National Weather Service.

- Tetens, O. (1930).
  Über einige meteorologische Begriffe. Zeitschrift für Geophysik, 6, 297–309.

- Thom, E. C. (1959). The discomfort index. Weatherwise, 12(2), 57–59.
---

## 🌲 Have feedback or suggestions?

Do you have an idea for improvement or found an error? Click the button below to open a new *issue* on GitHub and share your suggestions directly with us.


```{=html}
<a href='https://github.com/ByMaxAnjos/climasus4r/issues/new' class="github-feedback-btn">
  <svg xmlns="http://www.w3.org/2000/svg" width="18" height="18" viewBox="0 0 24 24" fill="currentColor">
    <path d="M12 0C5.37 0 0 5.37 0 12c0 5.31 3.435 9.795 8.205 11.385.6.105.825-.255.825-.57 0-.285-.015-1.23-.015-2.235-3.015.555-3.795-.735-4.035-1.41-.135-.345-.72-1.41-1.23-1.695-.42-.225-1.02-.78-.015-.795.945-.015 1.62.87 1.845 1.23 1.08 1.815 2.805 1.305 3.495.99.105-.78.42-1.305.765-1.605-2.67-.3-5.46-1.335-5.46-5.925 0-1.305.465-2.385 1.23-3.225-.12-.3-.54-1.53.12-3.18 0 0 1.005-.315 3.3 1.23.96-.27 1.98-.405 3-.405s2.04.135 3 .405c2.295-1.56 3.3-1.23 3.3-1.23.66 1.65.24 2.88.12 3.18.765.84 1.23 1.905 1.23 3.225 0 4.605-2.805 5.625-5.475 5.925.435.375.81 1.095.81 2.22 0 1.605-.015 2.895-.015 3.3 0 .315.225.69.825.57A12.02 12.02 0 0 0 24 12c0-6.63-5.37-12-12-12z"/>
  </svg>
  Open issue on GitHub
</a>
```
