---
title: "Case Study 3: Cardiovascular Mortality in Older Adults and Heatwaves in São Paulo"
subtitle: "Full SIM-DO + INMET + heatwaves pipeline | São Paulo | 2016–2023"
author:
  - name: "Dr. Max Anjos"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    code_download: true
    self_contained: true
    df_print: paged
---

```{css, echo=FALSE}
/* ── Variáveis de Cores Modernas ────────────────── */
:root {
  --primary-color: #1D9E75;
  --dark-green: #0f3d2e;
  --bg-light: #f8faf9;
  --text-main: #1a1a1a;
  --code-bg: #fdfdfd;
  --warning-orange: #ff9800;
  --info-blue: #2196f3;
  --pipeline-purple: #9b59b6;
}

html {
  scroll-behavior: smooth;
}

body {
  font-family: 'Source Serif 4', Georgia, serif;
  font-size: 18px;
  color: var(--text-main);
  background-color: #fff;
  line-height: 1.6;
}

/* ── Títulos com Inter-spacing ─────────────────── */
h1, h2, h3, h4 {
  font-family: 'Inter', sans-serif;
  font-weight: 700;
  color: var(--dark-green);
}

h1.title {
  font-size: 2.8em;
  border-bottom: 4px solid var(--primary-color);
  padding-bottom: 10px;
  margin-bottom: 1.5rem;
}

h2 {
  margin-top: 2rem;
  border-left: 4px solid var(--primary-color);
  padding-left: 1rem;
}

/* ── Code Blocks Estilizados ────────────────────── */
pre.r, pre.sourceCode {
  background-color: var(--code-bg);
  border: 1px solid #e1e1e1;
  border-left: 5px solid var(--primary-color) !important;
  border-radius: 8px;
  box-shadow: 0 2px 4px rgba(0,0,0,0.05);
  padding: 1rem;
}

/* ── Custom Callouts (Estilo Quarto) ────────────── */
.callout {
  padding: 1.2em 1.5em;
  margin: 1.5em 0;
  border-radius: 8px;
  border-left: 6px solid;
  background-color: var(--bg-light);
}

.callout-note {
  background-color: #f0f7ff;
  border-color: #007bff;
}

.callout-tip {
  background-color: #e8f5e9;
  border-color: #1D9E75;
}

.callout-warning {
  background-color: #fff3e0;
  border-color: #ff9800;
}

.callout-info {
  background-color: #e3f2fd;
  border-color: #2196f3;
}

.callout-important {
  background-color: #fef2e8;
  border-color: #ff6b6b;
}

/* ── Pipeline cards ───────────────────────────────── */
.pipeline-grid {
  display: grid;
  grid-template-columns: repeat(auto-fit, minmax(280px, 1fr));
  gap: 1.5rem;
  margin: 2rem 0;
}

.pipeline-card {
  background: white;
  border-radius: 12px;
  padding: 1.5rem;
  border: 1px solid #e1e1e1;
  transition: transform 0.2s ease, box-shadow 0.2s ease;
  cursor: pointer;
}

.pipeline-card:hover {
  transform: translateY(-4px);
  box-shadow: 0 8px 24px rgba(0,0,0,0.1);
}

.pipeline-icon {
  font-size: 2rem;
  margin-bottom: 1rem;
}

.pipeline-card h3 {
  margin-top: 0;
  color: var(--primary-color);
  font-size: 1.1rem;
}

.pipeline-card p {
  font-size: 0.85rem;
  color: var(--text-main);
  margin-bottom: 0;
}

.pipeline-badge {
  display: inline-block;
  background-color: var(--primary-color);
  color: white;
  font-size: 0.7rem;
  padding: 2px 8px;
  border-radius: 12px;
  margin-top: 0.5rem;
}

/* ── Table styling ───────────────────────────────── */
table {
  width: 100%;
  border-collapse: collapse;
  margin: 1.5rem 0;
}

th, td {
  padding: 12px;
  text-align: left;
  border-bottom: 1px solid #e1e1e1;
}

th {
  background-color: var(--bg-light);
  font-weight: 600;
  color: var(--dark-green);
}

tr:hover {
  background-color: var(--bg-light);
}

/* ── Workflow container ─────────────────────────── */
.workflow-container {
  background: linear-gradient(135deg, #f8faf7 0%, #ffffff 100%);
  border-radius: 12px;
  padding: 1.5rem;
  margin: 1.5rem 0;
  border: 1px solid rgba(16, 185, 129, 0.2);
  overflow-x: auto;
}

.workflow-steps {
  display: flex;
  flex-wrap: wrap;
  justify-content: center;
  gap: 0.5rem;
  font-family: 'Inter', monospace;
  font-size: 0.8rem;
}

.workflow-step {
  background: var(--primary-color);
  color: white;
  padding: 0.5rem 1rem;
  border-radius: 20px;
  white-space: nowrap;
}

.workflow-arrow {
  color: var(--primary-color);
  font-weight: bold;
  font-size: 1.2rem;
}

/* ── Comparison container ───────────────────────── */
.comparison-container {
  display: grid;
  grid-template-columns: 1fr 1fr;
  gap: 1.5rem;
  margin: 1.5rem 0;
}

.comparison-box {
  background: var(--bg-light);
  border-radius: 8px;
  padding: 1rem;
}

.comparison-box pre {
  margin: 0;
  font-size: 0.8rem;
}

/* ── Botão GitHub aprimorado ───────────────────── */
.github-feedback-btn {
  display: inline-flex;
  align-items: center;
  gap: 8px;
  background-color: #24292e;
  color: white;
  padding: 10px 20px;
  border-radius: 6px;
  text-decoration: none;
  font-family: sans-serif;
  transition: background-color 0.2s;
  margin-top: 1rem;
}

.github-feedback-btn:hover {
  background-color: #2c3e50;
  color: white;
  text-decoration: none;
}

/* ── Responsive ─────────────────────────────────── */
@media (max-width: 768px) {
  body {
    font-size: 16px;
  }

  h1.title {
    font-size: 2rem;
  }

  .workflow-steps {
    flex-direction: column;
    align-items: center;
  }

  .workflow-arrow {
    transform: rotate(90deg);
  }

  .pipeline-grid {
    grid-template-columns: 1fr;
  }

  .comparison-container {
    grid-template-columns: 1fr;
  }

  th, td {
    padding: 8px;
    font-size: 0.85rem;
  }
}

@media (prefers-reduced-motion: reduce) {
  .pipeline-card,
  .github-feedback-btn {
    transition: none;
  }
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = FALSE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.align = "center",
  out.width = "80%",
  fig.retina = 2
)
```

# Objective

This case study quantifies excess cardiovascular mortality in older adults aged 65 and over during heatwave periods in São Paulo, from 2016 to 2023. The pipeline integrates daily SIM-DO data with automatic heatwave detection using the WHO (World Health Organization) and WMO (World Meteorological Organization) criteria, through the `sus_climate_compute_heatwaves()` function. The final product is a daily dataset with heatwave flags by method, ready for time series modeling with control for seasonality and trend.

# Scientific Context

Cardiovascular disease is the leading cause of death in the Brazilian population aged 55 and over, accounting for more than 30% of deaths in metropolitan regions. During periods of extreme heat, the human body activates thermoregulatory mechanisms — peripheral vasodilation, tachycardia, and sweating — that increase cardiac demand. In older adults, whose thermoregulatory capacity and cardiac reserve are reduced, this physiological stress can precipitate acute myocardial infarction, heart failure, and stroke (Gasparrini et al., 2017). Multicountry studies estimate that the relative risk of cardiovascular death increases by 20–30% during heatwaves in middle-income country populations (Vicedo-Cabrera et al., 2021).

São Paulo is simultaneously the largest city in the Southern Hemisphere and one of the largest urban heat islands in South America. The average maximum summer temperature in the city has increased by 0.4–0.6°C per decade over the last 50 years (Marengo et al., 2020), with extreme heat episodes becoming increasingly frequent and intense. Intraurban heterogeneity — between the historic center and the peripheries, between tree-covered areas and industrial zones — increases uncertainty in exposure attribution at the individual level, but the state-level (UF) analysis used in this case study is adequate for estimating the population-level impact.

Heatwave detection uses two complementary criteria: the WHO criterion, which defines a heatwave as a maximum temperature above the 90th percentile of the reference series (1981–2010) for at least 3 consecutive days; and the WMO criterion, which uses the same logic but applied to daily mean temperature. Combining both criteria allows comparison of the sensitivity and specificity of each definition in capturing excess mortality.

# Data Used

| Source | System | Variables | State (UF) | Period |
|-------|---------|-----------|-----|---------|
| DATASUS | SIM-DO | Deaths by ICD-10 I00–I99 (cardiovascular) | SP | 2016–2023 |
| INMET | Automatic stations | Maximum temperature (tair_max_c), minimum temperature (tair_min_c), dry-bulb temperature (tair_dry_bulb_c) | SP | 2016–2023 |

```{=html}
<div class="callout callout-note">
  <strong>Climate reference period:</strong> The 1981–2010 period is the standard 30-year WMO baseline used to calculate climatological percentiles. It is available in INMET's historical series and is automatically used by <code>sus_climate_compute_heatwaves()</code> when <code>baseline_start</code> and <code>baseline_end</code> are specified.
</div>
```

# Full Pipeline

```{r pipeline-ec03, eval=FALSE}
library(climasus4r)

# --- 1. Health — cardiovascular deaths in older adults ---
df_saude <- sus_data_import(
  uf     = "SP",
  year   = 2016:2023,
  system = "SIM-DO",
  lang   = "pt"
) |>
  sus_data_clean_encoding(lang = "pt") |>
  sus_data_standardize(lang = "pt") |>
  sus_data_filter_cid(disease_group = "cardiovascular", lang = "pt") |>
  sus_data_create_variables(
    create_age_groups = TRUE,
    age_breaks        = c(0, 65, 75, 85, Inf),
    age_labels        = c("< 65", "65-74", "75-84", "85+"),
    lang              = "pt"
  ) |>
  sus_data_filter_demographics(
    age_range = c(65, Inf),
    lang      = "pt"
  ) |>
  sus_data_aggregate(
    time_unit = "day",
    group_by  = "age_group",
    lang      = "pt"
  )

# --- 2. Raw climate + gap filling ---
df_clima_bruto <- sus_climate_inmet(
  uf       = "SP",
  year     = 2016:2023,
  variable = c("tair_max_c", "tair_min_c", "tair_dry_bulb_c"),
  lang     = "pt"
) |>
  sus_climate_fill_inmet(method = "random_forest", lang = "pt")

# --- 3. Heatwave detection ---
df_clima_hw <- sus_climate_compute_heatwaves(
  df             = df_clima_bruto,
  method         = c("WHO", "WMO"),
  baseline_start = "1981-01-01",
  baseline_end   = "2010-12-31",
  percentile     = 90,
  min_duration   = 3L,
  lang           = "pt"
)

# --- 4. Health + climate integration with heatwave flags ---
df_final <- sus_climate_aggregate(
  df_health  = df_saude,
  df_climate = df_clima_hw$daily_flag,
  strategy   = "threshold_exceedance",
  group_by   = "age_group",
  lang       = "pt"
)
```

# Step-by-Step Explanation

## `sus_data_filter_cid()` with `disease_group = "cardiovascular"`

The `"cardiovascular"` group covers Chapter IX of ICD-10: I00–I99 (Diseases of the Circulatory System). This includes acute myocardial infarction (I21–I22), heart failure (I50), ischemic stroke (I63) and hemorrhagic stroke (I61), cardiac arrhythmias (I46–I49), cardiomyopathies (I42–I43), and ischemic heart disease (I20–I25). For heat studies, it is common to restrict the analysis to a more specific subgroup; additional filtering by ICD code can be done with `dplyr::filter()` after the standardization step.

## `time_unit = "day"` in aggregation — why daily resolution?

The acute effect of heat on cardiovascular mortality manifests over a short lag: DLNM studies estimate that the peak relative risk occurs at lag 0–2 days after exposure (Gasparrini et al., 2017). Weekly aggregation would lose this temporal dynamic. Daily resolution also allows the calculation of daily excess mortality relative to a seasonal baseline, a standard approach in extreme event surveillance studies.

## `sus_climate_compute_heatwaves()` — Automatic heatwave detection

This function is what sets this case study apart from Case Studies 1 and 2. It takes the filled daily climate dataset and returns a list with three elements:

- **`$events`**: a tibble with the detected heatwave events — start date, end date, duration in days, mean intensity (degrees above the percentile), and severity class (moderate: 3–4 days; severe: 5–6 days; extreme: 7+ days).
- **`$daily_flag`**: daily dataset with binary variables `hw_who_flag` and `hw_wmo_flag` (1 = day within a heatwave under the corresponding criterion, 0 = normal day).
- **`$annual_summary`**: annual summary with the number of events, total heatwave days, and the maximum temperature of the most intense event of each year.

The `baseline_start = "1981-01-01"` and `baseline_end = "2010-12-31"` arguments define the climate reference period (WMO climatological normal) from which the percentiles are calculated. `percentile = 90` sets the temperature threshold for the WHO criterion. `min_duration = 3L` requires at least 3 consecutive days above the threshold to characterize a heatwave.

```{=html}
<div class="callout callout-tip">
  <strong>WHO vs. WMO criteria:</strong> The WHO criterion uses daily <em>maximum</em> temperature > P90 for >= 3 consecutive days. The WMO criterion uses daily <em>mean</em> temperature > P90 for >= 3 days. The WHO criterion is more sensitive to daytime peaks; the WMO criterion better captures cumulative nighttime thermal stress, which is more relevant for older adults.
</div>
```

## `strategy = "threshold_exceedance"` in `sus_climate_aggregate()`

The `"threshold_exceedance"` strategy performs a temporal join between the health dataset and the `$daily_flag` returned by `sus_climate_compute_heatwaves()`. The `hw_who_flag` and `hw_wmo_flag` variables are added to the daily mortality dataset, allowing the user to directly use these flags as a binary covariate in GLM models, Poisson regression models with fixed effects for day of the week and Fourier terms for seasonality, or as a stratification variable in case-crossover studies.

# Expected Output

After the pipeline runs, `df_final` is a `climasus_df` object with the following main columns:

| Column | Type | Description |
|--------|------|-----------|
| `date` | Date | Date of death (daily resolution) |
| `age_group` | factor | Age group of older adults (65–74, 75–84, 85+) |
| `n_deaths` | integer | Daily count of cardiovascular deaths |
| `hw_who_flag` | integer | Daily heatwave flag (WHO criterion) |
| `hw_wmo_flag` | integer | Daily heatwave flag (WMO criterion) |
| `tair_max_c` | double | Daily maximum temperature (°C) |
| `tair_min_c` | double | Daily minimum temperature (°C) |

```{=html}
<div class="callout callout-tip">
  <strong>Exploring the detected events:</strong> Use <code>df_clima_hw$events</code> to obtain the list of all detected heatwaves, with start date, end date, duration, and intensity. Use <code>df_clima_hw$annual_summary</code> for the annual summary.
</div>
```

# Next Steps

With the integrated daily dataset, the subsequent statistical analyses most commonly used in the literature are:

- **Time series modeling with DLNM**: estimate the temperature-mortality exposure-response function with a lag distribution of 0 to 10 days, controlling for seasonality (Fourier terms), linear trend, and day of the week. See the [DLNM Modeling](mod-01-dlnm.html) tutorial.
- **Attributable mortality analysis**: estimate the absolute number of deaths attributable to extreme heat and to specific heatwaves, with uncertainty intervals via bootstrapping. See the [Attributable Mortality](mod-02-atribuivel.html) tutorial.
- **Case-crossover study**: an epidemiological design suited to estimating the effect of acute exposures (such as heatwaves) on rare outcomes, automatically controlling for long-period confounders.

# References

Gasparrini, A., Guo, Y., Sera, F., et al. (2017). Projections of temperature-related excess mortality under climate change scenarios. *Lancet Planetary Health*, 1(9), e360–e367.

Vicedo-Cabrera, A. M., Scovronick, N., Sera, F., et al. (2021). The burden of heat-related mortality attributable to recent human-induced climate change. *Nature Climate Change*, 11, 492–500.

Marengo, J. A., Camargo, A. P., Alves, L. M., et al. (2020). Trends in extreme rainfall and hydrogeometeorological disasters in the Metropolitan Area of Sao Paulo: a review. *Climatic Change*, 161, 225–240.

Nairn, J. R., & Fawcett, R. J. B. (2015). The excess heat factor: a metric for heatwave intensity and its use in classifying heatwave severity. *International Journal of Environmental Research and Public Health*, 12(1), 227–253.

Brasil, Ministério da Saúde (2022). *SIM — Sistema de Informações sobre Mortalidade*. Brasília: DATASUS. Available at: https://datasus.saude.gov.br/

---

```{=html}
<p><a href="ec-02-dengue-clima.html">&larr; Case Study 2</a> &nbsp;|&nbsp; <a href="ec-04-hospitalizacoes-frio.html">Next: Case Study 4 &rarr;</a></p>
```

---

## Have feedback or suggestions?

Do you have an idea for improvement or found an error? Click the button below to open a new *issue* on GitHub and share your suggestions directly with us.

```{=html}
<a href='https://github.com/ByMaxAnjos/climasus4r/issues/new' class="github-feedback-btn">
  <svg xmlns="http://www.w3.org/2000/svg" width="18" height="18" viewBox="0 0 24 24" fill="currentColor">
    <path d="M12 0C5.37 0 0 5.37 0 12c0 5.31 3.435 9.795 8.205 11.385.6.105.825-.255.825-.57 0-.285-.015-1.23-.015-2.235-3.015.555-3.795-.735-4.035-1.41-.135-.345-.72-1.41-1.23-1.695-.42-.225-1.02-.78-.015-.795.945-.015 1.62.87 1.845 1.23 1.08 1.815 2.805 1.305 3.495.99.105-.78.42-1.305.765-1.605-2.67-.3-5.46-1.335-5.46-5.925 0-1.305.465-2.805 5.625-5.475 5.925.435.375.81 1.095.81 2.22 0 1.605-.015 2.895-.015 3.3 0 .315.225.69.825.57A12.02 12.02 0 0 0 24 12c0-6.63-5.37-12-12-12z"/>
  </svg>
  Open issue on GitHub
</a>
```
