---
title: "Modeling 1: DLNM — Distributed Lag Non-Linear Model"
author: "climasus4r Team"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    code_download: true
    self_contained: true
    df_print: paged
---

```{css, echo=FALSE}
/* -- Variaveis de Cores Modernas -- */
:root {
  --primary-color: #1D9E75;
  --dark-green: #0f3d2e;
  --bg-light: #f8faf9;
  --text-main: #1a1a1a;
  --code-bg: #fdfdfd;
  --warning-orange: #ff9800;
  --info-blue: #2196f3;
}
html { scroll-behavior: smooth; }
body { font-family: 'Source Serif 4', Georgia, serif; font-size: 18px; color: var(--text-main); background-color: #fff; line-height: 1.6; }
h1, h2, h3, h4 { font-family: 'Inter', sans-serif; font-weight: 700; color: var(--dark-green); }
h1.title { font-size: 2.8em; border-bottom: 4px solid var(--primary-color); padding-bottom: 10px; margin-bottom: 1.5rem; }
h2 { margin-top: 2rem; border-left: 4px solid var(--primary-color); padding-left: 1rem; }
pre.r, pre.sourceCode { background-color: var(--code-bg); border: 1px solid #e1e1e1; border-left: 5px solid var(--primary-color) !important; border-radius: 8px; box-shadow: 0 2px 4px rgba(0,0,0,0.05); padding: 1rem; }
.callout { padding: 1.2em 1.5em; margin: 1.5em 0; border-radius: 8px; border-left: 6px solid; background-color: var(--bg-light); }
.callout-note { background-color: #f0f7ff; border-color: #007bff; }
.callout-tip { background-color: #e8f5e9; border-color: #1D9E75; }
.callout-warning { background-color: #fff3e0; border-color: #ff9800; }
.workflow-container { background: linear-gradient(135deg, #f8faf7 0%, #ffffff 100%); border-radius: 12px; padding: 1.5rem; margin: 1.5rem 0; border: 1px solid #dce8e1; overflow-x: auto; }
.workflow-steps { display: flex; flex-wrap: wrap; justify-content: center; gap: 0.5rem; font-family: 'Inter', monospace; font-size: 0.8rem; }
.workflow-step { background: var(--primary-color); color: white; padding: 0.5rem 1rem; border-radius: 20px; white-space: nowrap; }
.workflow-arrow { color: var(--primary-color); font-weight: bold; font-size: 1.2rem; }
.feature-grid { display: grid; grid-template-columns: repeat(auto-fit, minmax(280px, 1fr)); gap: 1.5rem; margin: 2rem 0; }
.feature-card { background: white; border-radius: 12px; padding: 1.5rem; border: 1px solid #e1e1e1; transition: transform 0.2s ease, box-shadow 0.2s ease; text-align: center; }
.feature-card:hover { transform: translateY(-4px); box-shadow: 0 8px 24px rgba(0,0,0,0.1); }
.feature-icon { font-size: 2rem; margin-bottom: 1rem; }
.feature-card h3 { margin-top: 0; color: var(--primary-color); font-size: 1.1rem; }
table { width: 100%; border-collapse: collapse; margin: 1.5rem 0; }
th, td { padding: 12px; text-align: left; border-bottom: 1px solid #e1e1e1; }
th { background-color: var(--bg-light); font-weight: 600; color: var(--dark-green); }
.github-feedback-btn { display: inline-flex; align-items: center; gap: 8px; background-color: #24292e; color: white; padding: 10px 20px; border-radius: 6px; text-decoration: none; font-family: sans-serif; transition: background-color 0.2s; margin-top: 1rem; }
.github-feedback-btn:hover { background-color: #2c3e50; color: white; text-decoration: none; }
@media (max-width: 768px) { body { font-size: 16px; } h1.title { font-size: 2rem; } .workflow-steps { flex-direction: column; align-items: center; } }
```


```{r setup, include=FALSE}
# ATENÇÃO — estratégia de pré-renderização.
# Este tutorial NÃO ajusta modelos durante a renderização. Todas as figuras e
# tabelas foram geradas UMA vez pelo script scripts/modelagem_01_dlnm.R e
# são apenas carregadas aqui (PNG via include_graphics; RDS via readRDS).
# Por isso, os chunks de ensino abaixo usam eval = FALSE: o código é mostrado,
# mas não é executado no momento do knit.
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)

# Helpers para exibir artefatos de forma robusta: se o arquivo ainda não foi
# gerado (script de geração não executado), mostram uma nota em vez de quebrar
# o knit. Assim a vignette renderiza a qualquer momento.
fig_ou_nota <- function(path, nota) {
  if (file.exists(path)) knitr::include_graphics(path)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
tbl_ou_nota <- function(path, nota, caption = NULL) {
  if (file.exists(path)) knitr::kable(readRDS(path), caption = caption)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
.nota_gerar <- "Artefato indisponível — rode `Rscript scripts/modelagem_01_dlnm.R` (a partir da raiz do pacote) para gerá-lo."
```

# Objective

By the end of this tutorial you will be able to **fit a DLNM
(Distributed Lag Non-linear Model)** to quantify the relationship
between temperature and daily mortality, interpret the **cumulative
exposure-response curve**, the **lag-response curve**, and the
**two-dimensional exposure × lag surface**, and become familiar with
all the visualizations available in `sus_mod_plot_dlnm()`.

The DLNM is the **cornerstone of climasus4r's biostatistics hat**. It
connects the daily health series (produced in Tutorials 1–4) to the
aligned climate exposure (Tutorials 7–8) and produces relative-risk
estimates ready for attributable-burden quantification (Modeling
Tutorial 2), multi-city meta-analysis (Modeling Tutorial 5), and
vulnerability analysis (Modeling Tutorial 9).

# Scientific context

Modern environmental epidemiology recognizes that the association
between temperature and mortality is **non-linear** and **lagged in
time**. Extreme heat kills within the first 1–2 days (acute effect);
cold kills over 1 to 3 weeks (delayed, cumulative effect). Modeling
only same-day temperature systematically underestimates the risk of
cold and can mask the winter mortality peak in subtropical climates.

**Gasparrini, Armstrong & Kenward (2010)** proposed the DLNM to solve
this problem with a *cross-basis*: the outer product of two spline
bases — one on the **exposure** dimension (captures the U/J shape: risk
rises with both cold and heat) and one on the **lag** dimension
(captures how the effect is distributed over the following days). The
R implementation is the `dlnm` package (**Gasparrini, 2011**), on top
of which `climasus4r` builds `sus_mod_dlnm()`.

The fitting procedure follows the best practices for time-series
regression in environmental epidemiology systematized by
**Bhaskaran et al. (2013)**: a log-linear quasi-Poisson model, control
of long-term trend and seasonality via a natural spline of time, and
relative risk centered at the median exposure as the reference value
(**Armstrong, 2006**).

The detailed formulas for the *crossbasis* and the cumulative RR are
presented in the **Formulas** section below.

> **Driving case.** We investigate **temperature and health outcomes
> in older adults (60+) in the São Paulo Metropolitan Region,
> 2014–2019** (respiratory diseases J00–J99). We join the **daily
> death series** (`dados/caso_serie.rds`, Tutorial 4) with the
> **temperature from INMET station A701** (`dados/caso_clima_estacao.rds`,
> Tutorial 7), build the distributed-lag structure, and fit the
> quasi-Poisson DLNM to estimate the exposure-response curve and the
> peak lag.

# Formulas

The DLNM models the expected number of events $Y_t$ on date $t$ as:

$$
\log\{E(Y_t)\} = \alpha
  + \sum_{\ell=0}^{L} s(x_{t-\ell},\, \ell;\,\boldsymbol{\beta})
  + \mathrm{ns}(t)
  + \gamma\,\mathrm{DOW}_t
$$

where $s(x_{t-\ell}, \ell;\,\boldsymbol{\beta})$ is the *cross-basis*
function — the outer product of a spline basis over the **exposure**
$x_{t-\ell}$ (temperature at lag $\ell$) and a spline basis over the
**lag** $\ell \in \{0, \ldots, L\}$, with coefficient vector
$\boldsymbol{\beta}$; $\mathrm{ns}(t)$ is a natural spline of time
controlling for seasonality and long-term trend; and
$\gamma\,\mathrm{DOW}_t$ controls for the day-of-week effect.

The **cumulative effect** of exposure $x$ relative to the reference
value $x_0$ (lag-0 median) is:

$$
RR_{\text{cum}}(x) = \exp\!\left\{
  \sum_{\ell=0}^{L} \bigl[s(x,\, \ell;\,\boldsymbol{\hat\beta})
    - s(x_0,\, \ell;\,\boldsymbol{\hat\beta})\bigr]
\right\}
$$

**Key parameters implemented by `sus_mod_dlnm()`:**

| Parameter | Notation | Controlled by |
|-----------|---------|---------------|
| Exposure basis | $s(\cdot; \boldsymbol{\beta})$ over $x$ | `argvar` |
| Lag basis | $s(\cdot; \boldsymbol{\beta})$ over $\ell$ | `arglag` |
| Lag window | $L$ | `lag_max` |
| Seasonality df | df of $\mathrm{ns}(t)$ | `dof_per_year` |
| Reference value | $x_0$ | `ref_value` |

# Data used

| Input | Origin | Content | Role |
|---------|--------|----------|-------|
| `dados/caso_serie.rds` | Tutorial 4 | Daily death series (older adults 60+) | Outcome `n_obitos` |
| `dados/caso_clima_estacao.rds` | Tutorial 7 | Daily temperature from station A701 | Exposure `tair_dry_bulb_c` |

The two series are joined by **date**. The exposure is then expanded
into 22 **distributed-lag** columns
(`tair_dry_bulb_c_lag0`, …, `tair_dry_bulb_c_lag21`), a structure
that in a real workflow is produced by
`sus_climate_aggregate(temporal_strategy = "distributed_lag", lag_days = 21)`.
The script guards each step with `file.exists()` and, if the inputs
are missing, generates a **minimal, realistic synthetic series** (with
a V-shaped temperature-mortality signal) so the code still runs
without the full pipeline.

# Complete pipeline

The blocks below reproduce exactly what
`scripts/modelagem_01_dlnm.R` executes.

```{r carregar}
library(climasus4r)
```

## Block 0 — Load the chained inputs

```{r bloco0-carregar}
# Caminhos das entradas encadeadas
path_saude <- file.path("vignettes-pt", "dados", "caso_serie.rds")
path_clima <- file.path("vignettes-pt", "dados", "caso_clima_estacao.rds")

# Série diária de saúde
saude_diaria <- NULL
if (file.exists(path_saude)) {
  caso_serie   <- readRDS(path_saude)
  saude_df     <- dplyr::as_tibble(caso_serie)
  saude_diaria <- saude_df |>
    dplyr::summarise(n_obitos = sum(n_obitos, na.rm = TRUE), .by = date) |>
    dplyr::arrange(date)
}

# Série diária de temperatura
clima_diario <- NULL
if (file.exists(path_clima)) {
  caso_clima   <- readRDS(path_clima)
  clima_diario <- caso_clima$clima_diario |>
    dplyr::transmute(date = as.Date(date), tair_dry_bulb_c) |>
    dplyr::arrange(date)
}

# Fallback sintético realista (série em V, 6 anos)
if (is.null(clima_diario) || is.null(saude_diaria)) {
  # ... veja scripts/modelagem_01_dlnm.R para a implementação completa
}

# Juntar por data
serie <- dplyr::inner_join(saude_diaria, clima_diario, by = "date") |>
  dplyr::arrange(date) |>
  tidyr::drop_na(n_obitos, tair_dry_bulb_c)
```

## Block 1 — Build the distributed-lag structure

In a complete `climasus4r` workflow, the `_lagN` columns are produced
by `sus_climate_aggregate(temporal_strategy = "distributed_lag", lag_days = 21)`.
Here, with the series already aligned by date, we generate the lags
directly:

```{r bloco1-lags}
VAR_TEMP <- "tair_dry_bulb_c"
LAG_MAX  <- 21L

df_lags <- serie
for (k in 0:LAG_MAX) {
  df_lags[[paste0(VAR_TEMP, "_lag", k)]] <- dplyr::lag(serie$tair_dry_bulb_c, n = k)
}

# Seleciona apenas: date, desfecho, colunas de lag
df_lags <- df_lags |>
  dplyr::select(
    date, n_obitos,
    dplyr::all_of(paste0(VAR_TEMP, "_lag", 0:LAG_MAX))
  )

# Promove a climasus_df com stage = "climate", type = "distributed_lag"
df_dl <- dplyr::as_tibble(df_lags)
attr(df_dl, "sus_meta") <- list(
  system  = "SIM",
  stage   = "climate",
  type    = "distributed_lag",
  backend = "tibble",
  history = "[derivado] serie diaria -> lags distribuidos tair_dry_bulb_c"
)
class(df_dl) <- c("climasus_df", class(df_dl))
```

## Block 2 — `sus_mod_dlnm()`: fitting the quasi-Poisson DLNM

```{r bloco2-dlnm}
fit <- sus_mod_dlnm(
  df           = df_dl,
  outcome_col  = "n_obitos",
  climate_col  = "tair_dry_bulb_c",
  lag_max      = 21L,
  covariates   = NULL,                       # confundidores adicionais (opcional)
  argvar       = list(fun = "ns", df = 4),   # spline natural, 4 gl exposição
  arglag       = list(fun = "ns", df = 3),   # spline natural, 3 gl lag
  family       = "quasipoisson",             # robusto à superdispersão
  ns_df        = NULL,                       # calculado automaticamente via dof_per_year
  dof_per_year = 4L,                         # gl/ano para sazonalidade (use 6 para Brasil subtropical)
  ref_value    = NULL,                       # mediana de lag-0 = valor de referência
  pred_at      = c(0.25, 0.50, 0.75, 0.90, 0.95, 0.99),
  alpha        = 0.05,
  lang         = "pt",
  verbose      = TRUE
)

# Inspecionar o resultado
print(fit)
fit$exposure_response   # RR cumulativo nos percentis pred_at
fit$lag_response        # RR por lag no percentil 75
fit$diagnostics         # razão de dispersão, Ljung-Box, AIC
```

## Block 3 — `sus_mod_plot_dlnm()`: all visualizations

```{r bloco3-plots}
# overall: curva exposição-resposta cumulativa (painel principal + histograma)
sus_mod_plot_dlnm(fit, type = "overall", lang = "pt")

# lag: curva lag-resposta no percentil 75 (RR específico e cumulativo)
sus_mod_plot_dlnm(fit, type = "lag", lang = "pt")

# surface: superfície 3-D bidimensional exposição × lag
# (interactive = FALSE usa contorno estático como fallback)
sus_mod_plot_dlnm(fit, type = "surface", interactive = FALSE, lang = "pt")

# contour: mapa de contorno 2-D log(RR) para publicação
sus_mod_plot_dlnm(fit, type = "contour", lang = "pt")

# slice: curvas de resposta por lag específico (valores > lag_max são descartados)
sus_mod_plot_dlnm(fit, type = "slice",
                  lags_at = c(0L, 3L, 7L, 14L, 21L), lang = "pt")

# distribution: histograma/densidade da exposição com percentis-chave
sus_mod_plot_dlnm(fit, type = "distribution", lang = "pt")

# series: série temporal desfecho (barras) + exposição lag-0 (eixo dual)
sus_mod_plot_dlnm(fit, type = "series", lang = "pt")

# output_type = "all": retorna lista nomeada $plot, $table, $data
out <- sus_mod_plot_dlnm(fit, type = "overall",
                         output_type = "all", lang = "pt")
out$table   # lista de 3 gt_tbl: especificação, curva RR, diagnósticos
out$data    # lista: $exposure_response, $lag_response, $data_daily, $diagnostics
```

## Block 4 — Save the fit for reuse (M2 and M8)

```{r bloco4-salvar}
# O objeto climasus_dlnm é salvo para alimentar
# Tutorial M2 (fração atribuível) e M8 (meta-análise multi-cidade).
# save_tbl() (definida em scripts/_setup_tutoriais.R) persiste em
# vignettes-pt/dados/mod_dlnm_fit.rds; o script também copia para
# dados/mod_dlnm_fit.rds (raiz do pacote) para encadeamento com M2 e M8.
save_tbl(fit, "mod_dlnm_fit")
saveRDS(fit, file.path("dados", "mod_dlnm_fit.rds"))
```

# Function characteristics

## `sus_mod_dlnm()`

Fits a DLNM from a `climasus_df` at stage `"climate"` / type
`"distributed_lag"`, produced by
`sus_climate_aggregate(temporal_strategy = "distributed_lag")`.
Builds the *crossbasis*, fits a GLM, and precomputes the `crosspred`
on a grid of 100 points between the 1st and 99th percentile of
exposure.

| Argument | Default | Description |
|-----------|--------|-----------|
| `df` | — | `climasus_df` with `date`, outcome, and `{var}_lagN` columns |
| `outcome_col` | `"n_obitos"` | Name of the daily outcome count column |
| `climate_col` | `NULL` | Exposure name base (without `_lagN`); auto-detected by priority if `NULL` |
| `lag_max` | `NULL` | Maximum lag; auto-detected from the largest `N` among `_lagN` columns |
| `covariates` | `NULL` | Additional confounders to include linearly in the GLM |
| `argvar` | `list(fun="ns", df=4)` | Basis for the exposure dimension of the *crossbasis* |
| `arglag` | `list(fun="ns", df=3)` | Basis for the lag dimension of the *crossbasis* |
| `family` | `"quasipoisson"` | GLM family; `"negbin"` is redirected to `"quasipoisson"` |
| `ns_df` | `NULL` | df for the temporal spline; `NULL` uses `dof_per_year × n_years` |
| `dof_per_year` | `4L` | df per year for seasonality (6 recommended for subtropical Brazil) |
| `ref_value` | `NULL` | Reference value (RR = 1); `NULL` uses the lag-0 median |
| `pred_at` | `c(.25,.5,.75,.9,.95,.99)` | Quantiles to report in the `$exposure_response` table |
| `alpha` | `0.05` | CI level (default: 95% CI) |
| `lang` | `"pt"` | Message language: `"pt"`, `"en"`, `"es"` |

**Return — `climasus_dlnm` object (S3 list):**

| Component | Content |
|------------|----------|
| `$model` | Fitted `glm` object (compatible with `dlnm::crosspred()`) |
| `$crossbasis` | `dlnm::crossbasis` object (exposure × lag basis) |
| `$pred` | `dlnm::crosspred` object — `allRRfit/low/high` (overall effect) and `matRRfit/low/high` (lag-specific) |
| `$exposure_response` | Tibble: `pct`, `exposure`, `rr`, `lo`, `hi` at the `pred_at` percentiles |
| `$lag_response` | Tibble: `lag`, `rr`, `lo`, `hi`, `rr_cum` — RR per lag at P75 |
| `$models` | 1-row summary tibble: `variable`, `n`, `family`, `lag_max`, `rr`, `lo`, `hi`, `lag_peak`, `disp_ratio`, `aic_poisson` |
| `$data_daily` | Tibble aggregated by date (outcome summed; climate averaged) used in the fit |
| `$diagnostics` | List: `disp_ratio`, `disp_category`, `autocorr_pval`, `has_autocorr`, `aic_poisson`, `deviance`, `df_residual` |
| `$meta` | List with all fit parameters plus the `sus_meta` history |

**Basis selection (`argvar` / `arglag`):**

| `argvar` | When to use |
|----------|-------------|
| `list(fun="ns", df=3)` | Smooth non-linearity; conservative |
| `list(fun="ns", df=4)` | U/J-shaped curve (heat + cold, **default**) |
| `list(fun="ns", df=5)` | Complex dose-response; large N |
| `list(fun="lin")` | Linear hypothesis test |

| `arglag` | When to use |
|----------|-------------|
| `list(fun="ns", df=3)` | Smooth decay; temperature-related mortality (**default**) |
| `list(fun="ns", df=4)` | Bimodal lag; cold effect |
| `list(fun="poly", df=2)` | Conservative; small samples |

**Seasonality control (`dof_per_year`):**

- `4` df/year: 1–2 year series; annual seasonality
- `6` df/year: subtropical Brazil; more pronounced seasonality
- `8` df/year: arboviruses; sub-seasonal variation

**Available S3 methods:** `print()`, `summary()`, `coef()`, `tidy()`
(returns a one-row tibble compatible with `dplyr::bind_rows()`).

---

## `sus_mod_plot_dlnm()`

Produces scientific visualizations and statistical tables from a
`climasus_dlnm` object. Implements all seven standard visualization
types used in DLNM epidemiology.

| Argument | Default | Description |
|-----------|--------|-----------|
| `fit` | — | `climasus_dlnm` object from `sus_mod_dlnm()` |
| `type` | `"overall"` | Visualization type (see table below) |
| `output_type` | `"plot"` | `"plot"` → ggplot2/plotly; `"table"` → gt; `"all"` → list `$plot, $table, $data` |
| `exposure_at` | `NULL` | Focal exposure for the `"lag"` plot (default: P75) |
| `lags_at` | `c(0,3,7,14,21)` | Lags shown in the `"slice"` type (clipped at `lag_max`) |
| `pred_at` | `c(.05,.10,.25,.50,.75,.90,.95,.99)` | Quantiles in the statistical table |
| `interactive` | `FALSE` | `TRUE` → `plotly::ggplotly()` / 3-D plotly surface |
| `color_palette` | `"npg"` | `ggsci` palette (others: `"lancet"`, `"jama"`, `"nejm"`) |
| `lang` | `"pt"` | Label language: `"pt"`, `"en"`, `"es"` |

### Visualization types (`type =`)

| `type` | What it shows | Reference |
|--------|-------------|------------|
| `"overall"` | Cumulative exposure-response curve (RR + 95% CI) with the exposure histogram in the bottom panel | Gasparrini et al. (2010) |
| `"lag"` | Lag-specific and cumulative RR at a given exposure level (default P75) | Gasparrini et al. (2010) |
| `"surface"` | Two-dimensional 3-D exposure × lag surface (`interactive = TRUE` → 3-D plotly; static → contour) | Gasparrini (2011) |
| `"contour"` | 2-D contour map (log(RR), divergent scale) for publication | Gasparrini et al. (2014) |
| `"slice"` | Overlaid response curves for specific lags (dose-response profiles by lag time) | Armstrong (2006) |
| `"distribution"` | Histogram + kernel density of exposure with markers at key percentiles | Bhaskaran et al. (2013) |
| `"series"` | Daily time series of the outcome (bars) and exposure (line, dual axis) | Bhaskaran et al. (2013) |

**Return (`output_type`):**

- `"plot"` → `ggplot` or `plotly` object
- `"table"` → list of `gt_tbl` (or tibble if `gt` is unavailable) with
  3 sections: (1) model specification, (2) RR curve at percentiles,
  (3) model diagnostics
- `"all"` → named list `$plot`, `$table`, `$data` (containing
  `$exposure_response`, `$lag_response`, `$data_daily`, `$diagnostics`)

# Illustrated results

## Time series: outcome and exposure

```{r fig-series, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/moddlnm_series.png", .nota_gerar)
```

**Figure 1.** Daily time series of deaths among older adults (green
bars) and mean temperature (orange line, scaled secondary axis), São
Paulo Metropolitan Region 2014–2019. The seasonal pattern is visible:
mortality peaks coincide with the subtropical winter (June–August),
when minimum daily temperatures drop below the thermal comfort
threshold. Exposure to extreme heat (austral summer,
December–February) produces a more acute effect.

## Temperature distribution

```{r fig-dist, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/moddlnm_distribution.png", .nota_gerar)
```

**Figure 2.** Distribution of daily temperature (lag 0) with key
percentiles marked (P10, P25, P50 — reference, P75, P90, P99). The
median is the reference value (RR = 1) of the DLNM. The P25–P75 range
represents "usual" thermal conditions; values above P90 and below P10
define the extreme heat and cold thresholds.

## Cumulative exposure-response curve (`type = "overall"`)

```{r fig-overall, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/moddlnm_overall.png", .nota_gerar)
```

**Figure 3.** Cumulative exposure-response curve (RR and 95% CI via
the delta method). The V-shaped pattern is characteristic of the
temperature-mortality association in subtropical Brazil: elevated
risk under moderate cold (delayed, cumulative effect) and, to a lesser
degree, under extreme heat (acute effect). The horizontal dashed line
marks RR = 1; the vertical dotted line marks the median exposure
(reference value). The colored points indicate the quantiles reported
in Table 1.

## Lag-response curve (`type = "lag"`)

```{r fig-lag, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/moddlnm_lag.png", .nota_gerar)
```

**Figure 4.** Lag-response curve for the 75th percentile of exposure.
The solid line shows the lag-specific RR (effect on each individual
day); the dashed line shows the cumulative RR accumulated up to that
lag. The **peak lag** — the day at which the lag-specific effect is
maximal — is reported in the model diagnostics.

## Contour map (`type = "contour"`)

```{r fig-contour, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/moddlnm_contour.png", .nota_gerar)
```

**Figure 5.** Contour map of log(RR) as a function of temperature
(y-axis) and lag day (x-axis). Warm colors (orange/red) indicate
excess risk; cool colors (green) indicate reduced risk relative to the
median. The map simultaneously reveals the **shape** (non-linearity
with respect to temperature) and the **temporal distribution** (how
the effect persists across lags). White isolines mark the ±0.1 and
±0.2 log(RR) levels.

## Curves by specific lag (`type = "slice"`)

```{r fig-slice, eval=TRUE, echo=FALSE}
fig_ou_nota("figuras/moddlnm_slice.png", .nota_gerar)
```

**Figure 6.** Exposure-response curves for selected lags (0, 3, 7, 14,
and 21 days). Each curve shows how the dose-response profile changes
over time. The extreme-heat effect is more pronounced at lags 0–3
(acute response); the moderate-cold effect persists up to lags 14–21
(delayed, distributed response).

## Model statistical summary

```{r tbl-resumo, eval=TRUE, echo=FALSE}
tbl_ou_nota("dados/moddlnm_resumo.rds", .nota_gerar,
            caption = "Tabela 1. Especificação e diagnósticos do DLNM ajustado.")
```

```{r tbl-expo, eval=TRUE, echo=FALSE}
tbl_ou_nota("dados/moddlnm_expo_resp.rds", .nota_gerar,
            caption = "Tabela 2. Curva exposição-resposta: RR cumulativo nos percentis da temperatura.")
```

**Table 2.** Cumulative Relative Risk (95% CI) for each percentile of
the temperature distribution, relative to the median (RR = 1). Values
above 1.00 indicate excess risk; values below 1.00 indicate relative
protection. Percentiles above P90 and below P10 indicate extreme
exposures.

# Interpretation and best practices

**Dispersion ratio ($\hat\phi$).** The quasi-Poisson is recommended
when $\hat\phi > 1{,}5$ (Armstrong, 2006). The dispersion ratio is
reported in `fit$diagnostics$disp_ratio`; category `"adequate"`
indicates $\hat\phi < 1{,}5$.

**Residual autocorrelation.** The Ljung-Box test (10 lags of the
deviance residuals) detects remaining autocorrelation after
seasonality control. If positive (`has_autocorr = TRUE`), increase
`dof_per_year` or add `covariates = "dow"` (day of week)
(Bhaskaran et al., 2013).

**Degrees of freedom of the temporal spline.** The general rule is
`dof_per_year = 4` for 1–2 years of data and `dof_per_year = 6` for
subtropical Brazil (more pronounced seasonality). For arboviruses with
sub-seasonal peaks, `dof_per_year = 8`. Visualize the seasonal residual
with `fit$data_daily` to check the fit.

**Reference value.** The default is the median exposure at lag-0. In
multi-city comparative studies, use a fixed common value
(`ref_value = 20.0`) to ensure comparability across exposure-response
curves.

**Chaining.** The `mod_dlnm_fit.rds` object (a `climasus_dlnm`
component) is the **input** for the tutorials:

- **M2 (Attributable Fraction):** `sus_mod_af(fit = fit, ...)`
- **M5 (Meta-analysis):** `sus_mod_pool(fits = list(rmsp = fit, ...), ...)`
- **M8 (Space-Time Bayes):** model with an offset based on the
  estimated RR

# References

- Armstrong, B. (2006). Models for the relationship between ambient
  temperature and daily deaths. *Epidemiology*, 17(6), 624–631.
  <https://doi.org/10.1097/01.ede.0000239732.50999.58>

- Bhaskaran, K., Gasparrini, A., Hajat, S., Smeeth, L., & Armstrong, B.
  (2013). Time series regression studies in environmental epidemiology.
  *International Journal of Epidemiology*, 42(4), 1187–1195.
  <https://doi.org/10.1093/ije/dyt092>

- Gasparrini, A., Armstrong, B., & Kenward, M. G. (2010). Distributed
  lag non-linear models. *Statistics in Medicine*, 29(21), 2224–2234.
  <https://doi.org/10.1002/sim.3940>

- Gasparrini, A. (2011). Distributed lag linear and non-linear models in
  R: the package dlnm. *Journal of Statistical Software*, 43(8), 1–20.
  <https://doi.org/10.18637/jss.v043.i08>

- Gasparrini, A., Armstrong, B., & Kenward, M. G. (2014). Reducing and
  meta-analysing estimates from distributed lag non-linear models.
  *BMC Medical Research Methodology*, 14, 70.
  <https://doi.org/10.1186/1471-2288-14-70>

- Saldanha, R. F., Bastos, R. R., & Barcellos, C. (2019). Microdatasus:
  pacote para download e pré-processamento de microdados do DATASUS.
  *Cadernos de Saúde Pública*, 35(9), e00032419.

---

# Navigation

- [Back to tutorial hub](tutorials.html)
- **Next:** [Modeling 2 — Attributable Fraction and Number](mod-02-atribuivel.html)

> **How to reproduce the figures.** Run once, from the package root:
> `Rscript scripts/modelagem_01_dlnm.R`
> The script saves the figures to `vignettes-pt/figuras/` (prefix
> `moddlnm_`), the tables to `vignettes-pt/dados/`, and the fitted
> object `mod_dlnm_fit.rds` (input for M2 and M8).
