---
title: "Modeling 08: Vulnerability Index and SWOT Synthesis"
author: "climasus4r Team"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    code_download: true
    self_contained: true
    df_print: paged
---

```{css, echo=FALSE}
/* -- Variaveis de Cores Modernas -- */
:root {
  --primary-color: #1D9E75;
  --dark-green: #0f3d2e;
  --bg-light: #f8faf9;
  --text-main: #1a1a1a;
  --code-bg: #fdfdfd;
  --warning-orange: #ff9800;
  --info-blue: #2196f3;
}
html { scroll-behavior: smooth; }
body { font-family: 'Source Serif 4', Georgia, serif; font-size: 18px; color: var(--text-main); background-color: #fff; line-height: 1.6; }
h1, h2, h3, h4 { font-family: 'Inter', sans-serif; font-weight: 700; color: var(--dark-green); }
h1.title { font-size: 2.8em; border-bottom: 4px solid var(--primary-color); padding-bottom: 10px; margin-bottom: 1.5rem; }
h2 { margin-top: 2rem; border-left: 4px solid var(--primary-color); padding-left: 1rem; }
pre.r, pre.sourceCode { background-color: var(--code-bg); border: 1px solid #e1e1e1; border-left: 5px solid var(--primary-color) !important; border-radius: 8px; box-shadow: 0 2px 4px rgba(0,0,0,0.05); padding: 1rem; }
.callout { padding: 1.2em 1.5em; margin: 1.5em 0; border-radius: 8px; border-left: 6px solid; background-color: var(--bg-light); }
.callout-note { background-color: #f0f7ff; border-color: #007bff; }
.callout-tip { background-color: #e8f5e9; border-color: #1D9E75; }
.callout-warning { background-color: #fff3e0; border-color: #ff9800; }
.workflow-container { background: linear-gradient(135deg, #f8faf7 0%, #ffffff 100%); border-radius: 12px; padding: 1.5rem; margin: 1.5rem 0; border: 1px solid #dce8e1; overflow-x: auto; }
.workflow-steps { display: flex; flex-wrap: wrap; justify-content: center; gap: 0.5rem; font-family: 'Inter', monospace; font-size: 0.8rem; }
.workflow-step { background: var(--primary-color); color: white; padding: 0.5rem 1rem; border-radius: 20px; white-space: nowrap; }
.workflow-arrow { color: var(--primary-color); font-weight: bold; font-size: 1.2rem; }
.feature-grid { display: grid; grid-template-columns: repeat(auto-fit, minmax(280px, 1fr)); gap: 1.5rem; margin: 2rem 0; }
.feature-card { background: white; border-radius: 12px; padding: 1.5rem; border: 1px solid #e1e1e1; transition: transform 0.2s ease, box-shadow 0.2s ease; text-align: center; }
.feature-card:hover { transform: translateY(-4px); box-shadow: 0 8px 24px rgba(0,0,0,0.1); }
.feature-icon { font-size: 2rem; margin-bottom: 1rem; }
.feature-card h3 { margin-top: 0; color: var(--primary-color); font-size: 1.1rem; }
table { width: 100%; border-collapse: collapse; margin: 1.5rem 0; }
th, td { padding: 12px; text-align: left; border-bottom: 1px solid #e1e1e1; }
th { background-color: var(--bg-light); font-weight: 600; color: var(--dark-green); }
.github-feedback-btn { display: inline-flex; align-items: center; gap: 8px; background-color: #24292e; color: white; padding: 10px 20px; border-radius: 6px; text-decoration: none; font-family: sans-serif; transition: background-color 0.2s; margin-top: 1rem; }
.github-feedback-btn:hover { background-color: #2c3e50; color: white; text-decoration: none; }
@media (max-width: 768px) { body { font-size: 16px; } h1.title { font-size: 2rem; } .workflow-steps { flex-direction: column; align-items: center; } }
```


```{r setup, include=FALSE}
# ATENÇÃO — estratégia de pré-renderização.
# Este tutorial NÃO executa modelagem durante a renderização. Todas as figuras e
# tabelas foram geradas UMA vez pelo script scripts/modelagem_08_vulnerabilidade_swot.R e
# são apenas carregadas aqui (PNG via include_graphics; RDS via readRDS).
# Por isso, os chunks de ensino abaixo usam eval = FALSE: o código é mostrado,
# mas não é executado no momento do knit.
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)

# Helpers robustos: se o arquivo não foi gerado, mostram uma nota em vez de
# quebrar o knit. Assim a vignette renderiza a qualquer momento.
fig_ou_nota <- function(path, nota) {
  if (file.exists(path)) knitr::include_graphics(path)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
tbl_ou_nota <- function(path, nota, caption = NULL) {
  if (file.exists(path)) knitr::kable(readRDS(path), caption = caption)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
.nota_gerar <- "Artefato indisponível — rode `Rscript scripts/modelagem_08_vulnerabilidade_swot.R` (a partir da raiz do pacote) para gerá-lo."

# Caminhos base (relativos à raiz do pacote, de onde o script é executado)
.fig <- function(n) file.path("figuras", paste0(n, ".png"))
.tbl <- function(n) file.path("dados",   paste0(n, ".rds"))
```

# Objective

By the end of this tutorial you will be able to:

1. **Compose a Composite Vulnerability Index (CVI)** following the IPCC AR6 three-pillar
   structure — climate Exposure, population Sensitivity, and Adaptive Capacity — from
   health-climate, socioeconomic, and health infrastructure indicators, using
   `sus_mod_vulnerability_index()`;
2. **Interpret the vulnerability ranking** across municipalities and measure territorial
   inequality with the Gini Coefficient and the Lorenz curve;
3. **Synthesize the results into a SWOT analysis** (Strengths, Weaknesses, Opportunities
   and Threats) oriented toward health-climate public policy, using `sus_mod_swot()`;
4. **Visualize and communicate** the results with the paired plotting functions
   `sus_mod_plot_vulnerability()` (ranking, pillars, Lorenz) and `sus_mod_plot_swot()`
   (radar, 2×2 matrix, bars).

This is the **eighth module in the Modeling series** of `climasus4r`. The main inputs
are the `climasus_af` object from Module M2 (`mod_af.rds`) and the `climasus_dlnm`
object from Module M1 (`mod_dlnm_fit.rds`). The main output is the `climasus_vi` object
saved as `mod_vi.rds` for reuse in final reports and risk communication.

---

# Scientific background

## Climate vulnerability according to the IPCC

The **IPCC's Fifth Assessment Report** (AR5, 2014) established the three-pillar
framework that defines vulnerability — later refined in AR6 (2022) — as a function of
three interdependent components:

- **Exposure**: the degree to which a system or population is affected by climate
  stressors (extreme temperature, heat waves, days above the 90th percentile of
  heat);
- **Sensitivity**: the level of response of a system to the stressor, conditioned by
  intrinsic factors such as the proportion of elderly people, prevalence of chronic
  disease, and the dose-response relationship estimated by DLNM models;
- **Adaptive Capacity**: the set of institutional, economic, and infrastructural
  resources that allow a system to modulate its impacts (human development index,
  treated water coverage, density of primary health care units).

The composite index formula, originally proposed by Turner et al. (2003) and adopted
in the social vulnerability literature (Cutter, Boruff & Shirley, 2003; Benmarhnia et
al., 2015), inverts the Adaptive Capacity pillar before combination, since higher
capacity means lower vulnerability:

$$
VI_i = \frac{w_E \cdot E_i + w_S \cdot S_i + w_{AC} \cdot (1 - AC_i)}{w_E + w_S + w_{AC}}
$$

where $E_i$, $S_i$, and $AC_i$ are the weighted averages of the normalized indicators
(min-max or z-score) of municipality $i$ for each pillar, and $w_E$, $w_S$, $w_{AC}$
are the inter-pillar weights. Missing indicators are handled through automatic
reweighting.

## Min-max and z-score normalization

**Min-max normalization** projects each indicator $x_{ij}$ onto the interval $[0, 1]$:

$$
\tilde{x}_{ij} = \frac{x_{ij} - \min_j(x_{ij})}{\max_j(x_{ij}) - \min_j(x_{ij})}
$$

**Z-score normalization** standardizes by mean and standard deviation, preserving the
original distribution — useful when indicators have very heterogeneous scales, but it
requires a back-transformation for direct interpretation.

## Territorial inequality and the Lorenz Curve

The **Lorenz Curve** is the classic instrument for visualizing the concentration of
any resource or burden across an ordered population. In the context of climate risk,
the curve accumulates the proportion of total CVI (y-axis) against the proportion of
municipalities (x-axis), from least to most vulnerable. The larger the area between
the curve and the diagonal of perfect equality, the higher the **Gini Coefficient** —
which ranges from 0 (perfect equality) to 1 (all risk concentrated in a single
municipality).

Climate risk concentration analyses have been applied at multi-city scale by
Gasparrini et al. (2017) and in national socio-environmental vulnerability studies in
Brazil by Confalonieri et al. (2009) and Marengo et al. (2022).

## SWOT analysis oriented toward health-climate policy

The SWOT framework (Andrews, 1971), widely used in strategic planning, has been
adapted to climate-health risk management. In this approach:

- **Strengths (S)** represent existing protective factors — high adaptive capacity,
  low total attributable fraction;
- **Weaknesses (W)** reflect current vulnerabilities — high population sensitivity,
  high heat- and cold-attributable fractions;
- **Opportunities (O)** identify windows for intervention — municipalities with a
  relatively low vulnerability index and moderate current exposure have room to
  advance prevention before reaching critical levels;
- **Threats (T)** signal emerging or already present risks — high climate exposure,
  high relative heat risk (DLNM), high attributable number.

---

# Data used

| Object | Source | Content |
|--------|--------|----------|
| `caso_espacial.rds` | Tutorial T5 | sf geometries of the RMSP municipalities |
| `caso_socio.rds` | Tutorial T6 | Socioeconomic indicators per municipality (HDI, % elderly, etc.) |
| `mod_dlnm_fit.rds` | Module M1 | `climasus_dlnm` object fitted for the RMSP |
| `mod_af.rds` | Module M2 | `climasus_af` object with total/heat/cold AF% |

Whenever any of these four files is missing (execution outside the full chain), the
pre-rendering script synthesizes realistic data with a warning — the Rmd renders
normally, displaying illustrative results.

---

# Complete pipeline

```{r carregar}
library(climasus4r)
library(dplyr)
```

## Load chained inputs

```{r carregar-entradas}
# Objeto climasus_af do Módulo M2
mod_af <- if (file.exists("vignettes-pt/dados/mod_af.rds")) {
  readRDS("vignettes-pt/dados/mod_af.rds")
} else {
  NULL  # fallback: SWOT rodará sem AF
}

# Objeto climasus_dlnm do Módulo M1
mod_dlnm <- if (file.exists("vignettes-pt/dados/mod_dlnm_fit.rds")) {
  readRDS("vignettes-pt/dados/mod_dlnm_fit.rds")
} else {
  NULL  # fallback: sem extração DLNM automática
}

# Indicadores socioeconômicos
caso_socio <- if (file.exists("vignettes-pt/dados/caso_socio.rds")) {
  readRDS("vignettes-pt/dados/caso_socio.rds")
} else {
  # Mínimo realista para o script funcionar sem a cadeia completa
  data.frame(
    city          = c("Sao Paulo", "Guarulhos", "Santo Andre", "Osasco", "Maua"),
    pct_idosos    = c(16.2, 12.4, 14.8, 13.1, 11.9),
    pct_doenca_cr = c(28.1, 24.3, 26.7, 22.8, 20.5),
    hdi           = c(0.845, 0.763, 0.815, 0.776, 0.742),
    pct_agua      = c(97.2, 88.4, 94.1, 90.3, 85.7),
    densidade_ubs = c(6.8, 3.2, 5.1, 4.0, 2.9)
  )
}
```

## Build the three IPCC pillars

Each pillar is supplied as a `data.frame` with a municipal identifier column
(`city_col = "city"`) and numeric indicator columns. Indicators missing for a given
municipality are filled with `NA` and excluded from the pillar's weighted average,
without forcing complete exclusion.

```{r pilares}
# ----- Pilar E: Exposição Climática ------------------------------------------
# Fonte: sus_climate_aggregate() (temperatura maxima media, dias de calor P90,
#        contagem de ondas de calor por ano)
exposure_df <- data.frame(
  city           = caso_socio$city,
  tmax_media     = c(29.8, 31.2, 28.9, 30.1, 31.8),  # grau Celsius
  dias_calor_p90 = c(52,   68,   45,   58,   72),     # dias/ano
  ondas_calor    = c(4,    8,    3,    6,    9)        # eventos/ano
)

# ----- Pilar S: Sensibilidade Populacional -----------------------------------
# Fonte: caso_socio.rds + extração automática de climasus_dlnm (argumento fits)
sensitivity_df <- caso_socio[, c("city", "pct_idosos", "pct_doenca_cr")]

# ----- Pilar AC: Capacidade Adaptativa ---------------------------------------
# Fonte: caso_socio.rds (HDI, cobertura hídrica, densidade UBS)
adaptive_capacity_df <- caso_socio[, c("city", "hdi", "pct_agua", "densidade_ubs")]
```

## Fit the Composite Vulnerability Index

`sus_mod_vulnerability_index()` accepts a `fits` argument (a named list of
`climasus_dlnm` objects) for automatic extraction of the metrics `dlnm_hot_rr`,
`dlnm_cold_rr`, and `dlnm_sensitivity_index`, which are automatically merged into the
sensitivity pillar.

```{r vulnerability-index}
# Opcional: lista nomeada de climasus_dlnm (um por cidade)
# fits_dlnm <- list(sao_paulo = mod_dlnm)  # ajuste por municipio

vi_fit <- sus_mod_vulnerability_index(
  exposure_df          = exposure_df,
  sensitivity_df       = sensitivity_df,
  adaptive_capacity_df = adaptive_capacity_df,
  fits                 = NULL,                # ou: fits_dlnm (lista climasus_dlnm)
  city_col             = "city",
  normalize            = "minmax",            # "minmax" ou "zscore"
  weights              = c(tmax_media = 2,    # dobrar peso da Tmax
                           pct_idosos = 2),   # dobrar peso da proporção de idosos
  pillar_weights       = c(exposure         = 1.5,
                           sensitivity      = 1.0,
                           adaptive_capacity = 1.0),
  hot_percentile       = 0.99,
  cold_percentile      = 0.01,
  lang                 = "pt",
  verbose              = TRUE
)

# Inspecionar o resultado
print(vi_fit)
```

The `climasus_vi` object contains:

| Element | Type | Content |
|----------|------|----------|
| `$vi_table` | tibble | CVI, pillar scores, rank, and percentile per city |
| `$pillar_scores` | tibble (long) | Raw scores for the three pillars per city |
| `$normalized_indicators` | tibble | Normalized indicators (prefixes `exp_`, `sen_`, `ac_`) |
| `$raw_indicators` | tibble | Original indicators without normalization |
| `$weights` | list | Intra-pillar and inter-pillar weights effectively used |
| `$gini_coefficient` | numeric | Gini of the CVI (0 = perfect equality) |
| `$meta` | list | Call parameters, counts, timing |

## Fit the SWOT Analysis

`sus_mod_swot()` aggregates climasus objects from multiple modules into normalized
(0–100) and categorized SWOT quadrants. The argument `breaks = c(25, 50, 75)` produces
four categories (Low / Medium / High / Very High) — the default `c(33, 66)` produces
three categories.

```{r swot}
swot_fit <- sus_mod_swot(
  vulnerability = vi_fit,
  af            = mod_af,    # climasus_af do Módulo M2 (ou NULL)
  burden        = NULL,      # climasus_burden do Módulo M2 (opcional)
  dlnm          = mod_dlnm,  # climasus_dlnm do Módulo M1 (broadcast)
  sensitivity   = NULL,      # climasus_sensitivity do Módulo M7 (opcional)
  score_type    = "both",    # "numeric", "categorical" ou "both"
  breaks        = c(25, 50, 75),
  labels        = c("Baixo", "Médio", "Alto", "Muito Alto"),  # 4 cats para 3 breaks
  city_col      = "city",
  lang          = "pt",
  verbose       = TRUE
)

# Visualizar tabela de escores
swot_fit$scores
```

---

# Function characteristics

## `sus_mod_vulnerability_index()` — IPCC Vulnerability Index

**Full signature:**

```{r assinatura-vi}
sus_mod_vulnerability_index(
  exposure_df          = NULL,
  sensitivity_df       = NULL,
  adaptive_capacity_df = NULL,
  fits                 = NULL,
  city_col             = "city",
  normalize            = "minmax",
  weights              = NULL,
  pillar_weights       = c(exposure = 1, sensitivity = 1, adaptive_capacity = 1),
  hot_percentile       = 0.99,
  cold_percentile      = 0.01,
  lang                 = "pt",
  verbose              = TRUE
)
```

**Key parameters:**

| Argument | Type | Description |
|-----------|------|-----------|
| `exposure_df` | `data.frame` | Indicators for the Exposure pillar (higher = more exposed) |
| `sensitivity_df` | `data.frame` | Indicators for the Sensitivity pillar (higher = more sensitive) |
| `adaptive_capacity_df` | `data.frame` | Indicators for the Adaptive Capacity pillar (higher = more adapted) |
| `fits` | named list of `climasus_dlnm` | Automatic extraction of heat/cold RR and sensitivity index |
| `city_col` | `character` | Name of the municipal identifier column (default `"city"`) |
| `normalize` | `"minmax"` or `"zscore"` | Indicator normalization method |
| `weights` | named vector | Intra-pillar weights by column name; unlisted columns get weight 1 |
| `pillar_weights` | named vector `c(exposure, sensitivity, adaptive_capacity)` | Inter-pillar weights; missing pillars automatically get weight 0 |
| `hot_percentile` / `cold_percentile` | numeric | Quantiles for RR extraction from the DLNM fits |

**Return — `climasus_vi` object:**

- `$vi_table`: tibble ordered by `vi_rank` (rank 1 = most vulnerable), with columns
  `vi_score` ∈ [0, 1], `exposure_score`, `sensitivity_score`,
  `adaptive_capacity_score`, `vi_rank`, `vi_percentile`.
- `$pillar_scores`: long tibble with `city`, `pillar`, `score` (raw score before the
  AC inversion).
- `$normalized_indicators` / `$raw_indicators`: wide tibbles with prefixes `exp_`,
  `sen_`, `ac_`.
- `$weights`: list with the intra-pillar and inter-pillar weights effectively used.
- `$gini_coefficient`: Gini of the CVI distribution.
- `$meta`: list of parameters, counts, and timestamp.

**Available S3 methods:** `print()`, `summary()`, `tidy()` (returns a flat tibble),
`coef()` (returns a named vector city → vi_score).

---

## `sus_mod_plot_vulnerability()` — CVI Visualizations

**Full signature:**

```{r assinatura-plot-vi}
sus_mod_plot_vulnerability(
  x,
  type        = c("ranking", "pillars", "lorenz"),
  output_type = c("plot", "table", "all"),
  interactive = FALSE,
  base_size   = 12L,
  save_plot   = NULL,
  lang        = c("pt", "en", "es"),
  verbose     = FALSE
)
```

**Plot types (`type=`):**

| `type` | Description | Data returned in `"table"` |
|--------|-----------|----------------------------|
| `"ranking"` | Horizontal lollipop: CVI per city, most vulnerable at the top | `$vi_table` |
| `"pillars"` | Horizontal stacked bars: contribution of E / S / AC per city | `$pillar_scores` |
| `"lorenz"` | Lorenz curve of CVI inequality with the Gini annotated | `$vi_table` |

**Return:** `ggplot` (`output_type = "plot"`), tibble (`"table"`), or a named list
`$plot`, `$table`, `$data` (`"all"`). With `interactive = TRUE`, returns a `plotly`
widget (requires the `plotly` package).

---

## `sus_mod_swot()` — Climate-Health SWOT Synthesis

**Full signature:**

```{r assinatura-swot}
sus_mod_swot(
  vulnerability = NULL,
  af            = NULL,
  burden        = NULL,
  dlnm          = NULL,
  sensitivity   = NULL,
  score_type    = c("both", "numeric", "categorical"),
  breaks        = c(33, 66),
  labels        = NULL,
  city_col      = "city",
  lang          = c("pt", "en", "es"),
  verbose       = TRUE
)
```

**Indicators per quadrant:**

| Quadrant | Source | Extracted indicators |
|-----------|-------|-----------------------|
| **S** Strengths | `climasus_vi` | `adaptive_capacity_score` (high capacity = strength) |
| **S** Strengths | `climasus_af` | 100 − total AF% (low burden = strength) |
| **S** Strengths | `climasus_burden` | Inverted burden rank |
| **W** Weaknesses | `climasus_vi` | `sensitivity_score` (high sensitivity = weakness) |
| **W** Weaknesses | `climasus_af` | Heat AF%; cold AF% |
| **W** Weaknesses | `climasus_sensitivity` | RR inequality across strata |
| **O** Opportunities | `climasus_vi` | 100 − CVI percentile (low percentile = window); current exposure inverted |
| **T** Threats | `climasus_vi` | `exposure_score`; `vi_score` |
| **T** Threats | `climasus_dlnm` | Heat RR at P95 (scaled 0–100) |
| **T** Threats | `climasus_burden` | Normalized attributable number |
| **T** Threats | `climasus_sensitivity` | Maximum RR across strata |

**Return — `climasus_swot` object:**

- `$scores`: tibble with one row per entity; columns `entity`, `S_score`,
  `W_score`, `O_score`, `T_score` (0–100) and, when `score_type %in% c("categorical","both")`,
  `S_cat`, `W_cat`, `O_cat`, `T_cat`; counts `n_S`, `n_W`, `n_O`, `n_T`.
- `$indicators`: long tibble with one row per (entity × quadrant × indicator):
  `entity`, `quadrant`, `ind_code`, `indicator`, `raw_value`, `norm_score`,
  `direction` (`"positive"` or `"negative"`).
- `$meta`: list with `n_entities`, `n_indicators`, `inputs_used`, `score_type`,
  `breaks`, `labels`, `lang`, `call_time`.

**S3 methods:** `print()`, `summary()`, `tidy()` (returns `$scores`).

---

## `sus_mod_plot_swot()` — SWOT Visualizations

**Full signature:**

```{r assinatura-plot-swot}
sus_mod_plot_swot(
  x,
  type        = c("radar", "matrix", "bar"),
  output_type = c("plot", "table", "all"),
  interactive = FALSE,
  entities    = NULL,
  top_n       = 3L,
  base_size   = 12L,
  save_plot   = NULL,
  lang        = c("pt", "en", "es"),
  verbose     = FALSE
)
```

**Plot types (`type=`):**

| `type` | Description | Recommended use |
|--------|-----------|-----------------|
| `"radar"` | Spider/radar chart of the four SWOT scores (0–100) per entity, with overlapping polygons for multi-city comparison | Comparing profiles across municipalities |
| `"matrix"` | Classic 2×2 chart: numeric score, category, and bullet points of the leading indicators per quadrant. Shows only one entity (the first in the `entities` selection) | Risk communication for a single municipality |
| `"bar"` | Grouped horizontal bars of the four SWOT scores per entity | Quick comparison across all municipalities |

**Argument `top_n`** (only for `type = "matrix"`): maximum number of indicator bullets
displayed per quadrant (default 3).

**Return:** `ggplot` or `plotly` (`output_type = "plot"`), tibble (`"table"`), or a
list `$plot`, `$table` (`"all"`).

---

# Reference formulas

## Composite Vulnerability Index (IPCC)

The central formula combines the three pillars with the Adaptive Capacity inversion:

$$
VI_i = \frac{w_E \cdot E_i + w_S \cdot S_i + w_{AC} \cdot (1 - AC_i)}{w_E + w_S + w_{AC}}
$$

where each pillar $P_i$ is the weighted average of the normalized indicators:

$$
P_i = \frac{\sum_j \omega_j \tilde{x}_{ij}}{\sum_j \omega_j}
$$

Missing indicators are excluded and the remaining weights are renormalized, partially
preserving municipalities with incomplete data.

## Min-max normalization

$$
\tilde{x}_{ij} = \frac{x_{ij} - \min_j(x_{ij})}{\max_j(x_{ij}) - \min_j(x_{ij})}
$$

When $\max_j - \min_j \approx 0$ (constant variable), the value is fixed at 0.5
(implemented in `.vi_minmax_col()`).

## Gini Coefficient

Given the vector $\mathbf{v}$ of $VI_i$ scores sorted in ascending order:

$$
G = \frac{2 \sum_{i=1}^{n} i \cdot v_i}{n \sum_{i=1}^{n} v_i} - \frac{n+1}{n}
$$

The Gini is reported in `$gini_coefficient` and annotated on the `type = "ranking"`
and `type = "lorenz"` plots.

---

# Illustrated results

## Vulnerability ranking

The lollipop chart orders municipalities from most (top) to least vulnerable, with
the CVI score on the horizontal axis. The Gini Coefficient in the subtitle quantifies
the territorial inequality of vulnerability.

```{r fig-ranking, echo=FALSE, eval=TRUE}
fig_ou_nota(.fig("modvi_ranking"),
            paste("Ranking de vulnerabilidade —", .nota_gerar))
```

**Figure 1.** Climate-health vulnerability ranking of the RMSP municipalities.
Min-max scores on the x-axis; the municipality at the top = most vulnerable (highest
CVI). Gini indicates the concentration of risk across municipalities.

---

## Decomposition of the IPCC pillars

The stacked bars allow identification of which pillar dominates the vulnerability of
each municipality: Exposure (red), Sensitivity (blue), or Adaptive Capacity (green).
Municipalities with balanced pillars indicate multifactorial risk.

```{r fig-pillars, echo=FALSE, eval=TRUE}
fig_ou_nota(.fig("modvi_pillars"),
            paste("Decomposição dos pilares IPCC —", .nota_gerar))
```

**Figure 2.** Decomposition of the three IPCC pillar scores by municipality, ordered
by total CVI (most vulnerable at the top).

---

## Lorenz curve of vulnerability

The Lorenz curve accumulates the proportion of total CVI (y-axis) as a function of
the proportion of municipalities (x-axis), from least to most vulnerable. The farther
from the diagonal (perfect equality), the greater the concentration of risk in a few
municipalities.

```{r fig-lorenz, echo=FALSE, eval=TRUE}
fig_ou_nota(.fig("modvi_lorenz"),
            paste("Curva de Lorenz da vulnerabilidade —", .nota_gerar))
```

**Figure 3.** Lorenz curve of the Composite Vulnerability Index in the RMSP. The
shaded area between the curve and the diagonal of perfect equality (dashed) is
proportional to the Gini Coefficient annotated on the plot.

---

## Vulnerability score table

```{r tbl-vi, echo=FALSE, eval=TRUE}
tbl_ou_nota(.tbl("modvi_vi_table"),
            paste("Tabela de escores —", .nota_gerar),
            caption = "Table 1. Vulnerability scores by municipality (top 10, decreasing order of vulnerability).")
```

---

## SWOT profile — Radar

The radar chart shows the four SWOT scores (0–100) for all municipalities
simultaneously. Larger polygons (especially in W and T) signal greater fragility;
wide polygons in S and O indicate protective assets and intervention opportunities.

```{r fig-radar, echo=FALSE, eval=TRUE}
fig_ou_nota(.fig("modswot_radar"),
            paste("Radar SWOT —", .nota_gerar))
```

**Figure 4.** SWOT profile by municipality in the RMSP. Axes: Strengths (top),
Opportunities (right), Threats (bottom), Weaknesses (left). Scale 0–100.

---

## SWOT 2×2 chart — Most vulnerable municipality

The classic 2×2 chart displays, for the municipality with the highest CVI, the
numeric score and category of each quadrant, together with the leading indicators
(with normalized score) that contributed the most to each quadrant.

```{r fig-matrix, echo=FALSE, eval=TRUE}
fig_ou_nota(.fig("modswot_matrix"),
            paste("Matriz SWOT —", .nota_gerar))
```

**Figure 5.** SWOT 2×2 chart for the most vulnerable municipality in the RMSP. An
ideal format for risk communication to municipal managers.

---

## SWOT scores — Comparative bars

```{r fig-bar, echo=FALSE, eval=TRUE}
fig_ou_nota(.fig("modswot_bar"),
            paste("Barras SWOT —", .nota_gerar))
```

**Figure 6.** Grouped SWOT scores by municipality. Allows simultaneous comparison of
Strengths, Weaknesses, Opportunities, and Threats across all municipalities in the
RMSP.

---

## SWOT score table

```{r tbl-swot, echo=FALSE, eval=TRUE}
tbl_ou_nota(.tbl("modswot_scores"),
            paste("Tabela SWOT —", .nota_gerar),
            caption = "Table 2. SWOT scores and categories by municipality in the RMSP.")
```

---

# References

Andrews, K. (1971). *The Concept of Corporate Strategy*. Dow Jones-Irwin.

Benmarhnia, T., Deguen, S., Kaufman, J. S., & Smargiassi, A. (2015).
Vulnerability to heat-related mortality: a systematic review, meta-analysis,
and meta-regression analysis. *Epidemiology*, 26(6), 781–793.
<https://doi.org/10.1097/EDE.0000000000000375>

Confalonieri, U., Marinho, D., Dias, H., & Malheiros, T. (2009).
O índice de vulnerabilidade aos riscos climáticos do Brasil.
*Parcerias Estratégicas*, 27, 161–172.

Cutter, S. L., Boruff, B. J., & Shirley, W. L. (2003).
Social vulnerability to environmental hazards.
*Social Science & Medicine*, 56(11), 2301–2322.
<https://doi.org/10.1016/S0277-9536(02)00605-X>

Gasparrini, A., & Armstrong, B. (2013).
Attributable risk from distributed lag models.
*BMC Medical Research Methodology*, 13, 3.
<https://doi.org/10.1186/1471-2288-13-3>

Gasparrini, A., et al. (2017).
Projections of temperature-related excess mortality under climate change scenarios.
*The Lancet Planetary Health*, 1(9), e360–e367.
<https://doi.org/10.1016/S2542-5196(17)30156-0>

IPCC. (2014). *Fifth Assessment Report — Working Group II: Impacts, Adaptation and
Vulnerability*. Part A: Global and Sectoral Aspects. Chapter 19: Emergent Risks
and Key Vulnerabilities. Cambridge University Press.
<https://doi.org/10.1017/CBO9781107415379>

IPCC. (2022). *Sixth Assessment Report — Working Group II: Impacts, Adaptation and
Vulnerability*. Chapter 16: Key Risks across Sectors and Regions.
Cambridge University Press.
<https://doi.org/10.1017/9781009325844>

Marengo, J. A., et al. (2022).
Vulnerability to climate change in Brazil: current status and future scenarios.
*International Journal of Climate Change Strategies and Management*, 14(2), 133–149.

Turner, B. L., et al. (2003).
A framework for vulnerability analysis in sustainability science.
*PNAS*, 100(14), 8074–8079.
<https://doi.org/10.1073/pnas.1231335100>

---

# Navigation

> **End of the Modeling section.** This is the final module in the Modeling series
> of `climasus4r`. The results produced here (`mod_vi.rds`) can be loaded into final
> reports and risk communication for managers and decision-makers.

| | |
|---:|:---|
| Previous module | [Modeling 07: Spatial and Spatio-Temporal Bayes](mod-07-bayes.html) |
| Tutorial hub | [Tutorial Index](tutorials.html) |

> **How to reproduce the figures.** Run once, from the package root:
> `Rscript scripts/modelagem_08_vulnerabilidade_swot.R`. The script saves the figures
> in `vignettes-pt/figuras/` (prefixes `modvi_` and `modswot_`), the tables in
> `vignettes-pt/dados/`, and the chained vulnerability object (`mod_vi.rds`).
