---
title: "Modeling 06: Spatial Models — Autocorrelation, Regression, and Scan Statistics"
author: "climasus4r Team"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    code_download: true
    self_contained: true
    df_print: paged
---

```{css, echo=FALSE}
/* -- Variaveis de Cores Modernas -- */
:root {
  --primary-color: #1D9E75;
  --dark-green: #0f3d2e;
  --bg-light: #f8faf9;
  --text-main: #1a1a1a;
  --code-bg: #fdfdfd;
  --warning-orange: #ff9800;
  --info-blue: #2196f3;
}
html { scroll-behavior: smooth; }
body { font-family: 'Source Serif 4', Georgia, serif; font-size: 18px; color: var(--text-main); background-color: #fff; line-height: 1.6; }
h1, h2, h3, h4 { font-family: 'Inter', sans-serif; font-weight: 700; color: var(--dark-green); }
h1.title { font-size: 2.8em; border-bottom: 4px solid var(--primary-color); padding-bottom: 10px; margin-bottom: 1.5rem; }
h2 { margin-top: 2rem; border-left: 4px solid var(--primary-color); padding-left: 1rem; }
pre.r, pre.sourceCode { background-color: var(--code-bg); border: 1px solid #e1e1e1; border-left: 5px solid var(--primary-color) !important; border-radius: 8px; box-shadow: 0 2px 4px rgba(0,0,0,0.05); padding: 1rem; }
.callout { padding: 1.2em 1.5em; margin: 1.5em 0; border-radius: 8px; border-left: 6px solid; background-color: var(--bg-light); }
.callout-note { background-color: #f0f7ff; border-color: #007bff; }
.callout-tip { background-color: #e8f5e9; border-color: #1D9E75; }
.callout-warning { background-color: #fff3e0; border-color: #ff9800; }
.workflow-container { background: linear-gradient(135deg, #f8faf7 0%, #ffffff 100%); border-radius: 12px; padding: 1.5rem; margin: 1.5rem 0; border: 1px solid #dce8e1; overflow-x: auto; }
.workflow-steps { display: flex; flex-wrap: wrap; justify-content: center; gap: 0.5rem; font-family: 'Inter', monospace; font-size: 0.8rem; }
.workflow-step { background: var(--primary-color); color: white; padding: 0.5rem 1rem; border-radius: 20px; white-space: nowrap; }
.workflow-arrow { color: var(--primary-color); font-weight: bold; font-size: 1.2rem; }
.feature-grid { display: grid; grid-template-columns: repeat(auto-fit, minmax(280px, 1fr)); gap: 1.5rem; margin: 2rem 0; }
.feature-card { background: white; border-radius: 12px; padding: 1.5rem; border: 1px solid #e1e1e1; transition: transform 0.2s ease, box-shadow 0.2s ease; text-align: center; }
.feature-card:hover { transform: translateY(-4px); box-shadow: 0 8px 24px rgba(0,0,0,0.1); }
.feature-icon { font-size: 2rem; margin-bottom: 1rem; }
.feature-card h3 { margin-top: 0; color: var(--primary-color); font-size: 1.1rem; }
table { width: 100%; border-collapse: collapse; margin: 1.5rem 0; }
th, td { padding: 12px; text-align: left; border-bottom: 1px solid #e1e1e1; }
th { background-color: var(--bg-light); font-weight: 600; color: var(--dark-green); }
.github-feedback-btn { display: inline-flex; align-items: center; gap: 8px; background-color: #24292e; color: white; padding: 10px 20px; border-radius: 6px; text-decoration: none; font-family: sans-serif; transition: background-color 0.2s; margin-top: 1rem; }
.github-feedback-btn:hover { background-color: #2c3e50; color: white; text-decoration: none; }
@media (max-width: 768px) { body { font-size: 16px; } h1.title { font-size: 2rem; } .workflow-steps { flex-direction: column; align-items: center; } }
```


```{r setup, include=FALSE}
# ATENÇÃO — estratégia de pré-renderização.
# Este tutorial NÃO executa modelos durante a renderização. Todas as figuras e
# tabelas foram geradas UMA vez pelo script scripts/modelagem_06_espacial.R e
# são apenas carregadas aqui (PNG via include_graphics; RDS via readRDS).
# Por isso, os chunks de ensino abaixo usam eval = FALSE: o código é mostrado,
# mas não é executado no momento do knit.
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)

# Helpers para exibir artefatos de forma robusta: se o arquivo ainda não foi
# gerado (script de geração não executado), mostram uma nota em vez de quebrar
# o knit. Assim a vignette renderiza a qualquer momento.
fig_ou_nota <- function(path, nota) {
  if (file.exists(path)) knitr::include_graphics(path)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
tbl_ou_nota <- function(path, nota, caption = NULL) {
  if (file.exists(path)) knitr::kable(readRDS(path), caption = caption)
  else knitr::asis_output(paste0("> *", nota, "*"))
}
.nota_gerar <- "Artefato indisponível — rode `Rscript scripts/modelagem_06_espacial.R` (a partir da raiz do pacote) para gerá-lo."
```

---

# Objective

By the end of this tutorial you will be able to:

1. Build a **Queen spatial weights matrix** from municipal polygons with
   `sus_mod_spatial_weights()`;
2. Compute and interpret the **global Moran's I** and **local LISA clusters**
   (HH/LL/HL/LH) with `sus_mod_spatial_moran()` and `sus_mod_plot_spatial_moran()`;
3. Fit and compare **spatial regression** models (SAR, SEM, SDM, SAC) with
   `sus_mod_spatial_reg()`, extracting coefficients and
   direct/indirect/total impacts;
4. Detect **geographic clusters of excess cases** using Kulldorff's scan
   statistic via `sus_mod_spatial_scan()` and `sus_mod_plot_spatial_scan()`.

This tutorial covers the **spatial analysis stage** of the climasus4r
pipeline — the one that precedes the Bayesian spatial models (Tutorial 07)
and follows the Machine Learning models (Tutorial 05).

---

# Scientific context

## Why does the spatial dimension matter in environmental health?

Health data collected in neighboring municipalities are rarely independent
of one another. Spatial dependence emerges from shared factors (climate,
pollution, access to services, social behavior) that vary smoothly in
space. Ignoring it produces inefficient estimators, incorrect confidence
intervals, and, ultimately, mistaken inferences about the
exposure–outcome association (Anselin, 1988; Bivand, Pebesma & Gomez-Rubio,
2013).

Classical spatial analysis in epidemiology involves three questions:

- **Is there spatial dependence?** (autocorrelation tests)
- **Where are the clusters?** (LISA, Kulldorff scan)
- **Does the effect of an exposure spread to neighboring municipalities?**
  (spillovers; spatial regression)

> **Note on CRS and units.** The functions `sus_mod_spatial_weights()` and
> `sus_mod_spatial_moran()` accept `sf` objects in any CRS, but the `snap`
> parameter of `poly2nb()` is interpreted **in the units of the CRS used**.
> For a geographic CRS (degrees, e.g. SIRGAS 2000 / EPSG:4674, the `geobr`
> default), the value `snap = 1e-3` corresponds to
> ≈ 0.001° ≈ 100 m, adequate for closing small gaps at the boundaries
> of IBGE municipal polygons. For CRS projected in meters (e.g.
> SIRGAS2000 / UTM), values between `1` and `100` meters are more
> appropriate.
>
> **MAUP (Modifiable Areal Unit Problem).** The results of this stage
> depend on the aggregation scale adopted (municipality, micro-region,
> state). A Moran's I calculated at the municipal level may differ
> significantly from one calculated at the micro-region level, even with
> the same raw data. This is due to MAUP (Openshaw & Taylor, 1979):
> the shape and size of spatial units influences every derived indicator.
> It is recommended to check the robustness of results at at least two
> scales before drawing inferences (Fotheringham & Wong, 1991).

## Moran's I

Global Moran's I (Moran, 1950) measures whether high (or low) values tend
to occur close to one another. It is defined as:

$$I = \frac{n}{\sum_{i}\sum_{j}w_{ij}} \frac{\sum_{i}\sum_{j}w_{ij}(x_i-\bar{x})(x_j-\bar{x})}{\sum_{i}(x_i-\bar{x})^2}$$

where $n$ is the number of spatial units, $\sum_{i}\sum_{j} w_{ij}$
is the total sum of spatial weights, $x_i$ is the value of the variable of
interest in unit $i$, $\bar{x}$ is the global mean, and $w_{ij}$ is the
spatial weight between units $i$ and $j$ (with $w_{ii} = 0$).

With **row-standardized** weights (`style = "W"`), each $w_{ij} = 1/n_i$,
where $n_i$ is the number of neighbors of $i$, and $S_0 = n$. Thus the
spatial lag $Wz_i = \sum_j w_{ij} z_j$ is simply the weighted average of
the values at the neighbors.

Values of $I$ close to $+1$ indicate **positive clustering** (spatial
similarity); $I \approx -1$ indicates **spatial dispersion** (units with
high values surrounded by low ones, and vice versa); $I \approx E[I] = -1/(n-1)$
indicates a random distribution.

Inference is performed via Monte Carlo permutation — `sus_mod_spatial_moran()`
uses `spdep::moran.mc()` for the global test and `spdep::localmoran_perm()`
for the local LISA tests.

## LISA — Local Indicators of Spatial Association

LISA indicators (Anselin, 1995) decompose the global I into individual
contributions from each unit, allowing identification of **hotspots** (HH:
high surrounded by high), **coldspots** (LL: low surrounded by low), and
**spatial outliers** (HL and LH). The classification uses the Moran diagram
($z_i$ vs. $Wz_i$, the spatial lag of $z$):

| Quadrant | $z_i$ | $Wz_i$ | Interpretation |
|-----------|-------|---------|---------------|
| HH | > 0 | > 0 | Hotspot |
| LL | < 0 | < 0 | Coldspot |
| HL | > 0 | < 0 | High outlier in a low region |
| LH | < 0 | > 0 | Low outlier in a high region |
| NS | — | — | Not significant |

## Spatial regression models

When spatial autocorrelation is confirmed, OLS models are inadequate.
climasus4r implements four model families via `spatialreg`
(Bivand & Piras, 2015):

| Model | Abbreviation | Parameter | Simplified equation |
|--------|-------|-----------|----------------------|
| Spatial Autoregressive | SAR | $\rho$ | $y = \rho W y + X\beta + \varepsilon$ |
| Spatial Error | SEM | $\lambda$ | $y = X\beta + u;\; u = \lambda W u + \varepsilon$ |
| Spatial Durbin | SDM | $\rho$ + $\theta$ | $y = \rho W y + X\beta + WX\theta + \varepsilon$ |
| Spatial Autocorrelation | SAC | $\rho$, $\lambda$ | $y = \rho W_1 y + X\beta + u;\; u = \lambda W_2 u + \varepsilon$ |

For SAR/SDM/SAC models, the total effect of a covariate is decomposed into
**direct impact** (own unit) + **indirect impact** (spillover to neighbors)
= **total impact** (LeSage & Pace, 2009).

## Kulldorff's scan statistic

Kulldorff's method (1997) imposes circular windows of varying radius over
the map and, for each candidate window $Z$, computes the likelihood ratio
(LR) under the Poisson model, **only when** the internal rate exceeds the
external rate ($C_{in}/E_{in} > C_{out}/E_{out}$):

$$
\text{LR}(Z) = \mathbb{1}\!\left[\frac{C_{in}}{E_{in}} > \frac{C_{out}}{E_{out}}\right]
\cdot
\left(\frac{C_{in}}{E_{in}}\right)^{\!C_{in}}
\left(\frac{C_{out}}{E_{out}}\right)^{\!C_{out}}
$$

where $C_{in}$ and $E_{in}$ are, respectively, the observed and expected
cases **inside** the window, and $C_{out}$ and $E_{out}$ the corresponding
values **outside** the window. The window with the highest LR is the
**most likely cluster**; its p-value is obtained by comparison with the
empirical distribution of maximum LR values under $n_{sim}$ Monte Carlo
permutations (under the null hypothesis of a uniform distribution of cases
proportional to population).

The function `sus_mod_spatial_scan()` delegates the computation to
`SpatialEpi::kulldorff()`, which requires geographic coordinates
(longitude/latitude, WGS 84). For this reason, the polygon centroids are
reprojected to EPSG:4326 internally before the calculation.

---

# Data used

The guiding case is: **"Temperature and health outcomes in older adults
(60+) in the São Paulo Metropolitan Region, 2014–2019"** (respiratory
causes J00–J99).

For this stage we use:

| Object | File | Content |
|--------|---------|----------|
| `rmsp_sf` | `dados/caso_espacial.rds` | RMSP `sf` with `n_obitos` and `pop` per municipality |
| `mod_weights` | `dados/mod_weights.rds` | Queen weights built in this tutorial |

If `caso_espacial.rds` is absent or incomplete, the generation script
synthesizes a minimal, realistic panel for the RMSP with 39 municipalities,
preserving plausible spatial gradients for didactic purposes.

> **IBGE municipality code.** climasus4r uses the **7-digit IBGE code** as
> the unique identifier for municipalities (e.g. `3550308` for São Paulo-SP).
> The first two digits identify the state (35 = São Paulo) and the
> following five identify the municipality within the state. Make sure
> that the `code_muni` column in your `sf` follows this pattern — `geobr`
> already provides it correctly. 6-digit codes (the old standard) must be
> converted before using the `sus_mod_spatial_*()` functions.


---

# Complete pipeline

Load the package and the data before any analysis. The paths below assume
the script is run from the **package root** (`climasus4r/`), which is
where the `scripts/modelagem_06_espacial.R` script saves its artifacts.

```{r carregar}
library(climasus4r)
library(dplyr)

# Dados espaciais da RMSP (gerados por scripts/modelagem_06_espacial.R)
# Caminhos relativos à raiz do pacote (não à pasta vignettes-pt/).
rmsp_sf <- if (file.exists("vignettes-pt/dados/caso_espacial.rds")) {
  obj <- readRDS("vignettes-pt/dados/caso_espacial.rds")
  # Converter para sf e agregar por município (média do período)
  if (requireNamespace("sf", quietly = TRUE) &&
      "geom" %in% names(obj) &&
      inherits(obj[["geom"]], "sfc")) {
    sf_obj <- sf::st_as_sf(obj)
    dplyr::group_by(sf_obj, code_muni) |>
      dplyr::summarise(
        n_obitos  = mean(n_obitos, na.rm = TRUE),
        name_muni = dplyr::first(name_muni),
        pop       = mean(pop, na.rm = TRUE),
        temp_med  = mean(temp_med, na.rm = TRUE),
        .groups   = "drop"
      )
  } else {
    message("Objeto sf não encontrado; verifique caso_espacial.rds.")
    NULL
  }
} else {
  message("Execute scripts/modelagem_06_espacial.R para gerar os dados.")
  NULL
}

# Calcular taxa de mortalidade por 100 mil habitantes
if (!is.null(rmsp_sf) && !"taxa_obitos" %in% names(rmsp_sf)) {
  rmsp_sf$taxa_obitos <- rmsp_sf$n_obitos / rmsp_sf$pop * 1e5
}
```

## Step 1 — Build the spatial weights matrix

**Queen vs. Rook contiguity.** Queen contiguity considers two
municipalities neighbors when their polygons **share a vertex or an
edge** (8 directions). Rook contiguity restricts neighborhood to
municipalities that **share an edge** (4 directions). For IBGE municipal
polygons, where municipalities may touch at only a single point, Queen
contiguity is more appropriate and is the package default (`queen = TRUE`).

**Row standardization (`style = "W"`).** With this normalization, each
weight $w_{ij} = 1/n_i$ (where $n_i$ is the number of neighbors of $i$),
so that $\sum_j w_{ij} = 1$ for every unit $i$ with neighbors. This makes
the spatial lag $Wz_i$ easy to interpret as an **average of the values at
the neighbors** and is the default style for Moran's I and SAR/SEM models.

```{r pesos}
mod_weights <- sus_mod_spatial_weights(
  municipalities = rmsp_sf,
  style          = "W",        # row-standardised: sum(w_ij) = 1 para cada i
  queen          = TRUE,       # contiguidade Queen (vértice ou aresta)
  snap           = 1e-3,       # tolerância para gaps (em graus, CRS SIRGAS2000)
  zero_policy    = TRUE,       # permite ilhas espaciais (sem vizinho)
  return_matrix  = FALSE,      # não retorna a matriz densa W (economia de memória)
  lang           = "pt",
  verbose        = TRUE
)
print(mod_weights)

# Salvar para reuso em tutoriais subsequentes (Tutorial 07 — Bayes Espacial)
save_tbl(mod_weights, "mod_weights")
```

## Step 2 — Global spatial autocorrelation + LISA

```{r moran}
# Dados tabulares sem geometria (sus_mod_spatial_moran aceita data.frame ou
# climasus_df; a coluna code_muni deve estar presente)
df_tabular <- sf::st_drop_geometry(rmsp_sf)

# A taxa já foi calculada no passo de carregamento; verificar e recalcular
# se necessário
if (!"taxa_obitos" %in% names(df_tabular)) {
  df_tabular$taxa_obitos <- df_tabular$n_obitos / df_tabular$pop * 1e5
}

moran_resultado <- sus_mod_spatial_moran(
  df           = df_tabular,
  outcome      = "taxa_obitos",  # coluna numérica de desfecho (arg: "outcome")
  W            = mod_weights,    # climasus_weights do passo anterior
  municipalities = NULL,         # NULL: usa todas as linhas na ordem de W
  permutations = 999L,
  alpha        = 0.05,
  adjust_p     = "fdr",          # correção FDR (Benjamini-Hochberg)
  lang         = "pt",
  verbose      = TRUE
)
print(moran_resultado)
```

```{r plot-moran-mapa}
# Mapa LISA de clusters (HH/LL/HL/LH/NS)
sus_mod_plot_spatial_moran(
  x              = moran_resultado,
  municipalities = rmsp_sf,
  type           = "map",
  alpha          = 0.05,
  lang           = "pt"
)
```

```{r plot-moran-scatter}
# Diagrama de dispersão de Moran (z vs. Wz)
sus_mod_plot_spatial_moran(
  x    = moran_resultado,
  type = "scatter",
  lang = "pt"
)
```

```{r plot-moran-both}
# Ambos os painéis lado a lado (requer patchwork)
sus_mod_plot_spatial_moran(
  x              = moran_resultado,
  municipalities = rmsp_sf,
  type           = "both",
  lang           = "pt"
)
```

## Step 3 — Spatial regression (SAR / SEM / SDM / SAC)

```{r reg-sar}
# Garantir que df_reg não contém geometria (sus_mod_spatial_reg descarta
# automaticamente, mas é boa prática passar um data.frame puro)
df_reg <- sf::st_drop_geometry(rmsp_sf)

# Verificar / calcular taxa
if (!"taxa_obitos" %in% names(df_reg)) {
  df_reg$taxa_obitos <- df_reg$n_obitos / df_reg$pop * 1e5
}

# Modelo SAR (Spatial Lag) — rho: spillover do desfecho entre municípios
# Estimado via spatialreg::lagsarlm() com método de ML exato ("eigen")
reg_sar <- sus_mod_spatial_reg(
  df          = df_reg,
  formula     = taxa_obitos ~ temp_med,
  W           = mod_weights,
  model       = "lag",      # SAR: y = rho * W*y + X*beta + epsilon
  method      = "eigen",    # ML exato; use "LU" ou "Chebyshev" para n > 2.000
  zero_policy = TRUE,
  lang        = "pt",
  verbose     = TRUE
)
print(reg_sar)

# Impactos direto/indireto/total (calculados internamente com R = 200 simulações)
# $impacts é NULL para modelos SEM (sem lag do desfecho)
reg_sar$impacts

# Diagnóstico: I de Moran nos resíduos — deve ser não-significativo se o modelo
# capturou a autocorrelação espacial adequadamente
reg_sar$moran_residuals
```

```{r reg-sem}
# Modelo SEM (Spatial Error) — lambda: erro espacialmente autocorrelacionado
# Indicado quando a autocorrelação é devida a variáveis omitidas espacialmente
# estruturadas (e.g. poluição não mensurada, gradientes socioeconômicos)
reg_sem <- sus_mod_spatial_reg(
  df      = df_reg,
  formula = taxa_obitos ~ temp_med,
  W       = mod_weights,
  model   = "error",   # SEM: y = X*beta + u; u = lambda * W*u + epsilon
  lang    = "pt",
  verbose = FALSE
)

# Comparação de AIC: menor AIC = melhor ajuste (penalizando complexidade)
# OLS ignora autocorrelação; SAR/SEM incorporam de formas distintas
data.frame(
  modelo = c("OLS", "SAR", "SEM"),
  AIC    = c(reg_sar$lm_aic, reg_sar$aic, reg_sem$aic)
)
```

```{r reg-sdm}
# Modelo SDM (Spatial Durbin) — rho + covariáveis defasadas espacialmente
# Estimado via lagsarlm(Durbin = TRUE): inclui W*X além do W*y
# Recomendado quando há razão para acreditar que covariáveis de municípios
# vizinhos também afetam o desfecho (e.g. temperatura regional)
reg_sdm <- sus_mod_spatial_reg(
  df      = df_reg,
  formula = taxa_obitos ~ temp_med,
  W       = mod_weights,
  model   = "durbin",  # SDM: y = rho*W*y + X*beta + W*X*theta + epsilon
  lang    = "pt",
  verbose = FALSE
)
reg_sdm$impacts   # direto, indireto e total por covariável
```

```{r reg-sac}
# Modelo SAC (rho + lambda) — modelo mais geral; lag do desfecho E erro espacial
# Estimado via spatialreg::sacsarlm(); usa a mesma matriz W para ambos
# os componentes por padrão
reg_sac <- sus_mod_spatial_reg(
  df      = df_reg,
  formula = taxa_obitos ~ temp_med,
  W       = mod_weights,
  model   = "sac",   # SAC: y = rho*W*y + X*beta + u; u = lambda*W*u + epsilon
  lang    = "pt",
  verbose = FALSE
)
reg_sac$rho    # parâmetro de lag do desfecho
reg_sac$lambda # parâmetro de erro espacial
```

## Step 4 — Cluster detection: Kulldorff scan

```{r scan}
# sus_mod_spatial_scan() requer um data.frame com code_muni, cases e population.
# Os centroides dos polígonos são reprojetados para WGS 84 (EPSG:4326)
# internamente para calcular as distâncias usadas pelas janelas circulares.
df_scan <- data.frame(
  code_muni  = as.character(rmsp_sf$code_muni),
  cases      = as.integer(round(rmsp_sf$n_obitos)),
  population = as.integer(rmsp_sf$pop),
  stringsAsFactors = FALSE
)

scan_resultado <- sus_mod_spatial_scan(
  df             = df_scan,
  cases          = "cases",
  population     = "population",
  municipalities = rmsp_sf,
  expected       = NULL,      # NULL → Poisson simples (taxa global × pop_i)
  max_pop_frac   = 0.5,       # janela máxima contém ≤ 50% da pop. total
  n_simulations  = 999L,      # para publicação use 9999L
  alpha          = 0.05,
  lang           = "pt",
  verbose        = TRUE
)
print(scan_resultado)

# Cluster mais provável: location_ids, RR, log_lik, p_value
scan_resultado$most_likely_cluster

# Clusters secundários (lista de listas, mesma estrutura)
scan_resultado$secondary_clusters
```

```{r plot-scan-rr}
# Mapa de Risco Relativo (escala contínua divergente)
sus_mod_plot_spatial_scan(
  x              = scan_resultado,
  municipalities = rmsp_sf,
  show_rr        = TRUE,
  alpha          = 0.05,
  lang           = "pt"
)
```

```{r plot-scan-cat}
# Mapa categórico (mais provável / secundário / fundo)
sus_mod_plot_spatial_scan(
  x              = scan_resultado,
  municipalities = rmsp_sf,
  show_rr        = FALSE,
  alpha          = 0.05,
  lang           = "pt"
)
```

---

# Function reference

## `sus_mod_spatial_weights()` — Spatial weights matrix

**Purpose.** Builds a `climasus_weights` object (a list subclass) with the
spatial neighborhood structure required by all other spatial models in
climasus4r.

**Main arguments:**

| Argument | Default | Meaning |
|-----------|--------|-------------|
| `municipalities` | — | `sf` with municipal polygons |
| `style` | `"W"` | Normalization: `"W"` (row-standardized), `"B"` (binary), `"C"`, `"U"`, `"S"`, `"minmax"` |
| `queen` | `TRUE` | `TRUE` = Queen (vertex or edge); `FALSE` = Rook (edge only) |
| `snap` | `NULL` → `1e-3` | Tolerance for boundary gaps |
| `zero_policy` | `TRUE` | Allow islands (municipalities with no neighbor) |
| `return_matrix` | `FALSE` | Return the dense $W$ matrix ($n \times n$) |

**Return value (`climasus_weights`):**

```
$listw      — objeto spdep::listw pronto para modelagem
$nb         — lista de vizinhos (spdep::nb)
$n_regions  — número total de regiões
$n_islands  — número de ilhas (sem vizinhos)
$island_ids — códigos das ilhas (vetor de caracteres, vazio se nenhuma)
$W          — matriz densa W n×n (NULL se return_matrix = FALSE)
$style      — estilo de normalização usado
$call       — chamada original
```

> **Practical tip.** For IBGE data with a geographic CRS (SIRGAS2000 /
> EPSG:4674), use `snap = 1e-3` (default). For polygons in a projected CRS
> (meters), adjust `snap` to 1–100 m. Municipalities with no neighbors
> (islands, e.g. Fernando de Noronha) do not cause an error with
> `zero_policy = TRUE`, but result in an `NA` spatial lag — check
> `$island_ids`. Use `style = "B"` (binary) only when you want to weight
> by the presence of a neighbor relationship without normalizing by the
> number of neighbors (e.g. in ICAR-type Bayesian models).

---

## `sus_mod_spatial_moran()` — Global Moran's I + LISA

**Purpose.** Tests global spatial autocorrelation via Moran's I (through
Monte Carlo permutation) and classifies each unit into a LISA quadrant.

**Main arguments:**

| Argument | Default | Meaning |
|-----------|--------|-------------|
| `df` | — | `data.frame` or `climasus_df` with `code_muni` and the outcome column |
| `outcome` | — | Name of the numeric outcome column (e.g. `"taxa_obitos"`) |
| `W` | — | `climasus_weights` object from `sus_mod_spatial_weights()` |
| `municipalities` | `NULL` | Vector of `code_muni` values to filter/reorder `df` |
| `permutations` | `999L` | Monte Carlo permutations (`nsim` in `spdep::moran.mc`) |
| `alpha` | `0.05` | Significance level for LISA classification |
| `adjust_p` | `"fdr"` | Multiple-testing correction: `"fdr"`, `"bonferroni"`, `"none"` |

**Return value (`climasus_spatial_moran`):**

```
$global       — data.frame com I, E.I, Var.I, Z.I, p_value, p_simulated
$local        — data.frame com code_muni, Ii, Z.Ii, p_raw, p_adj, quadrant
$n_HH         — número de hotspots (HH)
$n_LL         — número de coldspots (LL)
$n_HL         — número de outliers HL
$n_LH         — número de outliers LH
$outcome_name — nome da coluna de desfecho
$call         — chamada original
```

> **About `p_value` vs `p_simulated`.** The `p_value` in the `$global`
> slot is the asymptotic p-value (normal distribution). `p_simulated` is
> the Monte Carlo permutation p-value — preferable for small samples
> (n < 30 municipalities) or skewed distributions. For the local LISA
> tests (`$local`), the p-values are already conditional-permutation
> based (`spdep::localmoran_perm()`), corrected using the method specified
> in `adjust_p`.

---

## `sus_mod_plot_spatial_moran()` — LISA visualization

**Purpose.** Produces a LISA choropleth map and/or a Moran scatterplot from
a `climasus_spatial_moran` object.

**Available types (`type=`):**

| `type` | Description | Requires `municipalities` |
|--------|-----------|------------------------|
| `"map"` | HH/LL/HL/LH/NS choropleth map | Yes |
| `"scatter"` | $z$ vs. $Wz$ scatterplot | No |
| `"both"` | Both side by side via patchwork | Yes |

**Return value:** a `ggplot` object (or a named list `list(map, scatter)` if
patchwork is not available for `type = "both"`).

---

## `sus_mod_spatial_reg()` — Spatial regression

**Purpose.** Fits SAR/SEM/SDM/SAC maximum-likelihood models via the
`spatialreg` package, reporting coefficients, spatial parameters,
impacts, and residual Moran's I.

**Available models (`model=`):**

| `model` | Family | Spatial parameter | Estimated via |
|---------|---------|--------------------|--------------|
| `"lag"` | SAR | $\rho$ | `spatialreg::lagsarlm()` |
| `"error"` | SEM | $\lambda$ | `spatialreg::errorsarlm()` |
| `"durbin"` | SDM | $\rho$ + $\theta_{WX}$ | `spatialreg::lagsarlm(Durbin=TRUE)` |
| `"sac"` | SAC | $\rho$ + $\lambda$ | `spatialreg::sacsarlm()` |

**Main arguments:**

| Argument | Default | Meaning |
|-----------|--------|-------------|
| `df` | — | `data.frame` or `climasus_df` (`sf` geometry is discarded automatically) |
| `formula` | — | Standard R formula (`outcome ~ cov1 + cov2`) |
| `W` | — | `climasus_weights` object from `sus_mod_spatial_weights()` |
| `model` | `"lag"` | Model family (see table above) |
| `method` | `"eigen"` | Estimation method: `"eigen"` (exact ML), `"LU"`, `"Chebyshev"`, `"MC"` |
| `zero_policy` | `TRUE` | Allow islands (municipalities with no neighbors) |
| `lang` | `"pt"` | Message language |
| `verbose` | `TRUE` | Show progress messages |

**Return value (`climasus_spatial_reg`):**

```
$model           — tipo de modelo ("lag", "error", "durbin", "sac")
$model_label     — rótulo legível ("SAR (Spatial Lag)", ...)
$coefficients    — data.frame: term, estimate, std_error, z_value, p_value
$rho             — parâmetro de lag espacial ρ (NULL para SEM)
$lambda          — parâmetro de erro espacial λ (NULL para SAR/SDM)
$impacts         — data.frame: term, direct, indirect, total
                   (calculado com R = 200 via spatialreg::impacts();
                    NULL para SEM, onde os impactos = coeficientes OLS)
$aic             — AIC do modelo espacial
$lm_aic          — AIC do baseline OLS (mesma fórmula, mesmos dados)
$moran_residuals — lista: I, p_value, z (teste de Moran nos resíduos)
$fitted          — valores ajustados
$residuals       — resíduos do modelo
$call            — chamada original
```

> **Residual diagnostics.** The `$moran_residuals` element contains the
> result of `spdep::moran.test()` applied to the model's residuals. A
> **non-significant** p-value (> 0.05) indicates that the spatial model
> adequately absorbed the spatial autocorrelation. If the residual Moran's
> I remains significant even after SAR/SEM, consider the SDM (spatially
> lagged covariates) or a hierarchical Bayesian model (Tutorial 07).

---

## `sus_mod_spatial_scan()` — Kulldorff scan

**Purpose.** Applies Kulldorff's circular scan statistic to detect the most
likely cluster and secondary clusters of excess cases, with Monte Carlo
inference.

**Main arguments:**

| Argument | Default | Meaning |
|-----------|--------|-------------|
| `df` | — | `data.frame` with `code_muni`, `cases`, and `population` columns |
| `cases` | — | Name of the case-count column (integers ≥ 0) |
| `population` | — | Name of the population-at-risk column |
| `municipalities` | — | `sf` with polygons and `code_muni` |
| `expected` | `NULL` | Name of the expected-count column (NULL: global rate × pop_i) |
| `max_pop_frac` | `0.5` | Maximum fraction of the total population in a single window |
| `n_simulations` | `999L` | Monte Carlo simulations (use `9999L` for publication) |
| `alpha` | `0.05` | Significance level for secondary clusters |
| `lang` | `"pt"` | Message language |
| `verbose` | `TRUE` | Show progress messages |

**Return value (`climasus_spatial_scan`):**

```
$most_likely_cluster — lista: location_ids (code_muni), observed, expected,
                       RR (razão de verossimilhança de excesso), log_lik, p_value
$secondary_clusters  — lista de listas com a mesma estrutura
$n_clusters          — total de clusters significativos (incluindo o mais provável)
$data                — tibble: code_muni, cases, population, expected, in_mlc
$meta                — parâmetros: cases_col, population_col, max_pop_frac,
                       n_simulations, alpha, n_municipalities, call_time
$call                — chamada original
```

> **CRS and centroids.** The function `sus_mod_spatial_scan()` uses
> `sf::st_centroid()` after reprojecting the polygons to WGS 84 (EPSG:4326),
> since `SpatialEpi::kulldorff()` requires longitude/latitude coordinates.
> The `municipalities` object may be in any CRS — the reprojection is
> performed internally. The scan window is circular in geographic space
> (lat/lon), not in projected space.

---

## `sus_mod_plot_spatial_scan()` — Cluster map

**Purpose.** Renders a choropleth map of the Kulldorff scan results with
continuous or categorical RR encoding.

**Argument `show_rr`:**

| `show_rr` | Encoding | Scale |
|-----------|-------------|--------|
| `TRUE` (default) | Cluster Relative Risk (continuous) | Diverging gradient (navy–white–red, centered at RR = 1) |
| `FALSE` | Cluster label (categorical) | Discrete colors per cluster |

**Return value:** a `ggplot` object.

---

# Illustrated results

## Figure 1 — LISA cluster map of respiratory mortality (RMSP)

The map below shows the LISA classification of each RMSP municipality
based on the average respiratory mortality rate for the period 2014–2019.
Municipalities in **red (HH)** form spatial hotspots — they show an
elevated rate and are surrounded by neighbors with an equally elevated
rate.

```{r fig-lisa-mapa, eval=TRUE, echo=FALSE, fig.cap="**Figure 1.** LISA cluster map (local spatial autocorrelation) for the respiratory mortality rate in older adults in the RMSP, 2014–2019. HH = hotspot; LL = coldspot; HL/LH = spatial outlier; NS = not significant (FDR 5%)."}
fig_ou_nota(
  "figuras/modmoran_lisa_mapa.png",
  .nota_gerar
)
```

## Figure 2 — Moran scatterplot

The scatterplot positions each municipality in the plane ($z_i$, $Wz_i$):
the slope of the regression line corresponds to the global Moran's I.
Points in the upper-right (HH) and lower-left (LL) quadrants indicate
positive clustering; points in the cross quadrants (HL, LH) are spatial
outliers.

```{r fig-lisa-scatter, eval=TRUE, echo=FALSE, fig.cap="**Figure 2.** Moran scatterplot — standardized value of the mortality rate (z) versus spatial lag (Wz). The slope of the gray line corresponds to the global Moran's I. Points colored by LISA quadrant."}
fig_ou_nota(
  "figuras/modmoran_lisa_scatter.png",
  .nota_gerar
)
```

## Table 1 — Global Moran's I

```{r tbl-moran, eval=TRUE, echo=FALSE}
tbl_ou_nota(
  "dados/modmoran_moran_global.rds",
  .nota_gerar,
  caption = "**Table 1.** Global Moran's I for the respiratory mortality rate. p_simulated: Monte Carlo permutation p-value (999 permutations); p_value: asymptotic p-value."
)
```

## Figure 3 — Direct, indirect, and total impacts (SAR)

In the SAR model, the coefficient $\beta$ of a covariate **does not
directly represent the effect on the outcome** — because the matrix
$(I - \rho W)^{-1}$ propagates any shock through the entire neighbor
network. The impacts are decomposed into (LeSage & Pace, 2009):

- **Direct impact**: average effect of the covariate in the **municipality
  itself**;
- **Indirect impact** (spillover): average effect in **neighboring
  municipalities**, mediated by $\rho$ and the structure of $W$;
- **Total impact** = direct + indirect.

The package uses `spatialreg::impacts()` with R = 200 simulations
(Monte Carlo reference distribution) to build the confidence intervals.

```{r fig-sar-impactos, eval=TRUE, echo=FALSE, fig.cap="**Figure 3.** Decomposition of the direct, indirect, and total impacts of average temperature on the respiratory mortality rate. SAR (spatial lag) model, R = 200 simulations via spatialreg::impacts()."}
fig_ou_nota(
  "figuras/modmoran_sar_impactos.png",
  .nota_gerar
)
```

## Table 2 — SAR model coefficients

```{r tbl-sar, eval=TRUE, echo=FALSE}
tbl_ou_nota(
  "dados/modmoran_coef_sar.rds",
  .nota_gerar,
  caption = "**Table 2.** Coefficients of the SAR (spatial lag) model. std_error: standard error; z_value: z statistic; p_value: two-sided p-value (asymptotic t-tests)."
)
```

## Table 3 — AIC comparison (OLS / SAR / SEM)

The choice between OLS, SAR, and SEM can be guided by the AIC (Akaike
information criterion). The model with the lowest AIC (penalizing
complexity) is preferable. A substantial AIC reduction in SAR or SEM
relative to OLS confirms that spatial autocorrelation is relevant.

```{r tbl-aic, eval=TRUE, echo=FALSE}
tbl_ou_nota(
  "dados/modmoran_aic_comparacao.rds",
  .nota_gerar,
  caption = "**Table 3.** AIC comparison between OLS, SAR (spatial lag), and SEM (spatial error). Lower AIC = better fit penalized for complexity."
)
```

## Table 4 — Diagnostic: Moran's I on residuals (SAR)

After fitting the SAR model, Moran's I on the residuals should be
non-significant, indicating that spatial autocorrelation was adequately
captured. A residual Moran's I that remains significant suggests
insufficient specification and points toward more complex models (SDM or
hierarchical Bayes).

```{r tbl-moran-res, eval=TRUE, echo=FALSE}
tbl_ou_nota(
  "dados/modmoran_moran_residuos_sar.rds",
  .nota_gerar,
  caption = "**Table 4.** Moran's I on the residuals of the SAR model. z: standardized statistic; p_value: asymptotic p-value."
)
```

## Figure 4 — Relative Risk map (Kulldorff, continuous RR)

The map below shows the RR for each municipality belonging to a
significant cluster detected by the Kulldorff scan. Municipalities in
intense red belong to the cluster with the greatest excess risk.
Municipalities with no color (light gray) are part of the background —
they do not belong to any cluster.

```{r fig-scan-rr, eval=TRUE, echo=FALSE, fig.cap="**Figure 4.** Relative Risk (RR) map of the most likely cluster detected by the Kulldorff circular scan (999 simulations). Thick borders indicate municipalities in a significant cluster (α = 5%)."}
fig_ou_nota(
  "figuras/modmoran_scan_rr.png",
  .nota_gerar
)
```

## Figure 5 — Categorical cluster map (Kulldorff)

The categorical version visually distinguishes the **most likely
cluster** from **secondary clusters** and the background, facilitating
communication with health managers who need to identify priority
intervention regions.

```{r fig-scan-cat, eval=TRUE, echo=FALSE, fig.cap="**Figure 5.** Categorical map of spatial clusters (Kulldorff scan). Red = most likely cluster; other colors = secondary clusters; gray = background (no cluster)."}
fig_ou_nota(
  "figuras/modmoran_scan_cat.png",
  .nota_gerar
)
```

---

# References

Anselin, L. (1988). *Spatial Econometrics: Methods and Models*. Kluwer
Academic Publishers.

Anselin, L. (1995). Local indicators of spatial association — LISA.
*Geographical Analysis*, 27(2), 93–115.

Bivand, R. S., Pebesma, E., & Gomez-Rubio, V. (2013).
*Applied Spatial Data Analysis with R* (2nd ed.). Springer.

Bivand, R. S., & Piras, G. (2015). Comparing implementations of estimation
methods for spatial econometrics. *Journal of Statistical Software*,
63(18), 1–36.

Bivand, R. S., & Wong, D. W. S. (2018). Comparing implementations of
global indicators of spatial association. *TEST*, 27(3), 716–748.

Fotheringham, A. S., & Wong, D. W. S. (1991). The modifiable areal unit
problem in multivariate statistical analysis. *Environment and Planning A*,
23(7), 1025–1044.

Kulldorff, M. (1997). A spatial scan statistic. *Communications in
Statistics — Theory and Methods*, 26(6), 1481–1496.

Kulldorff, M., & Nagarwalla, N. (1995). Spatial disease clusters: detection
and inference. *Statistics in Medicine*, 14(8), 799–810.

LeSage, J., & Pace, R. K. (2009). *Introduction to Spatial Econometrics*.
CRC Press / Taylor & Francis.

Moran, P. A. P. (1950). Notes on continuous stochastic phenomena.
*Biometrika*, 37(1–2), 17–23.

Openshaw, S., & Taylor, P. J. (1979). A million or so correlation
coefficients: three experiments on the modifiable areal unit problem. In
N. Wrigley (Ed.), *Statistical Applications in the Spatial Sciences*
(pp. 127–144). Pion.

Tiefelsdorf, M., Griffith, D. A., & Boots, B. (1999). A variance-
stabilizing coding scheme for spatial link matrices. *Environment and
Planning A*, 31(1), 165–180.

---

# Navigation

- [Back to tutorial hub](tutorials.html)
- **Previous:** [Modeling 05 — Machine Learning](mod-05-ml.html)
- **Next:** [Modeling 07 — Spatial Bayes](mod-07-bayes.html)

> **How to reproduce the figures.** Run once, from the package root:
> `Rscript scripts/modelagem_06_espacial.R`. The script saves the figures
> to `vignettes-pt/figuras/`, the tables to `vignettes-pt/dados/`, and the
> `mod_weights.rds` object that Tutorial 07 (Spatial Bayes) will load.
