---
title: "PART 2: Integration of Climate and Health Data"
subtitle: "Practical Guide::sus_climate_aggregate()"
author:
  - name: "Max Anjos"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    code_download: true
    self_contained: true
    df_print: paged
---

```{css, echo=FALSE}
/* ── Variáveis de Cores Modernas ────────────────── */
:root {
  --primary-color: #1D9E75;
  --dark-green: #0f3d2e;
  --bg-light: #f8faf9;
  --text-main: #1a1a1a;
  --code-bg: #fdfdfd;
  --warning-orange: #ff9800;
  --info-blue: #2196f3;
}

html {
  scroll-behavior: smooth;
}

body {
  font-family: 'Source Serif 4', Georgia, serif;
  font-size: 18px;
  color: var(--text-main);
  background-color: #fff;
  line-height: 1.6;
}

/* ── Títulos com Inter-spacing ─────────────────── */
h1, h2, h3, h4 {
  font-family: 'Inter', sans-serif;
  font-weight: 700;
  color: var(--dark-green);
}

h1.title {
  font-size: 2.8em;
  border-bottom: 4px solid var(--primary-color);
  padding-bottom: 10px;
  margin-bottom: 1.5rem;
}

h2 {
  margin-top: 2rem;
  border-left: 4px solid var(--primary-color);
  padding-left: 1rem;
}

/* ── Code Blocks Estilizados ────────────────────── */
pre.r, pre.sourceCode {
  background-color: var(--code-bg);
  border: 1px solid #e1e1e1;
  border-left: 5px solid var(--primary-color) !important;
  border-radius: 8px;
  box-shadow: 0 2px 4px rgba(0,0,0,0.05);
  padding: 1rem;
}

/* ── Custom Callouts (Estilo Quarto) ────────────── */
.callout {
  padding: 1.2em 1.5em;
  margin: 1.5em 0;
  border-radius: 8px;
  border-left: 6px solid;
  background-color: var(--bg-light);
}

.callout-note {
  background-color: #f0f7ff;
  border-color: #007bff;
}

.callout-tip {
  background-color: #e8f5e9;
  border-color: #1D9E75;
}

.callout-warning {
  background-color: #fff3e0;
  border-color: #ff9800;
}

.callout-info {
  background-color: #e3f2fd;
  border-color: #2196f3;
}

/* ── Workflow container ─────────────────────────── */
.workflow-container {
  background: linear-gradient(135deg, #f8faf7 0%, #ffffff 100%);
  border-radius: 12px;
  padding: 1.5rem;
  margin: 1.5rem 0;
  border: 1px solid var(--fc-border);
  overflow-x: auto;
}

.workflow-steps {
  display: flex;
  flex-wrap: wrap;
  justify-content: center;
  gap: 0.5rem;
  font-family: 'Inter', monospace;
  font-size: 0.8rem;
}

.workflow-step {
  background: var(--primary-color);
  color: white;
  padding: 0.5rem 1rem;
  border-radius: 20px;
  white-space: nowrap;
}

.workflow-arrow {
  color: var(--primary-color);
  font-weight: bold;
  font-size: 1.2rem;
}

/* ── Feature cards ───────────────────────────────── */
.feature-grid {
  display: grid;
  grid-template-columns: repeat(auto-fit, minmax(280px, 1fr));
  gap: 1.5rem;
  margin: 2rem 0;
}

.feature-card {
  background: white;
  border-radius: 12px;
  padding: 1.5rem;
  border: 1px solid #e1e1e1;
  transition: transform 0.2s ease, box-shadow 0.2s ease;
  text-align: center;
}

.feature-card:hover {
  transform: translateY(-4px);
  box-shadow: 0 8px 24px rgba(0,0,0,0.1);
}

.feature-icon {
  font-size: 2rem;
  margin-bottom: 1rem;
}

.feature-card h3 {
  margin-top: 0;
  color: var(--primary-color);
  font-size: 1.1rem;
}

.feature-card p {
  font-size: 0.85rem;
  color: var(--text-main);
  margin-bottom: 0;
}

/* ── Table styling ───────────────────────────────── */
table {
  width: 100%;
  border-collapse: collapse;
  margin: 1.5rem 0;
}

th, td {
  padding: 12px;
  text-align: left;
  border-bottom: 1px solid #e1e1e1;
}

th {
  background-color: var(--bg-light);
  font-weight: 600;
  color: var(--dark-green);
}

tr:hover {
  background-color: var(--bg-light);
}

/* ── Figure styling ─────────────────────────────── */
.figure {
  margin: 1.5rem 0;
  text-align: center;
}

.figure img {
  border-radius: 8px;
  box-shadow: 0 4px 8px rgba(0,0,0,0.1);
  max-width: 100%;
  height: auto;
}

.figure .caption {
  font-size: 0.85rem;
  color: #666;
  margin-top: 8px;
  font-style: italic;
}

/* ── Botão GitHub aprimorado ───────────────────── */
.github-feedback-btn {
  display: inline-flex;
  align-items: center;
  gap: 8px;
  background-color: #24292e;
  color: white;
  padding: 10px 20px;
  border-radius: 6px;
  text-decoration: none;
  font-family: sans-serif;
  transition: background-color 0.2s;
  margin-top: 1rem;
}

.github-feedback-btn:hover {
  background-color: #2c3e50;
  color: white;
  text-decoration: none;
}

/* ── Responsive ─────────────────────────────────── */
@media (max-width: 768px) {
  body {
    font-size: 16px;
  }

  h1.title {
    font-size: 2rem;
  }

  .workflow-steps {
    flex-direction: column;
    align-items: center;
  }

  .workflow-arrow {
    transform: rotate(90deg);
  }

  .feature-grid {
    grid-template-columns: 1fr;
  }

  th, td {
    padding: 8px;
    font-size: 0.85rem;
  }
}

@media (prefers-reduced-motion: reduce) {
  .feature-card,
  .github-feedback-btn {
    transition: none;
  }
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = FALSE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.align = "center",
  out.width = "80%",
  fig.retina = 2
)
```

# Introduction

::: callout-tip
**Complete Tutorial** — This tutorial covers everything from the basic temporal strategies to the most advanced ones of the `sus_climate_aggregate()` function, the mathematical and epidemiological engine of `climasus4r` responsible for aligning health data (DATASUS) with meteorological time series (INMET).
:::

# Function Inputs

For the integration to happen, you need to provide two ready `climasus_df` objects:

| Argument | Description | `stage` Requirement in Metadata |
|-----------|-----------|-----------------------------------|
| `health_data` | DATASUS health data already spatially aggregated. | Must be at the `"spatial"` stage (output of `sus_spatial_join()`). |
| `climate_data` | INMET meteorological data with no gaps. | Must be at the `"climate"` stage (output of `sus_climate_inmet()` and/or `sus_climate_fill()`). |
| `climate_var` | Which climate variables to pull (e.g., `"tair_dry_bulb_c"`, `"rainfall_mm"`). Default: `"all"`. | - |
| `temporal_strategy` | One of the 10 strategies detailed in the previous tutorial. | - |
| `climate_region` | `"auto"`, `"tropical"`, `"subtropical"`, or `"temperate"`. Automatically adapts parameters. | - |
| `window_days` | Integer. Required for `moving_window`, `degree_days`, and the exceedance/weighting strategies. | - |
| `lag_days` | Integer (or vector). Required for `discrete_lag` and `distributed_lag`. | - |
| `offset_days` | Vector of 2 integers (e.g., `c(7, 14)`). Required for `offset_window`. | - |

::: callout-warning
**Attention:** The function will abort with messages in Portuguese, English, or Spanish (`lang` parameter) if the metadata indicates that the data has not gone through the correct pipeline stages. This prevents silent biases in your analysis.
:::

---

# Part 1: Health Data Preparation (SIM-DO)

## 1.1 Import and Cleaning of Death Data

The pipeline below processes data from the Mortality Information System (SIM) for the state of Amazonas in 2023.

```{r step1_health_data, eval=FALSE}
# ============================================================================
# COMPLETE HEALTH DATA PROCESSING PIPELINE
# ============================================================================
library(climasus4r)
df_sim <- sus_data_import(
  uf = "AM",                    # Amazonas
  year = 2023,                  # Full year
  system = "SIM-DO",            # SIM Death Certificate
  use_cache = TRUE,             # Use cache to avoid re-downloading
  verbose = TRUE
) |>
  # Encoding cleanup (accents, special characters)
  sus_data_clean_encoding() |>

  # Column standardization (names, types, formats)
  sus_data_standardize(lang = "pt") |>

  # CID filter - respiratory diseases (J00-J99)
  sus_data_filter_cid(disease_group = "respiratory") |>

  # Creation of derived variables
  sus_data_create_variables(
    create_age_groups = TRUE,                    # Age groups
    age_breaks = c(60, 70, 80, 90, Inf),        # Bins for the elderly
    age_labels = c("60-69", "70-79", "80-89", "90+"),
    create_calendar_vars = TRUE,                 # Month, week, day of week
    create_climate_vars = TRUE,                  # Derived climate variables
    climate_region = "Norte",                    # Northern region of Brazil
    lang = "pt"
  ) |>

  # Demographic filters
  sus_data_filter_demographics(
    sex = c("Masculino", "Feminino"),            # Both sexes
    age_range = c(65, Inf)                       # Elderly only (65+)
  ) |>

  # Temporal aggregation (daily)
  sus_data_aggregate(
    date_col = NULL,              # Uses death date automatically
    complete_dates = FALSE,       # Do not complete missing dates
    time_unit = "day",            # Daily aggregation
    lang = "pt"
  )

# View the object's metadata
sus_meta(df_sim)

# Statistical summary
summary(df_sim)
```

::: callout-info
**Interpretation of Results:**
- The `df_sim` object contains the daily count of deaths from respiratory diseases
- Target population: elderly (65+ years) in Amazonas
- Period: January 1 to December 31, 2023
:::

# Part 2: Spatial Aggregation

## 2.1 Spatial Join with the Municipal Mesh

The `sus_spatial_join()` function assigns each death the geographic coordinates of the municipality of occurrence.

```{r step2_spatial_join, eval=FALSE}
# ============================================================================
# SPATIAL AGGREGATION AT THE MUNICIPAL LEVEL
# ============================================================================

sf_sim_spatial <- sus_spatial_join(
  df = df_sim,
  level = "munic",              # Municipal level (vs. "state" or "health_region")
  lang = "pt"
)

# Check metadata (now with a spatial component)
sus_meta(sf_sim_spatial)

```

::: callout-note
**Note:** The object is now an `sf` (simple features) object with municipal geometry.
:::

---

# Part 3: Climate Data (INMET)

## 3.1 Download, Preprocessing, and Gap Filling

We download meteorological data from INMET stations in Amazonas.

```{r step3_climate_data, eval=FALSE}
# ============================================================================
# DOWNLOAD OF INMET DATA
# ============================================================================

df_inmet_raw <- sus_climate_inmet(
  years = 2023,                 # Same year as the health data
  uf = "AM"                     # Only stations in Amazonas
)

# Check downloaded stations
unique(df_inmet_raw$station_name)

# ============================================================================
# IMPUTATION OF MISSING DATA
# ============================================================================

df_inmet <- sus_climate_fill_inmet(
  df_inmet_raw,
  target_var = "all",           # Impute all climate variables
  parallel = TRUE,              # Parallel processing (faster)
  run_evaluation = FALSE        # Skip detailed evaluation
)

# To speed up the examples, we can save/read the processed data in the ultra-fast parquet format
# sus_data_export(df_inmet, "df_inmet_AM_2023_filled.parquet")
# sus_data_export(sf_sim_spatial, "df_sim_AM_2023_spatial.parquet")

# Load pre-processed data (if available)
# df_inmet <- sus_data_read("df_inmet_AM_2023_filled.parquet")
# sf_sim_spatial <- sus_data_read("df_sim_AM_2023_spatial.parquet")
```

---

# Part 4: Climate Aggregation Strategies with sus_climate_aggregate()

## 4.1 Exact Strategy (Exact Match)

**Objective:** Associate the temperature of the same day as the death.
**Application:** Acute events such as hyperthermia, stroke, heart attack.

```{r step4_exact, eval=FALSE}
# ============================================================================
# EXACT: Same-day temperature
# ============================================================================

df_exact <- sus_climate_aggregate(
  health_data = sf_sim_spatial,
  climate_data = df_inmet,
  climate_var = "tair_dry_bulb_c",      # Dry-bulb temperature (°C)
  temporal_strategy = "exact",
  lang = "pt",
  verbose = TRUE
)

# Check new column
names(df_exact)
summary(df_exact$tair_dry_bulb_c)

# Quick analysis: relationship between temperature and deaths
library(ggplot2)
df_exact %>%
    ggplot(aes(x = tair_dry_bulb_c, y = n_obitos)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "loess", color = "red") +
    labs(
      title = "Relação entre Temperatura e Óbitos (Mesmo Dia)",
      x = "Temperatura (°C)",
      y = "Número de Óbitos"
    ) +
    theme_minimal()

```

::: callout-tip
INMET Supported Variables
**Atmospheric Pressure**: patm_mb (Mean), patm_max_mb (Maximum), patm_min_mb (Minimum)

**Air Temperature**: tair_dry_bulb_c (Dry Bulb), tair_max_c (Maximum), tair_min_c (Minimum)

**Dew Point**: dew_tmean_c (Mean), dew_tmax_c (Maximum), dew_tmin_c (Minimum)

**Relative Humidity**: rh_mean_porc (Mean), rh_max_porc (Maximum), rh_min_porc (Minimum)

**Wind**: ws_2_m_s (Speed), ws_gust_m_s (Gust), wd_degrees (Direction)

**Miscellaneous**: rainfall_mm (Rainfall), sr_kj_m2 (Solar Radiation)
:::

## 4.2 Discrete Lag Strategy

**Objective:** Assess the effect of temperature at different lags, identifying the critical exposure period.

```{r step4_discrete_lag, eval=FALSE}
# ============================================================================
# DISCRETE LAG: Temperature 7, 14, and 21 days before
# ============================================================================

df_discrete_lag <- sus_climate_aggregate(
  health_data = sf_sim_spatial,
  climate_data = df_inmet,
  climate_var = "tair_dry_bulb_c",
  temporal_strategy = "discrete_lag",
  lag_days = c(7, 14, 21),          # Three distinct lags
  lang = "pt",
  verbose = TRUE
)

# Check created columns
names(df_discrete_lag)

# Compare correlations at different lags
correlations <- data.frame(
  lag = c(0, 7, 14, 21),
  correlation = c(
    cor(df_exact$tair_dry_bulb_c, df_exact$n_obitos, use = "complete.obs"),
    cor(df_discrete_lag$lag7_tair_dry_bulb_c, df_discrete_lag$n_obitos, use = "complete.obs"),
    cor(df_discrete_lag$lag14_tair_dry_bulb_c, df_discrete_lag$n_obitos, use = "complete.obs"),
    cor(df_discrete_lag$lag21_tair_dry_bulb_c, df_discrete_lag$n_obitos, use = "complete.obs")
  )
)

print(correlations)
```

::: callout-info
**Interpretation:** The correlation values between lag and temperature are purely illustrative and reflect the period used in this tutorial.
:::

---

## 4.3 Distributed Lag Strategy (Matrix for DLNM)

**Objective:** Build a lag matrix for DLNM models.
**Application:** Non-linear modeling of lagged effects.

```{r step4_distributed_lag, eval=FALSE}
# ============================================================================
# DISTRIBUTED LAG: Lag matrix from 0 to 21
# ============================================================================

df_distributed <- sus_climate_aggregate(
  health_data = sf_sim_spatial,
  climate_data = df_inmet,
  climate_var = "tair_dry_bulb_c",
  temporal_strategy = "distributed_lag",
  lag_days = 21,                    # Maximum lag (0 to 21)
  lang = "pt",
  verbose = TRUE
)

# Check matrix structure
head(df_distributed)

```

::: callout-info
**DLNM:** The generated matrix will serve as input for the DLNM model (the `dlnm` package). `climasus4r` will soon provide new modeling and results-visualization functionality.
:::
---

## 4.4 Moving Window Strategy

**Objective:** Average temperature over the last 14 days.
**Application:** Cumulative effect of prolonged exposure.

```{r step4_moving_window, eval=FALSE}
# ============================================================================
# MOVING WINDOW: Average of the last 14 days
# ============================================================================

df_moving <- sus_climate_aggregate(
  health_data = sf_sim_spatial,
  climate_data = df_inmet,
  climate_var = "tair_dry_bulb_c",
  temporal_strategy = "moving_window",
  window_days = 14,                 # 14-day window
  min_obs = 0.7,                    # Minimum 70% valid data
  lang = "pt",
  verbose = TRUE
)

# Compare daily temperature vs. moving average
df_compare <- df_exact %>%
    dplyr::select(date, tair_dry_bulb_c) %>%
    dplyr::left_join(
      df_moving %>% dplyr::select(date, tair_dry_bulb_c_mean_w14),
      by = "date", relationship = "many-to-many"
    ) %>%
    tidyr::pivot_longer(-date, names_to = "metric", values_to = "temperature")
  ggplot(df_compare, aes(x = date, y = temperature, color = metric)) +
    geom_line(alpha = 0.7) +
    labs(
      title = "Temperatura Diária vs. Média Móvel (14 dias)",
      x = "Data",
      y = "Temperatura (°C)",
      color = "Métrica"
    ) +
    theme_minimal()
```

---

## 4.5 Offset Window Strategy

**Objective:** Isolate the incubation period (excluding recent days).
**Application:** Dengue, Zika, and other diseases with an incubation period.

```{r step4_offset_window, eval=FALSE}
# ============================================================================
# OFFSET WINDOW: Average between 14 and 7 days before (ignores the last week)
# ============================================================================

df_offset <- sus_climate_aggregate(
  health_data = sf_sim_spatial,
  climate_data = df_inmet,
  climate_var = "tair_dry_bulb_c",
  temporal_strategy = "offset_window",
  offset_days = c(7, 14),           # Window [t-14, t-7]
  min_obs = 0.7,
  lang = "pt",
  verbose = TRUE
)

head(df_offset)
```

::: callout-note
**Interpretation:** This variable represents exposure over the period that excludes the week immediately preceding death, useful for isolating long-term effects from acute exposures.
:::

---

## 4.6 Degree Days Strategy

**Objective:** Accumulate heat above a base temperature.
**Application:** Vector development (Aedes aegypti), heat stress.


```{r step4_degree_days, eval=FALSE}
# ============================================================================
# DEGREE DAYS: Accumulated GDD over the last 21 days (base 20°C)
# ============================================================================

df_gdd <- sus_climate_aggregate(
  health_data = sf_sim_spatial,
  climate_data = df_inmet,
  temporal_strategy = "degree_days",
  window_days = 21,                 # Accumulation period
  temp_base = 20,                   # Base temperature (human)
  gdd_temp_var = "tair_dry_bulb_c",
  min_obs = 0.7,
  lang = "pt",
  verbose = TRUE
)

head(df_gdd)

# Example with base temperature for Aedes aegypti (11°C)
df_gdd_aedes <- sus_climate_aggregate(
  health_data = sf_sim_spatial,
  climate_data = df_inmet,
  temporal_strategy = "degree_days",
  window_days = 21,
  temp_base = 11,                   # Base for Aedes aegypti
  gdd_temp_var = "tair_dry_bulb_c",
  lang = "pt",
  verbose = TRUE
)

head(df_gdd_aedes)
```

### Thermal Thresholds by Species

| Organism | Tbase (temp_base) (°C) | GDD for Development | Application |
|-----------|------------|--------------------------|-----------|
| *Aedes aegypti* | 11 | 100-150 | Dengue transmission |
| *Anopheles gambiae* | 18 | 80-120 | Malaria transmission |
| *Triatoma infestans* | 16 | 200-300 | Chagas disease |
| *Human (stress)* | 20 | - | Cardiovascular morbidity |

---

## 4.7 Seasonal Strategy

**Objective:** Analyze seasonal patterns.
**Application:** Ecological studies, long-term trends.

```{r step4_seasonal, eval=FALSE}
# ============================================================================
# SEASONAL: Averages by season of the year
# ============================================================================

df_seasonal <- sus_climate_aggregate(
  health_data = sf_sim_spatial,
  climate_data = df_inmet,
  temporal_strategy = "seasonal",
  lang = "pt",
  verbose = TRUE
)

head(df_seasonal)
```

---

## 4.8 Threshold Exceedance Strategy

**Objective:** Count days with extreme temperatures.
**Application:** Heat waves, extreme precipitation events.

```{r step4_threshold, eval=FALSE}
# ============================================================================
# THRESHOLD EXCEEDANCE: Days with Tmax > 35°C over the last 7 days
# ============================================================================

df_threshold <- sus_climate_aggregate(
  health_data = sf_sim_spatial,
  climate_data = df_inmet,
  climate_var = "tair_max_c",
  temporal_strategy = "threshold_exceedance",
  window_days = 7,
  threshold_value = 35,
  threshold_direction = "above",    # Count days ABOVE the threshold
  min_obs = 0.7,
  lang = "pt",
  verbose = TRUE
)

head(df_threshold)

```

---

## 4.9 Weighted Window Strategy

**Objective:** Apply decreasing weights to the days in the window.
**Application:** Modeling biological decay of exposure.

```{r step4_weighted, eval=FALSE}
# ============================================================================
# WEIGHTED WINDOW: Weighted average with different weighting functions
# ============================================================================

# Case 1: Linear weights (automatic)
df_wlin <- sus_climate_aggregate(
  health_data = sf_sim_spatial,
  climate_data = df_inmet,
  climate_var = "tair_dry_bulb_c",
  temporal_strategy = "weighted_window",
  window_days = 7,
  weights = NULL,                   # Automatic linear decay
  min_obs = 0.7,
  lang = "pt",
  verbose = TRUE
)

head(df_wlin)

# Case 2: Exponential weights (fast decay)
lambda <- 0.3
W <- 7
pesos_exp <- exp(-lambda * seq(0, W))
pesos_exp <- pesos_exp / pesos_exp[1]  # Normalize (peak = 1)

df_wexp <- sus_climate_aggregate(
  health_data = sf_sim_spatial,
  climate_data = df_inmet,
  climate_var = "tair_dry_bulb_c",
  temporal_strategy = "weighted_window",
  window_days = W,
  weights = pesos_exp,
  min_obs = 0.7,
  lang = "pt",
  verbose = TRUE
)

head(df_wexp)

# Case 3: Gaussian weights (peak in the middle of the window)
mu <- 3          # Peak at day t-3
sigma <- 1.5     # Window width
pesos_gaus <- exp(-((seq(0, W) - mu)^2) / (2 * sigma^2))
pesos_gaus <- pesos_gaus / max(pesos_gaus)

df_wgaus <- sus_climate_aggregate(
  health_data = sf_sim_spatial,
  climate_data = df_inmet,
  climate_var = "tair_dry_bulb_c",
  temporal_strategy = "weighted_window",
  window_days = W,
  weights = pesos_gaus,
  min_obs = 0.7,
  lang = "pt",
  verbose = TRUE
)
head(df_wgaus)

```

---

## 🎉 Conclusion

The `sus_climate_aggregate()` function elevates `climasus4r` from a simple data-cleaning package into an **epidemiological inference tool**.

By aligning naming conventions with the R community standard (`moving_window`), providing a unique mathematical tool for incubation periods (`offset_window`), and introducing advanced strategies such as `threshold_exceedance` and `weighted_window`, this package ensures that your conclusions about "How climate affects health in Brazil" are methodologically sound and ready for publication in high-impact journals.

---

## 🌲 Have feedback or suggestions?

Do you have an idea for improvement or found a bug? Click the button below to open a new *issue* on GitHub and share your suggestions directly with us.

```{=html}
<a href='https://github.com/ByMaxAnjos/climasus4r/issues/new' class="github-feedback-btn">
  <svg xmlns="http://www.w3.org/2000/svg" width="18" height="18" viewBox="0 0 24 24" fill="currentColor">
    <path d="M12 0C5.37 0 0 5.37 0 12c0 5.31 3.435 9.795 8.205 11.385.6.105.825-.255.825-.57 0-.285-.015-1.23-.015-2.235-3.015.555-3.795-.735-4.035-1.41-.135-.345-.72-1.41-1.23-1.695-.42-.225-1.02-.78-.015-.795.945-.015 1.62.87 1.845 1.23 1.08 1.815 2.805 1.305 3.495.99.105-.78.42-1.305.765-1.605-2.67-.3-5.46-1.335-5.46-5.925 0-1.305.465-2.385 1.23-3.225-.12-.3-.54-1.53.12-3.18 0 0 1.005-.315 3.3 1.23.96-.27 1.98-.405 3-.405s2.04.135 3 .405c2.295-1.56 3.3-1.23 3.3-1.23.66 1.65.24 2.88.12 3.18.765.84 1.23 1.905 1.23 3.225 0 4.605-2.805 5.625-5.475 5.925.435.375.81 1.095.81 2.22 0 1.605-.015 2.895-.015 3.3 0 .315.225.69.825.57A12.02 12.02 0 0 0 24 12c0-6.63-5.37-12-12-12z"/>
  </svg>
  Open issue on GitHub
</a>
```
