---
title: "Pipelines with Smart Caching using targets"
subtitle: "Generate `_targets.R`, run with dynamic parallelism, and monitor with `sus_rap_targets()` and `sus_rap_make()`"
author:
  - name: "Dr. Max Anjos"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    code_download: true
    self_contained: true
    df_print: paged
---

```{css, echo=FALSE}
:root {
  --primary-color: #1D9E75;
  --dark-green: #0f3d2e;
  --bg-light: #f8faf9;
  --text-main: #1a1a1a;
  --code-bg: #fdfdfd;
  --warning-orange: #ff9800;
  --info-blue: #2196f3;
}
html { scroll-behavior: smooth; }
body { font-family: 'Source Serif 4', Georgia, serif; font-size: 18px; color: var(--text-main); background-color: #fff; line-height: 1.6; }
h1, h2, h3, h4 { font-family: 'Inter', sans-serif; font-weight: 700; color: var(--dark-green); }
h1.title { font-size: 2.8em; border-bottom: 4px solid var(--primary-color); padding-bottom: 10px; margin-bottom: 1.5rem; }
h2 { margin-top: 2rem; border-left: 4px solid var(--primary-color); padding-left: 1rem; }
pre.r, pre.sourceCode { background-color: var(--code-bg); border: 1px solid #e1e1e1; border-left: 5px solid var(--primary-color) !important; border-radius: 8px; box-shadow: 0 2px 4px rgba(0,0,0,0.05); padding: 1rem; }
.callout { padding: 1.2em 1.5em; margin: 1.5em 0; border-radius: 8px; border-left: 6px solid; background-color: var(--bg-light); }
.callout-note { background-color: #f0f7ff; border-color: #007bff; }
.callout-tip { background-color: #e8f5e9; border-color: #1D9E75; }
.callout-warning { background-color: #fff3e0; border-color: #ff9800; }
.workflow-container { background: linear-gradient(135deg, #f8faf7 0%, #ffffff 100%); border-radius: 12px; padding: 1.5rem; margin: 1.5rem 0; border: 1px solid #dce8e1; overflow-x: auto; }
.workflow-steps { display: flex; flex-wrap: wrap; justify-content: center; gap: 0.5rem; font-family: 'Inter', monospace; font-size: 0.8rem; }
.workflow-step { background: var(--primary-color); color: white; padding: 0.5rem 1rem; border-radius: 20px; white-space: nowrap; }
.workflow-arrow { color: var(--primary-color); font-weight: bold; font-size: 1.2rem; }
.feature-grid { display: grid; grid-template-columns: repeat(auto-fit, minmax(280px, 1fr)); gap: 1.5rem; margin: 2rem 0; }
.feature-card { background: white; border-radius: 12px; padding: 1.5rem; border: 1px solid #e1e1e1; transition: transform 0.2s ease, box-shadow 0.2s ease; text-align: center; }
.feature-card:hover { transform: translateY(-4px); box-shadow: 0 8px 24px rgba(0,0,0,0.1); }
.feature-icon { font-size: 2rem; margin-bottom: 1rem; }
.feature-card h3 { margin-top: 0; color: var(--primary-color); font-size: 1.1rem; }
table { width: 100%; border-collapse: collapse; margin: 1.5rem 0; }
th, td { padding: 12px; text-align: left; border-bottom: 1px solid #e1e1e1; }
th { background-color: var(--bg-light); font-weight: 600; color: var(--dark-green); }
.github-feedback-btn { display: inline-flex; align-items: center; gap: 8px; background-color: #24292e; color: white; padding: 10px 20px; border-radius: 6px; text-decoration: none; font-family: sans-serif; transition: background-color 0.2s; margin-top: 1rem; }
.github-feedback-btn:hover { background-color: #2c3e50; color: white; text-decoration: none; }
@media (max-width: 768px) { body { font-size: 16px; } h1.title { font-size: 2rem; } .workflow-steps { flex-direction: column; align-items: center; } }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  eval      = FALSE,
  warning   = FALSE,
  message   = FALSE,
  comment   = "#>",
  fig.align = "center",
  out.width = "90%",
  fig.retina = 2
)
```

# Introduction — Why use `targets`?

Epidemiological analyses with climate data are rarely linear. Over the course of a research cycle, it's common to change a CID-10 filter, swap the climate aggregation window, or add a new state to the dataset. Without a smart caching system, any of these changes forces a re-run of the **entire** pipeline — including costly steps such as the DATASUS download, encoding standardization, and the spatial join, which can take hours.

The `targets` package (Landau, 2021) solves this problem with **automatic dependency tracking**: each pipeline step is a *target*, and `targets` only re-executes the targets whose inputs changed since the last run. Intermediate results are cached in the `_targets/` directory, ready to be loaded instantly.

The `sus_rap_targets()` function bridges `climasus4r`'s RAP objects with the `targets` ecosystem, automatically generating the `_targets.R` file with all the necessary structure. `sus_rap_make()` then runs this pipeline reproducibly, with support for parallelism and monitoring.

```{=html}
<div class="workflow-container">
  <div class="workflow-steps">
    <span class="workflow-step">rap_object</span>
    <span class="workflow-arrow">→</span>
    <span class="workflow-step">sus_rap_targets()</span>
    <span class="workflow-arrow">→</span>
    <span class="workflow-step">_targets.R</span>
    <span class="workflow-arrow">→</span>
    <span class="workflow-step">sus_rap_make()</span>
    <span class="workflow-arrow">→</span>
    <span class="workflow-step">cached results</span>
  </div>
</div>
```

## Functions covered in this tutorial

This tutorial covers two exported functions from `climasus4r`:

| Function | Description |
|---|---|
| `sus_rap_targets()` | Generates the `_targets.R` file from a RAP object |
| `sus_rap_make()` | Runs the pipeline via `targets::tar_make()` |

# Prerequisites

## Installing dependencies

The `targets` package is listed under `Suggests` in `climasus4r`. Install it before using the functions in this tutorial:

```{r install-deps}
# Necessary dependencies
install.packages("targets")
install.packages("tarchetypes")  # optional: additional helpers
```

```{=html}
<div class="callout callout-warning">
**targets package required:** `targets` is listed under `Suggests` in climasus4r. Install it before using `sus_rap_targets()` and `sus_rap_make()`. `climasus4r` will automatically check for the package's availability when calling these functions.
</div>
```

## Starting point: the RAP object

The functions in this tutorial operate on a **RAP object** — a pipeline expression created with `climasus4r`'s functions. You can obtain it in two ways:

```{r prereq-rap}
library(climasus4r)

# Option A: read an existing RAP (created with sus_rap_export())
rap <- sus_rap_read("analise_dengue_nordeste.R")

# Option B: use a pipeline expression directly
pipeline <- quote(
  sus_data_import(uf = c("CE", "RN", "PB", "PE", "AL"),
                  years = 2018:2022, system = "SINAN") |>
    sus_data_clean_encoding() |>
    sus_data_standardize() |>
    sus_data_filter_cid(cid_group = "dengue") |>
    sus_data_aggregate(by = "month")
)
```

```{=html}
<div class="callout callout-note">
**About the RAP object:** RAP Tutorial 1 — *Export and Run* — explains in detail how to create, export, and load RAP objects with `sus_rap_export()` and `sus_rap_read()`. Refer to it if you need to build the object from scratch.
</div>
```

# `sus_rap_targets()` — Generating the `_targets.R`

The `sus_rap_targets()` function takes a `climasus4r` pipeline and generates a `_targets.R` file ready to be run with the `targets` package.

## Function signature

```{r sig-targets}
sus_rap_targets(
  pipeline,
  file_path  = "_targets.R",
  branching  = c("none", "uf", "year", "cross"),
  workers    = 1L,
  include_qc = TRUE,
  lang       = "pt",
  overwrite  = FALSE
)
```

## Basic usage

```{r basic-targets}
sus_rap_targets(
  pipeline   = pipeline,
  file_path  = "_targets.R",
  branching  = "none",
  workers    = 1L,
  include_qc = TRUE,
  lang       = "pt"
)
#> i Gerando _targets.R para 'Pipeline SINAN' (branching: none)...
#> v Pipeline targets criado: /path/to/_targets.R
```

The generated `_targets.R` file contains all the structure needed for `targets` to track dependencies and manage the cache. Here is a simplified excerpt:

```{r generated-targets-R}
# _targets.R -- Generated by climasus4r::sus_rap_targets()
# Pipeline: Pipeline SINAN
# Created : 2026-06-18 10:30:00

library(targets)
library(tarchetypes)
library(climasus4r)
library(dplyr)

options(clustermq.scheduler = "sequential")

list(
  tar_target(params_uf,        c("CE", "RN", "PB", "PE", "AL")),
  tar_target(params_years,     2018:2022),
  tar_target(params_system,    "SINAN"),
  tar_target(params_time_unit, "month"),
  tar_target(params_lang,      "pt"),

  tar_target(pipeline_sinan,
    climasus4r::sus_data_import(
      uf     = params_uf,
      year   = params_years,
      system = params_system,
      lang   = params_lang
    )
  ),

  tar_target(pipeline_sinan_qc,
    climasus4r::sus_data_quality_report(pipeline_sinan)
  ),

  NULL
)
```

```{=html}
<div class="callout callout-tip">
**include_qc = TRUE:** When enabled (the default), `sus_rap_targets()` automatically adds a quality-control target (`_qc`) after each main step. This records data integrity metrics in the cache at no additional execution cost.
</div>
```

## The `overwrite` parameter

By default, `overwrite = FALSE` protects existing files. If a `_targets.R` already exists at the given path, the function issues a warning and stops. Use `overwrite = TRUE` to replace it:

```{r overwrite}
# Replace an existing _targets.R
sus_rap_targets(
  pipeline  = pipeline,
  file_path = "_targets.R",
  overwrite = TRUE
)
#> ! Arquivo _targets.R existente sera substituido (overwrite = TRUE).
#> v Pipeline targets criado: /path/to/_targets.R
```

# The 4 Branching Strategies

The `branching` parameter defines how the pipeline is split into parallel *branches*. Each strategy has an ideal use case:

```{=html}
<div class="feature-grid">
  <div class="feature-card">
    <div class="feature-icon">1️⃣</div>
    <h3>"none"</h3>
    <p>Single analysis. All states and years processed together. Ideal for national analyses or rapid prototyping.</p>
  </div>
  <div class="feature-card">
    <div class="feature-icon">🗺️</div>
    <h3>"uf"</h3>
    <p>One branch per state (UF). Ideal for inter-state comparisons with independent caching per federative unit.</p>
  </div>
  <div class="feature-card">
    <div class="feature-icon">📅</div>
    <h3>"year"</h3>
    <p>One branch per year. Useful for long time series where past years rarely change after collection.</p>
  </div>
  <div class="feature-card">
    <div class="feature-icon">⚡</div>
    <h3>"cross"</h3>
    <p>Cartesian product of state x year. Maximum parallelism. Requires more workers and disk space.</p>
  </div>
</div>
```

## Branching by state (`"uf"`)

```{r branching-uf}
sus_rap_targets(
  pipeline  = pipeline,
  file_path = "_targets_uf.R",
  branching = "uf",
  workers   = 4L
)
#> i Gerando _targets.R para 'Pipeline SINAN' (branching: uf)...
#> i Estrategia: 1 branch por UF (5 UFs detectados)
#> v Pipeline targets criado: /path/to/_targets_uf.R
```

With `branching = "uf"`, the generated `_targets.R` uses `targets::tar_map()` to create an independent branch per state. The relevant excerpt of the generated file looks like this:

```{r branching-uf-generated}
# Excerpt of the generated _targets_uf.R
list(
  tar_target(params_uf,    c("CE", "RN", "PB", "PE", "AL")),
  tar_target(params_years, 2018:2022),

  tar_map(
    values = list(uf = c("CE", "RN", "PB", "PE", "AL")),
    tar_target(pipeline_sinan,
      climasus4r::sus_data_import(
        uf     = uf,
        year   = params_years,
        system = "SINAN"
      ),
      pattern = map(params_uf)
    )
  ),

  NULL
)
```

Each state has its own cache: if the data for `"CE"` changes, only the `"CE"` branch is re-executed. The others remain cached.

## Cross branching (`"cross"`)

```{r branching-cross}
sus_rap_targets(
  pipeline  = pipeline,
  file_path = "_targets_cross.R",
  branching = "cross",
  workers   = 8L
)
#> i Gerando _targets.R para 'Pipeline SINAN' (branching: cross)...
#> i Estrategia: produto cartesiano UF x ano (5 UFs x 5 anos = 25 branches)
#> ! Aviso: 25 branches com workers = 8 pode exigir bastante memoria RAM.
#> v Pipeline targets criado: /path/to/_targets_cross.R
```

```{=html}
<div class="callout callout-warning">
**Resource usage with "cross":** The cartesian product of state x year can generate many simultaneous branches. For 10 states and 10 years, that's 100 branches. Monitor RAM usage and adjust `workers` according to the resources available on the machine.
</div>
```

## Comparing the strategies

| Strategy | Branches | When to use | Recommended workers |
|---|---|---|---|
| `"none"` | 1 | Prototyping, single analyses | 1 |
| `"uf"` | 1 per state | State-level comparisons | n_states |
| `"year"` | 1 per year | Long series, historical data | n_years |
| `"cross"` | states × years | Complete space-time panels | n_states × n_years |

# `sus_rap_make()` — Running the Pipeline

The `sus_rap_make()` function runs the pipeline generated with `sus_rap_targets()` by calling `targets::tar_make()` internally, with integration into `climasus4r`'s metadata system.

## Function signature

```{r sig-make}
sus_rap_make(
  rap,
  workers   = 1L,
  reporter  = "verbose",
  branching = "none",
  lang      = "pt",
  ...
)
```

## Basic execution

```{r basic-make}
resultado <- sus_rap_make(
  rap      = rap,
  workers  = 1L,
  reporter = "verbose",
  lang     = "pt"
)
#> i Executando via targets::tar_make()...
#> * Target params_uf
#> * Target params_years
#> * Target params_system
#> * Target params_time_unit
#> * Target pipeline_sinan [10.3s]
#> * Target pipeline_sinan_qc [0.8s]
#> v Execucao targets concluida.
```

## Accessing the results

The object returned by `sus_rap_make()` is a list with three components:

```{r access-results}
# Access results
head(resultado$results)  # final pipeline data.frame
resultado$store          # path to the cache directory (_targets/)
resultado$rap            # original rap_object with updated metadata

# You can also use targets' native API
targets::tar_load(pipeline_sinan)
targets::tar_read("pipeline_sinan_qc")
targets::tar_visnetwork()  # visual dependency graph
```

```{=html}
<div class="callout callout-tip">
**Tip:** To see the pipeline's dependency graph, use `targets::tar_visnetwork()` after generating the `_targets.R`. This helps identify bottlenecks and optimize the order in which targets run.
</div>
```

## The power of caching: second run

On the second run of the same pipeline, `targets` checks the hash of each target. If nothing changed in the inputs, the target is marked `[skipped]` and loaded directly from the cache — without re-running the code:

```{r second-run}
# Second run: only what changed is re-executed
resultado2 <- sus_rap_make(rap, workers = 1L)
#> i Executando via targets::tar_make()...
#> * Target params_uf         [skipped]
#> * Target params_years      [skipped]
#> * Target params_system     [skipped]
#> * Target params_time_unit  [skipped]
#> * Target pipeline_sinan    [skipped]  # cache valido!
#> * Target pipeline_sinan_qc [skipped]  # cache valido!
#> v Execucao targets concluida.
```

This means that, when only the CID-10 filter is changed, only the steps starting from `sus_data_filter_cid()` are re-executed. The import and encoding steps remain cached.

## Running with parallelism

For pipelines with `branching = "uf"` or `"cross"`, parallelism is enabled with `workers > 1`. `climasus4r` uses the `future`/`furrr` framework internally:

```{r parallel-make}
# Run with 4 parallel workers
resultado_paralelo <- sus_rap_make(
  rap      = rap,
  workers  = 4L,
  reporter = "verbose",
  branching = "uf",
  lang     = "pt"
)
#> i Executando via targets::tar_make() com 4 workers (future::multisession)...
#> * Target pipeline_sinan_CE [3.1s]
#> * Target pipeline_sinan_RN [2.8s]  # executando em paralelo com CE
#> * Target pipeline_sinan_PB [3.4s]
#> * Target pipeline_sinan_PE [3.0s]  # executando em paralelo com PB
#> * Target pipeline_sinan_AL [2.5s]
#> v Execucao targets concluida em 7.2s (vs ~14s sequencial).
```

```{=html}
<div class="callout callout-note">
**Controlling the parallelism plan:** `climasus4r` never sets the `future` plan globally. If you have already set `future::plan(multisession, workers = 8)` before calling `sus_rap_make()`, that plan will be respected. The `workers` parameter in `sus_rap_make()` is used only when no custom plan is active.
</div>
```

# Workflow with `targets` + `renv`

For maximum reproducibility, combine `targets`'s smart caching with `renv`'s dependency management. This combination ensures that both the **package environment** and the **intermediate results** are reproducible:

```{r renv-workflow}
# 1. Initialize renv and create a snapshot of the dependencies
renv::init()
renv::snapshot()

# 2. Generate _targets.R with the appropriate branching strategy
sus_rap_targets(
  pipeline  = pipeline,
  file_path = "_targets.R",
  branching = "uf",
  workers   = 4L
)

# 3. Run the pipeline (uses the _targets.R generated in the previous step)
sus_rap_make(rap, workers = 4L)

# 4. Check the pipeline status at any time
targets::tar_progress()   # status of each target
targets::tar_outdated()   # outdated targets
targets::tar_visnetwork() # interactive dependency graph
```

```{=html}
<div class="callout callout-note">
**For complete projects:** Use `sus_rap_template()` (RAP Tutorial 3) which automatically creates the project structure with `renv`, `_targets.R`, Quarto, and GitHub Actions in a single call. Ideal for starting a project from scratch with all best practices already configured.
</div>
```

## Recommended project structure

A well-organized `climasus4r` project using `targets` and `renv` follows this directory structure:

```{r project-structure}
# Recommended structure for a climasus4r + targets project
# meu_projeto_dengue/
# ├── _targets.R              # generated by sus_rap_targets()
# ├── _targets/               # targets cache (do not version in git)
# ├── renv/                   # renv environment
# ├── renv.lock               # dependency snapshot (version this)
# ├── R/
# │   └── funcoes_aux.R       # project helper functions
# ├── data/
# │   └── dados_brutos/       # raw data (do not version)
# ├── output/
# │   ├── figuras/
# │   └── tabelas/
# └── relatorio.Rmd           # final report that loads from the cache
```

```{=html}
<div class="callout callout-tip">
**Git and the cache:** Add `_targets/` to your `.gitignore`. The cache can be gigabytes in size and should not be versioned. What should go into git is the `_targets.R` (generated by `sus_rap_targets()`) and the `renv.lock` — these two files allow any collaborator to reproduce the results from scratch.
</div>
```

# References

- Landau WM (2021). The targets R package: a dynamic Make-like function-oriented pipeline toolkit for reproducibility and high-performance computing. *Journal of Open Source Software*, 6(57), 2959. <https://doi.org/10.21105/joss.02959>

- Marwick B, Boettiger C, Mullen L (2018). Packaging data analytical work reproducibly using R (and friends). *The American Statistician*, 72(1), 80–88. <https://doi.org/10.1080/00031305.2017.1375986>

---

**Navigation:**
[Previous: RAP 1 - Export and Run](rap-01-exportar-executar.html) | [Next: RAP 3 - Share and Scaffolding](rap-03-compartilhar.html)

```{=html}
<div style="text-align: center; margin-top: 3rem; padding-top: 2rem; border-top: 2px solid #e1e1e1;">

<p><strong>climasus4r</strong> · Integrated health-climate-environment analysis in Brazil</p>

<p>Found an issue or have suggestions for this tutorial?</p>

<a href="https://github.com/ByMaxAnjos/climasus4r/issues" class="github-feedback-btn" target="_blank">
  &#x1F41B; Report on GitHub
</a>

</div>
```
