---
title: "Exporting and Running Reproducible Pipelines"
subtitle: "Serialize, read, inspect, and re-run analyses with `sus_rap_export()`, `sus_rap_read()`, `sus_rap_run()`, and `sus_rap_update()`"
author:
  - name: "Dr. Max Anjos"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    code_download: true
    self_contained: true
    df_print: paged
---

```{css, echo=FALSE}
/* -- Variaveis de Cores Modernas -- */
:root {
  --primary-color: #1D9E75;
  --dark-green: #0f3d2e;
  --bg-light: #f8faf9;
  --text-main: #1a1a1a;
  --code-bg: #fdfdfd;
  --warning-orange: #ff9800;
  --info-blue: #2196f3;
}
html { scroll-behavior: smooth; }
body { font-family: 'Source Serif 4', Georgia, serif; font-size: 18px; color: var(--text-main); background-color: #fff; line-height: 1.6; }
h1, h2, h3, h4 { font-family: 'Inter', sans-serif; font-weight: 700; color: var(--dark-green); }
h1.title { font-size: 2.8em; border-bottom: 4px solid var(--primary-color); padding-bottom: 10px; margin-bottom: 1.5rem; }
h2 { margin-top: 2rem; border-left: 4px solid var(--primary-color); padding-left: 1rem; }
pre.r, pre.sourceCode { background-color: var(--code-bg); border: 1px solid #e1e1e1; border-left: 5px solid var(--primary-color) !important; border-radius: 8px; box-shadow: 0 2px 4px rgba(0,0,0,0.05); padding: 1rem; }
.callout { padding: 1.2em 1.5em; margin: 1.5em 0; border-radius: 8px; border-left: 6px solid; background-color: var(--bg-light); }
.callout-note { background-color: #f0f7ff; border-color: #007bff; }
.callout-tip { background-color: #e8f5e9; border-color: #1D9E75; }
.callout-warning { background-color: #fff3e0; border-color: #ff9800; }
.workflow-container { background: linear-gradient(135deg, #f8faf7 0%, #ffffff 100%); border-radius: 12px; padding: 1.5rem; margin: 1.5rem 0; border: 1px solid #dce8e1; overflow-x: auto; }
.workflow-steps { display: flex; flex-wrap: wrap; justify-content: center; gap: 0.5rem; font-family: 'Inter', monospace; font-size: 0.8rem; }
.workflow-step { background: var(--primary-color); color: white; padding: 0.5rem 1rem; border-radius: 20px; white-space: nowrap; }
.workflow-arrow { color: var(--primary-color); font-weight: bold; font-size: 1.2rem; }
.feature-grid { display: grid; grid-template-columns: repeat(auto-fit, minmax(280px, 1fr)); gap: 1.5rem; margin: 2rem 0; }
.feature-card { background: white; border-radius: 12px; padding: 1.5rem; border: 1px solid #e1e1e1; transition: transform 0.2s ease, box-shadow 0.2s ease; text-align: center; }
.feature-card:hover { transform: translateY(-4px); box-shadow: 0 8px 24px rgba(0,0,0,0.1); }
.feature-icon { font-size: 2rem; margin-bottom: 1rem; }
.feature-card h3 { margin-top: 0; color: var(--primary-color); font-size: 1.1rem; }
table { width: 100%; border-collapse: collapse; margin: 1.5rem 0; }
th, td { padding: 12px; text-align: left; border-bottom: 1px solid #e1e1e1; }
th { background-color: var(--bg-light); font-weight: 600; color: var(--dark-green); }
.github-feedback-btn { display: inline-flex; align-items: center; gap: 8px; background-color: #24292e; color: white; padding: 10px 20px; border-radius: 6px; text-decoration: none; font-family: sans-serif; transition: background-color 0.2s; margin-top: 1rem; }
.github-feedback-btn:hover { background-color: #2c3e50; color: white; text-decoration: none; }
@media (max-width: 768px) { body { font-size: 16px; } h1.title { font-size: 2rem; } .workflow-steps { flex-direction: column; align-items: center; } }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  eval      = FALSE,
  warning   = FALSE,
  message   = FALSE,
  comment   = "#>",
  fig.align = "center",
  out.width = "90%",
  fig.retina = 2
)
```

# Introduction — What is a RAP?

## The reproducibility problem in environmental epidemiology

Public health and environmental epidemiology research in Brazil faces a
growing challenge: how can an analysis conducted today — say, on
respiratory-disease mortality in Sao Paulo during the winter of 2022 —
be reproduced, updated, and audited by another researcher three years
from now? With different package versions, updated databases, and new
municipalities to include?

**Reproducible Analytical Pipelines** (RAPs) are data science's answer
to this problem, both in government and academic settings. The concept
was formalized by the **UK Government Data Science** unit (2022) to
ensure that analyses published by public agencies are auditable,
updatable, and executable by any collaborator with access to the
correct environment.

In `climasus4r`, the RAP module serializes the entire analysis
pipeline — from importing DATASUS data to temporal aggregation — into
a portable text file (`.R`, `.Rmd`, or `.qmd`), with embedded
traceability metadata. This means you can:

- **Archive** the analysis alongside the article as supplementary material
- **Re-run** the same pipeline for new federative units or periods with
  a single command
- **Compare** two versions of the pipeline to identify divergences
- **Update** parameters without manually reopening and editing the file

```{=html}
<div class="feature-grid">

<div class="feature-card">
<div class="feature-icon">📦</div>
<h3>Portability</h3>
<p>The exported pipeline is a standard R text file. Any collaborator with <code>climasus4r</code> installed can re-run it.</p>
</div>

<div class="feature-card">
<div class="feature-icon">🔍</div>
<h3>Traceability</h3>
<p>Automatic metadata records the R version, packages, date/time, and platform at the moment of export.</p>
</div>

<div class="feature-card">
<div class="feature-icon">🔄</div>
<h3>Updatability</h3>
<p>Parameters such as state (UF), period, and DATASUS system can be overridden without editing the exported file.</p>
</div>

<div class="feature-card">
<div class="feature-icon">📊</div>
<h3>Comparability</h3>
<p>Two versions of the pipeline can be compared side by side to identify what changed between revisions.</p>
</div>

</div>
```

## The `rap_object` class

`sus_rap_export()` returns an object of the S3 class `rap_object`,
which is a named list with seven fields:

| Field | Type | Description |
|-------|------|-----------|
| `metadata` | list | R version, packages, timestamp, platform, random seed |
| `params` | list | Parameters detected in the pipeline (uf, years, system, etc.) |
| `steps` | list | Each pipeline step with its function and arguments |
| `structure` | list | Format, output type, total number of steps |
| `source` | character | Path of the exported file on disk |
| `format` | character | Export format (`"script"`, `"rmarkdown"`, etc.) |
| `raw` | character | Lines of the exported file (for direct inspection) |

```{=html}
<div class="callout callout-note">
**Note:** `rap_object` inherits from `list`, which means you can
access any field with `$` or `[[`. For example, `rap$params$uf`
returns the federative unit recorded at the time of export.
</div>
```

# Prerequisites

## Required packages

```{r prereqs-install}
# Instalacao do climasus4r (se ainda nao instalado)
# install.packages("remotes")
# remotes::install_github("ByMaxAnjos/climasus4r")

library(climasus4r)
library(dplyr)     # para manipulacao de dados
library(magrittr)  # para o operador %>%
```

## Building the example pipeline

In this tutorial we will use as our driving case an analysis of
**respiratory-disease mortality in the state of Sao Paulo, 2020-2022**,
imported from SIM-DO (Mortality Information System). The pipeline is
defined with `quote()`, which captures the expression without
evaluating it:

```{r pipeline-definicao}
pipeline <- quote(
  sus_data_import(uf = "SP", years = 2020:2022, system = "SIM-DO") |>
    sus_data_clean_encoding() |>
    sus_data_standardize() |>
    sus_data_filter_cid(cid_group = "respiratory") |>
    sus_data_aggregate(by = "month", lang = "pt")
)
```

```{=html}
<div class="callout callout-tip">
**Tip:** Using `quote()` is essential. It creates a language object
(`call`) that `sus_rap_export()` can inspect, parametrize, and
serialize. If you pass the result of a pipeline that has already been
executed, the RAP module will not be able to reconstruct the steps in
an auditable way.
</div>
```

## Checking the pipeline structure

```{r pipeline-verificacao}
# O objeto 'pipeline' e um objeto de linguagem do tipo 'call'
class(pipeline)
#> [1] "call"

# Visualizando a arvore de chamadas
lobstr::ast(!!pipeline)
#> o-`|>`
#> +-o-`|>`
#> | +-o-`|>`
#> | | +-o-`|>`
#> | | | +-sus_data_import(uf = "SP", years = 2020:2022, system = "SIM-DO")
#> | | | \-sus_data_clean_encoding()
#> | | \-sus_data_standardize()
#> | \-sus_data_filter_cid(cid_group = "respiratory")
#> \-sus_data_aggregate(by = "month", lang = "pt")
```

# `sus_rap_export()` — Serializing the Pipeline

## Full signature

```{r export-assinatura}
sus_rap_export(
  pipeline             = NULL,
  file_path            = NULL,
  format               = c("script", "rmarkdown", "quarto", "function"),
  include_metadata     = TRUE,
  include_validation   = TRUE,
  include_documentation = c("standard", "minimal", "comprehensive"),
  output_type          = c("analysis", "dashboard", "report"),
  lang                 = "pt",
  overwrite            = FALSE
)
```

## Exporting as an R script (default format)

The `"script"` format generates a self-contained `.R` file, with a
metadata header, a parameter block, a validation block, and the
pipeline itself. It is the recommended format for archiving and
re-running from the command line.

```{r export-script}
rap <- sus_rap_export(
  pipeline              = pipeline,
  file_path             = "analise_respiratoria_sp.R",
  format                = "script",
  include_metadata      = TRUE,
  include_validation    = TRUE,
  include_documentation = "standard",
  lang                  = "pt"
)
#> v Pipeline exportado para 'analise_respiratoria_sp.R'
#> i Formato      : script
#> i Etapas       : 5
#> i Parametros   : uf, years, system, cid_group, by, lang
#> i Metadados    : R 4.4.1 | climasus4r 0.1.0 | 2026-06-18 09:32:14
```

### Structure of the generated file

The exported file `analise_respiratoria_sp.R` has the following structure:

```{r export-script-conteudo}
# ============================================================
# RAP gerado por climasus4r::sus_rap_export()
# Timestamp : 2026-06-18 09:32:14 BRT
# R version : 4.4.1 (2024-06-14)
# climasus4r: 0.1.0
# Plataforma: aarch64-apple-darwin20
# ============================================================

# -- PARAMETROS (edite aqui para reutilizar o pipeline) ------
params <- list(
  uf        = "SP",
  years     = 2020:2022,
  system    = "SIM-DO",
  cid_group = "respiratory",
  by        = "month",
  lang      = "pt"
)

# -- VALIDACAO -----------------------------------------------
stopifnot(
  is.character(params$uf),
  all(params$years >= 1979),
  params$system %in% c("SIM-DO", "SIH", "SINAN", "SIA", "CNES", "SINASC")
)

# -- PIPELINE ------------------------------------------------
resultado <- sus_data_import(
    uf = params$uf, years = params$years, system = params$system
  ) |>
  sus_data_clean_encoding() |>
  sus_data_standardize() |>
  sus_data_filter_cid(cid_group = params$cid_group) |>
  sus_data_aggregate(by = params$by, lang = params$lang)
```

## Structure of the returned `rap_object`

```{r export-str}
str(rap)
#> List of 7
#>  $ metadata : list [r_version, packages, timestamp, platform, seed]
#>  $ params   : list [uf, years, system, cid_group, by, lang]
#>  $ steps    : list [5 pipeline steps]
#>  $ structure: list [type, input_params, output_params, total_steps]
#>  $ source   : chr "analise_respiratoria_sp.R"
#>  $ format   : chr "script"
#>  $ raw      : chr [linhas do arquivo exportado]
#>  - attr(*, "class")= chr [1:2] "rap_object" "list"

# Acessando campos individuais
rap$params$uf
#> [1] "SP"

rap$metadata$r_version
#> [1] "R version 4.4.1 (2024-06-14)"

rap$structure$total_steps
#> [1] 5
```

## Exporting as R Markdown

For publishing reproducible reports with narrative text, use
`format = "rmarkdown"`. The generated `.Rmd` file includes a YAML
header, configuration chunks, and the pipeline inside executable chunks.

```{r export-rmarkdown}
rap_rmd <- sus_rap_export(
  pipeline              = pipeline,
  file_path             = "relatorio_respiratorio_sp.Rmd",
  format                = "rmarkdown",
  include_metadata      = TRUE,
  include_validation    = TRUE,
  include_documentation = "comprehensive",
  output_type           = "report",
  lang                  = "pt"
)
#> v Pipeline exportado para 'relatorio_respiratorio_sp.Rmd'
#> i Formato      : rmarkdown
#> i Etapas       : 5
#> i output_type  : report
#> i Documentacao : comprehensive
```

## Exporting as Quarto

The `"quarto"` format generates a `.qmd` file compatible with the
Quarto system, ideal for modern technical documents with support for
Python, Julia, and multiple execution engines.

```{r export-quarto}
rap_qmd <- sus_rap_export(
  pipeline   = pipeline,
  file_path  = "analise_respiratoria_sp.qmd",
  format     = "quarto",
  output_type = "dashboard",
  lang       = "pt"
)
#> v Pipeline exportado para 'analise_respiratoria_sp.qmd'
#> i Formato: quarto | output_type: dashboard
```

## Exporting as an encapsulated function

The `"function"` format is best suited for integration into packages
or automation scripts. The pipeline is encapsulated in an R function
with all parameters as formal arguments, ready to be imported or
reused programmatically.

```{r export-function}
rap_fn <- sus_rap_export(
  pipeline   = pipeline,
  file_path  = "run_analise_respiratoria.R",
  format     = "function",
  lang       = "pt"
)
#> v Pipeline exportado para 'run_analise_respiratoria.R'
#> i Formato: function
#> i Funcao gerada: run_pipeline(uf, years, system, cid_group, by, lang)
```

The generated file contains a function like this:

```{r export-function-conteudo}
run_pipeline <- function(
    uf        = "SP",
    years     = 2020:2022,
    system    = "SIM-DO",
    cid_group = "respiratory",
    by        = "month",
    lang      = "pt"
) {
  sus_data_import(uf = uf, years = years, system = system) |>
    sus_data_clean_encoding() |>
    sus_data_standardize() |>
    sus_data_filter_cid(cid_group = cid_group) |>
    sus_data_aggregate(by = by, lang = lang)
}
```

## Documentation levels (`include_documentation`)

The `include_documentation` parameter controls the amount of comments
generated in the exported file:

| Level | Description |
|-------|-----------|
| `"minimal"` | Only the essential header and the parameter block |
| `"standard"` | Header, parameters, validation, and step comments |
| `"comprehensive"` | All of the above + a description of each function, expected types, and links to the documentation |

```{r export-documentacao-minimal}
# Exportacao minima -- apenas o essencial
rap_min <- sus_rap_export(
  pipeline              = pipeline,
  file_path             = "analise_minima.R",
  include_documentation = "minimal",
  lang                  = "pt"
)
```

```{r export-documentacao-comprehensive}
# Exportacao abrangente -- maxima rastreabilidade
rap_full <- sus_rap_export(
  pipeline              = pipeline,
  file_path             = "analise_completa.R",
  include_documentation = "comprehensive",
  lang                  = "pt"
)
```

```{=html}
<div class="callout callout-warning">
**Warning:** `overwrite = FALSE` is the default. If the file already
exists on disk, `sus_rap_export()` aborts with an explanatory error.
Use `overwrite = TRUE` to explicitly overwrite it, or specify a new path.
</div>
```

# `sus_rap_read()` — Reading It Back

Once exported, the `.R` file (or `.Rmd`/`.qmd`) can be read back into a
`rap_object` in any R session, even if the original pipeline no longer
exists in memory.

## Signature

```{r read-assinatura}
sus_rap_read(file_path, lang = "pt", validate = TRUE)
```

## Reading the exported file

```{r read-exemplo}
rap2 <- sus_rap_read("analise_respiratoria_sp.R", lang = "pt")
#> v Arquivo lido: 'analise_respiratoria_sp.R'
#> v Metadados validados
#> i R version  : 4.4.1 (exportado em 2026-06-18 09:32:14)
#> i climasus4r : 0.1.0
#> i Etapas     : 5
#> i Parametros : uf = "SP" | years = 2020:2022 | system = "SIM-DO"
```

## Printing the `rap_object`

The `print.rap_object()` method displays a formatted summary of the pipeline:

```{r read-print}
print(rap2)
#> -- RAP: analise_respiratoria_sp.R -----------------------
#> Formato   : script
#> Exportado : 2026-06-18 09:32:14 BRT
#> R version : 4.4.1 | climasus4r 0.1.0
#> Plataforma: aarch64-apple-darwin20
#>
#> -- Parametros ------------------------------------------
#> uf        : "SP"
#> years     : 2020 2021 2022
#> system    : "SIM-DO"
#> cid_group : "respiratory"
#> by        : "month"
#> lang      : "pt"
#>
#> -- Pipeline (5 etapas) ----------------------------------
#> [1] sus_data_import()
#> [2] sus_data_clean_encoding()
#> [3] sus_data_standardize()
#> [4] sus_data_filter_cid()
#> [5] sus_data_aggregate()
#> ---------------------------------------------------------
```

```{=html}
<div class="callout callout-note">
**Note:** The `validate = TRUE` argument (default) checks whether the
file being read contains the required metadata fields (`timestamp`,
`r_version`, `climasus4r`) and issues a warning if the file was edited
manually and its metadata is incomplete. Use `validate = FALSE` to
suppress this check.
</div>
```

# `sus_rap_inspect()` — Inspecting and Comparing

## Signature

```{r inspect-assinatura}
sus_rap_inspect(rap, rap2 = NULL, verbose = TRUE, lang = "pt")
```

## 5.1 Inspecting a single RAP

With just one `rap_object`, `sus_rap_inspect()` displays a detailed
summary of the pipeline: steps, parameters, metadata, and, with
`verbose = TRUE`, the raw code of the exported file.

```{r inspect-simples}
sus_rap_inspect(rap2, verbose = TRUE)
#> == Inspecao: analise_respiratoria_sp.R ==================
#>
#> METADADOS
#>   Timestamp : 2026-06-18 09:32:14 BRT
#>   R version : 4.4.1
#>   climasus4r: 0.1.0
#>   Plataforma: aarch64-apple-darwin20
#>   Semente   : nao definida
#>
#> ESTRUTURA
#>   Formato      : script
#>   Total etapas : 5
#>   Output type  : analysis
#>   Documentacao : standard
#>
#> PARAMETROS
#>   uf        = "SP"
#>   years     = 2020:2022
#>   system    = "SIM-DO"
#>   cid_group = "respiratory"
#>   by        = "month"
#>   lang      = "pt"
#>
#> PIPELINE
#>   [1] sus_data_import(uf, years, system)
#>   [2] sus_data_clean_encoding()
#>   [3] sus_data_standardize()
#>   [4] sus_data_filter_cid(cid_group)
#>   [5] sus_data_aggregate(by, lang)
#>
#> CODIGO (raw)
#>   [ver arquivo: analise_respiratoria_sp.R]
#> =========================================================
```

## 5.2 Comparing two RAPs (diff)

When passed two `rap_object`s, `sus_rap_inspect()` performs a
structured diff, identifying:

- Parameters that changed value
- Steps added or removed
- Discrepancies in metadata (package version, platform)

The scenario below compares the Sao Paulo pipeline with an equivalent
version exported for Rio de Janeiro:

```{r inspect-diff-criacao}
# Primeiro, exportamos o pipeline para o RJ
pipeline_rj <- quote(
  sus_data_import(uf = "RJ", years = 2020:2022, system = "SIM-DO") |>
    sus_data_clean_encoding() |>
    sus_data_standardize() |>
    sus_data_filter_cid(cid_group = "respiratory") |>
    sus_data_aggregate(by = "month", lang = "pt")
)

sus_rap_export(
  pipeline  = pipeline_rj,
  file_path = "analise_respiratoria_rj.R",
  lang      = "pt"
)

rap_rj <- sus_rap_read("analise_respiratoria_rj.R")
```

```{r inspect-diff}
sus_rap_inspect(rap2, rap_rj)
#> == Diff: [1] analise_respiratoria_sp.R vs [2] analise_respiratoria_rj.R ==
#>
#> PARAMETROS ALTERADOS
#>   uf        : "SP"  -->  "RJ"
#>
#> PARAMETROS IDENTICOS
#>   years, system, cid_group, by, lang
#>
#> ETAPAS
#>   Identicas : 5/5  (nenhuma etapa adicionada ou removida)
#>
#> METADADOS
#>   Timestamps : diferentes (2026-06-18 09:32:14  vs  2026-06-18 09:45:07)
#>   R version  : identicas (4.4.1)
#>   climasus4r : identicas (0.1.0)
#>
#> CONCLUSAO: pipelines equivalentes com parametro 'uf' distinto.
#> =========================================================
```

```{=html}
<div class="callout callout-tip">
**Tip:** The diff is especially useful for peer review. If a
collaborator modifies the pipeline and re-exports it, you can compare
the original RAP with the new one to understand exactly what changed,
without having to read the code line by line.
</div>
```

# `sus_rap_run()` — Re-running with New Parameters

## Signature

```{r run-assinatura}
sus_rap_run(rap, ..., envir = parent.frame(), dry_run = FALSE, lang = "pt")
```

The arguments passed via `...` override the RAP's parameters **only
for the current execution** — the file on disk is not modified. To
permanently modify the file, use `sus_rap_update()` (section 7).

## 6.1 Simulation with `dry_run = TRUE`

Before running the pipeline with new parameters, it is recommended to
check the code that would be executed using `dry_run = TRUE`. No
connection to DATASUS is opened, no data is downloaded.

```{r run-dry}
sus_rap_run(rap2, uf = "RJ", years = 2021:2023, dry_run = TRUE)
#> -- [dry_run] Codigo que seria executado ------------------
#>
#> sus_data_import(uf = "RJ", years = 2021:2023, system = "SIM-DO") |>
#>   sus_data_clean_encoding() |>
#>   sus_data_standardize() |>
#>   sus_data_filter_cid(cid_group = "respiratory") |>
#>   sus_data_aggregate(by = "month", lang = "pt")
#>
#> Parametros substituidos: uf = "RJ" | years = 2021:2023
#> Parametros mantidos    : system, cid_group, by, lang
#> ----------------------------------------------------------
#> i Nenhum dado foi baixado (dry_run = TRUE)
```

## 6.2 Actual execution

Once the code is confirmed, remove `dry_run = TRUE` to run the pipeline:

```{r run-efetivo}
resultado_rj <- sus_rap_run(
  rap2,
  uf    = "RJ",
  years = 2021:2023,
  lang  = "pt"
)
#> -- Executando RAP: analise_respiratoria_sp.R -------------
#> i Parametros substituidos: uf = "RJ" | years = 2021:2023
#> v [1/5] sus_data_import()     ... 148.392 registros importados
#> v [2/5] sus_data_clean_encoding() ... concluido
#> v [3/5] sus_data_standardize() ... concluido
#> v [4/5] sus_data_filter_cid()  ... 42.817 registros (cid_group = "respiratory")
#> v [5/5] sus_data_aggregate()   ... 36 linhas (mensal, 2021-2023)
#> -- Pipeline concluido em 4.2 segundos --------------------

class(resultado_rj)
#> [1] "climasus_df" "tbl_df"      "tbl"         "data.frame"
```

## 6.3 Overriding multiple parameters

You can override as many parameters as you want in a single call:

```{r run-multiplos-params}
resultado_mg_anual <- sus_rap_run(
  rap2,
  uf    = "MG",
  years = 2019:2023,
  by    = "year",
  lang  = "en"
)
#> -- Executando RAP: analise_respiratoria_sp.R -------------
#> i Parametros substituidos: uf = "MG" | years = 2019:2023 | by = "year"
#> v Pipeline concluido em 6.1 segundos
```

## 6.4 Controlling the execution environment

The `envir` argument defines in which environment the pipeline is
evaluated. The default `parent.frame()` is suitable for interactive
use. For automated executions in scripts or functions, pass
`envir = new.env()` to isolate the pipeline:

```{r run-envir}
env_isolado <- new.env(parent = baseenv())
resultado_iso <- sus_rap_run(
  rap2,
  uf   = "CE",
  envir = env_isolado
)
```

```{=html}
<div class="callout callout-warning">
**Warning:** `sus_rap_run()` requires an internet connection to
download data from DATASUS via `microdatasus`. Make sure the execution
environment has access to the DATASUS server before calling the function.
</div>
```

# `sus_rap_update()` — Updating Parameters in the File

## Signature

```{r update-assinatura}
sus_rap_update(rap, ..., backup = TRUE, lang = "pt")
```

While `sus_rap_run()` overrides parameters only for the current
execution (without touching the file), `sus_rap_update()`
**permanently edits** the parameter block in the file on disk. This is
useful when you want the "official" version of the pipeline to reflect
a permanent change in scope.

## Updating the state (UF) and the period

```{r update-uf-years}
sus_rap_update(
  rap2,
  uf    = c("RJ", "MG"),
  years = 2021:2023,
  lang  = "pt"
)
#> i Backup criado: analise_respiratoria_sp.R.bak (2026-06-18 10:02:31)
#> v Parametros atualizados em 'analise_respiratoria_sp.R':
#>   uf    : "SP"          -->  c("RJ", "MG")
#>   years : 2020:2022     -->  2021:2023
#> i Parametros nao alterados: system, cid_group, by, lang
```

## The parameter block before and after

**Before** the update:

```{r update-antes}
# -- PARAMETROS (edite aqui para reutilizar o pipeline) ------
params <- list(
  uf        = "SP",
  years     = 2020:2022,
  system    = "SIM-DO",
  cid_group = "respiratory",
  by        = "month",
  lang      = "pt"
)
```

**After** the update:

```{r update-depois}
# -- PARAMETROS (edite aqui para reutilizar o pipeline) ------
# Atualizado por sus_rap_update() em 2026-06-18 10:02:31
params <- list(
  uf        = c("RJ", "MG"),
  years     = 2021:2023,
  system    = "SIM-DO",
  cid_group = "respiratory",
  by        = "month",
  lang      = "pt"
)
```

## Automatic backup

By default, `backup = TRUE` creates a copy of the original file before
any modification, with a `.bak` suffix and timestamp. This ensures
that you never lose the previous version of the pipeline:

```{r update-backup}
# O backup e criado automaticamente:
# analise_respiratoria_sp.R.bak.2026-06-18_10-02-31

# Para restaurar manualmente:
file.copy(
  "analise_respiratoria_sp.R.bak.2026-06-18_10-02-31",
  "analise_respiratoria_sp.R",
  overwrite = TRUE
)
```

```{=html}
<div class="callout callout-tip">
**Versioning tip:** For projects under Git, combine
`sus_rap_update()` with regular commits. The history of the exported
file is the most robust traceability mechanism available, especially
for external audits.
</div>
```

## Reloading the updated RAP

After the update, re-read the file to obtain a `rap_object` with the
new parameters reflected in memory:

```{r update-recarregar}
rap2_atualizado <- sus_rap_read("analise_respiratoria_sp.R")

# Confirme a atualizacao
rap2_atualizado$params$uf
#> [1] "RJ" "MG"

rap2_atualizado$params$years
#> [1] 2021 2022 2023
```

# Complete Workflow

The full life cycle of a RAP in `climasus4r` passes through six
stages. The diagram below illustrates the typical sequence of use:

```{=html}
<div class="workflow-container">
  <div class="workflow-steps">
    <span class="workflow-step">quote(pipeline)</span>
    <span class="workflow-arrow">&#x2192;</span>
    <span class="workflow-step">sus_rap_export()</span>
    <span class="workflow-arrow">&#x2192;</span>
    <span class="workflow-step">sus_rap_read()</span>
    <span class="workflow-arrow">&#x2192;</span>
    <span class="workflow-step">sus_rap_inspect()</span>
    <span class="workflow-arrow">&#x2192;</span>
    <span class="workflow-step">sus_rap_run()</span>
    <span class="workflow-arrow">&#x2192;</span>
    <span class="workflow-step">sus_rap_update()</span>
  </div>
</div>
```

## Full reference script

The code below brings together all the steps in a single reference script:

```{r fluxo-completo}
library(climasus4r)

# 1. Definir o pipeline com quote()
pipeline <- quote(
  sus_data_import(uf = "SP", years = 2020:2022, system = "SIM-DO") |>
    sus_data_clean_encoding() |>
    sus_data_standardize() |>
    sus_data_filter_cid(cid_group = "respiratory") |>
    sus_data_aggregate(by = "month", lang = "pt")
)

# 2. Exportar para disco
rap <- sus_rap_export(
  pipeline              = pipeline,
  file_path             = "analise_respiratoria_sp.R",
  format                = "script",
  include_documentation = "standard",
  lang                  = "pt"
)

# 3. Reler de disco (simula colaborador recebendo o arquivo)
rap2 <- sus_rap_read("analise_respiratoria_sp.R")

# 4. Inspecionar
sus_rap_inspect(rap2, verbose = TRUE)

# 5. Simular execucao com novos parametros (sem baixar dados)
sus_rap_run(rap2, uf = "RJ", years = 2021:2023, dry_run = TRUE)

# 6. Executar de verdade
resultado <- sus_rap_run(rap2, uf = "RJ", years = 2021:2023)

# 7. Atualizar parametros no arquivo para escopo ampliado
sus_rap_update(rap2, uf = c("RJ", "MG"), years = 2021:2023)
```

## Quick reference table

| Function | Input | Output | Modifies disk? |
|--------|---------|-------|-----------------|
| `sus_rap_export()` | `quote(pipeline)` | `rap_object` | Yes (creates file) |
| `sus_rap_read()` | file path | `rap_object` | No |
| `sus_rap_inspect()` | 1 or 2 `rap_object` | console output | No |
| `sus_rap_run()` | `rap_object` | `climasus_df` | No |
| `sus_rap_update()` | `rap_object` + params | `rap_object` (invisible) | Yes (edits file) |

# References

- Wilson G, Bryan J, Cranston K, Kitzes J, Nederbragt L, Teal TK (2017).
  Good enough practices in scientific computing.
  *PLOS Computational Biology*, 13(6), e1005510.
  <https://doi.org/10.1371/journal.pcbi.1005510>

- Marwick B, Boettiger C, Mullen L (2018).
  Packaging data analytical work reproducibly using R (and friends).
  *The American Statistician*, 72(1), 80-88.
  <https://doi.org/10.1080/00031305.2017.1375986>

- UK Government Data Science (2022).
  *Reproducible Analytical Pipelines (RAP) Strategy*.
  Cabinet Office, His Majesty's Government.
  <https://analysisfunction.civilservice.gov.uk/support/reproducible-analytical-pipelines/>

- Anjos M (2026). *climasus4r: Integrated health-climate-environment
  analysis in Brazil*. R package version 0.1.0.
  <https://github.com/ByMaxAnjos/climasus4r>

---

**Navigation:**
[Previous: Thematic Tutorials](tutorials.html) | [Next: RAP 2 - Pipelines with targets](rap-02-targets.html)

```{=html}
<div style="text-align: center; margin-top: 3rem; padding-top: 2rem; border-top: 2px solid #e1e1e1;">

<p><strong>climasus4r</strong> · Integrated health-climate-environment analysis in Brazil</p>

<p>Found an issue or have suggestions for this tutorial?</p>

<a href="https://github.com/ByMaxAnjos/climasus4r/issues" class="github-feedback-btn" target="_blank">
  &#x1F41B; Report on GitHub
</a>

</div>
```
