---
title: "Reproducible Analytical Pipelines with climasus4r"
subtitle: "A Complete Guide to `sus_export_raps` and `sus_read_rap`"
author:
  - name: "Max Anjos"
    affiliation: "INCT-CONEXÃO Amazônia - FIOCRUZ-RO "
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
    theme: flatly
    highlight: kate
    code_folding: show
    fig_caption: true
    df_print: paged
    css: |
      /* ── Typography ─────────────────────────────────── */
      @import url('https://fonts.googleapis.com/css2?family=Inter:wght@400;500;600&family=Source+Serif+4:ital,wght@0,400;0,600;1,400&family=JetBrains+Mono:wght@400;500&display=swap');

      body {
        font-family: 'Source Serif 4', Georgia, serif;
        font-size: 17px;
        line-height: 1.85;
        color: #1a1a1a;
        max-width: 900px;
      }

      h1, h2, h3, h4, h5 {
        font-family: 'Inter', 'Helvetica Neue', sans-serif;
        font-weight: 600;
        letter-spacing: -0.02em;
        margin-top: 2em;
      }

      h1.title {
        font-size: 2.4em;
        color: #0f3d2e;
        border-bottom: 3px solid #1D9E75;
        padding-bottom: 0.3em;
        margin-bottom: 0.2em;
      }

      h3.subtitle { color: #555; font-weight: 400; margin-top: 0.1em; }

      code, pre {
        font-family: 'JetBrains Mono', 'Fira Code', monospace;
        font-size: 0.84em;
      }

      /* ── Callout boxes ──────────────────────────────── */
      .callout {
        border-radius: 6px;
        padding: 14px 20px;
        margin: 1.6em 0;
        border-left-width: 4px;
        border-left-style: solid;
      }
      .callout-title {
        font-family: 'Inter', sans-serif;
        font-weight: 600;
        font-size: 0.85em;
        text-transform: uppercase;
        letter-spacing: 0.06em;
        margin-bottom: 6px;
      }
      .callout-objective  { background: #e8f5f0; border-color: #1D9E75; }
      .callout-objective .callout-title { color: #0f6e56; }
      .callout-tip        { background: #eef6ff; border-color: #378ADD; }
      .callout-tip .callout-title { color: #185FA5; }
      .callout-warning    { background: #fff8ee; border-color: #EF9F27; }
      .callout-warning .callout-title { color: #854F0B; }
      .callout-important  { background: #fef2f2; border-color: #E24B4A; }
      .callout-important .callout-title { color: #A32D2D; }
      .callout-concept    { background: #f3f0fe; border-color: #7F77DD; }
      .callout-concept .callout-title { color: #3C3489; }

      /* ── Function reference pills ───────────────────── */
      .fn-ref {
        display: inline-block;
        background: #0f3d2e;
        color: #a8f0d8 !important;
        font-family: 'JetBrains Mono', monospace;
        font-size: 0.80em;
        padding: 2px 10px;
        border-radius: 4px;
        font-weight: 500;
        text-decoration: none;
      }

      /* ── Pipeline flow badge ────────────────────────── */
      .flow-step {
        display: inline-flex;
        align-items: center;
        gap: 8px;
        background: #f0faf6;
        border: 1px solid #9FE1CB;
        border-radius: 6px;
        padding: 6px 14px;
        font-family: 'Inter', sans-serif;
        font-size: 0.83em;
        font-weight: 500;
        color: #0f6e56;
        margin: 3px 4px;
      }
      .flow-step .step-num {
        background: #1D9E75;
        color: white;
        border-radius: 50%;
        width: 20px;
        height: 20px;
        display: inline-flex;
        align-items: center;
        justify-content: center;
        font-size: 0.78em;
        font-weight: 600;
        flex-shrink: 0;
      }

      /* ── Comparison table ───────────────────────────── */
      .compare-table { width: 100%; border-collapse: collapse; margin: 1.2em 0; font-family: 'Inter', sans-serif; font-size: 0.88em; }
      .compare-table th { background: #0f3d2e; color: white; padding: 10px 14px; text-align: left; font-weight: 500; }
      .compare-table td { padding: 9px 14px; border-bottom: 1px solid #e8e8e8; vertical-align: top; }
      .compare-table tr:nth-child(even) td { background: #f8faf9; }
      .compare-table .check { color: #1D9E75; font-weight: 600; }
      .compare-table .cross { color: #E24B4A; }

      /* ── Output block ───────────────────────────────── */
      .r-output {
        background: #1a2332;
        color: #c9d6e3;
        border-radius: 6px;
        padding: 14px 18px;
        font-family: 'JetBrains Mono', monospace;
        font-size: 0.82em;
        line-height: 1.7;
        margin: 0.5em 0 1.2em;
        overflow-x: auto;
      }
      .r-output .ok    { color: #5DCAA5; }
      .r-output .warn  { color: #EF9F27; }
      .r-output .note  { color: #888; font-style: italic; }

      /* ── Misc ───────────────────────────────────────── */
      .lead { font-size: 1.12em; color: #333; line-height: 1.7; }
      .section-intro { color: #444; font-size: 1.05em; }
      hr { border: none; border-top: 1px solid #e0e0e0; margin: 2.5em 0; }
      .toc-float { max-height: 90vh; overflow-y: auto; }
      pre.r { border-left: 3px solid #1D9E75 !important; }
      a { color: #0f6e56; }
      a:hover { color: #1D9E75; }
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = FALSE,      # set TRUE in a live session with climasus4r installed
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.align = "center",
  out.width = "100%"
)
```

<!-- ── Hero band ───────────────────────────────────────────── -->

<p class="lead">
Every epidemiological analysis starts with the same problem: raw data, manual steps,
and results that are difficult to reproduce three months later — let alone by a
colleague on another continent. The `climasus4r` package solves this by turning your
pipeline into a **shareable, re-executable artifact** with a single function call.
</p>

<p class="lead">
This tutorial walks you through the full lifecycle: from writing your first pipeline,
to exporting it as a reproducible script, to reading it back, inspecting it,
and running it with updated parameters — without touching the original file.
</p>

---

# Why Reproducible Analytical Pipelines?

Before writing any code, it is worth understanding what problem we are actually solving.

## The reproducibility crisis in health research

Imagine a common scenario in a public health department:

> *A researcher downloads mortality data from DATASUS, cleans it in Excel, applies
> some filters, and produces a chart for a ministry report. Six months later, a
> colleague asks: "Can you redo this for 2024 data?" The original analyst has left.
> Nobody knows which filters were applied. The analysis cannot be reproduced.*

This is not a failure of individual researchers — it is a structural failure. The
analysis existed only as a series of manual steps in someone's memory.

**Reproducible Analytical Pipelines (RAPs)** solve this by encoding every step of
an analysis as executable, versioned code, with enough metadata that anyone — or
any machine — can reproduce the exact result.

<div class="callout callout-concept">
<div class="callout-title">What is a RAP?</div>
A Reproducible Analytical Pipeline is a fully automated, documented, and version-
controlled workflow that transforms raw data into analytical outputs. Running it
again, on any machine, by any researcher, produces the same result — or clearly
documents why it does not.
</div>

## What `climasus4r` adds

The `climasus4r` package implements RAPs specifically for Brazilian health and climate
data (DATASUS, SINAN, INMET). The two functions at the heart of this tutorial are:

<br>
<span class="fn-ref">sus_export_raps()</span> &nbsp; — &nbsp; converts a live pipeline into a portable file with embedded metadata

<span class="fn-ref">sus_read_rap()</span> &nbsp;&nbsp;&nbsp; — &nbsp; reads that file back into a live R object

Together with <span class="fn-ref">sus_inspect_rap()</span>, <span class="fn-ref">sus_run_rap()</span>, and <span class="fn-ref">sus_update_rap()</span> they form a
complete **export → share → read → run** lifecycle.

---

# Setup and Installation

```{r installation}
# Install climasus4r from GitHub (requires remotes)
# install.packages("remotes")
remotes::install_github("ByMaxAnjos/climasus4r")

# Core dependencies used throughout this tutorial
library(climasus4r)
library(dplyr)
library(glue)
```

<div class="callout callout-tip">
<div class="callout-title">Tip — reproducibility from the start</div>
Pin your package versions with `renv` so that collaborators get the same
environment you used when exporting:

```r
renv::init()     # once, at project start
renv::snapshot() # after any install
```
</div>

---

# The `climasus4r` Pipeline Vocabulary

Before exporting anything you need to know how pipelines are built. Every
`climasus4r` workflow follows the same grammar:

<br>
<span class="flow-step"><span class="step-num">1</span> Import</span>
<span class="flow-step"><span class="step-num">2</span> Clean encoding</span>
<span class="flow-step"><span class="step-num">3</span> Standardise</span>
<span class="flow-step"><span class="step-num">4</span> Create variables</span>
<span class="flow-step"><span class="step-num">5</span> Filter</span>
<span class="flow-step"><span class="step-num">6</span> Aggregate</span>
<span class="flow-step"><span class="step-num">7</span> Export</span>
<br><br>

Each step is a `sus_*` function. The pipe operator (`|>` or `%>%`) chains them
into a single readable expression. Here is a minimal example:

```{r minimal-pipeline}
df <- sus_data_import(
    uf     = "SP",
    year   = 2022,
    system = "SIM-DO",
    lang   = "pt"
  ) |>
  sus_data_clean_encoding(lang = "pt") |>
  sus_data_standardize(lang = "pt")   |>
  sus_data_filter_cid(disease_group = "cardiovascular", lang = "pt") |>
  sus_data_aggregate(time_unit = "month", lang = "pt")
```

<div class="callout callout-tip">
<div class="callout-title">Read a pipeline like a sentence</div>
"Take São Paulo mortality records for 2022, fix the character encoding, standardise
column names, keep only cardiovascular causes, then count by month."
Each pipe step adds one clause. If any clause is hard to read, the function name
probably needs a better parameter.
</div>

---

# `sus_export_raps()` — Exporting a Pipeline

## What it does

<span class="fn-ref">sus_export_raps()</span> takes a pipeline expression and writes it to disk as a
self-contained, annotated file. That file carries:

- The full pipeline code, ready to run
- Metadata: R version, package versions, date, platform, random seed
- A configurable `params <- list(...)` block at the top so future users can adapt
  the pipeline without reading the whole script
- Optional data validation and quality-report sections
- Roxygen2 documentation (for `format = "function"`)

## Function signature

```{r signature-export}
sus_export_raps(
  pipeline,                           # the pipeline expression or object
  file_path             = NULL,       # where to write; NULL returns the content
  format                = "script",   # "script" | "rmarkdown" | "quarto" | "function"
  include_metadata      = TRUE,       # embed session info in the header
  include_validation    = TRUE,       # add stopifnot() + NA checks to output
  include_documentation = "standard", # "minimal" | "standard" | "comprehensive"
  lang                  = "pt",       # "pt" | "en"
  output_type           = "analysis", # "analysis" | "dashboard" | "report"
  overwrite             = FALSE       # guard against accidental overwrites
)
```

## Format options in detail

<table class="compare-table">
<thead>
<tr>
<th>format</th>
<th>Output file</th>
<th>Best for</th>
<th>Renders?</th>
</tr>
</thead>
<tbody>
<tr>
<td><code>"script"</code></td>
<td><code>.R</code></td>
<td>Pipeline automation, cron jobs, CI/CD</td>
<td>—</td>
</tr>
<tr>
<td><code>"rmarkdown"</code></td>
<td><code>.Rmd</code></td>
<td>Reports, peer review, collaborative research</td>
<td class="check">✓ HTML / PDF / Word</td>
</tr>
<tr>
<td><code>"quarto"</code></td>
<td><code>.qmd</code></td>
<td>Modern publications, websites, presentations</td>
<td class="check">✓ HTML / PDF / RevealJS</td>
</tr>
<tr>
<td><code>"function"</code></td>
<td><code>.R</code></td>
<td>Package development, reusable modules</td>
<td>—</td>
</tr>
</tbody>
</table>

## Example 1 — Standalone R script

The simplest and most portable export. Use this for scheduled analyses, server
jobs, or any situation where you want a single file that runs top-to-bottom.

```{r export-script}
# Define the pipeline
pipeline <- quote(
  sus_data_import(
    uf     = c("SP", "RJ", "MG"),
    year   = 2020:2023,
    system = "SIM-DO",
    lang   = "pt"
  ) |>
  sus_data_clean_encoding(lang = "pt")       |>
  sus_data_standardize(lang = "pt")          |>
  sus_data_filter_cid(
    disease_group = "respiratory",
    lang          = "pt"
  ) |>
  sus_create_variables(
    create_age_groups = TRUE,
    age_breaks        = c(0, 1, 5, Inf),
    age_labels        = c("< 1 yr", "1–4 yr", "5+ yr"),
    lang              = "pt"
  ) |>
  sus_data_filter_demographics(
    age_range = c(0, 5),
    lang      = "pt"
  ) |>
  sus_data_aggregate(
    time_unit = "month",
    group_by  = "age_group",
    lang      = "pt"
  )
)

# Export as a standalone R script
sus_export_raps(
  pipeline              = pipeline,
  file_path             = "pipeline_respiratory_paediatric.R",
  format                = "script",
  include_metadata      = TRUE,
  include_validation    = TRUE,
  include_documentation = "comprehensive",
  lang                  = "pt"
)
```

<div class="r-output">
<span class="ok">🔍 Iniciando exportação | R version 4.4.1 (2024-06-14)</span><br>
<span class="ok">📋 Etapas detectadas no pipeline: 7</span><br>
<span class="ok">⚙️  Gerando conteúdo no formato: 'script'</span><br>
<span class="ok">✅ Pipeline exportado com sucesso para: pipeline_respiratory_paediatric.R</span>
</div>

### What the exported file looks like

The resulting `pipeline_respiratory_paediatric.R` opens with a reproducibility
header, then a params block the reader can edit, then the pipeline itself wrapped
in a `tryCatch`, followed by validation and `sessionInfo()`:

```{r show-exported-script, eval=FALSE, echo=TRUE}
# ============================================================================
# Pipeline Reprodutível — Pipeline Mortalidade por Doenças Respiratórias
# ============================================================================
# Versão R  : R version 4.4.1 (2024-06-14)
# Data      : 2025-01-15
# Pacote    : climasus4r 0.1.0
# Plataforma: linux
# Pacotes carregados no momento da exportação:
# climasus4r: 0.1.0
# dplyr: 1.1.4
# ============================================================================

# ── 0. Reprodutibilidade ────────────────────────────────────────────────────
set.seed(42L)

# ── 1. Dependências ──────────────────────────────────────────────────────────
required_pkgs <- c("climasus4r", "dplyr", "glue")
missing_pkgs <- required_pkgs[
  !vapply(required_pkgs, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1L))
]
if (length(missing_pkgs) > 0L) {
  stop(paste("Pacotes ausentes:", paste(missing_pkgs, collapse = ", ")))
}
library(climasus4r); library(dplyr); library(glue)

# ── 2. Parâmetros configuráveis (EDITE AQUI) ─────────────────────────────────
params <- list(
  uf         = c("SP", "RJ", "MG"),
  years      = 2020:2023,
  system     = "SIM-DO",
  time_unit  = "month",
  output_dir = "resultados",
  lang       = "pt",
  seed       = 42L
)

# ── 3. Execução do pipeline ───────────────────────────────────────────────────
# ...
```

<div class="callout callout-tip">
<div class="callout-title">Tip — the params block is the handover point</div>
When you share the exported file, collaborators only need to look at the
`params <- list(...)` block. Every dial they might want to turn is there.
They never need to read the pipeline code itself to adapt the analysis.
</div>

## Example 2 — RMarkdown report

Use this when the analysis needs to be **communicated**, not just reproduced.
The exported `.Rmd` includes a YAML header, a params table rendered with
`knitr::kable()`, the pipeline code, and optional visualisations.

```{r export-rmarkdown}
sus_export_raps(
  pipeline              = pipeline,
  file_path             = "report_respiratory_paediatric.Rmd",
  format                = "rmarkdown",
  include_metadata      = TRUE,
  include_validation    = TRUE,
  include_documentation = "comprehensive",
  output_type           = "report",
  lang                  = "pt"
)
```

## Example 3 — Quarto document

Identical intent to RMarkdown but uses the modern Quarto format, which supports
code annotations, cross-references, and multi-format output from a single source.

```{r export-quarto}
sus_export_raps(
  pipeline    = pipeline,
  file_path   = "report_respiratory_paediatric.qmd",
  format      = "quarto",
  output_type = "dashboard",
  lang        = "en"          # produces English comments and messages
)
```

## Example 4 — Reusable function

The most powerful format for package developers or research teams building
shared analysis libraries. The exported file contains a fully documented,
Roxygen2-annotated `run_pipeline()` function with `stopifnot()` argument
validation and an `invisible()` return.

```{r export-function}
sus_export_raps(
  pipeline              = pipeline,
  file_path             = "run_respiratory_paediatric.R",
  format                = "function",
  include_validation    = TRUE,
  include_documentation = "comprehensive",
  lang                  = "pt"
)
```

The resulting function can be called like this:

```{r call-exported-function}
# Source the exported function
source("run_respiratory_paediatric.R")

# Run with the original parameters
df_sp <- run_pipeline()

# Override any parameter at call time
df_rj <- run_pipeline(
  uf        = "RJ",
  years     = 2021:2024,
  time_unit = "week"
)
```

<div class="callout callout-tip">
<div class="callout-title">Tip — preview before writing</div>
Pass `file_path = NULL` to return the content as a character vector without
writing any file. This is useful for inspection or for piping into a custom
writer:

```r
content <- sus_export_raps(pipeline, format = "script")
cat(content[1:30], sep = "\n")   # preview the first 30 lines
```
</div>

## The `overwrite` guard

By default, `sus_export_raps()` refuses to overwrite an existing file:

```{r overwrite-demo}
# Second call to the same path raises an error by default
sus_export_raps(pipeline, "pipeline_respiratory_paediatric.R")
#> Error: O arquivo 'pipeline_respiratory_paediatric.R' já existe.
#>        Use overwrite = TRUE para sobrescrever.

# Explicitly allow overwriting when you intend to
sus_export_raps(pipeline, "pipeline_respiratory_paediatric.R", overwrite = TRUE)
```

This guard protects against accidentally clobbering an exported file that a
collaborator may have already started editing.

---

# `sus_read_rap()` — Reading a Pipeline Back

## What it does

<span class="fn-ref">sus_read_rap()</span> is the mirror of `sus_export_raps()`. It reads any exported
file — `.R`, `.Rmd`, or `.qmd` — and reconstructs a live **`rap_object`** in your
R session. No re-running, no re-downloading: everything is parsed from the file
metadata and structure.

```{r signature-read}
sus_read_rap(
  file_path,          # path to the exported file
  lang     = "pt",    # language for log messages
  validate = TRUE     # run integrity checks on the parsed object
)
```

## The `rap_object` structure

Every `rap_object` is an S3 list with six fields:

```{r rap-object-structure, eval=FALSE}
rap$metadata   # r_version, packages, timestamp, platform, seed
rap$params     # uf, years, system, time_unit, lang, output_dir
rap$steps      # list of pipeline steps with function names and arguments
rap$structure  # derived: type, input_params, output_params, total_steps
rap$source     # normalised path to the file on disk
rap$format     # "script" | "rmarkdown" | "quarto" | "function"
rap$raw        # character vector — the raw file content
```

## Reading the exported file

```{r read-basic}
rap <- sus_read_rap(
  file_path = "pipeline_respiratory_paediatric.R",
  lang      = "pt",
  validate  = TRUE
)
```

<div class="r-output">
<span class="ok">📖 Lendo RAP: /projects/epidemio/pipeline_respiratory_paediatric.R</span><br>
<span class="note">   Formato detectado: script</span><br>
<span class="ok">✅ Leitura OK | 7 etapas | uf=c("SP", "RJ", "MG") | years=2020:2023</span>
</div>

```{r print-rap}
# The default print method gives a compact overview
print(rap)
```

<div class="r-output">
&lt;rap_object&gt;<br>
&nbsp;&nbsp;source : /projects/epidemio/pipeline_respiratory_paediatric.R<br>
&nbsp;&nbsp;format : script<br>
&nbsp;&nbsp;type   : Mortalidade por Doenças Respiratórias<br>
&nbsp;&nbsp;steps  : 7<br>
&nbsp;&nbsp;uf     : SP, RJ, MG<br>
&nbsp;&nbsp;years  : 2020:2023
</div>

## Accessing fields directly

```{r access-fields}
# Which UFs does this pipeline cover?
rap$params$uf
#> [1] "SP" "RJ" "MG"

# What year range?
rap$params$years
#> [1] 2020 2021 2022 2023

# What package version was used when this was exported?
rap$metadata$package_version
#> [1] "0.1.0"

# What R version?
rap$metadata$r_version
#> [1] "R version 4.4.1 (2024-06-14)"

# List all pipeline step names in order
vapply(rap$steps, `[[`, "function_name", FUN.VALUE = character(1L))
#> [1] "sus_data_import"              "sus_data_clean_encoding"
#> [3] "sus_data_standardize"         "sus_data_filter_cid"
#> [5] "sus_create_variables"         "sus_data_filter_demographics"
#> [7] "sus_data_aggregate"
```

<div class="callout callout-tip">
<div class="callout-title">Tip — use `validate = FALSE` for legacy files</div>
If you're reading a hand-edited or older file that lacks some expected metadata
fields, `validate = FALSE` makes the parser more permissive. You'll still get a
`rap_object` back — just with `NULL` in the missing fields rather than an error.
</div>

---

# Inspecting, Running, and Updating

`sus_read_rap()` is the entry point to three further operations. Each one is a
single function call.

## `sus_inspect_rap()` — understand what you have

```{r inspect-single}
sus_inspect_rap(rap, verbose = TRUE)
```

<div class="r-output">
── RAP object summary ──────────────────────────────────────────<br>
&nbsp;&nbsp;Source        &nbsp; /projects/epidemio/pipeline_respiratory_paediatric.R<br>
&nbsp;&nbsp;Format        &nbsp; script<br>
&nbsp;&nbsp;Type          &nbsp; Mortalidade por Doenças Respiratórias<br>
&nbsp;&nbsp;R version     &nbsp; R version 4.4.1 (2024-06-14)<br>
&nbsp;&nbsp;Package       &nbsp; climasus4r 0.1.0<br>
&nbsp;&nbsp;Created       &nbsp; 2025-01-15 09:14:22 -03<br>
&nbsp;&nbsp;Seed          &nbsp; 42<br>
<br>
&nbsp;&nbsp;Params:<br>
&nbsp;&nbsp;uf             &nbsp; SP, RJ, MG<br>
&nbsp;&nbsp;years          &nbsp; 2020:2023<br>
&nbsp;&nbsp;system         &nbsp; SIM-DO<br>
&nbsp;&nbsp;time_unit      &nbsp; month<br>
<br>
&nbsp;&nbsp;Pipeline steps (7):<br>
&nbsp;&nbsp;&nbsp;&nbsp;1. sus_data_import              uf="SP","RJ","MG" year=2020:2023 system="SIM-DO"<br>
&nbsp;&nbsp;&nbsp;&nbsp;2. sus_data_clean_encoding      lang="pt"<br>
&nbsp;&nbsp;&nbsp;&nbsp;3. sus_data_standardize         lang="pt"<br>
&nbsp;&nbsp;&nbsp;&nbsp;4. sus_data_filter_cid          disease_group="respiratory"<br>
&nbsp;&nbsp;&nbsp;&nbsp;5. sus_create_variables         create_age_groups=TRUE<br>
&nbsp;&nbsp;&nbsp;&nbsp;6. sus_data_filter_demographics age_range=0:5<br>
&nbsp;&nbsp;&nbsp;&nbsp;7. sus_data_aggregate           time_unit="month"
</div>

### Comparing two pipeline versions with diff mode

Pass two `rap_object`s to see exactly what changed between versions:

```{r inspect-diff}
rap_v1 <- sus_read_rap("pipeline_respiratory_paediatric_v1.R")
rap_v2 <- sus_read_rap("pipeline_respiratory_paediatric_v2.R")

sus_inspect_rap(rap_v1, rap_v2)
```

<div class="r-output">
── Diff: RAP 1 vs RAP 2 ──────────────────────────────────────<br>
<br>
&nbsp;&nbsp;Params changed:<br>
&nbsp;&nbsp;uf        &nbsp;&nbsp; SP, RJ, MG &nbsp;&nbsp;→&nbsp;&nbsp; SP, RJ, MG, ES, BA<br>
&nbsp;&nbsp;years     &nbsp;&nbsp; 2020:2023  &nbsp;&nbsp;→&nbsp;&nbsp; 2020:2024<br>
<br>
&nbsp;&nbsp;Steps added in RAP 2:<br>
&nbsp;&nbsp;+ sus_data_quality_report<br>
<br>
&nbsp;&nbsp;Package version drift:<br>
&nbsp;&nbsp;climasus4r  &nbsp;&nbsp; 0.1.0 → 0.2.0<br>
&nbsp;&nbsp;dplyr       &nbsp;&nbsp; 1.1.3 → 1.1.4
</div>

<div class="callout callout-important">
<div class="callout-title">Why version drift matters</div>
A `climasus4r` minor version bump may carry changed defaults in any `sus_*`
function. If two researchers run nominally "the same" pipeline on different
package versions, their outputs may differ. The diff surface this
immediately — before anyone wastes time debugging discrepancies in results.
</div>

## `sus_run_rap()` — re-execute with optional overrides

```{r run-original}
# Re-run with the original parameters
df_original <- sus_run_rap(rap)
```

```{r run-overrides}
# Override specific parameters without touching the file
df_nordeste <- sus_run_rap(
  rap,
  uf    = c("CE", "PE", "BA", "MA"),
  years = 2022:2024
)
```

```{r run-dry}
# Audit the reconstructed call before executing anything
sus_run_rap(rap, uf = "AM", dry_run = TRUE)
```

<div class="r-output">
── [dry_run] Código que seria executado ──────────────────────<br>
sus_data_import(uf = "AM", year = 2020:2023, system = "SIM-DO", lang = "pt", parallel = TRUE)<br>
|&gt; sus_data_clean_encoding(lang = "pt")<br>
|&gt; sus_data_standardize(lang = "pt")<br>
|&gt; sus_data_filter_cid(disease_group = "respiratory", lang = "pt")<br>
|&gt; sus_create_variables(create_age_groups = TRUE)<br>
|&gt; sus_data_filter_demographics(age_range = 0:5, lang = "pt")<br>
|&gt; sus_data_aggregate(time_unit = "month", lang = "pt")
</div>

<div class="callout callout-tip">
<div class="callout-title">Tip — always dry-run before a long download</div>
DATASUS requests for multiple states and years can take several minutes.
`dry_run = TRUE` costs nothing and lets you verify that the reconstructed
call looks exactly right before committing.
</div>

## `sus_update_rap()` — patch the file in place

When you want to change parameters and persist that change back to disk:

```{r update-rap}
# Add two more states and extend the year range
rap_updated <- sus_update_rap(
  rap,
  uf    = 'c("SP", "RJ", "MG", "ES", "BA")',   # verbatim R syntax
  years = "2020:2024",
  backup = TRUE   # creates pipeline_respiratory_paediatric.R.bak automatically
)
```

<div class="r-output">
<span class="ok">💾 Backup criado: pipeline_respiratory_paediatric.R.bak</span><br>
<span class="ok">  ✎ uf = c("SP", "RJ", "MG", "ES", "BA")</span><br>
<span class="ok">  ✎ years = 2020:2024</span><br>
<span class="ok">✅ Arquivo atualizado: pipeline_respiratory_paediatric.R</span><br>
<span class="ok">✅ Leitura OK | 7 etapas | uf=c("SP","RJ","MG","ES","BA") | years=2020:2024</span>
</div>

<div class="callout callout-warning">
<div class="callout-title">Pass values as strings</div>
`sus_update_rap()` writes values **verbatim** into the R source file.
Pass the string that should appear in the file, not the R object:

```r
# Correct
sus_update_rap(rap, uf = 'c("SP", "RJ")')

# Wrong — writes the evaluated R vector, not valid R syntax in the file
sus_update_rap(rap, uf = c("SP", "RJ"))
```
</div>

---

# The Full Lifecycle — A Complete Worked Example

This section assembles everything into a single, end-to-end workflow. Follow it
as a template for any new research project.

## Step 1 — Write the pipeline

```{r full-pipeline}
library(climasus4r)
library(dplyr)

# ── Pipeline: dengue seasonality, Amazon region, 2015–2020 ──────────────────
pipeline_dengue <- quote(
  sus_data_import(
    uf     = "AM",
    year   = 2015:2020,
    system = "SINAN-DENGUE",
    lang   = "pt"
  ) |>
  sus_data_clean_encoding(lang = "pt")   |>
  sus_data_standardize(lang = "pt")      |>
  sus_create_variables(
    create_calendar_vars = TRUE,
    lang                 = "pt"
  ) |>
  sus_data_aggregate(
    time_unit = "season",
    lang      = "pt"
  )
)
```

## Step 2 — Export as a reproducible script

```{r full-export}
sus_export_raps(
  pipeline              = pipeline_dengue,
  file_path             = "pipeline_dengue_amazon.R",
  format                = "script",
  include_metadata      = TRUE,
  include_validation    = TRUE,
  include_documentation = "comprehensive",
  lang                  = "pt"
)
```

## Step 3 — Commit to version control

```{bash git-commit, eval=FALSE}
git add pipeline_dengue_amazon.R
git commit -m "feat: dengue seasonality pipeline, AM 2015-2020"
git push
```

Now any collaborator with access to the repository can reproduce your exact
analysis.

## Step 4 — Collaborator reads and inspects the pipeline

```{r collab-read}
# On a colleague's machine, after cloning the repository:
rap <- sus_read_rap("pipeline_dengue_amazon.R", lang = "pt")

# Get a full picture of what this pipeline does
sus_inspect_rap(rap, verbose = TRUE)
```

## Step 5 — Run with extended scope

```{r collab-run}
# Re-run for additional years without editing the file
df_extended <- sus_run_rap(
  rap,
  years = 2015:2023,   # original was 2015:2020
  dry_run = FALSE
)
```

## Step 6 — Persist the change and re-export as an RMarkdown report

```{r collab-update-export}
# Update the file to reflect the new scope
rap_v2 <- sus_update_rap(rap, years = "2015:2023")

# Produce a full report for publication
sus_export_raps(
  pipeline    = rap_v2,
  file_path   = "report_dengue_amazon_2015_2023.Rmd",
  format      = "rmarkdown",
  output_type = "report",
  lang        = "pt"
)
```

## Step 7 — Audit the difference between versions

```{r audit-diff}
rap_v1 <- sus_read_rap("pipeline_dengue_amazon.R")
rap_v2 <- sus_read_rap("pipeline_dengue_amazon_v2.R")

sus_inspect_rap(rap_v1, rap_v2)
# → shows years changed 2015:2020 → 2015:2023, all else identical
```

<div class="callout callout-objective">
<div class="callout-title">What this workflow gives you</div>
Every decision in the analysis is encoded in a file. The file carries its own
provenance metadata. Any collaborator — now or five years from now — can read
it, understand it, and re-execute it. This is what reproducible research looks
like in practice.
</div>

---

# Advanced Patterns

## Pattern 1 — Batch export across multiple disease groups

```{r batch-export}
disease_groups <- c("cardiovascular", "respiratory", "heat_related")

for (grp in disease_groups) {
  pipeline_expr <- bquote(
    sus_data_import(
      uf     = "SP",
      year   = 2020:2023,
      system = "SIM-DO",
      lang   = "pt"
    ) |>
    sus_data_clean_encoding(lang = "pt") |>
    sus_data_standardize(lang = "pt")   |>
    sus_data_filter_cid(
      disease_group = .(grp),
      lang          = "pt"
    ) |>
    sus_data_aggregate(time_unit = "month", lang = "pt")
  )

  sus_export_raps(
    pipeline  = pipeline_expr,
    file_path = glue("pipeline_{grp}_sp_2020_2023.R"),
    format    = "script",
    lang      = "pt"
  )
}
```

## Pattern 2 — Reading multiple RAPs and comparing them

```{r multi-read}
# Read all pipeline files in a directory
rap_files <- list.files("pipelines/", pattern = "\\.R$", full.names = TRUE)
rap_list  <- lapply(rap_files, sus_read_rap, validate = FALSE)

# Summarise all at once
summaries <- lapply(rap_list, function(r) {
  list(
    file       = basename(r$source),
    type       = r$structure$type,
    uf         = paste(r$params$uf, collapse = ", "),
    years      = paste(range(r$params$years), collapse = "–"),
    n_steps    = length(r$steps),
    pkg_version = r$metadata$package_version
  )
})

# Render as a summary table
do.call(rbind, lapply(summaries, as.data.frame))
```

## Pattern 3 — Chaining the full lifecycle in one expression

```{r chained}
# Read → update → run — no intermediate objects needed
result <- sus_read_rap("pipeline_dengue_amazon.R") |>
  sus_update_rap(years = "2015:2023") |>
  sus_run_rap()
```

## Pattern 4 — Checking for version drift before a large run

```{r version-guard}
rap <- sus_read_rap("pipeline_cardiovascular_sp.R")

current_version <- as.character(packageVersion("climasus4r"))
exported_version <- rap$metadata$package_version

if (!identical(current_version, exported_version)) {
  warning(glue(
    "Package version mismatch: pipeline was exported with climasus4r ",
    "{exported_version}, current session has {current_version}. ",
    "Results may differ. Consider re-exporting or pinning with renv."
  ))
}
```

---

# Reference — Function Quick Guide

<table class="compare-table">
<thead>
<tr>
<th>Function</th>
<th>Input</th>
<th>Output</th>
<th>Primary use</th>
</tr>
</thead>
<tbody>
<tr>
<td><code>sus_export_raps()</code></td>
<td>pipeline expression</td>
<td>file on disk</td>
<td>Author — freeze a pipeline for sharing</td>
</tr>
<tr>
<td><code>sus_read_rap()</code></td>
<td>file path</td>
<td><code>rap_object</code></td>
<td>Consumer — load a pipeline into R</td>
</tr>
<tr>
<td><code>sus_inspect_rap()</code></td>
<td>1 or 2 <code>rap_object</code>s</td>
<td>printed summary / diff</td>
<td>Audit — understand or compare pipelines</td>
</tr>
<tr>
<td><code>sus_run_rap()</code></td>
<td><code>rap_object</code> + overrides</td>
<td><code>data.frame</code></td>
<td>Execute — re-run with optional new params</td>
</tr>
<tr>
<td><code>sus_update_rap()</code></td>
<td><code>rap_object</code> + patches</td>
<td>updated <code>rap_object</code></td>
<td>Evolve — patch params without re-exporting</td>
</tr>
</tbody>
</table>

---

# Troubleshooting

<table class="compare-table">
<thead>
<tr><th>Symptom</th><th>Likely cause</th><th>Fix</th></tr>
</thead>
<tbody>
<tr>
<td><code>rap$params$uf</code> is <code>NULL</code></td>
<td>File was hand-edited and the params block syntax changed</td>
<td>Check that the <code>params &lt;- list(...)</code> block follows the expected indented key = value, format</td>
</tr>
<tr>
<td><code>sus_run_rap()</code> produces different results than the original run</td>
<td>Package version drift; seed not set before sampling</td>
<td>Check <code>sus_inspect_rap()</code> for version drift; ensure <code>set.seed()</code> appears before any sampling step</td>
</tr>
<tr>
<td><code>sus_update_rap()</code> key not found warning</td>
<td>Key name typo, or params block was renamed in the file</td>
<td>Check <code>names(rap$params)</code> for the exact key names</td>
</tr>
<tr>
<td><code>sus_export_raps()</code> file exists error</td>
<td>Output file already present</td>
<td>Pass <code>overwrite = TRUE</code>, or choose a new path</td>
</tr>
<tr>
<td>Steps count is lower than expected after <code>sus_read_rap()</code></td>
<td>Some <code>sus_*</code> calls not recognised by the parser</td>
<td>Ensure function names match the known <code>sus_*</code> vocabulary; custom wrapper functions won't be detected automatically</td>
</tr>
</tbody>
</table>

---

# Closing Notes

`sus_export_raps()` and `sus_read_rap()` are not just convenience wrappers — they
encode a philosophy: **every analytical decision should be inspectable, shareable,
and reproducible by default**. In a field where data quality, methodology transparency,
and auditability matter as much as the results themselves, this is not a luxury.

The RAP lifecycle described in this tutorial maps directly onto the requirements of
modern open science: FAIR data principles, pre-registration, reproducibility checks
during peer review, and long-term archival of analytical code alongside results.

By adopting this workflow, you are not just making your own work easier to reproduce.
You are making it easier for everyone who comes after you — collaborators, reviewers,
replicators, and future-you — to build on what you built.

<div class="callout callout-objective">
<div class="callout-title">Three things to do right now</div>

1. Take a pipeline you have written recently and export it with
   `sus_export_raps(..., format = "script")`. Open the file and read the params block.

2. Send the file to a colleague. Ask them to run
   `sus_read_rap()` followed by `sus_inspect_rap()` and tell you what they see.

3. Try `sus_run_rap(..., dry_run = TRUE)` on the `rap_object` before executing anything.
   Get comfortable auditing the reconstructed call.

These three steps are the entire workflow. Everything else is variation on this theme.
</div>

---

<div style="text-align:center; color:#888; font-size:0.85em; margin-top:3em; font-family:'Inter',sans-serif;">
Built with `climasus4r` &nbsp;·&nbsp;
<a href="https://github.com/ByMaxAnjos/climasus4r">github.com/ByMaxAnjos/climasus4r</a> &nbsp;·&nbsp;
Report issues via GitHub Issues
</div>

```{r session-info, echo=FALSE, eval=TRUE}
utils::sessionInfo()
```
